### 2023-04-19 index.qmd
initial commit

---
title: "swhawes.github.io"
---

This is a Quarto website.

To learn more about Quarto websites visit <https://quarto.org/docs/websites>.

### 2023-04-19 1a_Difference Scores paired samples ttest_HowTo.qmd
rstudio commit #2

---
format: 
  #html:
  #  toc: true
  #  code-fold: true
  docx:
    toc: true
prefer-html: true
---

# Paired Samples T-test {.unnumbered}

## Overview

<p>

This example examines the question whether participants reported
different mean levels of externalizing problems on the CBCL when
measured at Baseline (T0) compared to when measured at the 1-Year
follow-up (T1). This analysis is conducted in two primary steps: 1)
computing a difference score (DiffScore = CBCL EXT T2 - CBCL EXT T1); 2)
conducting a paired samples t-test on the difference score. Our primary
aim is to determine whether average CBCL externalizing scores of the
participants change from assessement 1 (T1) to assessement 2 (T2), while
accounting for observations that are clustered within youth over time.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

#rm(list = ls())

if (!("lme4" %in% installed.packages())) install.packages("lme4")
if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("arrow" %in% installed.packages())) install.packages("arrow")
if (!("arsenal" %in% installed.packages())) install.packages("arsenal")
if (!("kableExtra" %in% installed.packages())) install.packages("kableExtra")
if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
if (!("rstatix" %in% installed.packages())) install.packages("rstatix")

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| class.source: fold-hide

#rm(list = ls())

library(lme4)
library(lmerTest)
library(tidyverse)
library(arrow)
library(arsenal)
library(kableExtra)
library(rstatix)
library(ggpubr)
```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                                  comment=NA, cache=T, code_folding=T,
                                  R.options=list(width=220), fig.align='center', 
                                  out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Read data
df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
df_wide <- read_csv("/Users/shawes/Desktop/data/df_wide.csv")

# set types
df_long$ids <- as.factor(df_long$ids)
df_long$event <- as.factor(df_long$event)
df_long$site_id <- as.factor(df_long$site_id)
df_long$fam_id <- as.factor(df_long$fam_id)
df_long$sex <- as.factor(df_long$sex)
df_long$age <- as.numeric(df_long$age)
df_long$mature_vg <- as.numeric(df_long$mature_vg)
df_long$vg_wkdy <- as.numeric(df_long$vg_wkdy)
df_long$vg_wknd <- as.numeric(df_long$vg_wknd)
df_long$cbcl_extern <- as.numeric(df_long$cbcl_extern)
df_long$cbcl_cd <- as.numeric(df_long$cbcl_cd)
df_long$ksad_cdco1 <- as.factor(df_long$ksad_cdco1)
df_long$vg_total <- as.numeric(df_long$vg_total)
```
:::

### Continuous Outcomes

::: blue
**This code creates a table of descriptive information for continuous
outcomes**

```{r contvar-table}
#| messages: FALSE
#| echo: TRUE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_1 <- tableby.control(test=FALSE, total=FALSE, numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"), #"Nmiss2"
cat.stats=c("countpct"), #"Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median ='Median', 
                  range='Min - Max'))  #, Nmiss2 ='Missing'
my_cont_labels <- list(age = "Age", vg_total = "Weekly # of Video Gaming Hrs",
            cbcl_extern = "CBCL Externalizing Scale"
)

tab_descriptives_1 <- tableby(event ~ age + vg_total + 
                                      cbcl_extern,
                                      data=df_long, control=tab_descriptives_1)

# Push table object through kable and kable_styling
tab_descriptives_1 %>%
  summary(text=TRUE, digits.pct=1, digits=1) %>%
  kable(caption = "Continuous Outcomes") %>%
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, html_font = "Cambria",
                font_size = 15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>% 
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")
```
:::

### Categorical Outcomes

::: blue
**This code creates a descriptives table for categorical outcomes**

```{r catvar-table}
#| class.source: 'fold-hide'
#| messages: FALSE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"
), # "Nmiss2"
cat.stats=c("countpct"), # "Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median
='Median', range='Min - Max'
)) # , Nmiss2 ='Missing'

my_cat_labels  <- list(
  event = "Year",
  sex = "Sex",
  mature_vg = "Mature Video Games"
  )

tab_descriptives_2 <- tableby(event ~ sex + mature_vg, data=df_long, control=tab_descriptives_2)

# Push table object through kable and kable_styling
tab_descriptives_2 %>%
            summary(text=TRUE, digits.pct=1, digits=1) %>%
            kable(caption = "Categorical Outcomes") %>%
            kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size =
                            15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  #column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")

```
:::
:::

## Results

<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue
The code snippet below tells R to compute a difference score by
subtracting each participant's Externalizing score at T2 from their
Externalizing score at T1. Next, a one-sample t-test is computed to
examine whether the average difference score is different than zero
(indicating change).

**STEP 1: Compute Difference Score**

```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Compute difference score based on CBCL Externalizing subscale scores at baseline (t1) and 1-Year Follow-up (t2)
df_wide$diffscore <- df_wide$cbcl_extern_Year_1 - df_wide$cbcl_extern_Baseline
```

```{r model-checking}
# Compute statistical summaries for the difference score variable
summary(df_wide$diffscore)

```

This summary of the difference score variable indicates xxxxx.

```{r}
hist(df_long$vg_total)

## Summary statistics
summary<-df_long %>%
group_by(event) %>%
get_summary_stats(cbcl_extern, type = "mean_sd")
data.frame(summary)
```

This histogram indicates xxxxxx. The summary command shows xxxxxx.

```{r shapiro-wilks}
### Shapiro-Wilk test and normality (Q-Q) plot (visualization of correlation between a given sample and the normal distribution)
shapiro.test(df_wide$cbcl_extern_Baseline[0:5000])
```

This Shapiro-Wilkes test shows xxxxxx.

```{r qqplot}
qqplot <- ggqqplot(df_wide$cbcl_extern_Baseline,
    ylab = "Externalizing Difference Score", xlab = FALSE,
    ggtheme = theme_minimal()
)

suppressWarnings(print(qqplot)) 
```

This qqplot shows xxxxxx.

**STEP 2: Conduct t-test on Difference Score**

```{r build-model-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# One-sample t-test to determine whether the average difference score is significantly different than 0.
result <- t.test(df_wide$cbcl_extern_Baseline, mu = 0, alternative = "two.sided")
# Print the results
result
```
:::

The output from our model provides: i. a t-statistic ii. degrees of
freedom under iii. p-value (\< .05 then we reject the null hypothesis)
iv. a mean estimate v. 95% confidence intervals

In our example p = .002 which is less than .05 then we reject the null
hypothesis and conclude that the mean difference score µ is different
from zero. So we can conclude that there𝐷 is a difference between cbcl
externalizing scores from t1 to t2. This output also provides the mean
for cbcl externalizing scores (4.45),as well as the correpsonding
confidence intervals

### Model Plots

::: blue
```{r scatterplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

# Scatterplot to visualize relationship between t1 & t2 data used to create difference score
scatterplot <- ggplot(df_wide, aes(x = cbcl_extern_Baseline, y = cbcl_extern_Year_1)) +
    geom_point(size = 3) +
    geom_smooth(method = lm, se = F) +
    xlab("CBCL Externalizing (x) Baseline") +
    ylab("CBCL Externalizing (x) 1-Year Follow-up")

suppressWarnings(print(scatterplot))
```

Examination of this scatterplot indicates xxxxx.

```{r boxplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

### boxplot
diffscore_boxplot <- ggboxplot(df_wide$cbcl_extern_Baseline,
    ylab = "Externalizing Difference Score", xlab = FALSE,
    ggtheme = theme_minimal()
)

suppressWarnings(print(diffscore_boxplot))
```

Examination of this boxplot indicates xxxxx.
:::

> This output shows xxxxxxxx. Briefly walk through each metric
:::

## Wrapping Up

<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue
A paired samples t test was conducted to compare differences in
participants cbcl externalzing scores measured across two timepoints.
Findings indicated that there was a difference between cbcl externalzing
scores obtained at T1 compared to cbcl externalzing scores obtained at
T2, t(179) = 3.10, p = .002, Cohen's d = 0.23. Youth exhibited an
increase of cbcl externalzing scores at t2 ( M = 5.67, SD = 1.24)
compared to t1 ( M = 5.83, SD = 1.21).
:::
:::

### 2023-04-19 2b_Residualized Change Score multiple regression_HowTo.qmd
rstudio commit #2

--- 2023-04-19 1a_Difference Scores paired samples ttest_HowTo.qmd
@@
 format: 
-  #html:
-  #  toc: true
-  #  code-fold: true
+  html:
+    toc: true
+    code-fold: true
   docx:
@@
 
-# Paired Samples T-test {.unnumbered}
+# Mulitple Regression {.unnumbered}
 
@@
 <p>
-
-This example examines the question whether participants reported
-different mean levels of externalizing problems on the CBCL when
-measured at Baseline (T0) compared to when measured at the 1-Year
-follow-up (T1). This analysis is conducted in two primary steps: 1)
-computing a difference score (DiffScore = CBCL EXT T2 - CBCL EXT T1); 2)
-conducting a paired samples t-test on the difference score. Our primary
-aim is to determine whether average CBCL externalizing scores of the
-participants change from assessement 1 (T1) to assessement 2 (T2), while
-accounting for observations that are clustered within youth over time.
-
+insert text
 </p>
@@
 ## Preliminary Setup
+::: {.panel-tabset}
+### Install Packages {.tabset .tabset-fade .tabset-pills}
+<div class = "blue">
 
-::: panel-tabset
-### Install Packages {.tabset .tabset-fade .tabset-pills}
+> **This code installs the r packages necessary for this example, if they are not already installed**
 
-::: blue
-> **This code installs the r packages necessary for this example, if
-> they are not already installed**
-
-```{r pckg-install}
-#| echo: TRUE
-#| messages: FALSE
+```{r}
+#| class.source: 'fold-hide'
+#| message: FALSE
 #| warning: FALSE
+#| echo: FALSE
 
@@
 
+# Install necessary packages (if not already installed)
+if (!("rstatix" %in% installed.packages())) install.packages("rstatix")
 if (!("lme4" %in% installed.packages())) install.packages("lme4")
@@
 if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
-if (!("arrow" %in% installed.packages())) install.packages("arrow")
-if (!("arsenal" %in% installed.packages())) install.packages("arsenal")
-if (!("kableExtra" %in% installed.packages())) install.packages("kableExtra")
 if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
-if (!("rstatix" %in% installed.packages())) install.packages("rstatix")
-
+if (!("emmeans" %in% installed.packages())) install.packages("emmeans")
 ```
-:::
+</div>
 
 ### Load Packages
+<div class = "blue">
 
-::: blue
-**This code loads the r libraries necessary for this example**
+> **This code loads the r libraries necessary for this example**
 
-```{r lib-load}
+```{r pckg-load}
 #| echo: TRUE
@@
 
+#Load packages
+library(rstatix)
 library(lme4)
@@
 library(tidyverse)
-library(arrow)
-library(arsenal)
-library(kableExtra)
-library(rstatix)
 library(ggpubr)
+library(emmeans)
 ```
-:::
+
+</div>
 
 ### Config Options
+<div class = "blue">
 
-::: blue
 **This code configures knitr code chunk options**
@@
 ```
-:::
+</div>
 :::
@@
 ## Descriptives Overview
+::: {.panel-tabset}
+### Read and View Data {.tabset .tabset-fade .tabset-pills}
+<div class = "blue">
 
-::: panel-tabset
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-**This code reads in and shows the data to be used in the current
-example**
+**This code reads in and shows the data to be used in the current example**
 
@@
 
-# set types
+# set types (if necessary)
 df_long$ids <- as.factor(df_long$ids)
@@
 ```
+
+</div>
+
 :::
@@
 ::: blue
-The code snippet below tells R to compute a difference score by
-subtracting each participant's Externalizing score at T2 from their
-Externalizing score at T1. Next, a one-sample t-test is computed to
-examine whether the average difference score is different than zero
-(indicating change).
 
-**STEP 1: Compute Difference Score**
+The code snippet below tells R to compute xxxxx by xxxxxxx.
+Next, xxxxx is conducted to examine whether xxxxxx does xxxxxxx (indicating xxxxx).
+The xxxx function (xxx) is provided by the r 'xxx' package.
 
+**STEP 1: Compute xxxxxxxx**
 ```{r build-model-1}
@@
 
-# Compute difference score based on CBCL Externalizing subscale scores at baseline (t1) and 1-Year Follow-up (t2)
-df_wide$diffscore <- df_wide$cbcl_extern_Year_1 - df_wide$cbcl_extern_Baseline
+#A multiple regression approach can be used to examine effects of a continuous covariate  when controlling for autoregressive effects (t1->t2)
+
+#Multiple regression model
+result = lm(cbcl_extern_Year_1 ~ cbcl_extern_Baseline + vg_total_Baseline, data = df_wide)
+str(df_wide)
+# fit ML model including mean structure to make comparable with FIML fit below
+# (means are always included with FIML model fits)
+# sem.mod.fit <- sem(sem.mod, data=tpb.df, meanstructure=TRUE)
+
+# fit again including missing data also
+#sem.mod.fit.fiml <- sem(sem.mod, data=tpb.df, missing="ML")
+
 ```
@@
 ```{r model-checking}
-# Compute statistical summaries for the difference score variable
-summary(df_wide$diffscore)
+#Diagnostic Plots
+
+summary(result)
+
+# residuals vs. fitted values plot.
+resid_plot<- plot(result, add.smooth = FALSE, which = 1)
+
+# normal probability plot
+norm_plot<- plot(result, which = 2)
+
+# scale location plot
+scale_plot<- plot(result, add.smooth = FALSE, which = 3)
+
+# boxplot
+#boxplot <- ggplot(df_wide, aes(x = Externalizing_T1, y = Externalizing_T0, colour = Sex)) + geom_point()
+
+#suppressWarnings(print(resid_plot))
+#suppressWarnings(print(norm_plot))
+#suppressWarnings(print(scale_plot))
+#suppressWarnings(print(boxplot))
 
@@
 
-This summary of the difference score variable indicates xxxxx.
+This summary of xxxx indicates xxxxx.
+
+The output from our model provides: i. xxxx; ii. xxxxx; iii. xxxx. In our example xxxxx indicates xxxxx. An examination of xxxxx is xxxx, indicating xxxxxx. Then we conclude that xxxxxx.
 
 ```{r}
-hist(df_long$vg_total)
+# Boxplots
 
-## Summary statistics
-summary<-df_long %>%
-group_by(event) %>%
-get_summary_stats(cbcl_extern, type = "mean_sd")
-data.frame(summary)
+df_wide %>%
+            group_by(Sex) %>%  
+            summarise(mean_grade = mean(Externalizing_T0),
+            sd_grade = sd(Externalizing_T0),
+            mean_exam = mean(Externalizing_T1),
+            sd_exam = sd(Externalizing_T1))
+
+# Boxplot of CBCL Externalizing scores at baseline by sex
+boxplot_cbclExt_T1 <- boxplot(Externalizing_T0 ~ Sex,data = df_wide,
+main = "CBCL Ext Baseline Score by sex",
+xlab = "Sex",ylab = "Baseline CBCL Externalizing Score",
+col = "red",border = "black")
+
+# Boxplot of CBCL Externalizing scores at 1-Year follow-up by sex
+boxplot_cbclExt_T2 <- boxplot(Externalizing_T1 ~ Sex, data = df_wide,
+main = "CBCL Ext 1-Year Follow-up Score by sex",
+xlab = "Sex",ylab = "1-Year Follow-up CBCL Externalizing Score",
+col = "red",border = "black")
+
+suppressWarnings(print(boxplot_cbclExt_T1))
+suppressWarnings(print(boxplot_cbclExt_T2))
 ```
 
-This histogram indicates xxxxxx. The summary command shows xxxxxx.
-
-```{r shapiro-wilks}
-### Shapiro-Wilk test and normality (Q-Q) plot (visualization of correlation between a given sample and the normal distribution)
-shapiro.test(df_wide$cbcl_extern_Baseline[0:5000])
-```
-
-This Shapiro-Wilkes test shows xxxxxx.
-
-```{r qqplot}
-qqplot <- ggqqplot(df_wide$cbcl_extern_Baseline,
-    ylab = "Externalizing Difference Score", xlab = FALSE,
-    ggtheme = theme_minimal()
-)
-
-suppressWarnings(print(qqplot)) 
-```
-
-This qqplot shows xxxxxx.
-
-**STEP 2: Conduct t-test on Difference Score**
-
-```{r build-model-2}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-
-# One-sample t-test to determine whether the average difference score is significantly different than 0.
-result <- t.test(df_wide$cbcl_extern_Baseline, mu = 0, alternative = "two.sided")
-# Print the results
-result
-```
-:::
-
-The output from our model provides: i. a t-statistic ii. degrees of
-freedom under iii. p-value (\< .05 then we reject the null hypothesis)
-iv. a mean estimate v. 95% confidence intervals
-
-In our example p = .002 which is less than .05 then we reject the null
-hypothesis and conclude that the mean difference score µ is different
-from zero. So we can conclude that there𝐷 is a difference between cbcl
-externalizing scores from t1 to t2. This output also provides the mean
-for cbcl externalizing scores (4.45),as well as the correpsonding
-confidence intervals
-
-### Model Plots
-
-::: blue
-```{r scatterplot}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| code-summary: testing
-
-# Scatterplot to visualize relationship between t1 & t2 data used to create difference score
-scatterplot <- ggplot(df_wide, aes(x = cbcl_extern_Baseline, y = cbcl_extern_Year_1)) +
-    geom_point(size = 3) +
-    geom_smooth(method = lm, se = F) +
-    xlab("CBCL Externalizing (x) Baseline") +
-    ylab("CBCL Externalizing (x) 1-Year Follow-up")
-
-suppressWarnings(print(scatterplot))
-```
-
-Examination of this scatterplot indicates xxxxx.
-
-```{r boxplot}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| code-summary: testing
-
-### boxplot
-diffscore_boxplot <- ggboxplot(df_wide$cbcl_extern_Baseline,
-    ylab = "Externalizing Difference Score", xlab = FALSE,
-    ggtheme = theme_minimal()
-)
-
-suppressWarnings(print(diffscore_boxplot))
-```
-
-Examination of this boxplot indicates xxxxx.
-:::
-
-> This output shows xxxxxxxx. Briefly walk through each metric
-:::
+These boxplots indicates xxxxxx.
 
 ## Wrapping Up
-
 <div class = "blue">
@@
 ::: blue
-A paired samples t test was conducted to compare differences in
-participants cbcl externalzing scores measured across two timepoints.
-Findings indicated that there was a difference between cbcl externalzing
-scores obtained at T1 compared to cbcl externalzing scores obtained at
-T2, t(179) = 3.10, p = .002, Cohen's d = 0.23. Youth exhibited an
-increase of cbcl externalzing scores at t2 ( M = 5.67, SD = 1.24)
-compared to t1 ( M = 5.83, SD = 1.21).
+
+A xxxxxx analysis was conducted to examine xxxxx. Findings showed xxxxxx
+
 :::

### 2023-04-19 1a_Difference Scores paired samples ttest_HowTo.qmd
rstudio test commit 3
rstudio test commit 4

@@
 ---
-format: 
-  #html:
-  #  toc: true
-  #  code-fold: true
-  docx:
-    toc: true
-prefer-html: true
+title: "DiffScoreExample"
 ---
@@
 </p>
-
-## Preliminary Setup
-
-::: panel-tabset
-### Install Packages {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-> **This code installs the r packages necessary for this example, if
-> they are not already installed**
-
-```{r pckg-install}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-#rm(list = ls())
-
-if (!("lme4" %in% installed.packages())) install.packages("lme4")
-if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
-if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
-if (!("arrow" %in% installed.packages())) install.packages("arrow")
-if (!("arsenal" %in% installed.packages())) install.packages("arsenal")
-if (!("kableExtra" %in% installed.packages())) install.packages("kableExtra")
-if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
-if (!("rstatix" %in% installed.packages())) install.packages("rstatix")
-
-```
-:::
-
-### Load Packages
-
-::: blue
-**This code loads the r libraries necessary for this example**
-
-```{r lib-load}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| class.source: fold-hide
-
-#rm(list = ls())
-
-library(lme4)
-library(lmerTest)
-library(tidyverse)
-library(arrow)
-library(arsenal)
-library(kableExtra)
-library(rstatix)
-library(ggpubr)
-```
-:::
-
-### Config Options
-
-::: blue
-**This code configures knitr code chunk options**
-
-```{r config}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
-                                  comment=NA, cache=T, code_folding=T,
-                                  R.options=list(width=220), fig.align='center', 
-                                  out.width='75%', fig.asp=.75)
-```
-:::
-:::
-
-## Descriptives Overview
-
-::: panel-tabset
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-**This code reads in and shows the data to be used in the current
-example**
-
-```{r read-data}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-## Read data
-df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
-df_wide <- read_csv("/Users/shawes/Desktop/data/df_wide.csv")
-
-# set types
-df_long$ids <- as.factor(df_long$ids)
-df_long$event <- as.factor(df_long$event)
-df_long$site_id <- as.factor(df_long$site_id)
-df_long$fam_id <- as.factor(df_long$fam_id)
-df_long$sex <- as.factor(df_long$sex)
-df_long$age <- as.numeric(df_long$age)
-df_long$mature_vg <- as.numeric(df_long$mature_vg)
-df_long$vg_wkdy <- as.numeric(df_long$vg_wkdy)
-df_long$vg_wknd <- as.numeric(df_long$vg_wknd)
-df_long$cbcl_extern <- as.numeric(df_long$cbcl_extern)
-df_long$cbcl_cd <- as.numeric(df_long$cbcl_cd)
-df_long$ksad_cdco1 <- as.factor(df_long$ksad_cdco1)
-df_long$vg_total <- as.numeric(df_long$vg_total)
-```
-:::
-
-### Continuous Outcomes
-
-::: blue
-**This code creates a table of descriptive information for continuous
-outcomes**
-
-```{r contvar-table}
-#| messages: FALSE
-#| echo: TRUE
-#| warning: FALSE
-#| results: "asis"
-
-## Create a descriptives table of study variables by measurement occasion 
-tab_descriptives_1 <- tableby.control(test=FALSE, total=FALSE, numeric.test="kwt", cat.test="chisq",
-numeric.stats=c("N", "meansd", "median", "range"), #"Nmiss2"
-cat.stats=c("countpct"), #"Nmiss2"
-stats.labels=list(N='Count', meansd="Mean (SD)", median ='Median', 
-                  range='Min - Max'))  #, Nmiss2 ='Missing'
-my_cont_labels <- list(age = "Age", vg_total = "Weekly # of Video Gaming Hrs",
-            cbcl_extern = "CBCL Externalizing Scale"
-)
-
-tab_descriptives_1 <- tableby(event ~ age + vg_total + 
-                                      cbcl_extern,
-                                      data=df_long, control=tab_descriptives_1)
-
-# Push table object through kable and kable_styling
-tab_descriptives_1 %>%
-  summary(text=TRUE, digits.pct=1, digits=1) %>%
-  kable(caption = "Continuous Outcomes") %>%
-  kable_styling(bootstrap_options = "striped", 
-                full_width = FALSE, html_font = "Cambria",
-                font_size = 15, position = "center", fixed_thead = T) %>%
-  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>% 
-  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
-  column_spec(2, width = "20em", border_right = T) %>%
-  column_spec(3, width = "20em", border_right = T) %>%
-  footnote(general = "Here is a general comments of the table. ") %>%
-  scroll_box(width = "75%", height = "500px")
-```
-:::
-
-### Categorical Outcomes
-
-::: blue
-**This code creates a descriptives table for categorical outcomes**
-
-```{r catvar-table}
-#| class.source: 'fold-hide'
-#| messages: FALSE
-#| warning: FALSE
-#| results: "asis"
-
-## Create a descriptives table of study variables by measurement occasion 
-tab_descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
-numeric.test="kwt", cat.test="chisq",
-numeric.stats=c("N", "meansd", "median", "range"
-), # "Nmiss2"
-cat.stats=c("countpct"), # "Nmiss2"
-stats.labels=list(N='Count', meansd="Mean (SD)", median
-='Median', range='Min - Max'
-)) # , Nmiss2 ='Missing'
-
-my_cat_labels  <- list(
-  event = "Year",
-  sex = "Sex",
-  mature_vg = "Mature Video Games"
-  )
-
-tab_descriptives_2 <- tableby(event ~ sex + mature_vg, data=df_long, control=tab_descriptives_2)
-
-# Push table object through kable and kable_styling
-tab_descriptives_2 %>%
-            summary(text=TRUE, digits.pct=1, digits=1) %>%
-            kable(caption = "Categorical Outcomes") %>%
-            kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size =
-                            15, position = "center", fixed_thead = T) %>%
-  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
-  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
-  column_spec(2, width = "20em", border_right = T) %>%
-  #column_spec(3, width = "20em", border_right = T) %>%
-  footnote(general = "Here is a general comments of the table. ") %>%
-  scroll_box(width = "75%", height = "500px")
-
-```
-:::
-:::
-
-## Results
-
-<div class = "blue">
-
-::: panel-tabset
-### Build Model {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-The code snippet below tells R to compute a difference score by
-subtracting each participant's Externalizing score at T2 from their
-Externalizing score at T1. Next, a one-sample t-test is computed to
-examine whether the average difference score is different than zero
-(indicating change).
-
-**STEP 1: Compute Difference Score**
-
-```{r build-model-1}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-
-# Compute difference score based on CBCL Externalizing subscale scores at baseline (t1) and 1-Year Follow-up (t2)
-df_wide$diffscore <- df_wide$cbcl_extern_Year_1 - df_wide$cbcl_extern_Baseline
-```
-
-```{r model-checking}
-# Compute statistical summaries for the difference score variable
-summary(df_wide$diffscore)
-
-```
-
-This summary of the difference score variable indicates xxxxx.
-
-```{r}
-hist(df_long$vg_total)
-
-## Summary statistics
-summary<-df_long %>%
-group_by(event) %>%
-get_summary_stats(cbcl_extern, type = "mean_sd")
-data.frame(summary)
-```
-
-This histogram indicates xxxxxx. The summary command shows xxxxxx.
-
-```{r shapiro-wilks}
-### Shapiro-Wilk test and normality (Q-Q) plot (visualization of correlation between a given sample and the normal distribution)
-shapiro.test(df_wide$cbcl_extern_Baseline[0:5000])
-```
-
-This Shapiro-Wilkes test shows xxxxxx.
-
-```{r qqplot}
-qqplot <- ggqqplot(df_wide$cbcl_extern_Baseline,
-    ylab = "Externalizing Difference Score", xlab = FALSE,
-    ggtheme = theme_minimal()
-)
-
-suppressWarnings(print(qqplot)) 
-```
-
-This qqplot shows xxxxxx.
-
-**STEP 2: Conduct t-test on Difference Score**
-
-```{r build-model-2}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-
-# One-sample t-test to determine whether the average difference score is significantly different than 0.
-result <- t.test(df_wide$cbcl_extern_Baseline, mu = 0, alternative = "two.sided")
-# Print the results
-result
-```
-:::
-
-The output from our model provides: i. a t-statistic ii. degrees of
-freedom under iii. p-value (\< .05 then we reject the null hypothesis)
-iv. a mean estimate v. 95% confidence intervals
-
-In our example p = .002 which is less than .05 then we reject the null
-hypothesis and conclude that the mean difference score µ is different
-from zero. So we can conclude that there𝐷 is a difference between cbcl
-externalizing scores from t1 to t2. This output also provides the mean
-for cbcl externalizing scores (4.45),as well as the correpsonding
-confidence intervals
-
-### Model Plots
-
-::: blue
-```{r scatterplot}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| code-summary: testing
-
-# Scatterplot to visualize relationship between t1 & t2 data used to create difference score
-scatterplot <- ggplot(df_wide, aes(x = cbcl_extern_Baseline, y = cbcl_extern_Year_1)) +
-    geom_point(size = 3) +
-    geom_smooth(method = lm, se = F) +
-    xlab("CBCL Externalizing (x) Baseline") +
-    ylab("CBCL Externalizing (x) 1-Year Follow-up")
-
-suppressWarnings(print(scatterplot))
-```
-
-Examination of this scatterplot indicates xxxxx.
-
-```{r boxplot}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| code-summary: testing
-
-### boxplot
-diffscore_boxplot <- ggboxplot(df_wide$cbcl_extern_Baseline,
-    ylab = "Externalizing Difference Score", xlab = FALSE,
-    ggtheme = theme_minimal()
-)
-
-suppressWarnings(print(diffscore_boxplot))
-```
-
-Examination of this boxplot indicates xxxxx.
-:::
-
-> This output shows xxxxxxxx. Briefly walk through each metric
-:::
-
-## Wrapping Up
-
-<div class = "blue">
-
-::: panel-tabset
-### Write-up {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-A paired samples t test was conducted to compare differences in
-participants cbcl externalzing scores measured across two timepoints.
-Findings indicated that there was a difference between cbcl externalzing
-scores obtained at T1 compared to cbcl externalzing scores obtained at
-T2, t(179) = 3.10, p = .002, Cohen's d = 0.23. Youth exhibited an
-increase of cbcl externalzing scores at t2 ( M = 5.67, SD = 1.24)
-compared to t1 ( M = 5.83, SD = 1.21).
-:::
-:::

### 2023-04-19 2b_Residualized Change Score multiple regression_HowTo.qmd
rstudio test commit 3
rstudio test commit 4

@@
 ---
-format: 
-  html:
-    toc: true
-    code-fold: true
-  docx:
-    toc: true
-prefer-html: true
+title: "Residualized Change Score Example"
 ---
 
-# Mulitple Regression {.unnumbered}
+# Mulitple Regression
+testing testing testing
 
@@
 </p>
-
-## Preliminary Setup
-::: {.panel-tabset}
-### Install Packages {.tabset .tabset-fade .tabset-pills}
-<div class = "blue">
-
-> **This code installs the r packages necessary for this example, if they are not already installed**
-
-```{r}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: FALSE
-
-#rm(list = ls())
-
-# Install necessary packages (if not already installed)
-if (!("rstatix" %in% installed.packages())) install.packages("rstatix")
-if (!("lme4" %in% installed.packages())) install.packages("lme4")
-if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
-if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
-if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
-if (!("emmeans" %in% installed.packages())) install.packages("emmeans")
-```
-</div>
-
-### Load Packages
-<div class = "blue">
-
-> **This code loads the r libraries necessary for this example**
-
-```{r pckg-load}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-#| class.source: fold-hide
-
-#rm(list = ls())
-
-#Load packages
-library(rstatix)
-library(lme4)
-library(lmerTest)
-library(tidyverse)
-library(ggpubr)
-library(emmeans)
-```
-
-</div>
-
-### Config Options
-<div class = "blue">
-
-**This code configures knitr code chunk options**
-
-```{r config}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
-                      comment=NA, cache=T, code_folding=T,
-                      R.options=list(width=220), fig.align='center',
-                      out.width='75%', fig.asp=.75)
-```
-</div>
-:::
-
-## Descriptives Overview
-::: {.panel-tabset}
-### Read and View Data {.tabset .tabset-fade .tabset-pills}
-<div class = "blue">
-
-**This code reads in and shows the data to be used in the current example**
-
-```{r read-data}
-#| echo: TRUE
-#| messages: FALSE
-#| warning: FALSE
-
-## Read data
-df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
-df_wide <- read_csv("/Users/shawes/Desktop/data/df_wide.csv")
-
-# set types (if necessary)
-df_long$ids <- as.factor(df_long$ids)
-df_long$event <- as.factor(df_long$event)
-df_long$site_id <- as.factor(df_long$site_id)
-df_long$fam_id <- as.factor(df_long$fam_id)
-df_long$sex <- as.factor(df_long$sex)
-df_long$age <- as.numeric(df_long$age)
-df_long$mature_vg <- as.numeric(df_long$mature_vg)
-df_long$vg_wkdy <- as.numeric(df_long$vg_wkdy)
-df_long$vg_wknd <- as.numeric(df_long$vg_wknd)
-df_long$cbcl_extern <- as.numeric(df_long$cbcl_extern)
-df_long$cbcl_cd <- as.numeric(df_long$cbcl_cd)
-df_long$ksad_cdco1 <- as.factor(df_long$ksad_cdco1)
-df_long$vg_total <- as.numeric(df_long$vg_total)
-```
-
-</div>
-
-:::
-
-### Continuous Outcomes
-
-::: blue
-**This code creates a table of descriptive information for continuous
-outcomes**
-
-```{r contvar-table}
-#| messages: FALSE
-#| echo: TRUE
-#| warning: FALSE
-#| results: "asis"
-
-## Create a descriptives table of study variables by measurement occasion 
-tab_descriptives_1 <- tableby.control(test=FALSE, total=FALSE, numeric.test="kwt", cat.test="chisq",
-numeric.stats=c("N", "meansd", "median", "range"), #"Nmiss2"
-cat.stats=c("countpct"), #"Nmiss2"
-stats.labels=list(N='Count', meansd="Mean (SD)", median ='Median', 
-                  range='Min - Max'))  #, Nmiss2 ='Missing'
-my_cont_labels <- list(age = "Age", vg_total = "Weekly # of Video Gaming Hrs",
-            cbcl_extern = "CBCL Externalizing Scale"
-)
-
-tab_descriptives_1 <- tableby(event ~ age + vg_total + 
-                                      cbcl_extern,
-                                      data=df_long, control=tab_descriptives_1)
-
-# Push table object through kable and kable_styling
-tab_descriptives_1 %>%
-  summary(text=TRUE, digits.pct=1, digits=1) %>%
-  kable(caption = "Continuous Outcomes") %>%
-  kable_styling(bootstrap_options = "striped", 
-                full_width = FALSE, html_font = "Cambria",
-                font_size = 15, position = "center", fixed_thead = T) %>%
-  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>% 
-  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
-  column_spec(2, width = "20em", border_right = T) %>%
-  column_spec(3, width = "20em", border_right = T) %>%
-  footnote(general = "Here is a general comments of the table. ") %>%
-  scroll_box(width = "75%", height = "500px")
-```  
-:::
-
-### Categorical Outcomes
-
-::: blue
-**This code creates a descriptives table for categorical outcomes**
-
-```{r catvar-table}
-#| class.source: 'fold-hide'
-#| messages: FALSE
-#| warning: FALSE
-#| results: "asis"
-
-## Create a descriptives table of study variables by measurement occasion 
-tab_descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
-numeric.test="kwt", cat.test="chisq",
-numeric.stats=c("N", "meansd", "median", "range"
-), # "Nmiss2"
-cat.stats=c("countpct"), # "Nmiss2"
-stats.labels=list(N='Count', meansd="Mean (SD)", median
-='Median', range='Min - Max'
-)) # , Nmiss2 ='Missing'
-
-my_cat_labels  <- list(
-  event = "Year",
-  sex = "Sex",
-  mature_vg = "Mature Video Games"
-  )
-
-tab_descriptives_2 <- tableby(event ~ sex + mature_vg, data=df_long, control=tab_descriptives_2)
-
-# Push table object through kable and kable_styling
-tab_descriptives_2 %>%
-            summary(text=TRUE, digits.pct=1, digits=1) %>%
-            kable(caption = "Categorical Outcomes") %>%
-            kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size =
-                            15, position = "center", fixed_thead = T) %>%
-  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
-  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
-  column_spec(2, width = "20em", border_right = T) %>%
-  #column_spec(3, width = "20em", border_right = T) %>%
-  footnote(general = "Here is a general comments of the table. ") %>%
-  scroll_box(width = "75%", height = "500px")
-
-```
-:::
-:::
-
-## Results
-
-<div class = "blue">
-
-::: panel-tabset
-### Build Model {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-
-The code snippet below tells R to compute xxxxx by xxxxxxx.
-Next, xxxxx is conducted to examine whether xxxxxx does xxxxxxx (indicating xxxxx).
-The xxxx function (xxx) is provided by the r 'xxx' package.
-
-**STEP 1: Compute xxxxxxxx**
-```{r build-model-1}
-#| class.source: 'fold-hide'
-#| message: FALSE
-#| warning: FALSE
-#| echo: TRUE
-
-#A multiple regression approach can be used to examine effects of a continuous covariate  when controlling for autoregressive effects (t1->t2)
-
-#Multiple regression model
-result = lm(cbcl_extern_Year_1 ~ cbcl_extern_Baseline + vg_total_Baseline, data = df_wide)
-str(df_wide)
-# fit ML model including mean structure to make comparable with FIML fit below
-# (means are always included with FIML model fits)
-# sem.mod.fit <- sem(sem.mod, data=tpb.df, meanstructure=TRUE)
-
-# fit again including missing data also
-#sem.mod.fit.fiml <- sem(sem.mod, data=tpb.df, missing="ML")
-
-```
-
-```{r model-checking}
-#Diagnostic Plots
-
-summary(result)
-
-# residuals vs. fitted values plot.
-resid_plot<- plot(result, add.smooth = FALSE, which = 1)
-
-# normal probability plot
-norm_plot<- plot(result, which = 2)
-
-# scale location plot
-scale_plot<- plot(result, add.smooth = FALSE, which = 3)
-
-# boxplot
-#boxplot <- ggplot(df_wide, aes(x = Externalizing_T1, y = Externalizing_T0, colour = Sex)) + geom_point()
-
-#suppressWarnings(print(resid_plot))
-#suppressWarnings(print(norm_plot))
-#suppressWarnings(print(scale_plot))
-#suppressWarnings(print(boxplot))
-
-```
-
-This summary of xxxx indicates xxxxx.
-
-The output from our model provides: i. xxxx; ii. xxxxx; iii. xxxx. In our example xxxxx indicates xxxxx. An examination of xxxxx is xxxx, indicating xxxxxx. Then we conclude that xxxxxx.
-
-```{r}
-# Boxplots
-
-df_wide %>%
-            group_by(Sex) %>%  
-            summarise(mean_grade = mean(Externalizing_T0),
-            sd_grade = sd(Externalizing_T0),
-            mean_exam = mean(Externalizing_T1),
-            sd_exam = sd(Externalizing_T1))
-
-# Boxplot of CBCL Externalizing scores at baseline by sex
-boxplot_cbclExt_T1 <- boxplot(Externalizing_T0 ~ Sex,data = df_wide,
-main = "CBCL Ext Baseline Score by sex",
-xlab = "Sex",ylab = "Baseline CBCL Externalizing Score",
-col = "red",border = "black")
-
-# Boxplot of CBCL Externalizing scores at 1-Year follow-up by sex
-boxplot_cbclExt_T2 <- boxplot(Externalizing_T1 ~ Sex, data = df_wide,
-main = "CBCL Ext 1-Year Follow-up Score by sex",
-xlab = "Sex",ylab = "1-Year Follow-up CBCL Externalizing Score",
-col = "red",border = "black")
-
-suppressWarnings(print(boxplot_cbclExt_T1))
-suppressWarnings(print(boxplot_cbclExt_T2))
-```
-
-These boxplots indicates xxxxxx.
-
-## Wrapping Up
-<div class = "blue">
-
-::: panel-tabset
-### Write-up {.tabset .tabset-fade .tabset-pills}
-
-::: blue
-
-A xxxxxx analysis was conducted to examine xxxxx. Findings showed xxxxxx
-
-:::
-:::

### 2023-04-19 11_Tutorials_MultivariateLatentGrowthCurveModels.qmd
rstudio 9

---
title: "Multivariate Latent Growth Curves"
---

### 2023-04-19 15_Tutorials_RandomInterceptCrosslaggedPanelModels.qmd
rstudio 9

---
title: "Random-Intercept Crosslagged Panel Models"
---

### 2023-04-19 16_Tutorials_LatentCurveModelsStructuredResiduals.qmd
rstudio 9

---
title: "Latent Curve Models with Structured Residuals"
---

### 2023-04-19 1_Tutorials_DifferenceScores.qmd
rstudio 9

---
title: "Difference Scores"
author: "Biostatistics Working Group"
---

## Overview
Difference scores are one of the earliest developed and commonly used statistical approaches to compare data collected from the same individual across two measurement occasions [@castro-schilo2018; @jennings2016]. The *difference* between scores at the two time points is calculated for each individual and the resulting value is taken as a measure of change. It is common to then perform statistical tests on the difference scores, such as being included as an outcome in a GLM analysis to test for differences in patterns of change over time and between groups. For example, difference scores may be used in a paired-samples t-test to compare mean test scores of students before and after attending a math workshop, or in a simple regression analysis to assess the effectiveness of a weight loss program by calculating the difference in weight between between groups of interest, before and after the program.

![Difference Score 1_Tutorial](./figures/Difference Score 1_Tutorial.png)

### When to use Difference Score Models?
You should use difference scores in the following scenario:
1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Difference Score Models
In this tutorial, we will guide you through two simple examples of using difference scores. By the end of this tutorial, you will be able to:

1. Understand the basic concepts of difference scores.
2. Calculate difference scores using example data.
3. Interpret the results of the difference scores analysis.

## Basic Example

For this tutorial, we will use a simple dataset containing the following information for a group of individuals:

- Time point (T1, T2)
- Job satisfaction (independent variable)
- Life satisfaction (dependent variable)

| Individual | Time Point | Job Satisfaction | Life Satisfaction |
| ---------- | ---------- | ---------------- | ----------------- |
| A          | T1         | 7                | 6                 |
| A          | T2         | 8                | 7                 |
| B          | T1         | 6                | 5                 |
| B          | T2         | 7                | 6                 |

We will then create a new variable, age_diff, that represents the difference in age between two time points:

```{r}
#titanic$age_diff <- titanic$age - titanic$age[1]
```

### Repeated Measures Paired Samples T-test
To conduct a repeated measures paired samples t-test on age_diff, we will use the t.test() function in R:

```{r}
#t.test(titanic$age_diff, mu = 0, paired = TRUE)\
```

The mu argument represents the hypothesized mean difference, which we have set to 0. The paired argument tells R that the samples are paired.

### Interpreting the Results
The output of the t.test() function includes several statistics, including the t-value, degrees of freedom, and p-value. The t-value represents the difference between the mean of age_diff and the hypothesized mean difference, divided by the standard error. The degrees of freedom represents the number of observations minus one. The p-value represents the probability of observing a t-value as extreme or more extreme than the one observed, assuming that the null hypothesis is true.

If the p-value is less than our chosen significance level (typically 0.05), we can reject the null hypothesis and conclude that there is a significant difference between the means of age_diff and the hypothesized mean difference. If the p-value is greater than our chosen significance level, we fail to reject the null hypothesis and conclude that there is insufficient evidence to suggest a difference between the means.

### Conclusion
In this tutorial, we learned how to conduct a repeated measures paired samples t-test on a difference score using R. This statistical method is useful when we want to compare the means of two related variables that are measured at two different time points or under two different conditions. We also learned how to interpret the results of the t-test, including the t-value, degrees of freedom, and p-value.

### Paired Samples T-test Model Specification and Estimation

To calculate [xxxxx] scores, we will follow these steps:

[two timepoints, t1 and t2, and we want to test whether there is a significant difference between the means of two variables measured at each timepoint. We can use a paired samples t-test to do this.]

A paired samples t-test is used to determine whether there is a significant difference between two related variables. For example, you may be interested in whether there is a significant difference in anxiety levels between participants before and after an intervention. To conduct a paired samples t-test in R, we can use the t.test() function.

Let's use a hypothetical example to illustrate how to conduct a paired samples t-test in R. In this example, we have data from 30 participants who completed a pre-test and post-test on anxiety levels measured on a 10-point scale. We want to determine if there was a significant difference in anxiety levels before and after an intervention.

```{r}
# Generate example data
pre_test <- rnorm(30, mean = 6, sd = 1)
post_test <- rnorm(30, mean = 4, sd = 1)
```

```{r}
# Conduct paired samples t-test
t.test(pre_test, post_test, paired = TRUE)
```

The t.test() function returns the t-value, degrees of freedom, and p-value for the paired samples t-test. In this example, the p-value is less than 0.05, indicating that there is a significant difference in anxiety levels before and after the intervention.

### Simple Regression on a Difference Score
A simple regression on a difference score is used to determine the relationship between the difference scores of two related variables and a third variable. For example, you may be interested in whether there is a relationship between the difference in anxiety levels before and after an intervention and a participant's age. To conduct a simple regression on a difference score in R, we can use the lm() function.

Let's continue with the hypothetical example from the paired samples t-test. We want to determine if there is a relationship between the difference in anxiety levels before and after the intervention and the participants' age. We will create a new variable diff_score to represent the difference in anxiety levels.

```{r}
# Calculate difference score
diff_score <- post_test - pre_test

# Generate example age data
age <- rnorm(30, mean = 35, sd = 5)

# Conduct simple regression on difference score and age
model <- lm(diff_score ~ age)
summary(model)
```

The lm() function returns the intercept, slope, and p-value for the simple regression on the difference score. In this example, the p-value is greater than 0.05, indicating that there is no significant relationship between the difference in anxiety levels and age.

#### Interpreting the Results
For the paired samples t-test, the output from the t.test() function provides the t-value, degrees of freedom, and p-value. The t-value represents the size of the difference between the means of the two related variables relative to the variation within the data. The degrees of freedom represent the number of observations minus the number of variables being compared. The p-value represents the probability of obtaining a result as extreme as the one observed, assuming that the null hypothesis is true.

For the simple regression on the difference score, the output from the lm() function provides the intercept, slope,

### 2023-04-19 2_Tutorials_ResidualizedChangeScores.qmd
rstudio 9

---
title: "Residualized Change Scores"
author: "Biostatistics Working Group"
---

## Overview
Residualized change scores are a statistical method used to assess the degree of change in a variable while controlling for its initial level. The approach involves calculating the residualized change scores by regressing post-treatment scores on pre-treatment scores and using the resulting residuals as the measure of change. This estimated change score adjusts for baseline scores while ignoring any prior group assignments/differences. The residualized change score is often included in subsequent analysis, such as an examination of intervention effects in pre-test/post-test designs [@Kisbu-Sakarya2013]. This method is useful when there is a high correlation between the pre-treatment and post-treatment scores, which can make it difficult to determine the true degree of change. Research studies have shown that this method can provide more accurate and reliable results compared to other methods of measuring change, such as raw change scores or gain scores (Kenny, 1979; Rogosa & Willett, 1985).

Residualized change scores are a method used to analyze change in a variable over time, while accounting for the initial level of that variable. By using residualized change scores, we can examine the extent to which changes in one variable are associated with changes in another variable, controlling for the initial level of the variables.

![Residualized Change Score 1_Tutorial](./figures/Residualized Change Score.png)

xxxxxx
### When to use Residualized Change Score Models?
You should use residualized change scores in the following scenario:
1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Residualized Change Score Models
In this tutorial, we will introduce the concept of residualized change scores and guide you through a simple example using a small dataset. By the end of this tutorial, you will be able to:

1. Understand the basic concepts of residualized change scores.
2. Calculate residualized change scores using example data.
3. Interpret the results of the residualized change scores analysis.

## Basic Example
For this tutorial, we will use a simple dataset containing the following information for a group of individuals:

- Time point (T1, T2)
- Job satisfaction (independent variable)
- Life satisfaction (dependent variable)

| Individual | Time Point | Job Satisfaction | Life Satisfaction |
| ---------- | ---------- | ---------------- | ----------------- |
| A          | T1         | 7                | 6                 |
| A          | T2         | 8                | 7                 |
| B          | T1         | 6                | 5                 |
| B          | T2         | 7                | 6                 |

### Model Specification and Estimation

To calculate residualized change scores, we will follow these steps:

1. Calculate the raw change scores for each variable by subtracting the T1 value from the T2 value.
2. Fit a regression model with the change score of the dependent variable (life satisfaction) as the outcome and the change score of the independent variable (job satisfaction) and the initial level of the dependent variable (T1 life satisfaction) as predictors.
3. Extract the residuals from the regression model. These residuals represent the residualized change scores.

Using the example dataset, we can calculate the residualized change scores for life satisfaction as follows:

1. Calculate raw change scores:

   | Individual | Job Satisfaction Change | Life Satisfaction Change |
   | ---------- | ----------------------- | ------------------------ |
   | A          | 1                       | 1                        |
   | B          | 1                       | 1                        |

2. Fit the regression model:

   Life Satisfaction Change = b0 + b1 * Job Satisfaction Change + b2 * T1 Life Satisfaction

3. Extract residuals:

   | Individual | Residualized Change Score |
   | ---------- | ------------------------- |
   | A          | e1                        |
   | B          | e2                        |

### Interpreting the Results

Residualized change scores represent the change in the dependent variable (life satisfaction) after accounting for the initial level of that variable and the change in the independent variable (job satisfaction). By examining the association between the residualized change scores of two variables, we can gain insights into how changes in one variable are related to changes in another variable, controlling for initial levels.

### 2023-04-19 3_Tutorials_LinearMixedModels.qmd
rstudio 9

---
title: "Linear Mixed Models"
author: "Biostatistics Working Group"
---

## Overview
Linear mixed models (LMMs) are a powerful statistical tool that allows the analysis of complex data structures that contain both fixed and random effects (West et al., 2015). LMMs are widely used in various fields, including social sciences, biology, and engineering, due to their ability to handle hierarchical data structures and account for within-subject correlations (Bates et al., 2015). LMMs are an extension of the general linear model (GLM), where both fixed and random effects can be included in the model, making them more flexible and robust (Pinheiro & Bates, 2000). These models are particularly useful when analyzing longitudinal data, where measurements are taken repeatedly over time, and correlations between observations must be accounted for (Singer & Willett, 2003). Specifically, the LMM framework accounts for these dependencies among data by extending the general regression "fixed effects" model to allow both, fixed and random effects. This approach simultaneously models an overall sample mean trajectory (fixed effect) and subject-specific (random) effects that vary randomly about the sample mean trajectory. It is this "mixture" of fixed and random effects from which these models derive their name.

[+add primary diagram/figure]

### When to use Longitudinal Linear Mixed Models (LMMs)?
You should use LMMs in the following scenario:
1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Longitudinal Linear Mixed Models

In this tutorial, we will introduce the concept of longitudinal linear mixed models (LLMMs) and guide you through a simple example using a small dataset. By the end of this tutorial, you will be able to:

1. Understand the basic concepts of longitudinal linear mixed models.
2. Fit a longitudinal linear mixed model using example data in R.
3. Interpret the results of the LLMM analysis.

## Basic Example

For this tutorial, we will use a simple dataset containing the following information for a group of individuals:

- Subject ID
- Time (in years)
- Outcome variable (e.g., a measure of cognitive performance)

| Subject ID | Time | Outcome |
|------------|------|---------|
| 1          | 0    | 100     |
| 1          | 1    | 105     |
| 1          | 2    | 110     |
| 2          | 0    | 95      |
| 2          | 1    | 100     |

### Model Specification and Estimation

We will use the `lme4` package in R to fit a longitudinal linear mixed model. First, install and load the required package:

```{r}
if (!("lme4" %in% installed.packages())) install.packages("lme4")
library(lme4)
```

Next, create a data frame with the example data:

```{r}
data <- data.frame(
  subject_id = c(1, 1, 1, 2, 2),
  time = c(0, 1, 2, 0, 1),
  outcome = c(100, 105, 110, 95, 100)
)
```

Fit the LLMM with a random intercept for each subject and a fixed effect of time:

```{r}
model <- lmer(outcome ~ time + (1 | subject_id), data = data)
```

### Interpreting the Results
To interpret the results of the LLMM analysis, we can use the summary() function in R to display the estimated fixed effects and random effects:

```{r}
summary(model)
```

The output will show the estimated fixed effect of time, which represents the average change in the outcome variable over time across subjects. The random effects estimates show the variation in the intercept (initial level) of the outcome variable across subjects. By examining the fixed and random effects, we can gain insights into how the outcome variable changes over time and how this change differs between subjects.

### 2023-04-19 4_Tutorials_SignedRankTest.qmd
rstudio 9

---
title: "Signed-Rank Test"
author: "Biostatistics Working Group"
---

## Overview
The longitudinal signed-rank test is a nonparametric alternative to the paired t-test and is appropriate when the data are not normally distributed or when the assumptions of the paired t-test are violated. The test is based on the signed-rank of the differences between the paired observations, and it tests the null hypothesis that the median of the differences is zero. The repeated-measures design allows for the assessment of within-subject changes over time or under different conditions. The signed-rank test is robust to outliers and does not assume a normal distribution of the differences. This method has been shown to have good statistical power and efficiency, particularly when the sample size is small or the distribution is heavily skewed (Erceg-Hurn & Mirosevich, 2008; Bakdash & Marusich, 2017; Garcia-Berthou & Alcaraz, 2004). The signed-rank test is a useful tool for analyzing paired data when the assumption of normality is violated or when the data is highly skewed or contains outliers.

[+add diagrams/figures]

### When to use a Signed-Rank Test?
You should use a Wilcoxon Signed-Rank Test in the following scenario:
1.	You want to know if two groups are different on your variable of interest
2.	Your variable of interest is continuous
3.	You have two and only two groups
4.	You have independent samples
5.	You have a skewed variable of interest

### Getting Started with Longitudinal Signed-Rank Test

In this tutorial, we will introduce the concept of the longitudinal signed-rank test and guide you through a simple example using a small dataset. By the end of this tutorial, you will be able to:

1. Understand the basic concepts of the longitudinal signed-rank test.
2. Perform a longitudinal signed-rank test using example data in R.
3. Interpret the results of the test.

## Basic Example

For this tutorial, we will use a simple dataset containing the following information for a group of individuals:

- Subject ID
- Time point 1 (T1) measurement
- Time point 2 (T2) measurement

| Subject ID | T1  | T2  |
|------------|-----|-----|
| 1          | 100 | 105 |
| 2          | 95  | 100 |
| 3          | 110 | 112 |

### Model Specification and Estimation

We will use the `wilcox.test()` function in R to perform a longitudinal signed-rank test. First, create a data frame with the example data:

```{r}
data <- data.frame(
  subject_id = c(1, 2, 3),
  t1 = c(100, 95, 110),
  t2 = c(105, 100, 112)
)
```

Perform the longitudinal signed-rank test using the wilcox.test() function with the paired = TRUE argument:

```{r}
test_result <- wilcox.test(data$t1, data$t2, paired = TRUE)
```

### Interpreting the Results
To interpret the results of the longitudinal signed-rank test, examine the p-value from the test_result object:

```{r}
test_result$p.value
```

The p-value represents the probability of observing the data if there is no difference between the measurements at T1 and T2. If the p-value is less than a predetermined significance level (e.g., 0.05), we can reject the null hypothesis that there is no difference between the two time points and conclude that there is a significant difference in the measurements between T1 and T2.

### 2023-04-19 5_Tutorials_MarginalModels.qmd
rstudio 9

---
title: "Marginal Models"
author: "Biostatistics Working Group"
---

## Overview
Marginal models are a statistical method used to analyze longitudinal or clustered data. The marginal model estimates the average effect of the independent variables on the outcome while accounting for the within-subject correlation, and allows for the estimation of population-averaged effects, in contrast to subject-specific effects estimated by random-effects models (Verbeke & Molenberghs, 2000). The method is similar to the GEE approach but uses a different estimation technique and does not account for subject-specific effects. Marginal models have been shown to be robust to non-normality and non-constant variance, and can handle unbalanced or unequally spaced data (Fitzmaurice et al., 2011). The term marginal in this context is used to emphasize that the model for the mean response at each occasion depends only on the covariates of interest, and not on any random effects or previous responses.

[+add diagrams/figures]

### When to use Longitudinal Marginal Models
You should use xxxxxxx in the following scenario:
1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Longitudinal Marginal Models

In this tutorial, we will introduce the concept of longitudinal marginal models and guide you through a simple example using a small dataset. By the end of this tutorial, you will be able to:

1. Understand the basic concepts of longitudinal marginal models.
2. Fit a longitudinal marginal model using example data in R.
3. Interpret the results of the model.

## Basic Example

For this tutorial, we will use a simple dataset containing the following information for a group of individuals:

- Subject ID
- Time (in years)
- Binary outcome variable (e.g., presence or absence of a particular condition)

| Subject ID | Time | Outcome |
|------------|------|---------|
| 1          | 0    | 0       |
| 1          | 1    | 1       |
| 1          | 2    | 1       |
| 2          | 0    | 0       |
| 2          | 1    | 1       |

### Model Specification and Estimation

We will use the `geepack` package in R to fit a longitudinal marginal model. First, install and load the required package:

```{r}
if (!("lme4" %in% installed.packages())) install.packages("geepack")
library(geepack)
```

Next, create a data frame with the example data:

```{r}
data <- data.frame(
  subject_id = c(1, 1, 1, 2, 2),
  time = c(0, 1, 2, 0, 1),
  outcome = c(0, 1, 1, 0, 1)
)
```

Fit the longitudinal marginal model with a binary outcome using a logit link function:

```{r}
model <- geeglm(outcome ~ time, data = data, id = subject_id, family = binomial(link = "logit"), corstr = "exchangeable")
```

### Interpreting the Results
To interpret the results of the longitudinal marginal model, we can use the summary() function in R to display the estimated population-averaged effects:

```{r}
summary(model)
```

The output will show the estimated effect of time on the binary outcome variable, expressed as an odds ratio. By examining the odds ratio and its associated p-value, we can determine if the change in the outcome variable over time is significant at the population level.

### 2023-04-19 6_Tutorials_GeneralizedEstimatingEquations.qmd
rstudio 9

---
title: "Generalized Estimating Equations"
author: "Biostatistics Working Group"
---

[*add diagrams/figures]

## Overview
Generalized estimating equations (GEEs) are a statistical method used to analyze longitudinal or clustered data, while accounting for within-subject correlation. The goal GEEs are to make inferences about population-averaged effects (controlling for within-subject correlations), rather than individual subject-level effects. GEEs can handle a wide range of outcome distributions, including binary, count, and continuous data (Fitzmaurice et al., 2011; Hardin & Hilbe, 2012). This method is an extension of the generalized linear model (Diggle et al., 2002), shows good statistical power and efficiency, is robust to non-normality and non-constant variance and can handle unbalanced or unequally spaced data.

### When to use Longitudinal Generalized Estimating Equations (GEEs)?
You should use xxxxxxx in the following scenario:
1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Longitudinal Generalized Estimating Equations (GEEs)

In this tutorial, we will introduce the concept of longitudinal generalized estimating equations (GEEs) and guide you through a simple example using a small dataset. By the end of this tutorial, you will be able to:

1. Understand the basic concepts of longitudinal GEEs.
2. Fit a GEE model using example data in R.
3. Interpret the results of the model.

### Basic Example

For this tutorial, we will use a simple dataset containing the following information for a group of individuals:

- Subject ID
- Time (in years)
- Binary outcome variable (e.g., presence or absence of a particular condition)

| Subject ID | Time | Outcome |
|------------|------|---------|
| 1          | 0    | 0       |
| 1          | 1    | 1       |
| 1          | 2    | 1       |
| 2          | 0    | 0       |
| 2          | 1    | 1       |

### Model Specification and Estimation

We will use the `geepack` package in R to fit a longitudinal GEE model. First, install and load the required package:

```{r}
if (!("geepack" %in% installed.packages())) install.packages("geepack")
library(geepack)
```
Next, create a data frame with the example data:

```{r}
data <- data.frame(
  subject_id = c(1, 1, 1, 2, 2),
  time = c(0, 1, 2, 0, 1),
  outcome = c(0, 1, 1, 0, 1)
)
```

Fit the longitudinal GEE model with a binary outcome using a logit link function and an exchangeable correlation structure:

```{r}
model <- geeglm(outcome ~ time, data = data, id = subject_id, family = binomial(link = "logit"), corstr = "exchangeable")
```

### Interpreting the Results
To interpret the results of the longitudinal GEE model, we can use the summary() function in R to display the estimated population-averaged effects:

```{r}
summary(model)
```
The output will show the estimated effect of time on the binary outcome variable, expressed as an odds ratio. By examining the odds ratio and its associated p-value, we can determine if the change in the outcome variable over time is significant at the population level.

### 2023-04-19 7_Tutorials_GeneralizedLinearMixedEffectsModels.qmd
rstudio 9

---
title: "Generalized Linear Mixed Effects Models"
author: "Biostatistics Working Group"
---

## Overview
Generalized linear mixed-effects models (GLMMs) are a statistical method used to analyze longitudinal or clustered data that accounts for within-subject correlation and allows for the estimation of subject-specific effects (Pinheiro & Bates, 2000; Fitzmaurice et al., 2011). This method extends the generalized linear mixed-effects model by modeling both fixed-effects (population-averaged effects) and random-effects (subject-specific deviations), while accounting for within-subject correlations. The method can handle unbalanced or unequally spaced data, and can accommodate various outcome distributions, including binary, count, and continuous data. GLMMs have been shown to have good statistical power and efficiency, and can be used to model complex data structures, including crossed and nested random effects. GLMMs have been shown to outperform other methods, such as GEE, in terms of statistical power and efficiency (Hardin & Hilbe, 2012).

[+add diagrams/figures]

### When to use Longitudinal Generalized Linear Mixed Effects Models (GLMMs) Models?
You should use xxxxxxx in the following scenario:
1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Longitudinal Generalized Linear Mixed Effects Models (GLMMs)

In this tutorial, we will introduce the concept of longitudinal generalized linear mixed effects models (GLMMs) and guide you through a simple example using a small dataset. By the end of this tutorial, you will be able to:

1. Understand the basic concepts of longitudinal GLMMs.
2. Fit a GLMM using example data in R.
3. Interpret the results of the model.

## Basic Example

For this tutorial, we will use a simple dataset containing the following information for a group of individuals:

- Subject ID
- Time (in years)
- Binary outcome variable (e.g., presence or absence of a particular condition)

| Subject ID | Time | Outcome |
|------------|------|---------|
| 1          | 0    | 0       |
| 1          | 1    | 1       |
| 1          | 2    | 1       |
| 2          | 0    | 0       |
| 2          | 1    | 1       |

### Model Specification and Estimation

We will use the `lme4` package in R to fit a longitudinal GLMM. First, install the required package, if not already installed:

```{r}
if (!("lme4" %in% installed.packages())) install.packages("lme4")
library(lme4)
```

Next, create a data frame with the example data:

```{r}
data <- data.frame(
  subject_id = factor(c(1, 1, 1, 2, 2)),
  time = c(0, 1, 2, 0, 1),
  outcome = c(0, 1, 1, 0, 1)
)
```
Fit the longitudinal GLMM with a binary outcome using a logit link function and random intercepts for each subject:

```{r}
model <- glmer(outcome ~ time + (1 | subject_id), data = data, family = binomial(link = "logit"))
```

### Interpreting the Results
To interpret the results of the longitudinal GLMM, we can use the summary() function in R to display the estimated fixed effects and random effects:

```{r}
summary(model)
```

The output will show the estimated fixed effects of time on the binary outcome variable, expressed as an odds ratio. By examining the odds ratio and its associated p-value, we can determine if the change in the outcome variable over time is significant at the population level. Additionally, the output will display information about the random effects, such as the variance of the random intercepts for subjects.

### 2023-04-19 8_Tutorials_AutoregressiveCrosslaggedPanelModels.qmd
rstudio 9

---
title: "Autoregressive Crosslagged Panel Models"
---

### 2023-04-20 10_Tutorials_LatentGrowthCurveModels.qmd
tutorials update 1

---
title: "Latent Growth Curve Models"
author: "Biostatistics Working Group"
---

## Overview
In this tutorial, we'll introduce Latent Growth Curve Models (LGCM), a powerful technique for analyzing longitudinal data. Latent Growth Curve Models allow researchers to examine individual differences in growth trajectories and the factors that might influence these trajectories. We'll walk through an example using a dataset with three time points and a sample size of 250 participants.

[+add diagrams/figures]

### When to use Latent Growth Curve Models?
You should use xxxxxxx in the following scenario:

1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Latent Growth Curve Models

#### Prerequisites
To follow along with this tutorial, you'll need to have R installed on your computer, as well as the "lavaan" package for fitting latent variable models.

You can install the "lavaan" package using the following code:

```{r}
if (!("lavaan" %in% installed.packages())) {
  install.packages("lavaan")
}
```

## Basic Example 
For this tutorial, we'll use a hypothetical dataset called "data" with the following variables:

subject_id: A unique identifier for each participant
score_t1: The measurement of a variable of interest at Time 1
score_t2: The measurement of the same variable at Time 2
score_t3: The measurement of the same variable at Time 3
First, let's load the "lavaan" package and create a dataset for this example:

```{r}
library(lavaan)

# Create a sample dataset with 250 participants
set.seed(42) # For reproducibility
subject_id <- 1:250
score_t1 <- rnorm(250, mean = 20, sd = 5)
score_t2 <- score_t1 + rnorm(250, mean = 2, sd = 3)
score_t3 <- score_t2 + rnorm(250, mean = 3, sd = 4)
data <- data.frame(subject_id, score_t1, score_t2, score_t3)
```

### Model Specification and Estimation
Now, let's specify the Latent Growth Curve Model. In this example, we'll estimate the intercept (initial level) and slope (rate of change) for the variable of interest, as well as the variances of the intercept and slope.

```{r}
# Define the Latent Growth Curve Model
model <- '
  # Latent variables
    i =~ 1 * score_t1 + 1 * score_t2 + 1 * score_t3
    s =~ 0 * score_t1 + 1 * score_t2 + 2 * score_t3

  # Means
    i ~ mu_i
    s ~ mu_s

  # Variances
    i ~~ var_i * i
    s ~~ var_s * s

  # Covariance
    i ~~ cov_is * s
'
```

```{r, eval = FALSE}
# Fit the model
fit <- sem(model, data = data, missing = "FIML")
```

### Interpreting the Results
We can now examine the results of our Latent Growth Curve Model. The main parameters of interest are:

mu_i: The mean intercept (initial level) of the variable of interest
mu_s: The mean slope (rate of change) of the variable of interest
var_i: The variance of the intercept, which reflects individual differences in initial levels
var_s: The variance of the slope, which reflects individual differences in rates

### 2023-04-20 11_Tutorials_MultivariateLatentGrowthCurveModels.qmd
tutorials update 1

@@
 title: "Multivariate Latent Growth Curves"
+author: "Biostatistics Working Group"
 ---
+
+## Overview
+In this tutorial, we will introduce the Multivariate Latent Growth Curve Model (MLGCM), also known as Parallel Process Latent Growth Curve Model.
+These models allow us to analyze multiple growth processes simultaneously, examining the relationships between the initial levels and growth rates of multiple variables over time.
+
+We will use the "lavaan" package in R to fit a MLGCM with two variables, each measured at three timepoints. We will also provide a brief overview of interpreting the results.
+
+[+add diagrams/figures]
+
+### When to use Multivariate Latent Growth Curve Models?
+You should use xxxxxxx in the following scenario:
+
+1.	You want to know xxxxx
+2.	Your variable xxxxx
+3.	You have xxxxx
+
+### Getting Started with Multivariate Latent Growth Curve Models
+
+## Basic Example
+```{r}
+# Load necessary packages
+if (!("lavaan" %in% installed.packages())) install.packages("lavaan")
+library(lavaan)
+
+# Generate sample data
+set.seed(42)
+n <- 250
+time1_var1 <- rnorm(n, mean = 10, sd = 3)
+time2_var1 <- time1_var1 + rnorm(n, mean = 2, sd = 2)
+time3_var1 <- time2_var1 + rnorm(n, mean = 2, sd = 2)
+
+time1_var2 <- rnorm(n, mean = 5, sd = 2)
+time2_var2 <- time1_var2 + rnorm(n, mean = 1, sd = 1)
+time3_var2 <- time2_var2 + rnorm(n, mean = 1, sd = 1)
+
+# Combine into a data frame
+data <- data.frame(time1_var1, time2_var1, time3_var1,
+                   time1_var2, time2_var2, time3_var2)
+```
+
+### Model Specification and Estimation
+```{r}
+# Specify the multivariate latent growth curve model
+model <- '
+  # Intercept and slope factors for variable 1
+  i_var1 =~ 1*time1_var1 + 1*time2_var1 + 1*time3_var1
+  s_var1 =~ 0*time1_var1 + 1*time2_var1 + 2*time3_var1
+
+  # Intercept and slope factors for variable 2
+  i_var2 =~ 1*time1_var2 + 1*time2_var2 + 1*time3_var2
+  s_var2 =~ 0*time1_var2 + 1*time2_var2 + 2*time3_var2
+'
+```
+
+```{r}
+# Estimate the model
+fit <- lavaan::sem(model, data = data)
+```
+
+```{r}
+summary(fit, fit.measures = TRUE)
+```
+
+### Interpreting the Results
+In the output, you will see the estimates for the factor loadings, intercepts, and slopes for both variables.
+These estimates describe the initial levels and growth rates for each variable.
+You will also see the variances and covariances of the latent intercept and slope factors, which provide information about individual differences in initial levels and growth rates, as well as the relationships between the initial levels and growth rates across the two variables.
+
+Pay attention to the model fit indices (e.g., CFI, TLI, RMSEA, and SRMR) to evaluate how well the model fits the data.
+Good model fit is indicated by CFI and TLI values close to or greater than 0.95, RMSEA values close to or smaller than 0.06, and SRMR values close to or smaller than xxx

### 2023-04-20 12_Tutorials_LatentTransitionAnalysis.qmd
tutorials update 1

---
title: "Latent Transition Analysis"
author: "Biostatistics Working Group"
---

## Overview
In this tutorial, we will introduce Latent Transition Analysis (LTA), a statistical method used to analyze the movement of individuals between latent (unobserved) statuses over time. LTA is particularly useful when dealing with categorical data and allows us to estimate the probabilities of transitioning between latent statuses.

We will use the "tidyLPA" package in R to fit a simple LTA model with four timepoints and three indicators per timepoint. We will also provide a brief overview of interpreting the results.

[+add diagrams/figures]

### When to use Latent Transition Analysis?
You should use xxxxxxx in the following scenario:

1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Latent Transition Analysis

## Basic Example

```{r}
# Load necessary packages
if (!("tidyLPA" %in% installed.packages())) install.packages("tidyLPA")
library(tidyLPA)
```

```{r}
set.seed(42)
n <- 250

# Generate three indicators for time 1
time1_ind1 <- rbinom(n, size = 1, prob = 0.7)
time1_ind2 <- rbinom(n, size = 1, prob = 0.6)
time1_ind3 <- rbinom(n, size = 1, prob = 0.5)

# Generate three indicators for time 2
time2_ind1 <- rbinom(n, size = 1, prob = 0.6)
time2_ind2 <- rbinom(n, size = 1, prob = 0.5)
time2_ind3 <- rbinom(n, size = 1, prob = 0.4)

# Generate three indicators for time 3
time3_ind1 <- rbinom(n, size = 1, prob = 0.5)
time3_ind2 <- rbinom(n, size = 1, prob = 0.4)
time3_ind3 <- rbinom(n, size = 1, prob = 0.3)

# Generate three indicators for time 4
time4_ind1 <- rbinom(n, size = 1, prob = 0.4)
time4_ind2 <- rbinom(n, size = 1, prob = 0.3)
time4_ind3 <- rbinom(n, size = 1, prob = 0.2)
```

# Combine into a data frame
```{r}
data <- data.frame(time1_ind1, time1_ind2, time1_ind3,
                   time2_ind1, time2_ind2, time2_ind3,
                   time3_ind1, time3_ind2, time3_ind3,
                   time4_ind1, time4_ind2, time4_ind3)
```

### Model Specification and Estimation
```{r}
# Estimate the LTA model
fit <- tidyLPA::estimate_profiles(data, 2, variances = "equal", covariances = "zero")
```

### Interpreting the Results
In the output, you will find the estimated class probabilities for each timepoint. These probabilities represent the proportion of individuals belonging to each latent status at each timepoint. You will also see the estimated transition probabilities, which represent the probabilities of transitioning between latent statuses from one timepoint to the next.

You can visualize the results using the plot() function to better understand the transitions between latent statuses:

```{r, eval = FALSE}
plot(fit)
```

Remember to xxxxx

### 2023-04-20 13_Tutorials_GrowthMixtureModels.qmd
tutorials update 1

---
title: "Growth Mixture Models"
author: "Biostatistics Working Group"
---

## Overview
Growth Mixture Models (GMMs) are a type of statistical model that aims to identify distinct subgroups or classes within a population based on their growth trajectories. In this tutorial, we will walk you through a simple example using a simulated dataset and demonstrate how to fit a Growth Mixture Model using the `lcmm` package in R.

[+add diagrams/figures]

### When to use Growth Mixture Models (GMMs)?
You should use xxxxxxx in the following scenario:

1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Growth Mixture Models (GMMs)

#### Prerequisites
To follow along with this tutorial, you will need to have R and RStudio installed on your computer. Additionally, you will need to install the `lcmm` package, which provides functions for fitting Growth Mixture Models.

## Basic Example 
```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
# Install the lcmm package if not already installed
if (!("lcmm" %in% installed.packages())) {
  install.packages("lcmm")
}
```

Load the lcmm package
```{r}
library(lcmm)
```

Simulating the Data
For this tutorial, we will use a simulated dataset with three growth trajectory classes, where each class represents a different pattern of change over time. The dataset will consist of 300 individuals with four repeated measurements (time points) for each individual.

```{r}
set.seed(123)
n <- 300
timepoints <- 4

# Simulate the data for each class
class1 <- data.frame(id = 1:(n / 3),
                     time = rep(1:timepoints, each = n / 3),
                     class = 1,
                     value = rnorm(n * timepoints / 3, mean = 1 + 0.5 * (1:timepoints), sd = 0.5))

class2 <- data.frame(id = (n / 3 + 1):(2 * n / 3),
                     time = rep(1:timepoints, each = n / 3),
                     class = 2,
                     value = rnorm(n * timepoints / 3, mean = 2 - 0.3 * (1:timepoints), sd = 0.5))

class3 <- data.frame(id = (2 * n / 3 + 1):n,
                     time = rep(1:timepoints, each = n / 3),
                     class = 3,
                     value = rnorm(n * timepoints / 3, mean = 3 + 0.1 * (1:timepoints), sd = 0.5))
```

```{r}
# Combine the data from all classes
data <- rbind(class1, class2, class3)
```

### Model Specification and Estimation
Now that we have our simulated data, we can fit a Growth Mixture Model. In this example, we will assume that there are three latent classes. However, in practice, the number of classes is usually unknown and must be determined through model comparisons or other methods.
```{r, eval=FALSE}
# Fit the Growth Mixture Model
gmm <- hlme(value ~ time,
            random = ~ time,
            subject = "id",
            mixture = ~ time,
            data = data,
            ng = 3,
            B = list(value = c(1, 0, 0, 0)
```

### Interpreting the Results
xxxxxx

### 2023-04-20 14_Tutorials_StateTraitModels.qmd
tutorials update 1

---
title: "State-Trait Models"
author: "Biostatistics Working Group"
---

## Overview
State-Trait Models are a type of statistical model used to separate the variance in a given variable into two components: a stable, trait-like component and a time-specific, state-like component. In this tutorial, we will walk you through a simple example using a simulated dataset and demonstrate how to fit a State-Trait Model using the `lavaan` package in R.

[+add diagrams/figures]

### When to use State-Trait Models?
You should use xxxxxxx in the following scenario:

1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with State-Trait Models

#### Prerequisites
To follow along with this tutorial, you will need to have R and RStudio installed on your computer. Additionally, you will need to install the `lavaan` package, which provides functions for fitting various structural equation models, including State-Trait Models.

## Basic Example 
```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
# Install the lavaan package if not already installed
if (!("lavaan" %in% installed.packages())) {
  install.packages("lavaan")
}
```

# Load the lavaan package
```{r}
library(lavaan)
```

## Simulating the Data
For this tutorial, we will use a simulated dataset with 300 individuals, each measured at four time points. The dataset will contain a variable with both state-like and trait-like components.

```{r}
set.seed(123)
n <- 300
timepoints <- 4

# Simulate the trait-like component
trait <- rnorm(n, mean = 100, sd = 10)

# Simulate the state-like component
state <- matrix(rnorm(n * timepoints, mean = 0, sd = 5), nrow = n, ncol = timepoints)

# Combine trait and state components
data <- data.frame(id = rep(1:n, each = timepoints),
                   time = rep(1:timepoints, times = n),
                   value = c(t(trait + state)))
```

### Model Specification and Estimation
Now that we have our simulated data, we can fit a State-Trait Model using the lavaan package. The model will estimate the trait-like and state-like components for each individual.

```{r}
# Define the model
model <- '
  # Trait component
  trait =~ 1 * value_t1 + 1 * value_t2 + 1 * value_t3 + 1 * value_t4

  # State component
  state =~ value_t1 + value_t2 + value_t3 + value_t4

  # Residual variances
  value_t1 ~~ value_t1
  value_t2 ~~ value_t2
  value_t3 ~~ value_t3
  value_t4 ~~ value_t4
'
```

```{r}
# Reshape the data to wide format
data_wide <- reshape(data, idvar = "id", timevar = "time", direction = "wide")
```

```{r, eval=FALSE}
# Fit the State-Trait Model
fit <- sem(model, data = data_wide)
```

### Interpreting the Results
Now that we have estimated our State-Trait Model, we can interpret the results. The key parameters of interest in this model are the trait-like and state-like components, which are the variances attributed to each component.

```{r, eval=FALSE}
# Obtain the results
summary(fit)
```

### 2023-04-20 15_Tutorials_RandomInterceptCrosslaggedPanelModels.qmd
tutorials update 1

@@
 title: "Random-Intercept Crosslagged Panel Models"
+author: "Biostatistics Working Group"
 ---
+
+## Overview
+Random-Intercept Cross-Lagged Panel Models (RI-CLPM) are a type of structural equation model used to analyze longitudinal data with repeated measures of multiple variables.
+In this tutorial, we will walk you through a simple example using a simulated dataset and demonstrate how to fit a RI-CLPM using the `lavaan` package in R.
+
+[+add diagrams/figures]
+
+### When to use Random-Intercept Cross-Lagged Panel Models?
+You should use xxxxxxx in the following scenario:
+
+1.	You want to know xxxxx
+2.	Your variable xxxxx
+3.	You have xxxxx
+
+### Getting Started with Random-Intercept Cross-Lagged Panel Models
+
+#### Prerequisites
+To follow along with this tutorial, you will need to have R and RStudio installed on your computer.
+Additionally, you will need to install the `lavaan` package, which provides functions for fitting various structural equation models, including RI-CLPMs.
+
+```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
+# Install the lavaan package if not already installed
+if (!("lavaan" %in% installed.packages())) {
+  install.packages("lavaan")
+}
+# Load the lavaan package
+library(lavaan)
+```
+
+## Basic Example 
+Simulating the Data
+For this tutorial, we will use a simulated dataset with 500 individuals, each measured at five time points. The dataset will contain two variables, X and Y, with cross-lagged effects.
+
+```{r}
+set.seed(123)
+n <- 500
+timepoints <- 5
+
+# Simulate the data
+data <- data.frame(id = rep(1:n, each = timepoints),
+                   time = rep(1:timepoints, times = n),
+                   X = rnorm(n * timepoints, mean = 0, sd = 1),
+                   Y = rnorm(n * timepoints, mean = 0, sd = 1))
+
+# Introduce cross-lagged effects
+for (i in 2:timepoints) {
+  data$X[data$time == i] <- 0.3 * data$Y[data$time == (i - 1)] + data$X[data$time == i]
+  data$Y[data$time == i] <- 0.2 * data$X[data$time == (i - 1)] + data$Y[data$time == i]
+}
+```
+
+### Model Specification and Estimation
+Now that we have our simulated data, we can fit a Random-Intercept Cross-Lagged Panel Model using the lavaan package.
+
+```{r, eval=FALSE}
+# Define the model
+model <- '
+  # Random intercepts
+  i_X =~ 1 * X_t1 + 1 * X_t2 + 1 * X_t3 + 1 * X_t4 + 1 * X_t5
+  i_Y =~ 1 * Y_t1 + 1 * Y_t2 + 1 * Y_t3 + 1 * Y_t4 + 1 * Y_t5
+
+  # Cross-lagged effects
+  Y_t2 ~ beta_YX * X_t1
+  Y_t3 ~ beta_YX * X_t2
+  Y_t4 ~ beta_YX * X_t3
+  Y_t5 ~ beta_YX * X_t4
+
+  X_t2 ~ beta_XY * Y_t1
+  X_t3 ~ beta_XY *
+```
+
+### Interpreting the Results
+xxxxxx

### 2023-04-20 16_Tutorials_LatentCurveModelsStructuredResiduals.qmd
tutorials update 1

@@
 title: "Latent Curve Models with Structured Residuals"
+author: "Biostatistics Working Group"
 ---
+
+## Overview
+In this tutorial, we will explore Latent Curve Models with Structured Residuals (LCM-SR).
+LCM-SR allows for the separation of between-person and within-person components of individual change over time.
+By using structured residuals, we can account for time-varying covariates and interactions between variables at both the within-person and between-person levels.
+
+[+add diagrams/figures]
+
+### When to use Latent Curve Models with Structured Residuals?
+You should use xxxxxxx in the following scenario:
+
+1.	You want to know xxxxx
+2.	Your variable xxxxx
+3.	You have xxxxx
+
+### Getting Started with Latent Curve Models with Structured Residuals
+
+#### Prerequisites
+You need to have the `lavaan` and `semTools` packages installed in R to perform LCM-SR. Install them using the following commands:
+
+```{r}
+if (!("lavaan" %in% installed.packages())) install.packages("lavaan")
+if (!("semTools" %in% installed.packages())) install.packages("semTools")
+```
+
+## Basic Example 
+For this tutorial, we will use simulated data with 500 individuals and 5 timepoints. The data will include a time-varying covariate, covariate, and an outcome variable, outcome.
+
+```{r}
+set.seed(123)
+library(lavaan)
+library(semTools)
+
+# Simulate data
+n <- 500
+timepoints <- 5
+
+id <- rep(1:n, each = timepoints)
+time <- rep(1:timepoints, n)
+
+covariate <- rnorm(n * timepoints, mean = 0, sd = 1)
+outcome <- rnorm(n * timepoints, mean = 0, sd = 1)
+
+data <- data.frame(id, time, covariate, outcome)
+```
+
+### Model Specification and Estimation
+To specify an LCM-SR, we will use the lavaan syntax to define the measurement model and the structural model.
+We will model the linear growth of the outcome variable while accounting for the effect of the time-varying covariate.
+
+```{r}
+model <- '
+  # Latent variables
+  intercept =~ 1 * outcome_t1 + 1 * outcome_t2 + 1 * outcome_t3 + 1 * outcome_t4 + 1 * outcome_t5
+  slope =~ 0 * outcome_t1 + 1 * outcome_t2 + 2 * outcome_t3 + 3 * outcome_t4 + 4 * outcome_t5
+
+  # Time-varying covariate effect on outcome
+  outcome_t1 ~ c1 * covariate_t1
+  outcome_t2 ~ c2 * covariate_t2
+  outcome_t3 ~ c3 * covariate_t3
+  outcome_t4 ~ c4 * covariate_t4
+  outcome_t5 ~ c5 * covariate_t5
+
+  # Residual variances and covariances
+  outcome_t1 ~~ r1 * outcome_t1
+  outcome_t2 ~~ r2 * outcome_t2
+  outcome_t3 ~~ r3 * outcome_t3
+  outcome_t4 ~~ r4 * outcome_t4
+  outcome_t5 ~~ r5 * outcome_t5
+
+  # Latent variable variances
+  intercept ~~ i_var * intercept
+  slope ~~ s_var * slope
+
+  # Latent variable covariances
+  intercept ~~ i_s_cov * slope
+'
+```
+
+Model Estimation
+Now we will estimate the model using the lavaan function sem().
+
+```{r, eval=FALSE}
+# Reshape data to wide format
+wide_data <- spread(data, key = time, value = outcome, sep
+```
+### Interpreting the Results
+xxxxx

### 2023-04-20 8_Tutorials_AutoregressiveCrosslaggedPanelModels.qmd
tutorials update 1

@@
 title: "Autoregressive Crosslagged Panel Models"
+author: "Biostatistics Working Group"
 ---
+
+## Overview
+xxxx
+Autoregressive cross-lagged panel models (ACPMs) are a type of statistical model used to analyze longitudinal data and examine the temporal relationships between two or more variables over time.
+These models help to disentangle within-person effects from between-person effects and can estimate both autoregressive effects (influence of a variable on itself over time) and cross-lagged effects (influence of one variable on another variable over time).
+
+autoregressive cross-lagged panel models, commonly known as CLPMs.
+CLPMs are a statistical method used to analyze longitudinal data, particularly when we want to examine the reciprocal relationships between two or more variables over time.
+In essence, CLPMs allow us to investigate whether one variable at an earlier time point can predict another variable at a later time point, and vice versa.
+
+CLPMs are useful when we want to go beyond simple correlations between variables at different time points and instead model the dynamic interplay between them.
+Specifically, CLPMs account for the possibility that the relationship between two variables at one time point may influence their relationship at a subsequent time point.
+These models can also account for autocorrelation, meaning that the same variable measured over time is correlated with itself.
+
+[+add diagrams/figures]
+
+### When to use Autoregressive Cross-Lagged Panel Models (ACPMs)?
+You should use xxxxxxx in the following scenario:
+
+1.	You want to know xxxxx
+2.	Your variable xxxxx
+3.	You have xxxxx
+
+### Getting Started with Autoregressive Cross-Lagged Panel Models (ACPMs)
+In this tutorial, we will introduce the concept of autoregressive cross-lagged panel models (ACPMs) and guide you through a simple example using a larger dataset.
+By the end of this tutorial, you will be able to:
+
+1. Understand the basic concepts of ACPMs.
+2. Fit an ACPM using example data in R.
+3. Interpret the results of the model.
+
+## Basic Example
+For this tutorial, we will use a simulated dataset containing the following information for a group of 100 individuals:
+
+- Subject ID
+- Time (in years)
+- Variable 1 (e.g., stress level)
+- Variable 2 (e.g., job satisfaction)
+
+```{r}
+set.seed(123)  # For reproducibility
+n_subjects <- 100
+n_timepoints <- 3
+subject_ids <- factor(rep(1:n_subjects, each = n_timepoints))
+time <- rep(0:(n_timepoints - 1), times = n_subjects)
+variable1 <- rnorm(n_subjects * n_timepoints, mean = 3, sd = 1)
+variable2 <- rnorm(n_subjects * n_timepoints, mean = 3, sd = 1)
+
+data <- data.frame(subject_id = subject_ids, time = time, variable1 = variable1, variable2 = variable2)
+```
+
+### Model Specification and Estimation
+We will use the lavaan package in R to fit an ACPM. First, install the required package, if not already installed:
+
+```{r}
+if (!("lavaan" %in% installed.packages())) install.packages("lavaan")
+library(lavaan)
+```
+Define the ACPM with three timepoints:
+
+```{r}
+model <- 
+  # Autoregressive paths
+  variable1_t2 ~ a1 * variable1_t1
+  variable1_t1 ~ a2 * variable1_t0
+  variable2_t2 ~ a3 * variable2_t1
+  variable2_t1 ~ a4 * variable2_t0
+
+  # Cross-lagged paths
+  variable1_t2 ~ b1 * variable2_t1
+  variable1_t1 ~ b2 * variable2_t0
+  variable2_t2 ~ b3 * variable1_t1
+  variable2_t1 ~ b4 * variable1_t0
+
+```
+Fit the ACPM using the lavaan function:
+
+```{r}
+#fit <- lavaan(model, data = data, missing = )
+
+```
+
+### Interpreting the Results
+xxxxxxxx

### 2023-04-20 9_Tutorials_LatentChangeScoresModels.qmd
tutorials update 1

---
title: "Latent Change Score Models"
author: "Biostatistics Working Group"
---

## Overview
xxx
In this tutorial, we'll introduce Latent Change Score Models (LCSM), a powerful technique for analyzing longitudinal data. Latent Change Score Models allow researchers to examine individual differences in change over time, as well as the factors that might influence these changes. We'll walk through an example using a dataset with three time points and a larger sample size of 250 participants.
xxx

[+add diagrams/figures]

### When to use Latent Change Score Models?
You should use xxxxxxx in the following scenario:

1.	You want to know xxxxx
2.	Your variable xxxxx
3.	You have xxxxx

### Getting Started with Latent Change Score Models

#### Prerequisites
To follow along with this tutorial, you'll need to have R installed on your computer, as well as the "lavaan" package for fitting latent variable models.

You can install the "lavaan" package using the following code:

```{r}
if (!("lavaan" %in% installed.packages())) {
  install.packages("lavaan")
}
```

## Basic Example
For this tutorial, we'll use a hypothetical dataset called "data" with the following variables:

subject_id: A unique identifier for each participant
score_t1: The measurement of a variable of interest at Time 1
score_t2: The measurement of the same variable at Time 2
score_t3: The measurement of the same variable at Time 3
First, let's load the "lavaan" package and create a larger dataset for this example:

```{r}
library(lavaan)

# Create a sample dataset with 250 participants
set.seed(42) # For reproducibility
subject_id <- 1:250
score_t1 <- rnorm(250, mean = 20, sd = 5)
score_t2 <- score_t1 + rnorm(250, mean = 2, sd = 3)
score_t3 <- score_t2 + rnorm(250, mean = 3, sd = 4)
data <- data.frame(subject_id, score_t1, score_t2, score_t3)
```

### Model Specification and Estimation
Now, let's specify the Latent Change Score Model. In this example, we'll estimate the mean change in the variable of interest from Time 1 to Time 2 and from Time 2 to Time 3, as well as the variances of the change scores.
```

```{r, eval = FALSE}
# Define the Latent Change Score Model
model <- '
  # Latent variables
    delta12 =~ 1 * score_t2 - 1 * score_t1
    delta23 =~ 1 * score_t3 - 1 * score_t2

  # Means
    delta12 ~ mu_delta12
    delta23 ~ mu_delta23

  # Variances
    delta12 ~~ var_delta12 * delta12
    delta23 ~~ var_delta23 * delta23
`
```

```{r, eval = FALSE}
# Fit the model
fit <- sem(model, data = data, missing = "FIML")
```

### Interpreting the Results
We can now examine the results of our Latent Change Score Model. The main parameters of interest are:

mu_delta12: The mean change in the variable of interest from Time 1 to Time 2
mu_delta23: The mean change in the variable of interest from Time 2 to Time 3
var_delta12: The variance of the change scores from Time 1 to Time 2, which reflects individual differences in change
var_delta23: The variance of the change xxxx

### 2023-04-20 index.qmd
rstudio test commit 3
tutorials update 1

@@
 ---
-title: "swhawes.github.io"
+title: "Temporary Landing Page for Longitudinal Project"
 ---
 
-This is a Quarto website.
-
-To learn more about Quarto websites visit <https://quarto.org/docs/websites>.
+This is github pages test site for the longitudinal analysis project. The information included within the site is for testing purposes only and contents may be inaccurate.

### 2023-04-20 manuscript.qmd
manuscript draft formatting

---
title: Manuscript Working Draft

format: 
  pdf: 
    geometry: 
      - top=20mm
      - left=20mm
  docx: default
prefer-html: true
#custom-reference-doc.docx
---

# Introduction
\label{sec:headings}
The Adolescent Brain Cognitive Development (ABCD) Study is the largest long-term investigation of neurodevelopment and child health in the United States. Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018]. At its heart, the study is designed to chart the course of human development across multiple, interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative developmental outcomes. Central to achieving these goals is the ABCD Study’s® commitment to an open science framework designed to facilitate access to and sharing of scientific knowledge by espousing practices that increase openness, integrity, and reproducibility of scientific research (e.g., public data releases). In this sense, the ABCD Study® is a collaboration with the larger research community, with the rich longitudinal nature of the ABCD Study dataset allowing researchers to perform a variety of analyses of both methodological and substantive interest. Together, this presents a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes emerge and unfold across critical periods of development.

[section still in progress...]

## The ABCD Study® Data
Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians. The study sample was recruited from household populations in defined catchment areas for each of the 21 study sites across the United States (information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at the ABCD Study website). The ABCD Study is collecting longitudinal data on a rich variety of outcomes that will enable the construction of realistically-complex   etiological models by incorporating factors from many domains simultaneously. Each new wave of data collection provides the building blocks for conducting probing longitudinal analyses that allow us to characterize normative development, identify variables that presage deviations from prototypic development, and assess a range of outcomes associated with variables of interest. This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use [@xxxxx], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020]. Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward. A limited number of assessments are collected in semi-annual telephone interviews between in-person visits. Data are publicly released on an annual basis through the NIMH Data Archive. By necessity, the study’s earliest data releases were cross-sectional (i.e., the baseline data), however, the most recent public data release (NDA Release 4.0) contains data collected across three annual assessments, including two imaging assessments (baseline and year 2 follow-up visits).

## Organization of current manuscript
The rich longitudinal nature of the ABCD Study dataset will allow researchers to perform analyses of both methodological and substantive interest. This report describes methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents. The manuscript is organized as follows: xxxxxxxxx.

# Part I: Longitudinal Research: Basic Concepts and Considerations
\label{sec:headings}
There are several important concepts to consider when conducting longitudinal analyses in a developmental context. These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represent the same phenomenon or different ones. Further, in the case of developmentally focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child’s life course (and not just characterize growth during this period but also assesses the durability/chronicity of prior effects/consequences)  but also brings with it greater methodological opportunities to exploit the statistical properties of longitudinal data in the furtherance of critical scientific questions. That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows, assuming we have assessed the “right” variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest. Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.

**Vulnerable periods**. Development normatively progresses from less mature to more mature levels of functioning. However, unique epochs and experiences can alter the course of this idealized form of development. Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [@xxxx]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans)[@xxxx]. These examples highlight the importance of considering the role of vulnerable periods – temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli (e.g. cannabis exposure) on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.

**Developmental disturbances**. Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times environmental pressures will tend to suppress stability and disrupt the orderly stochastic process of normative development (e.g., xxx-xxx). This situation reflects a developmental disturbance in that the normal course of development is “disturbed” for some time by some time-limited process. In such cases, we might find that prediction of behavior in the period of the disturbance is reduced and/or, similarly, the behavior exhibited during the disturbance might have less predictive power with respect to distal outcomes compared to the behavior exhibited before and following the disrupted period. That is, once the environmental stimuli are removed (or the individual is removed from the environment), individual differences are again more freely expressed and the autoregressive effects increase to levels similar to those before entering the environment.

**Developmental snares and cascade effects**. Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next. For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect, creating a “snare” that retards psychosocial development. Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified. Although conceptually distinct from vulnerable periods, both of these types of developmental considerations highlight the importance of viewing behavior in the context of development and the importance of attempting to determine how various developmental pathways unfold.

**Distinguishing developmental change from experience effects**. One can often observe systematic changes over time in a variable of interest and assume this change is attributable to development. To this point, cognitive abilities (e.g, verbal ability, problem-solving) normatively grow earlier in development and often decline in late life (e.g., memory, speed of processing). However, the observed patterns of growth and decline often differ between cross-sectional vs. longitudinal effects [@salthouse2014] where subjects gain increasing experience with the assessment with each successive measurement occasion. Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD [@sullivan2017] and highlight the need to consider these effects and address them analytically. In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to “learning” the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before). Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; see @french2010a] and this phenomenon has been well documented in research on structured diagnostic interviews [@robins1985]. While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artefactually elevated [see @shrout2018a]. Some designs (specifically, accelerated longitudinal designs) are especially well suited for discovering these effects and modeling them. While ABCD was not designed as an accelerated longitudinal design, the variability in age at the time of baseline recruitment (9 years, 0 months to 10 years, 11 months) allows some measures, collected every year, to be conceptualized as an accelerated longitudinal design. Moreover, it is possible that in later waves, patterns of longitudinal missing data will allow some analyses to assess the confounded effects of age and the number of prior assessments. However, ABCD is fundamentally a single-cohort, longitudinal design, where a number of prior assessments and age are highly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.

# Part II Longitudinal Data: Interpretation / Issues / Pitfalls & Assumption
\label{sec:headings}
**Defining Features of Longitudinal Data Analysis.**
The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed. That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the “right” variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .

## Modeling Data Across Two Time Points versus Three or More Time Points.
Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions. Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1). At the same time, data based on two assessments are inherently limited on multiple fronts. As [@rogosa1982] noted approximately forty years ago, “Two waves of data are better than one, but maybe not much better”. These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for model identification and accurate parameter estimation. It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (since linear, straight line change is the only possible form, given two assessments; see [@duncan2009]). Research designs that include three or more time points allow for increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of inferential research. To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level. This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield2021. It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling. Indeed, several contemporary longitudinal models designed to isolate within-individual variability [e.g., the Latent Curve Model with Structured Residuals; @curran2014a] require at least four assessments to parameterize fully and, more generally, increasingly accurate parameter estimates are obtained as more assessment occasions are used [@duncan2009].

## Types of stability and change
If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it’s the characterization of stability and change as the organism traverses the life course. There are a few different ways to think of stability (and change). Consider we measure the height of all youth in a 6th-grade class, once in the fall at the beginning of the school year and once again in the spring at the end of the school year. A common first step may be to compare the class’s average height values obtained at these two different measurement occasions. This comparison of the average scores for the same group of individuals at multiple time points is referred to as “mean-level” stability as it provides information about continuity and change in the group level of an outcome of interest (e.g., height) over time. Another type of stability involves calculating the correlation between the values obtained at different time points (e.g., ‘height in the fall’ with ‘height in the spring’). This type of “rank-order” stability evaluates between-individual change by focusing on the degree to which individuals retain their relative placement in a group across time. Consider, someone who is the shortest person in their class in 6th grade may grow considerably over the school year (i.e., exhibit mean level change), but remain the shortest person among their classmates. That is, the individual is manifesting a type of rank-order stability. Both types of stability and change are important. Mean-level change in certain traits might help to explain why, in general, individuals are particularly vulnerable to social influences at some ages more than others; rank order change might help to quantify the extent to which certain characteristics of the individual are more trait-like. For example, in some areas of development, there is considerable mean-level change that occurs over time (e.g., changes in Big 5 personality traits), but relatively high rank-order stability. Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of “within-individual” change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest.

There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021; xxxxx et al. (xxxx). Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011]. Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures. As a result, experts increasingly caution about the need to xxxxxxxx [@xxxxx]. Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021]. Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a]. It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].

## Model Assumptions
Many statistical models assume certain characteristics about the data to which they are being applied. As an example, common assumptions of parametric statistical models include normality, linearity, and equality of variances. These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model’s assumptions can substantively alter the interpretation of results. Similarly, statistical models employed in the analyses of longitudinal data often entail a range of assumptions that must be closely inspected. One central issue for repeated measurements on an individual is how to account for the correlated nature of the data; another common feature of longitudinal data is heterogeneous variability; that is, the variance of the response changes over the duration of the study. Traditional techniques, such as a standard regression or ANOVA model, assume residuals are independent and thus are inappropriate for designs that assess (for example) the same individuals across time. That is, given the residuals are no longer independent, the standard errors from the models are biased and can produce misleading inferential results. Although there are formal tests of independence for time series data (e.g., the Durbin-Watson statistic; Durbin & Watson, 1950), more commonly independence is assumed to be violated in study designs with repeated assessments. Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model the covariance structure of said data.

## Covariance Structures
Statistical models for longitudinal data include two main components to account for assumptions that are commonly violated when working with repeated measures data: a model for the covariance among repeated measures (both the correlations among pairs of repeated measures on an individual and the variability of the responses on different occasions), coupled with a model for the mean response and its dependence on covariates (eg, treatment group in the context of clinical trials). This allows for the specification of a range of so-called covariance structures, each with its own set of tradeoffs between model fit and parsimony [e.g., see @kincaid2005].

## Accounting for Correlated Data
As an example, one alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design. As the name implies, the structure assumes a subsequent measurement occasion (e.g., assessment at Wave 2) is regressed onto (that is, is predicted by) a prior measurement occasion (e.g., assessment at Wave 1). The most common type of autoregressive design is the AR(1), where assessments at time T + 1 are regressed on assessments at Time T. Identical to compound symmetry, this model assumes the variances are homogenous across time. Diverting from compound symmetry, this model assumes the correlations between repeated assessments decline exponentially across time rather than remaining constant. For example, per the AR(1) structure, if the correlation between Time 1 and Time 2 data is thought to be .5, then the correlation between Time 1 and Time 3 data would be assumed to be .5*.5 = .25, and the correlation between Time 1 and Time 4 data would be assumed to be .5*.5*.5 = .125. As with compound symmetry, the basic AR(1) model is parsimonious in that it only requires two parameters (the variance of the assessments and the autoregressive coefficient). Notably, the assumption of constant autoregressive relations between assessments is often relaxed in commonly employed designs that use autoregressive modeling (e.g., cross-lagged panel models [CLPM]). These designs still typically assume an AR(1) process (e.g., it is sufficient to regress the Time 3 assessment onto the Time 2 assessment and is not necessary to also regress the Time 3 assessment onto the Time 1 assessment, which would result in an AR(2) process). However, the magnitude of these relations is often allowed to differ across different AR(1) pairs of assessment (e.g., the relation between Time 1 and Time 2 can be different from the relation between Time 2 and Time 3). These more commonly employed models also often relax the assumption of equal variances of the repeated assessments. Although the AR(1) structure may involve a more realistic set of assumptions compared to compound symmetry, in that the AR(1) model allows for diminishing correlations across time, the basic AR(1) model, as well as autoregressive models more generally, can also suffer from several limitations in contexts that are common in prospective designs. In particular, recent work demonstrates that if a construct being assessed prospectively across time is trait-like in nature, then autoregressive relations fail to adequately account for this trait-like structure, with the downstream consequence that estimates derived from models based on AR structures (such as the CLPM) can be misleading and fail to adequately demarcate between- vs. within-person sources of variance [@hamaker2015].

## Linear vs non-linear models
Identification of optimal statistical models and appropriate mathematical functions requires an understanding of the type of data being used. Repeated assessments can be based on either continuous or discrete measures. Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., a single item measuring the level of agreement to a statement on a three-point scale including the categories of “disagree”, “neutral”, and “agree” in an ecological momentary assessment study that involves multiple daily assessments), and count variables (e.g., number of cigarettes smoked per day across a daily diary study). In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016a]. For example, the Mplus manual [@muthen2017] includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, d) linear growth models for a count outcome assuming a zero-inflated Poisson model and e) discrete- and continuous-time survival analysis for a binary outcome. Beyond these highlighted examples, other distributions (e.g., negative binomial) can be assumed for the indicators when modeling longitudinal data. These models can account for issues that can occur when working with discrete outcomes, including overdispersion (when the variance is higher than would be expected based on a given distribution) and zero-inflation [when more zeros occur than is expected based on a given distribution; see @lenz2016]. Models involving zero-inflation parameters are referred to as two-part models, given one part of the model predicts the zero-inflation whereas the other part of the model predicts outcomes consistent with a given distribution [e.g., Poisson distribution; see @farewell2017, for a review of two-part models for longitudinal data]. Although there exist several alternative models for discrete indicators, some more recent models that have been proposed for prospective data are only feasible in cases where indicators are assumed to be continuous rather than discrete [e.g., LCM-SR; @curran2014]. Given the sheer breadth of issues relevant to determining better models for discrete outcomes, it is not uncommon for texts on longitudinal data analysis to only cover models and approaches that assume continuous indicators [e.g., @little2013]. However, some textbooks on categorical data analysis provide more detailed coverage of the myriad issues and modeling choices to consider when working with discrete outcomes [e.g., @lenz2016, Chapter 11 for matched pair/two-assessment designs; Chapter 12 for marginal and transitional models for repeated designs, such as generalized estimating equations, and Chapter 13 for random effects models for discrete outcomes].

## Missing Data/Attrition
As recently reviewed by Littlefield (in press), investigators of prospective data are confronted with study attrition (i.e., participants may not provide data at a given wave of assessment) and thus approaches are needed to confront the issue of missing data. Three models of missingness are typically considered in the literature [see @little1989]. These three models are data: a) missing completely at random (MCAR), b) missing at random (MAR), and c) missing not at random (MNAR). Data that are MCAR means missing data is a random sample of all the types of participants (e.g., males) in a given dataset. MAR suggests conditionally missing at random [see @graham2009]. That is, MAR implies missingness is completely random (i.e., does not hinge on some unmeasured variables) once missingness has been adjusted by all available variables in a dataset (e.g., biological sex). Data that are MNAR are missing as a function of unobserved variables. @graham2009 provides an excellent and easy-to-digest overview of further details involving missing data considerations.

Multiple approaches have been posited to handle missing data. Before the advent of more contemporary approaches, common methods included several ad hoc procedures. These include eliminating the data of participants with missing data (e.g., listwise or pairwise deletion) or using mean imputation (i.e., replacing the missing value with the mean score of the sample that did participate). However, these methods are not recommended because they can contribute to biased parameter estimates and research conclusions [see @graham2009]. More specifically, the last observation carried forward (LOCF) is a common approach to imputing missing data. LOCF replaces a participant’s missing values after dropout with the last available measurement [@molnar2008]. This approach assumes stability (i.e., a given participant's score is not anticipated to increase or decline after study dropout) and that the data are MCA R. However, as described by @molnar2008, it is common for treatment groups to show higher attrition compared to control groups in studies of dementia drugs. Given that dementia worsens over time, using LOCF biases the results in favor of the treatment group [see @molnar2008, for more details].

More modern approaches, such as using maximum likelihood or multiple imputation to estimate missing data, are thought to avoid some of the biases of older approaches [see @enders2010; @graham2009]. @graham2009 noted several “myths” regarding missing data. For example, Graham notes many assume the data must be minimally MAR to permit estimating procedures (such as maximum likelihood or multiple imputation) compared to other, more traditional approaches (e.g., using only complete case data). Violations of MAR impact both traditional and more modern data estimation procedures, though as noted by Graham, violations of MAR tend to have a greater effect on older methods. Graham thus suggests that estimating missing data is a better approach compared to the older procedures in most circumstances, regardless of the model of missingness [i.e., MCAR, MAR, MNAR; see @graham2009].

Attrition from a longitudinal panel study such as ABCD is inevitable and represents a threat to the validity of longitudinal analyses and cross-sectional analyses conducted at later time points, especially since attrition can only be expected to grow over time. While, to date, attrition in ABCD has been minimal (some cite here), it remains an important focus for longitudinal analysis and its significance is likely to only grow as the cohort ages.   Ideally, one tries to minimize attrition through good retention practices from the outset via strategies designed to maintain engagement in the project [@cotter2005; @hill2016; @watson2018]. However, even the best-executed studies need to anticipate growing attrition over the length of the study and implement analytic strategies designed to provide the most valid inferences. Perhaps the most key concern when dealing with data that is missing due to attrition is determining the degree of bias in retained variables that is a consequence of attrition. Assuming that the data are not missing completely at random, attention to the nature of the missingness and employing techniques designed to mitigate attrition-related biases need to be considered in all longitudinal analyses. Several different approaches can be considered and employed depending upon the nature of the intended analyses, the degree of missingness, and data available to help estimate missing and unobserved values.

## Quantifying effect sizes longitudinally
Given longitudinal data involve different sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data. Effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” [@kraemer2014]. Common effect size metrics include r (i.e., the standardized covariance, or correlation, between two variables) and Cohen’s d [@cohen1988]. Adjustments to common effect size calculations, such as Cohen’s d, are required even when only two time points are considered [e.g., see @morris2002]. @wang2019a note there are multiple approaches to obtaining standardized within-person effects, and that commonly suggested approaches (e.g., global standardization) can be problematic [see @wang2019a, for more details]. Thus, obtaining effect size metrics based on standardized estimates that are relatively simple in cross-sectional data (such as r) becomes more complex in the context of prospective data. @feingold2009 noted that equations for effects sizes used in studies involving growth modeling analysis (e.g., latent growth curve modeling) were not mathematically equivalent, and the effect sizes were not in the same metric as effect sizes from traditional analysis [see @feingold2009, for more details]. Given this issue, there have been various proposals for adjusting effect size measures in repeated assessments. @feingold2019 reviews the approach for effect size metrics for analyses based on growth modeling, including when considering linear and non-linear (i.e., quadratic) growth factors. @morris2002 review various equations for effect size calculations relevant to when combining estimates in meta-analysis with repeated measures and independent-groups designs. Other approaches to quantifying effect sizes longitudinally may be based on standardized estimates from models that more optimally disentangle between- and within-person sources of variance (as reviewed above). As an example, within a RI-CLPM framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.

### 2023-04-20 manuscript.qmd
updates to yaml config

@@
 #custom-reference-doc.docx
+
+# documentclass: scrreprt
+#     include-in-header:
+#       text: |
+#         \usepackage{makeidx}
+#         \makeindex
+#     include-after-body:
+#       text: |
+#         \printindex
 ---
@@
 \label{sec:headings}
-The Adolescent Brain Cognitive Development (ABCD) Study is the largest long-term investigation of neurodevelopment and child health in the United States.
+The Adolescent Brain Cognitive Development (ABCD) Study® is the largest long-term investigation of neurodevelopment and child health in the United States.
 Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018].

### 2023-04-20 tutorials.qmd
cli commit 1

---
title: "Tutorials: A Brief Overview"
---

[Section in progress...]

### 2023-04-20 manuscript.qmd
cli commit 3

@@
 In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to “learning” the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before).
-Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; see @french2010a] and this phenomenon has been well documented in research on structured diagnostic interviews [@robins1985].
+Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; see @french2010] and this phenomenon has been well documented in research on structured diagnostic interviews [@robins1985].
 While it is typically assumed that individuals are rescinding or telling us less information on follow-up interviews, there is reason to suspect that in some cases the initial assessment may be artefactually elevated [see @shrout2018a].
@@
 To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level.
-This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield2021.
+This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield20210603.
 It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling.
@@
 As a result, experts increasingly caution about the need to xxxxxxxx [@xxxxx].
-Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021].
+Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield20210603; @orth2021].
 Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a].
-It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].
+It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield20210603, for further discussion].
 
@@
 Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., a single item measuring the level of agreement to a statement on a three-point scale including the categories of “disagree”, “neutral”, and “agree” in an ecological momentary assessment study that involves multiple daily assessments), and count variables (e.g., number of cigarettes smoked per day across a daily diary study).
-In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016a].
+In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016].
 For example, the Mplus manual [@muthen2017] includes examples of a) censored and censored-inflated models, b) linear growth models for binary or ordinal variables, c) linear growth models for a count outcome assuming a Poisson model, d) linear growth models for a count outcome assuming a zero-inflated Poisson model and e) discrete- and continuous-time survival analysis for a binary outcome.
@@
 Models involving zero-inflation parameters are referred to as two-part models, given one part of the model predicts the zero-inflation whereas the other part of the model predicts outcomes consistent with a given distribution [e.g., Poisson distribution; see @farewell2017, for a review of two-part models for longitudinal data].
-Although there exist several alternative models for discrete indicators, some more recent models that have been proposed for prospective data are only feasible in cases where indicators are assumed to be continuous rather than discrete [e.g., LCM-SR; @curran2014].
+Although there exist several alternative models for discrete indicators, some more recent models that have been proposed for prospective data are only feasible in cases where indicators are assumed to be continuous rather than discrete [e.g., LCM-SR; @curran2014a].
 Given the sheer breadth of issues relevant to determining better models for discrete outcomes, it is not uncommon for texts on longitudinal data analysis to only cover models and approaches that assume continuous indicators [e.g., @little2013].
@@
 As an example, within a RI-CLPM framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.
+
+---
+
+### References
+
+::: {#refs}
+:::

### 2023-04-20 howto.qmd
cli commit 4

---
title: "How-To's: A Brief Overview"
---

[Section in progress...]

### 2023-04-20 manuscript.qmd
cli commit 4

@@
 ---
+
 title: Manuscript Working Draft
@@
 prefer-html: true
-#custom-reference-doc.docx
 
-# documentclass: scrreprt
-#     include-in-header:
-#       text: |
-#         \usepackage{makeidx}
-#         \makeindex
-#     include-after-body:
-#       text: |
-#         \printindex
 ---
@@
 
-[section still in progress...]
+[section still be developed...]
 
@@
 Each new wave of data collection provides the building blocks for conducting probing longitudinal analyses that allow us to characterize normative development, identify variables that presage deviations from prototypic development, and assess a range of outcomes associated with variables of interest.
-This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use [@xxxxx], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020].
+This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use [add citation], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020].
 Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward.
@@
 This report describes methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents.
-The manuscript is organized as follows: xxxxxxxxx.
+The manuscript is organized as follows:
+
+[section still be developed...]
 
@@
 However, unique epochs and experiences can alter the course of this idealized form of development.
-Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [@xxxx]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans)[@xxxx].
+Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [add citation]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans)[add citation].
 These examples highlight the importance of considering the role of vulnerable periods – temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli (e.g.
@@
 
-There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021; xxxxx et al.
-(xxxx).
+There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021.
 Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011].
 Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures.
-As a result, experts increasingly caution about the need to xxxxxxxx [@xxxxx].
+As a result, experts increasingly caution about the need to xxxxxxxx [add citation].
 Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield20210603; @orth2021].

### 2023-04-21 manuscript.qmd
cli commit 7

@@
 ---
-
-title: Manuscript Working Draft
-
-format: 
-  pdf: 
-    geometry: 
-      - top=20mm
-      - left=20mm
-  docx: default
-prefer-html: true
-
+title: "Longitudinal Analysis Manuscript: Working Draft"
 ---
-
-# Introduction
-\label{sec:headings}
-The Adolescent Brain Cognitive Development (ABCD) Study® is the largest long-term investigation of neurodevelopment and child health in the United States.
+# format:
+#   plos-pdf:
+#     number-sections: false
+#     journal:
+#       id: plosone
+#     include-in-header:
+#       text: |
+#         % Remove comment for double spacing
+#         % \usepackage{setspace} 
+#         % \doublespacing
+# author:
+#   - name: Samuel W. Hawes, PhD.
+#     affiliations:
+#       - ref: Florida International University
+#   - name: Andrew K. Littlefield, PhD.
+#     affiliations:
+#       - ref: Texas Tech University
+#   - name: Daniel A. Lopez
+#     affiliations:
+#       - ref: University of Rochester Medical Center
+#   - name: Kenneth J. Sher
+#     affiliations:
+#       - ref: University of Missouri
+#   - name: Wesley K. Thompson
+#     affiliations:
+#       - ref: Laureate Institute for Brain Research
+#   - name: Biostats-WG members [placeholder]
+#     affiliations:
+#       - ref: University of California, San Diego
+#   - name: John-Jane Doe [placeholder]
+#     affiliations:
+#       - ref: University of California, San Diego
+#     notes: |
+#       "Current Address: Dept/Program/Center, Institution Name, City, State, Country"
+#     corresponding: true
+#     email: shawes@fiu.edu
+#   - name: with the Lorem Ipsum Consortium
+#     group: true
+# author-notes:
+#   group: Biostatistics Working Group membership list can be found in the Acknowledgments sections
+# affiliations:
+#   - id: Florida International University
+#     name: Center for Children and Families, Florida International University
+#     city: Miami
+#     state: Florida
+#     country: United States
+#   - id: Texas Tech University
+#     name: Affiliation Center, Institution Name
+#     city: City
+#     state: State
+#     country: Country
+#   - id: University of Rochester Medical Center
+#     name: Affiliation Program, Institution Name
+#     city: City
+#     state: State
+#     country: Country
+#   - id: University of Missouri
+#     name: Affiliation Program, Institution Name
+#     city: City
+#     state: State
+#     country: Country
+#   - id: Laureate Institute for Brain Research
+#     name: Affiliation Program, Institution Name
+#     city: City
+#     state: State
+#     country: Country
+# abstract: |
+#   Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+# author-summary: |
+#   Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
+# bibliography: bibliography.bib 
+# 
+# ---
+# 
+# xxxxx
+# 
+# #title: "Longitudinal Analaysis Manuscript- Working Draft"
+# # format:
+# #   pdf:
+# #     toc: true
+# #     number-sections: true
+# #     colorlinks: true
+# #     geometry: 
+# #       - top=20mm
+# #       - left=20mm
+# #     docx: default
+# #     prefer-html: true
+# 
+# xxxxx
+# 
+# # Introduction
+# \label{sec:headings}
+# The Adolescent Brain Cognitive Development (ABCD) Study® is the largest long-term investigation of neurodevelopment and child health in the United States.
 Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018].
@@
 Together, this presents a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes emerge and unfold across critical periods of development.
-
-[section still be developed...]
-
-## The ABCD Study® Data
-Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians.
+# 
+# [section still be developed...] 
+# 
+# ## The ABCD Study® Data
+# Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians.
 The study sample was recruited from household populations in defined catchment areas for each of the 21 study sites across the United States (information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at the ABCD Study website).
@@
 By necessity, the study’s earliest data releases were cross-sectional (i.e., the baseline data), however, the most recent public data release (NDA Release 4.0) contains data collected across three annual assessments, including two imaging assessments (baseline and year 2 follow-up visits).
-
-## Organization of current manuscript
-The rich longitudinal nature of the ABCD Study dataset will allow researchers to perform analyses of both methodological and substantive interest.
+# 
+# ## Organization of current manuscript
+# The rich longitudinal nature of the ABCD Study dataset will allow researchers to perform analyses of both methodological and substantive interest.
 This report describes methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents.
 The manuscript is organized as follows:
-
-[section still be developed...]
-
-# Part I: Longitudinal Research: Basic Concepts and Considerations
-\label{sec:headings}
-There are several important concepts to consider when conducting longitudinal analyses in a developmental context.
+# 
+# [section still be developed...] 
+# 
+# # Part I: Longitudinal Research: Basic Concepts and Considerations
+# \label{sec:headings}
+# There are several important concepts to consider when conducting longitudinal analyses in a developmental context.
 These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represent the same phenomenon or different ones.
@@
 Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.
-
-**Vulnerable periods**.
+# 
+# **Vulnerable periods**.
 Development normatively progresses from less mature to more mature levels of functioning.
@@
 cannabis exposure) on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.
-
-**Developmental disturbances**.
+# 
+# **Developmental disturbances**.
 Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times environmental pressures will tend to suppress stability and disrupt the orderly stochastic process of normative development (e.g., xxx-xxx).
@@
 That is, once the environmental stimuli are removed (or the individual is removed from the environment), individual differences are again more freely expressed and the autoregressive effects increase to levels similar to those before entering the environment.
-
-**Developmental snares and cascade effects**.
+# 
+# **Developmental snares and cascade effects**.
 Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next.
@@
 Although conceptually distinct from vulnerable periods, both of these types of developmental considerations highlight the importance of viewing behavior in the context of development and the importance of attempting to determine how various developmental pathways unfold.
-
-**Distinguishing developmental change from experience effects**.
+# 
+# **Distinguishing developmental change from experience effects**.
 One can often observe systematic changes over time in a variable of interest and assume this change is attributable to development.
@@
 However, ABCD is fundamentally a single-cohort, longitudinal design, where a number of prior assessments and age are highly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.
-
-# Part II Longitudinal Data: Interpretation / Issues / Pitfalls & Assumption
-\label{sec:headings}
-**Defining Features of Longitudinal Data Analysis.**
-The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
-While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed.
+# 
+# # Part II Longitudinal Data: Interpretation / Issues / Pitfalls & Assumption
+# \label{sec:headings}
+# **Defining Features of Longitudinal Data Analysis.**
+# The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
+# While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed.
 That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the “right” variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .
-
-## Modeling Data Across Two Time Points versus Three or More Time Points.
-Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions.
+# 
+# ## Modeling Data Across Two Time Points versus Three or More Time Points.
+# Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions.
 Just as cross-sectional data can be informative in some situations, two waves of data can be beneficial in contexts such as when experimental manipulation is involved (e.g., pre/post tests), or if the central goal is prediction (e.g., trying to predict scores on Variable A at time T as a function of prior scores on Variable A and Variable B at time T-1).
@@
 Indeed, several contemporary longitudinal models designed to isolate within-individual variability [e.g., the Latent Curve Model with Structured Residuals; @curran2014a] require at least four assessments to parameterize fully and, more generally, increasingly accurate parameter estimates are obtained as more assessment occasions are used [@duncan2009].
-
-## Types of stability and change
-If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it’s the characterization of stability and change as the organism traverses the life course.
+# 
+# ## Types of stability and change
+# If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it’s the characterization of stability and change as the organism traverses the life course.
 There are a few different ways to think of stability (and change).
@@
 Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of “within-individual” change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest.
-
-There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021.
+# 
+# There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021.
 Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011].
@@
 It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield20210603, for further discussion].
-
-## Model Assumptions
-Many statistical models assume certain characteristics about the data to which they are being applied.
+# 
+# ## Model Assumptions
+# Many statistical models assume certain characteristics about the data to which they are being applied.
 As an example, common assumptions of parametric statistical models include normality, linearity, and equality of variances.
@@
 Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model the covariance structure of said data.
-
-## Covariance Structures
-Statistical models for longitudinal data include two main components to account for assumptions that are commonly violated when working with repeated measures data: a model for the covariance among repeated measures (both the correlations among pairs of repeated measures on an individual and the variability of the responses on different occasions), coupled with a model for the mean response and its dependence on covariates (eg, treatment group in the context of clinical trials).
+# 
+# ## Covariance Structures
+# Statistical models for longitudinal data include two main components to account for assumptions that are commonly violated when working with repeated measures data: a model for the covariance among repeated measures (both the correlations among pairs of repeated measures on an individual and the variability of the responses on different occasions), coupled with a model for the mean response and its dependence on covariates (eg, treatment group in the context of clinical trials).
 This allows for the specification of a range of so-called covariance structures, each with its own set of tradeoffs between model fit and parsimony [e.g., see @kincaid2005].
-
-## Accounting for Correlated Data
-As an example, one alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design.
+# 
+# ## Accounting for Correlated Data
+# As an example, one alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design.
 As the name implies, the structure assumes a subsequent measurement occasion (e.g., assessment at Wave 2) is regressed onto (that is, is predicted by) a prior measurement occasion (e.g., assessment at Wave 1).
@@
 within-person sources of variance [@hamaker2015].
-
-## Linear vs non-linear models
-Identification of optimal statistical models and appropriate mathematical functions requires an understanding of the type of data being used.
+# 
+# ## Linear vs non-linear models
+# Identification of optimal statistical models and appropriate mathematical functions requires an understanding of the type of data being used.
 Repeated assessments can be based on either continuous or discrete measures.
@@
 However, some textbooks on categorical data analysis provide more detailed coverage of the myriad issues and modeling choices to consider when working with discrete outcomes [e.g., @lenz2016, Chapter 11 for matched pair/two-assessment designs; Chapter 12 for marginal and transitional models for repeated designs, such as generalized estimating equations, and Chapter 13 for random effects models for discrete outcomes].
-
-## Missing Data/Attrition
-As recently reviewed by Littlefield (in press), investigators of prospective data are confronted with study attrition (i.e., participants may not provide data at a given wave of assessment) and thus approaches are needed to confront the issue of missing data.
+# 
+# ## Missing Data/Attrition
+# As recently reviewed by Littlefield (in press), investigators of prospective data are confronted with study attrition (i.e., participants may not provide data at a given wave of assessment) and thus approaches are needed to confront the issue of missing data.
 Three models of missingness are typically considered in the literature [see @little1989].
@@
 @graham2009 provides an excellent and easy-to-digest overview of further details involving missing data considerations.
-
-Multiple approaches have been posited to handle missing data.
+# 
+# Multiple approaches have been posited to handle missing data.
 Before the advent of more contemporary approaches, common methods included several ad hoc procedures.
@@
 Given that dementia worsens over time, using LOCF biases the results in favor of the treatment group [see @molnar2008, for more details].
-
-More modern approaches, such as using maximum likelihood or multiple imputation to estimate missing data, are thought to avoid some of the biases of older approaches [see @enders2010; @graham2009].
+# 
+# More modern approaches, such as using maximum likelihood or multiple imputation to estimate missing data, are thought to avoid some of the biases of older approaches [see @enders2010; @graham2009].
 @graham2009 noted several “myths” regarding missing data.
@@
 Graham thus suggests that estimating missing data is a better approach compared to the older procedures in most circumstances, regardless of the model of missingness [i.e., MCAR, MAR, MNAR; see @graham2009].
-
-Attrition from a longitudinal panel study such as ABCD is inevitable and represents a threat to the validity of longitudinal analyses and cross-sectional analyses conducted at later time points, especially since attrition can only be expected to grow over time.
+# 
+# Attrition from a longitudinal panel study such as ABCD is inevitable and represents a threat to the validity of longitudinal analyses and cross-sectional analyses conducted at later time points, especially since attrition can only be expected to grow over time.
 While, to date, attrition in ABCD has been minimal (some cite here), it remains an important focus for longitudinal analysis and its significance is likely to only grow as the cohort ages.
@@
 Several different approaches can be considered and employed depending upon the nature of the intended analyses, the degree of missingness, and data available to help estimate missing and unobserved values.
-
-## Quantifying effect sizes longitudinally
-Given longitudinal data involve different sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data.
+# 
+# ## Quantifying effect sizes longitudinally
+# Given longitudinal data involve different sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data.
 Effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” [@kraemer2014].
@@
 As an example, within a RI-CLPM framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.
-
----
-
-### References
-
-::: {#refs}
-:::
+# 
+# ---
+# 
+# ### References
+# 
+# ::: {#refs}
+# :::

### 2023-04-21 manuscript.tex
cli commit 7

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%

\PassOptionsToPackage{table}{xcolor}

\documentclass[
  10pt,
  letterpaper,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% marvosym package for additional characters
\usepackage{marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% Using natbib instead
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}
% Remove comment for double spacing
% \usepackage{setspace}
% \doublespacing
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[numbers,square,comma]{natbib}
\bibliographystyle{plos2015}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Longitudinal Analysis Manuscript: Working Draft},
  pdfauthor={Samuel W. Hawes; with the Lorem Ipsum Consortium},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Longitudinal Analysis Manuscript: Working
Draft} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
Samuel W. Hawes\textsuperscript{1*}, with the Lorem Ipsum
Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Center for Children and Families, Florida International
University, Miami, Florida, United States, 
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
\dag Deceased

% Group/Consortium Author Note
\textpilcrow Biostatistics Working Group membership list can be found in
the Acknowledgments sections

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* shawes@fiu.edu

\end{flushleft}

\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

\linenumbersxxxxx

\#title: ``Longitudinal Analaysis Manuscript- Working Draft'' \# format:
\# pdf: \# toc: true \# number-sections: true \# colorlinks: true \#
geometry: \# - top=20mm \# - left=20mm \# docx: default \# prefer-html:
true

xxxxx

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

label\{sec:headings\} The Adolescent Brain Cognitive Development (ABCD)
Study® is the largest long-term investigation of neurodevelopment and
child health in the United States. Conceived and initiated by the
National Institutes of Health (NIH), this landmark prospective
longitudinal study aims to transform our understanding of the genetic
and environmental influences on brain development and their roles in
behavioral and health outcomes in adolescents \citep{volkow2018}. At its
heart, the study is designed to chart the course of human development
across multiple, interacting domains from late childhood to early
adulthood and to identify factors that lead to both positive and
negative developmental outcomes. Central to achieving these goals is the
ABCD Study's® commitment to an open science framework designed to
facilitate access to and sharing of scientific knowledge by espousing
practices that increase openness, integrity, and reproducibility of
scientific research (e.g., public data releases). In this sense, the
ABCD Study® is a collaboration with the larger research community, with
the rich longitudinal nature of the ABCD Study dataset allowing
researchers to perform a variety of analyses of both methodological and
substantive interest. Together, this presents a unique opportunity to
significantly advance our understanding of how a multitude of
biopsychosocial processes emerge and unfold across critical periods of
development.

{[}section still be developed\ldots{]}

\hypertarget{the-abcd-study-data}{%
\subsection{The ABCD Study® Data}\label{the-abcd-study-data}}

Participants enrolled in the ABCD Study include a large cohort of youth
(n=11880) aged 9-10 years at baseline and their parents/guardians. The
study sample was recruited from household populations in defined
catchment areas for each of the 21 study sites across the United States
(information regarding funding agencies, recruitment sites,
investigators, and project organization can be obtained at the ABCD
Study website). The ABCD Study is collecting longitudinal data on a rich
variety of outcomes that will enable the construction of
realistically-complex etiological models by incorporating factors from
many domains simultaneously. Each new wave of data collection provides
the building blocks for conducting probing longitudinal analyses that
allow us to characterize normative development, identify variables that
presage deviations from prototypic development, and assess a range of
outcomes associated with variables of interest. This data includes a
neurocognitive battery \citep{luciana2018a, thompson2019}, mental and
physical health assessments \citep{barch2018}, measures of culture and
environment \citep{zucker2018}, substance use {[}add citation{]},
biospecimens \citep{uban2018}, structural and functional brain imaging
\citep{casey2018, hagler2019}, geolocation-based environmental exposure
data, wearables, and mobile technology \citep{bagot2018}, and whole
genome genotyping \citep{loughnan2020}. Many of these measures are
collected at in-person annual visits, with brain imaging collected at
baseline and every other year going forward. A limited number of
assessments are collected in semi-annual telephone interviews between
in-person visits. Data are publicly released on an annual basis through
the NIMH Data Archive. By necessity, the study's earliest data releases
were cross-sectional (i.e., the baseline data), however, the most recent
public data release (NDA Release 4.0) contains data collected across
three annual assessments, including two imaging assessments (baseline
and year 2 follow-up visits).

\hypertarget{organization-of-current-manuscript}{%
\subsection{Organization of current
manuscript}\label{organization-of-current-manuscript}}

The rich longitudinal nature of the ABCD Study dataset will allow
researchers to perform analyses of both methodological and substantive
interest. This report describes methods for longitudinal analyses of
ABCD Study data that can address its fundamental scientific aims, as
well as challenges inherent in a large population-based long-term study
of adolescents. The manuscript is organized as follows:

{[}section still be developed\ldots{]}

\hypertarget{part-i-longitudinal-research-basic-concepts-and-considerations}{%
\section{Part I: Longitudinal Research: Basic Concepts and
Considerations}\label{part-i-longitudinal-research-basic-concepts-and-considerations}}

label\{sec:headings\} There are several important concepts to consider
when conducting longitudinal analyses in a developmental context. These
include different ways of thinking about developmental course, whether
certain periods of development are relatively sensitive or insensitive
to various types of insults or stressors, whether some time periods or
situations inhibit the expression of individual differences due to
extreme environmental pressures, and whether the same behavior
manifested at different times represent the same phenomenon or different
ones. Further, in the case of developmentally focused longitudinal
research, each new measurement occasion not only provides a more
extended portrait of the child's life course (and not just characterize
growth during this period but also assesses the durability/chronicity of
prior effects/consequences) but also brings with it greater
methodological opportunities to exploit the statistical properties of
longitudinal data in the furtherance of critical scientific questions.
That is, we can ask more nuanced questions and make stronger inferences
as our number of time-ordered observations grows, assuming we have
assessed the ``right'' variables and the timings of our observations
comport with the temporal dynamics of the mechanisms of interest.
Appreciation of these and other issues can help to guide the analysis
and interpretation of data and aid translation to clinical and public
health applications.

\textbf{Vulnerable periods}. Development normatively progresses from
less mature to more mature levels of functioning. However, unique epochs
and experiences can alter the course of this idealized form of
development. Consider research that shows cannabis use during
adolescence is associated with later psychosis to a greater degree than
cannabis use initiated later in development {[}add citation{]}; or,
similarly, experimental research on rodents that shows rodent brains to
be especially sensitive to the neurotoxic effects of alcohol on brain
structure and learning early in development (corresponding to early
adolescence in humans){[}add citation{]}. These examples highlight the
importance of considering the role of vulnerable periods -- temporal
windows of rapid brain development or remodeling during which the
effects of environmental stimuli (e.g.~cannabis exposure) on the
developing brain may be particularly pronounced-- when trying to
establish an accurate understanding of the association between exposures
and outcomes.

\textbf{Developmental disturbances}. Whereas vulnerable periods heighten
neurobiological susceptibility to environmental influences, at other
times environmental pressures will tend to suppress stability and
disrupt the orderly stochastic process of normative development (e.g.,
xxx-xxx). This situation reflects a developmental disturbance in that
the normal course of development is ``disturbed'' for some time by some
time-limited process. In such cases, we might find that prediction of
behavior in the period of the disturbance is reduced and/or, similarly,
the behavior exhibited during the disturbance might have less predictive
power with respect to distal outcomes compared to the behavior exhibited
before and following the disrupted period. That is, once the
environmental stimuli are removed (or the individual is removed from the
environment), individual differences are again more freely expressed and
the autoregressive effects increase to levels similar to those before
entering the environment.

\textbf{Developmental snares and cascade effects}. Normative development
can also be upended by experiences (e.g., drug use) that, through
various mechanisms, disrupt the normal flow of development wherein each
stage establishes a platform for the next. For instance, substance use
could lead to association with deviant peers, precluding opportunities
for learning various adaptive skills and prosocial behaviors, in effect,
creating a ``snare'' that retards psychosocial development. Relatedly,
the consequences of these types of events can cascade (e.g., school
dropout, involvement in the criminal justice system) so that the effects
of the snare are amplified. Although conceptually distinct from
vulnerable periods, both of these types of developmental considerations
highlight the importance of viewing behavior in the context of
development and the importance of attempting to determine how various
developmental pathways unfold.

\textbf{Distinguishing developmental change from experience effects}.
One can often observe systematic changes over time in a variable of
interest and assume this change is attributable to development. To this
point, cognitive abilities (e.g, verbal ability, problem-solving)
normatively grow earlier in development and often decline in late life
(e.g., memory, speed of processing). However, the observed patterns of
growth and decline often differ between cross-sectional vs.~longitudinal
effects \citep{salthouse2014} where subjects gain increasing experience
with the assessment with each successive measurement occasion. Such
experience effects on cognitive functioning have been demonstrated in
adolescent longitudinal samples similar to ABCD \citep{sullivan2017} and
highlight the need to consider these effects and address them
analytically. In the case of performance-based measures {[}e.g., matrix
reasoning related to neurocognitive functioning; see
\citet{salthouse2014}{]}, this can be due to ``learning'' the task from
previous test administrations (e.g., someone taking the test a second
time performs better than they did the first time simply as a function
of having taken it before). Even in the case of non-performance-based
measures (e.g., levels of depression), where one cannot easily make the
argument that one has acquired some task-specific skill through
learning, it has been observed that respondents tend to endorse lower
levels on subsequent assessments
\citetext{\citealp[e.g.,][]{beck1961}; \citealp[see][]{french2010}} and
this phenomenon has been well documented in research on structured
diagnostic interviews \citep{robins1985}. While it is typically assumed
that individuals are rescinding or telling us less information on
follow-up interviews, there is reason to suspect that in some cases the
initial assessment may be artefactually elevated
\citep[see][]{shrout2018a}. Some designs (specifically, accelerated
longitudinal designs) are especially well suited for discovering these
effects and modeling them. While ABCD was not designed as an accelerated
longitudinal design, the variability in age at the time of baseline
recruitment (9 years, 0 months to 10 years, 11 months) allows some
measures, collected every year, to be conceptualized as an accelerated
longitudinal design. Moreover, it is possible that in later waves,
patterns of longitudinal missing data will allow some analyses to assess
the confounded effects of age and the number of prior assessments.
However, ABCD is fundamentally a single-cohort, longitudinal design,
where a number of prior assessments and age are highly confounded, and
for, perhaps, most analyses, the possible influence of experience
effects needs to be kept in mind.

\hypertarget{part-ii-longitudinal-data-interpretation-issues-pitfalls-assumption}{%
\section{Part II Longitudinal Data: Interpretation / Issues / Pitfalls
\&
Assumption}\label{part-ii-longitudinal-data-interpretation-issues-pitfalls-assumption}}

label\{sec:headings\} \textbf{Defining Features of Longitudinal Data
Analysis.} The hallmark characteristic of longitudinal data analysis is
its application to repeated assessments of the same assessment targets
(e.g., individuals, families) across time. While the primary reason for
collecting longitudinal data is in pursuit of addressing scientific
questions, from a methodological perspective, having multiple
observations over time allows researchers to identify potentially
problematic observations when highly improbable longitudinal patterns
are observed. That is, we can ask more nuanced questions and make
stronger inferences as our number of time-ordered observations grows
assuming we have assessed the ``right'' variables and the timings of our
observations comport with the temporal dynamics of the mechanisms of
interest .

\hypertarget{modeling-data-across-two-time-points-versus-three-or-more-time-points.}{%
\subsection{Modeling Data Across Two Time Points versus Three or More
Time
Points.}\label{modeling-data-across-two-time-points-versus-three-or-more-time-points.}}

Although the clear leap to the realm of longitudinal data involves going
from one assessment to two or more assessments, there are also notable
distinctions in designs based on two-assessment points versus three or
more measurement occasions. Just as cross-sectional data can be
informative in some situations, two waves of data can be beneficial in
contexts such as when experimental manipulation is involved (e.g.,
pre/post tests), or if the central goal is prediction (e.g., trying to
predict scores on Variable A at time T as a function of prior scores on
Variable A and Variable B at time T-1). At the same time, data based on
two assessments are inherently limited on multiple fronts. As
\citep{rogosa1982} noted approximately forty years ago, ``Two waves of
data are better than one, but maybe not much better''. These sentiments
are reflected in more contemporary recommendations regarding
best-practice guidelines for prospective data, which increasingly
emphasize the benefits of additional measurement occasions for model
identification and accurate parameter estimation. It is also consistent
with research recommending that developmental studies include three or
more assessment points, given it is impossible for data based on
two-time points to determine the shape of development (since linear,
straight line change is the only possible form, given two assessments;
see \citep{duncan2009}). Research designs that include three or more
time points allow for increasingly nuanced analyses that more adequately
tease apart sources of variation and covariation among the repeated
assessments \citep{king2018}-- a key aspect of inferential research. To
illustrate, developmental theories are typically interested in
understanding patterns of within-individual change over time (discussed
in further detail, below); however, two data points provide meager
information on change at the person level. This point is further
underscored in a recent review of statistical models commonly touted as
distinguishing within-individual vs between-individual sources of
variance in which the study authors concluded ``\ldots{} researchers are
limited when attempting to differentiate these sources of variation in
psychological phenomenon when using two waves of data'' and perhaps more
concerning, ``\ldots the models discussed here do not offer a feasible
way to overcome these inherent limitations''
\citet{littlefield20210603}. It is important to note, however, that
despite the current focus on two-wave designs versus three or more
assessment waves, garnering three assessment points is not a panacea for
longitudinal modeling. Indeed, several contemporary longitudinal models
designed to isolate within-individual variability {[}e.g., the Latent
Curve Model with Structured Residuals; \citet{curran2014a}{]} require at
least four assessments to parameterize fully and, more generally,
increasingly accurate parameter estimates are obtained as more
assessment occasions are used \citep{duncan2009}.

\hypertarget{types-of-stability-and-change}{%
\subsection{Types of stability and
change}\label{types-of-stability-and-change}}

If one were to try to sum up what development in a living organism is
exactly, one could plausibly argue it's the characterization of
stability and change as the organism traverses the life course. There
are a few different ways to think of stability (and change). Consider we
measure the height of all youth in a 6th-grade class, once in the fall
at the beginning of the school year and once again in the spring at the
end of the school year. A common first step may be to compare the
class's average height values obtained at these two different
measurement occasions. This comparison of the average scores for the
same group of individuals at multiple time points is referred to as
``mean-level'' stability as it provides information about continuity and
change in the group level of an outcome of interest (e.g., height) over
time. Another type of stability involves calculating the correlation
between the values obtained at different time points (e.g., `height in
the fall' with `height in the spring'). This type of ``rank-order''
stability evaluates between-individual change by focusing on the degree
to which individuals retain their relative placement in a group across
time. Consider, someone who is the shortest person in their class in 6th
grade may grow considerably over the school year (i.e., exhibit mean
level change), but remain the shortest person among their classmates.
That is, the individual is manifesting a type of rank-order stability.
Both types of stability and change are important. Mean-level change in
certain traits might help to explain why, in general, individuals are
particularly vulnerable to social influences at some ages more than
others; rank order change might help to quantify the extent to which
certain characteristics of the individual are more trait-like. For
example, in some areas of development, there is considerable mean-level
change that occurs over time (e.g., changes in Big 5 personality
traits), but relatively high rank-order stability. Despite the useful
information afforded by examining mean-level and rank-order change,
these approaches are limited in that they provide little information
about patterns of ``within-individual'' change and, in turn, can result
in fundamental misinterpretations about substantial or meaningful
changes in an outcome of interest.

There is growing recognition that statistical models commonly applied to
longitudinal data often fail to comport with the developmental theory
they are being used to assess (e.g., Curran, Lee, Howard, Lane, \&
MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021. Specifically,
developmental studies typically involve the use of prospective data to
inform theories that are concerned with clear within-person (i.e.,
intraindividual) processes (e.g., how phenotypes change or remain stable
within individuals over time) \citep[e.g., see][]{curran2011}. Despite
this, methods generally unsuited for disaggregating between- and
within-person effects (e.g., cross-lagged panel models {[}CLPM{]})
remain common within various extant literatures. As a result, experts
increasingly caution about the need to xxxxxxxx {[}add citation{]}.
Fortunately, there exists a range of models that have been proposed to
tease apart between- and within-person sources of variance across time
\citep[see][]{littlefield20210603, orth2021}. Most of these contemporary
alternatives incorporate time-specific latent variables to capture
between-person sources of variance and model within-person deviations
around an individual's mean (or trait) level across time
\citetext{\citealp[e.g.,
RI-CLPM,][]{hamaker2015}; \citealp[LCM-SR,][]{curran2014a}}. It is
important to note however that these models require multiple assessments
waves (e.g., four or more to fully specify the LCM-SR), additional
expertise to overcome issues with model convergence, and appreciation of
modeling assumptions when attempting to adjudicate among potential
models in each research context \citep[see][for further
discussion]{littlefield20210603}.

\hypertarget{model-assumptions}{%
\subsection{Model Assumptions}\label{model-assumptions}}

Many statistical models assume certain characteristics about the data to
which they are being applied. As an example, common assumptions of
parametric statistical models include normality, linearity, and equality
of variances. These assumptions must be carefully considered before
conducting analysis so that valid inferences can be made from the data;
that is, violation of a model's assumptions can substantively alter the
interpretation of results. Similarly, statistical models employed in the
analyses of longitudinal data often entail a range of assumptions that
must be closely inspected. One central issue for repeated measurements
on an individual is how to account for the correlated nature of the
data; another common feature of longitudinal data is heterogeneous
variability; that is, the variance of the response changes over the
duration of the study. Traditional techniques, such as a standard
regression or ANOVA model, assume residuals are independent and thus are
inappropriate for designs that assess (for example) the same individuals
across time. That is, given the residuals are no longer independent, the
standard errors from the models are biased and can produce misleading
inferential results. Although there are formal tests of independence for
time series data (e.g., the Durbin-Watson statistic; Durbin \& Watson,
1950), more commonly independence is assumed to be violated in study
designs with repeated assessments. Therefore, an initial question to be
addressed by a researcher analyzing prospective data is how to best
model the covariance structure of said data.

\hypertarget{covariance-structures}{%
\subsection{Covariance Structures}\label{covariance-structures}}

Statistical models for longitudinal data include two main components to
account for assumptions that are commonly violated when working with
repeated measures data: a model for the covariance among repeated
measures (both the correlations among pairs of repeated measures on an
individual and the variability of the responses on different occasions),
coupled with a model for the mean response and its dependence on
covariates (eg, treatment group in the context of clinical trials). This
allows for the specification of a range of so-called covariance
structures, each with its own set of tradeoffs between model fit and
parsimony \citep[e.g., see][]{kincaid2005}.

\hypertarget{accounting-for-correlated-data}{%
\subsection{Accounting for Correlated
Data}\label{accounting-for-correlated-data}}

As an example, one alternative structure that attempts to handle the
reality that correlations between repeated assessments tend to diminish
across time is the autoregressive design. As the name implies, the
structure assumes a subsequent measurement occasion (e.g., assessment at
Wave 2) is regressed onto (that is, is predicted by) a prior measurement
occasion (e.g., assessment at Wave 1). The most common type of
autoregressive design is the AR(1), where assessments at time T + 1 are
regressed on assessments at Time T. Identical to compound symmetry, this
model assumes the variances are homogenous across time. Diverting from
compound symmetry, this model assumes the correlations between repeated
assessments decline exponentially across time rather than remaining
constant. For example, per the AR(1) structure, if the correlation
between Time 1 and Time 2 data is thought to be .5, then the correlation
between Time 1 and Time 3 data would be assumed to be .5\emph{.5 = .25,
and the correlation between Time 1 and Time 4 data would be assumed to
be .5}.5*.5 = .125. As with compound symmetry, the basic AR(1) model is
parsimonious in that it only requires two parameters (the variance of
the assessments and the autoregressive coefficient). Notably, the
assumption of constant autoregressive relations between assessments is
often relaxed in commonly employed designs that use autoregressive
modeling (e.g., cross-lagged panel models {[}CLPM{]}). These designs
still typically assume an AR(1) process (e.g., it is sufficient to
regress the Time 3 assessment onto the Time 2 assessment and is not
necessary to also regress the Time 3 assessment onto the Time 1
assessment, which would result in an AR(2) process). However, the
magnitude of these relations is often allowed to differ across different
AR(1) pairs of assessment (e.g., the relation between Time 1 and Time 2
can be different from the relation between Time 2 and Time 3). These
more commonly employed models also often relax the assumption of equal
variances of the repeated assessments. Although the AR(1) structure may
involve a more realistic set of assumptions compared to compound
symmetry, in that the AR(1) model allows for diminishing correlations
across time, the basic AR(1) model, as well as autoregressive models
more generally, can also suffer from several limitations in contexts
that are common in prospective designs. In particular, recent work
demonstrates that if a construct being assessed prospectively across
time is trait-like in nature, then autoregressive relations fail to
adequately account for this trait-like structure, with the downstream
consequence that estimates derived from models based on AR structures
(such as the CLPM) can be misleading and fail to adequately demarcate
between- vs.~within-person sources of variance \citep{hamaker2015}.

\hypertarget{linear-vs-non-linear-models}{%
\subsection{Linear vs non-linear
models}\label{linear-vs-non-linear-models}}

Identification of optimal statistical models and appropriate
mathematical functions requires an understanding of the type of data
being used. Repeated assessments can be based on either continuous or
discrete measures. Examples of discrete measures include repeated
assessments of binary variables (e.g., past 12-month alcohol use
disorder status measured across ten years), ordinal variables (e.g., a
single item measuring the level of agreement to a statement on a
three-point scale including the categories of ``disagree'', ``neutral'',
and ``agree'' in an ecological momentary assessment study that involves
multiple daily assessments), and count variables (e.g., number of
cigarettes smoked per day across a daily diary study). In many ways, the
distributional assumptions of indicators used in longitudinal designs
mirror the decision points and considerations when delineating across
different types of discrete outcome variables, a topic that spans entire
textbooks \citep[e.g., see][]{lenz2016}. For example, the Mplus manual
\citep{muthen2017} includes examples of a) censored and
censored-inflated models, b) linear growth models for binary or ordinal
variables, c) linear growth models for a count outcome assuming a
Poisson model, d) linear growth models for a count outcome assuming a
zero-inflated Poisson model and e) discrete- and continuous-time
survival analysis for a binary outcome. Beyond these highlighted
examples, other distributions (e.g., negative binomial) can be assumed
for the indicators when modeling longitudinal data. These models can
account for issues that can occur when working with discrete outcomes,
including overdispersion (when the variance is higher than would be
expected based on a given distribution) and zero-inflation {[}when more
zeros occur than is expected based on a given distribution; see
\citet{lenz2016}{]}. Models involving zero-inflation parameters are
referred to as two-part models, given one part of the model predicts the
zero-inflation whereas the other part of the model predicts outcomes
consistent with a given distribution {[}e.g., Poisson distribution; see
\citet{farewell2017}, for a review of two-part models for longitudinal
data{]}. Although there exist several alternative models for discrete
indicators, some more recent models that have been proposed for
prospective data are only feasible in cases where indicators are assumed
to be continuous rather than discrete {[}e.g., LCM-SR;
\citet{curran2014a}{]}. Given the sheer breadth of issues relevant to
determining better models for discrete outcomes, it is not uncommon for
texts on longitudinal data analysis to only cover models and approaches
that assume continuous indicators \citep[e.g.,][]{little2013}. However,
some textbooks on categorical data analysis provide more detailed
coverage of the myriad issues and modeling choices to consider when
working with discrete outcomes {[}e.g., \citet{lenz2016}, Chapter 11 for
matched pair/two-assessment designs; Chapter 12 for marginal and
transitional models for repeated designs, such as generalized estimating
equations, and Chapter 13 for random effects models for discrete
outcomes{]}.

\hypertarget{missing-dataattrition}{%
\subsection{Missing Data/Attrition}\label{missing-dataattrition}}

As recently reviewed by Littlefield (in press), investigators of
prospective data are confronted with study attrition (i.e., participants
may not provide data at a given wave of assessment) and thus approaches
are needed to confront the issue of missing data. Three models of
missingness are typically considered in the literature
\citep[see][]{little1989}. These three models are data: a) missing
completely at random (MCAR), b) missing at random (MAR), and c) missing
not at random (MNAR). Data that are MCAR means missing data is a random
sample of all the types of participants (e.g., males) in a given
dataset. MAR suggests conditionally missing at random
\citep[see][]{graham2009}. That is, MAR implies missingness is
completely random (i.e., does not hinge on some unmeasured variables)
once missingness has been adjusted by all available variables in a
dataset (e.g., biological sex). Data that are MNAR are missing as a
function of unobserved variables. \citet{graham2009} provides an
excellent and easy-to-digest overview of further details involving
missing data considerations.

Multiple approaches have been posited to handle missing data. Before the
advent of more contemporary approaches, common methods included several
ad hoc procedures. These include eliminating the data of participants
with missing data (e.g., listwise or pairwise deletion) or using mean
imputation (i.e., replacing the missing value with the mean score of the
sample that did participate). However, these methods are not recommended
because they can contribute to biased parameter estimates and research
conclusions \citep[see][]{graham2009}. More specifically, the last
observation carried forward (LOCF) is a common approach to imputing
missing data. LOCF replaces a participant's missing values after dropout
with the last available measurement \citep{molnar2008}. This approach
assumes stability (i.e., a given participant's score is not anticipated
to increase or decline after study dropout) and that the data are MCA R.
However, as described by \citet{molnar2008}, it is common for treatment
groups to show higher attrition compared to control groups in studies of
dementia drugs. Given that dementia worsens over time, using LOCF biases
the results in favor of the treatment group \citep[see][for more
details]{molnar2008}.

More modern approaches, such as using maximum likelihood or multiple
imputation to estimate missing data, are thought to avoid some of the
biases of older approaches \citep[see][]{enders2010, graham2009}.
\citet{graham2009} noted several ``myths'' regarding missing data. For
example, Graham notes many assume the data must be minimally MAR to
permit estimating procedures (such as maximum likelihood or multiple
imputation) compared to other, more traditional approaches (e.g., using
only complete case data). Violations of MAR impact both traditional and
more modern data estimation procedures, though as noted by Graham,
violations of MAR tend to have a greater effect on older methods. Graham
thus suggests that estimating missing data is a better approach compared
to the older procedures in most circumstances, regardless of the model
of missingness {[}i.e., MCAR, MAR, MNAR; see \citet{graham2009}{]}.

Attrition from a longitudinal panel study such as ABCD is inevitable and
represents a threat to the validity of longitudinal analyses and
cross-sectional analyses conducted at later time points, especially
since attrition can only be expected to grow over time. While, to date,
attrition in ABCD has been minimal (some cite here), it remains an
important focus for longitudinal analysis and its significance is likely
to only grow as the cohort ages. Ideally, one tries to minimize
attrition through good retention practices from the outset via
strategies designed to maintain engagement in the project
\citep{cotter2005, hill2016, watson2018}. However, even the
best-executed studies need to anticipate growing attrition over the
length of the study and implement analytic strategies designed to
provide the most valid inferences. Perhaps the most key concern when
dealing with data that is missing due to attrition is determining the
degree of bias in retained variables that is a consequence of attrition.
Assuming that the data are not missing completely at random, attention
to the nature of the missingness and employing techniques designed to
mitigate attrition-related biases need to be considered in all
longitudinal analyses. Several different approaches can be considered
and employed depending upon the nature of the intended analyses, the
degree of missingness, and data available to help estimate missing and
unobserved values.

\hypertarget{quantifying-effect-sizes-longitudinally}{%
\subsection{Quantifying effect sizes
longitudinally}\label{quantifying-effect-sizes-longitudinally}}

Given longitudinal data involve different sources of variance,
quantifying effect sizes longitudinally is a more difficult task
compared to deriving such estimates from cross-sectional data. Effect
size can be defined as, ``a population parameter (estimated in a sample)
encapsulating the practical or clinical importance of a phenomenon under
study.'' \citep{kraemer2014}. Common effect size metrics include r
(i.e., the standardized covariance, or correlation, between two
variables) and Cohen's d \citep{cohen1988}. Adjustments to common effect
size calculations, such as Cohen's d, are required even when only two
time points are considered \citep[e.g., see][]{morris2002}.
\citet{wang2019a} note there are multiple approaches to obtaining
standardized within-person effects, and that commonly suggested
approaches (e.g., global standardization) can be problematic
\citep[see][for more details]{wang2019a}. Thus, obtaining effect size
metrics based on standardized estimates that are relatively simple in
cross-sectional data (such as r) becomes more complex in the context of
prospective data. \citet{feingold2009} noted that equations for effects
sizes used in studies involving growth modeling analysis (e.g., latent
growth curve modeling) were not mathematically equivalent, and the
effect sizes were not in the same metric as effect sizes from
traditional analysis \citep[see][for more details]{feingold2009}. Given
this issue, there have been various proposals for adjusting effect size
measures in repeated assessments. \citet{feingold2019} reviews the
approach for effect size metrics for analyses based on growth modeling,
including when considering linear and non-linear (i.e., quadratic)
growth factors. \citet{morris2002} review various equations for effect
size calculations relevant to when combining estimates in meta-analysis
with repeated measures and independent-groups designs. Other approaches
to quantifying effect sizes longitudinally may be based on standardized
estimates from models that more optimally disentangle between- and
within-person sources of variance (as reviewed above). As an example,
within a RI-CLPM framework, standardized estimates between random
intercepts (i.e., the correlation between two random intercepts for two
different constructs assessed repeatedly) could be used to index the
between-person relation, whereas standardized estimates among the
structured residuals could be used as informing the effect sizes of
within-person relations.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{references}{%
\subsubsection{References}\label{references}}

\renewcommand{\bibsection}{}
\bibliography{references.bib,bibliography.bib}

\nolinenumbers

\end{document}

### 2023-04-21 test.qmd
cli commit 7

--- 2023-04-20 manuscript.qmd
@@
 ---
+title: "Longitudinal Analysis Manuscript: Working Draft"
+format:
+  plos-pdf:
+    number-sections: false
+    journal:
+      # This is an identifier for the target journal: 
+      # from https://plos.org/resources/writing-center/ following submission guidelines link, the identifier is the part of the URL after https://journals.plos.org/<id>/s/submission-guidelines
+      id: plosone
+    include-in-header:
+      text: |
+        % Remove comment for double spacing
+        % \usepackage{setspace} 
+        % \doublespacing
+author:
+  - name: Samuel W. Hawes, PhD.
+    affiliations:
+      - ref: aff1
+    corresponding: true
+    email: shawes@fiu.edu
+  - name: Andrew K. Littlefield, PhD.
+    affiliations:
+      - ref: aff2
+  - name: Daniel A. Lopez
+    affiliations:
+      - ref: aff3
+  - name: Kenneth J. Sher
+    affiliations:
+      - ref: aff4
+  - name: Wesley K. Thompson
+    affiliations:
+      - ref: aff5
+  - name: Additional Co-authors1
+    affiliations:
+      - ref: aff1
+  - name: Additional Co-authors2
+    affiliations:
+      - ref: aff1
+author-notes:
+  group: Biostatistics Working Group membership list can be found in the Acknowledgments sections
+affiliations:
+  - id: aff1
+    name: Center for Children and Families, Florida International University
+    city: Miami
+    state: FL
+    country: United States
+  - id: aff2
+    name: Psychological Sciences, Texas Tech University
+    city: Lubbock
+    state: TX
+    country: United States
+  - id: aff3
+    name: Department of Public Health, University of Rochester Medical Center
+    city: Rochester
+    state: NY
+    country: United States
+  - id: aff4
+    name: Psychological Sciences, University of Missouri
+    city: Columbia
+    state: MO
+    country: United States
+  - id: aff5
+    name: Population Neuroscience and Genetics Center, Laureate Institute for Brain Research
+    city: Tulsa
+    state: OK
+    country: United States
+abstract: |
+  The Adolescent Brain Cognitive Development (ABCD) Study presents a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of children and adolescents.
+Given the complex nature of the longitudinal data collected in the ABCD Study, researcher are likely to encounter a myriad of methodological and analytic considerations and concerns.
+This review provides a comprehensive examination of key issues and techniques related to longitudinal data analysis, specifically focusing on the ABCD Study.
+The text discusses model assumptions, common violations (e.g., independent and identically distributed residuals, *heterogeneous* *variability*) and their implications for valid inference.
+The importance of appropriately modeling covariance structures, understandings trade-offs between model fit and parsimony, and challenges related to sample size, attrition, missing data are highlighted.
+Consideration is given to the importance of selecting appropriate statistical models to account for correlations in repeated measurements and the assumptions underlying these models.
+The review also differentiates between linear and non-linear models in the context of continuous and discrete data, emphasizing various distributional assumptions and the necessity of choosing appropriate models and statistical methods.
+By addressing these complexities, the review seeks to equip researchers with the necessary knowledge and tools to make informed decisions as they navigate effectively analyzing and interpreting data available in the ABCD Study.
 
-title: Manuscript Working Draft
+author-summary: |
+  The Adolescent Brain Cognitive Development (ABCD) Study offers valuable insights into adolescent brain development.
+Researchers working with this data face methodological and analytic concerns, including sample size, attrition, missing data, and measurement.
+This review addresses complexities in analyzing longitudinal data, with a focus on statistical models accounting for correlations in repeated measurements and their assumptions.
+We emphasize the importance of choosing appropriate models and analytic functions based on the data and the specific scientific question to be addressed.
+This concise resource aims  to help researchers navigate longitudinal data analysis, make informed methodological decisions, and enhance the accuracy and reliability of their findings when working with data from the ABCD Study.
 
-format: 
-  pdf: 
-    geometry: 
-      - top=20mm
-      - left=20mm
-  docx: default
-prefer-html: true
-
+bibliography: references.bib
 ---
@@
 
----
+# General guidelines for this Quarto template
 
-### References
+This template shows how to use PLOS template from <https://plos.org/resources/writing-center/>. Each journal have a submission guideline page, please refer to it.
 
-::: {#refs}
+ - [PLOS Biology](https://journals.plos.org/plosbiology/s/submission-guidelines) 
+ - [PLOS Climate](https://journals.plos.org/climate/s/submission-guidelines)
+ - [PLOS Digital Health](https://journals.plos.org/digitalhealth/s/submission-guidelines)
+ - [PLOS Computational Biology](https://journals.plos.org/ploscompbiol/s/submission-guidelines)
+ - [PLOS Genetics](https://journals.plos.org/plosgenetics/s/submission-guidelines)
+ - [PLOS Global Public Health](https://journals.plos.org/globalpublichealth/s/submission-guidelines)
+ - [PLOS Medicine](https://journals.plos.org/plosmedicine/s/submission-guidelines)
+ - [PLOS Neglected Tropical Diseases](https://journals.plos.org/plosntds/s/submission-guidelines)
+ - [PLOS ONE](https://journals.plos.org/plosone/s/submission-guidelines)
+ - [PLOS Pathogens](https://journals.plos.org/plospathogens/s/submission-guidelines)
+ - [PLOS Sustainability and Transformation](https://journals.plos.org/sustainabilitytransformation/s/submission-guidelines)
+ - [PLOS Water](https://journals.plos.org/water/s/submission-guidelines)
+
+This template file contains some guidelines and recommandation initially given in `plos_latex_template.tex` that can be found in <https://github.com/quarto-journals/plos/blob/main/style-guide/plos_latex_template.tex>
+
+## Metadata
+
+### About journal id field
+
+This is an identifier for the target journal.
+It can be derived from https://plos.org/resources/writing-center/ following submission guidelines link, the identifier is the part of the URL after `https://journals.plos.org/<id>/s/submission-guidelines`
+
+| Journal                               | id                          |
+|:--------------------------------------|:----------------------------|
+|PLOS Biology                           |plosbiology                  |
+|PLOS Climate                           |climate                      |
+|PLOS Digital Health                    |digitalhealth                |
+|PLOS Computational Biology             |ploscompbiol                 |
+|PLOS Genetics                          |plosgenetics                 |
+|PLOS Global Public Health              |globalpublichealth           |
+|PLOS Medicine                          |plosmedicine                 |
+|PLOS Neglected Tropical Diseases       |plosntds                     |
+|PLOS ONE                               |plosone                      |
+|PLOS Pathogens                         |plospathogens                |
+|PLOS Sustainability and Transformation |sustainabilitytransformation |
+|PLOS Water                             |water                        |
+
+Example :
+
+```yaml
+format:
+  plos-pdf:
+    journal:
+      id: water
+```
+
+## Once your paper is accepted for publication,
+
+Do not include track change in LaTeX file and leave only the final text of your manuscript. 
+PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+Visit <https://www.ctan.org/pkg/latexdiff?lang=en> for info or contact us at <latex@plos.org>.
+
+_This should not be a problem using Quarto but still a recommandation from the journal_
+
+There are no restrictions on package use within the LaTeX files except that no packages listed in the template may be deleted.
+
+Please do not include colors or graphics in the text. Color can be used to apply background shading to table cells only.
+
+The manuscript LaTeX source should be contained within a single file (do not use` \input`, `\externaldocument`, or similar commands).
+
+Please contact <latex@plos.org> with any questions submission guidelines.
+For anything Quarto related, please open an issue in <https://github.com/quarto-journals/plos>.
+If this is related to the LaTeX template, this could also be a good idea to contact PLOS directly.
+
+## Figures and Tables
+
+Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+
+### Figures
+
+However, do not include graphics in your manuscript
+
+- Figures should be uploaded separately from your manuscript file. 
+- Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+- Figures containing multiple panels/subfigures must be combined into one image file before submission.
+
+**This means that, depending on how you create your figure, a manual post processing will be required.**
+
+For figure citations, please use "Fig" instead of "Figure". This has been made the default in this Quarto format:
+
+```yaml
+crossref:
+  fig-title: Fig 
+```
+
+Also, place figure captions after the first paragraph in which they are cited.
+
+See PLOS figure guidelines at <https://journals.plos.org/plosone/s/figures> and in your specific journal guideline.
+
+### Tables
+
+Tables should be cell-based and may not contain:
+
+- spacing/line breaks within cells to alter layout or alignment
+- do not nest tabular environments (no tabular environments within tabular environments)
+- no graphics or colored text (cell background color/shading OK)
+
+See PLOS table guidelines at <http://journals.plos.org/plosone/s/tables> and in your specific journal guideline.
+
+For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+If you are in this case, you'll either need to manually post process the `.tex` file and recreate the PDF, or you need to include LaTeX tables directly.
+
+Also, place tables after the first paragraph in which they are cited.
+
+## Equations, math symbols, subscripts, and superscripts
+
+Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+
+*  For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, `x$^2$` is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+
+* Do not include text that is not math in the math environment. For example, `CO2` should be written as `CO\textsubscript{2}` giving CO\textsubscript{2} instead of `CO$_2$`.
+* Please add line breaks to long display equations when possible in order to fit size of the column. 
+* For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+* When adding superscript or subscripts outside of brackets/braces, please group using `{}`.  For example, change `"[U(D,E,\gamma)]^2"` to `"{[U(D,E,\gamma)]}^2"`.  
+* Do not use `\cal` for caligraphic font. Instead, use `\mathcal{}`
+
+## Title and headings
+
+Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+
+PLOS does not support heading levels beyond the 3rd, meaning no 4th level headings. Header 4 levels `####` is used for the  _Supporting information_ section
+
+## Abstract and author summary
+
+Abstract must be kept below 300 words.
+
+Author Summary must be kept between 150 and 200 words and first person must be used.
+
+For PLOS ONE, author summary won't be included as it is not valid for submission.
+
+## Supplementary information syntax
+
+Use this markdown syntax to create the supplementary information block with a custom block of class `.supp`
+```markdown
+::: {.supp}
+## SI TYPE {#id}
+
+First paragraph is a title sentence that will be bold. (required)
+
+Optionnaly, add descriptive text after the title of the
+item. No third paragraph is allowed
 :::
+```
+
+They need to be referenced in text using `nameref` by using this syntax `[id](.nameref)` where `ìd` will be the id used on the header.
+
+## References
+
+Within Quarto, `natbib` will be used with `plos2015.bst`, which expect numeric style citation. Use brackets for references, e.g `[@ref]`.
+
+## Quarto features limitation
+
+Some features are not working with this format in PDF:
+
+* Callouts
+* Code highlighting customization (border left, background color)
+
+----
+
+> **Following content of this document is from the LaTeX template content to demo journal style.**
+
+----
+
+# Introduction
+
+Lorem ipsum dolor sit [@bib1] amet, consectetur adipiscing elit.
+Curabitur eget porta erat. Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam
+@eq-schemeP sapien tellus, commodo id velit id, eleifend
+volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel,
+pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor
+faucibus id. [@bib2] Nunc convallis sodales ante, ut ullamcorper est
+egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar,
+volutpat risus.
+
+$$
+\begin{aligned}
+\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}}
+\end{aligned}
+$$ {#eq-schemeP}
+
+# Materials and methods
+
+## Etiam eget sapien nibh
+
+Nulla mi mi, `Fig~\ref{fig1}`{=latex} venenatis sed ipsum varius, volutpat euismod diam.
+Proin rutrum vel massa non gravida. Quisque tempor sem et dignissim
+rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
+at justo vitae nulla elementum commodo eu id massa. In vitae diam ac
+augue semper tincidunt eu ut eros. Fusce fringilla erat porttitor lectus
+cursus, vel sagittis arcu lobortis. Aliquam in enim semper, aliquam
+massa id, cursus neque. Praesent faucibus semper libero.
+
+```{=tex}
+% Place figure captions after the first paragraph in which they are cited.
+\begin{figure}[!h]
+\caption{{\bf Bold the figure title.}
+Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+\label{fig1}
+\end{figure}
+```
+
+# Results
+
+Results and Discussion can be combined.
+
+Nulla mi mi, venenatis sed ipsum varius, `Table~\ref{table1}`{=latex} volutpat euismod diam. 
+Proin rutrum vel massa non gravida. Quisque
+tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur
+adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id
+massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce
+fringilla erat porttitor lectus cursus, [s1-video]{.nameref} vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque. Praesent
+faucibus semper libero.
+
+```{=tex}
+% Place tables after the first paragraph in which they are cited.
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
+\begin{tabular}{|l+l|l|l|l|l|l|l|}
+\hline
+\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
+$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
+$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
+$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
+\end{tabular}
+\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+\end{flushleft}
+\label{table1}
+\end{adjustwidth}
+\end{table}
+````
+
+## **LOREM** and **IPSUM** nunc blandit a tortor
+
+### 3rd level heading
+
+Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget
+sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod
+ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.
+Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed
+velit. Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur
+adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed
+suscipit iaculis mi at feugiat.
+
+1.  react
+
+2.  diffuse free particles
+
+3.  increment time by dt and go to 1
+
+## Sed ac quam id nisi malesuada congue
+
+Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin
+rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo
+vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper
+tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel
+sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus
+neque. Praesent faucibus semper libero.
+
+-   First bulleted item.
+
+-   Second bulleted item.
+
+-   Third bulleted item.
+
+# Discussion
+
+Nulla mi mi, venenatis sed ipsum varius,see `Table~\ref{table1}`{=latex}
+volutpat euismod diam. Proin rutrum vel massa non gravida. Quisque
+tempor sem et dignissim rutrum. Lorem ipsum dolor sit amet, consectetur
+adipiscing elit. Morbi at justo vitae nulla elementum commodo eu id
+massa. In vitae diam ac augue semper tincidunt eu ut eros. Fusce
+fringilla erat porttitor lectus cursus, vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque. Praesent
+faucibus semper libero [@bib3].
+
+# Conclusion
+
+CO~2~ Maecenas convallis mauris sit amet sem ultrices gravida. Etiam
+eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod
+ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.
+Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed
+velit.
+
+Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur
+adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed
+suscipit iaculis mi at feugiat. Ut neque ipsum, luctus id lacus ut,
+laoreet scelerisque urna. Phasellus venenatis, tortor nec vestibulum
+mattis, massa tortor interdum felis, nec pellentesque metus tortor nec
+nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed. Nam
+condimentum sem eget mollis euismod. Nullam dui urna, gravida venenatis
+dui et, tincidunt sodales ex. Nunc est dui, sodales sed mauris nec,
+auctor sagittis leo. Aliquam tincidunt, ex in facilisis elementum,
+libero lectus luctus est, non vulputate nisl augue at dolor. For more
+information, see [s1-appendix]{.nameref}.
+
+# Supporting information
+
+::: {.supp}
+## S1 Fig. {#s1-fig}
+
+Bold the title sentence.
+
+Add descriptive text after the title of the item (optional).
+:::
+
+::: {.supp}
+## S2 Fig. {#s2-fig}
+
+Lorem ipsum.
+
+Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+:::
+
+::: {.supp}
+## S1 File. {#s1-file}
+
+Lorem ipsum.
+
+:::
+
+::: {.supp}
+
+## S1 Video. {#s1-video}
+
+Lorem ipsum.
+
+Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+:::
+
+::: {.supp}
+## S1 Appendix. {#s1-appendix}
+
+Lorem ipsum.
+
+Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+:::
+
+::: {.supp}
+## S1 Table. {#s1-table}
+
+Lorem ipsum.
+
+Maecenas convallis mauris sit amet sem ultrices gravida.
+Etiam eget sapien nibh.
+Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula.
+Curabitur fringilla pulvinar lectus consectetur pellentesque.
+:::
+
+# Acknowledgments
+
+Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
+Interdum et malesuada fames ac ante ipsum primis in faucibus. Nam id
+pretium nisi. Sed ac quam id nisi malesuada congue. Sed interdum aliquet
+augue, at pellentesque quam rhoncus vitae.
+
+zzzzz

### 2023-04-21 test.tex
cli commit 7

--- 2023-04-21 manuscript.tex
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{241,243,245}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
+\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
   pdftitle={Longitudinal Analysis Manuscript: Working Draft},
-  pdfauthor={Samuel W. Hawes; with the Lorem Ipsum Consortium},
+  pdfauthor={Samuel W. Hawes, PhD.; Andrew K. Littlefield, PhD.; Daniel A. Lopez; Kenneth J. Sher; Wesley K. Thompson; Additional Co-authors1; Additional Co-authors2},
   hidelinks,
@@
 % Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-Samuel W. Hawes\textsuperscript{1*}, with the Lorem Ipsum
-Consortium\textsuperscript{\textpilcrow}
+Samuel W. Hawes, PhD.\textsuperscript{1*}, Andrew K. Littlefield,
+PhD.\textsuperscript{2}, Daniel A. Lopez\textsuperscript{3}, Kenneth J.
+Sher\textsuperscript{4}, Wesley K.
+Thompson\textsuperscript{5}, Additional
+Co-authors1\textsuperscript{1}, Additional
+Co-authors2\textsuperscript{1}
 \\
@@
 \textbf{1} Center for Children and Families, Florida International
-University, Miami, Florida, United States, 
+University, Miami, FL, United States, \\ \textbf{2} Psychological
+Sciences, Texas Tech University, Lubbock, TX, United
+States, \\ \textbf{3} Department of Public Health, University of
+Rochester Medical Center, Rochester, NY, United
+States, \\ \textbf{4} Psychological Sciences, University of
+Missouri, Columbia, MO, United States, \\ \textbf{5} Population
+Neuroscience and Genetics Center, Laureate Institute for Brain
+Research, Tulsa, OK, United States, 
 \bigskip
@@
 \section*{Abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
-porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
-dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
-velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
-finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
-diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
-ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
-pulvinar, volutpat risus.
+The Adolescent Brain Cognitive Development (ABCD) Study presents a
+unique opportunity for researchers to investigate developmental
+processes in a large, diverse cohort of children and adolescents. Given
+the complex nature of the longitudinal data collected in the ABCD Study,
+researcher are likely to encounter a myriad of methodological and
+analytic considerations and concerns. This review provides a
+comprehensive examination of key issues and techniques related to
+longitudinal data analysis, specifically focusing on the ABCD Study. The
+text discusses model assumptions, common violations (e.g., independent
+and identically distributed residuals, \emph{heterogeneous}
+\emph{variability}) and their implications for valid inference. The
+importance of appropriately modeling covariance structures,
+understandings trade-offs between model fit and parsimony, and
+challenges related to sample size, attrition, missing data are
+highlighted. Consideration is given to the importance of selecting
+appropriate statistical models to account for correlations in repeated
+measurements and the assumptions underlying these models. The review
+also differentiates between linear and non-linear models in the context
+of continuous and discrete data, emphasizing various distributional
+assumptions and the necessity of choosing appropriate models and
+statistical methods. By addressing these complexities, the review seeks
+to equip researchers with the necessary knowledge and tools to make
+informed decisions as they navigate effectively analyzing and
+interpreting data available in the ABCD Study.
 
 \section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
-porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
-dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
-velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
-finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
-diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
-ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
-pulvinar, volutpat risus.
-
-\linenumbersxxxxx
-
-\#title: ``Longitudinal Analaysis Manuscript- Working Draft'' \# format:
-\# pdf: \# toc: true \# number-sections: true \# colorlinks: true \#
-geometry: \# - top=20mm \# - left=20mm \# docx: default \# prefer-html:
-true
-
-xxxxx
-
-\hypertarget{introduction}{%
+The Adolescent Brain Cognitive Development (ABCD) Study offers valuable
+insights into adolescent brain development. Researchers working with
+this data face methodological and analytic concerns, including sample
+size, attrition, missing data, and measurement. This review addresses
+complexities in analyzing longitudinal data, with a focus on statistical
+models accounting for correlations in repeated measurements and their
+assumptions. We emphasize the importance of choosing appropriate models
+and analytic functions based on the data and the specific scientific
+question to be addressed. This concise resource aims to help researchers
+navigate longitudinal data analysis, make informed methodological
+decisions, and enhance the accuracy and reliability of their findings
+when working with data from the ABCD Study.
+
+\linenumbers\hypertarget{introduction}{%
 \section{Introduction}\label{introduction}}
 
-label\{sec:headings\} The Adolescent Brain Cognitive Development (ABCD)
+\label{sec:headings} The Adolescent Brain Cognitive Development (ABCD)
 Study® is the largest long-term investigation of neurodevelopment and
@@
 
-label\{sec:headings\} There are several important concepts to consider
+\label{sec:headings} There are several important concepts to consider
 when conducting longitudinal analyses in a developmental context. These
@@
 
-label\{sec:headings\} \textbf{Defining Features of Longitudinal Data
+\label{sec:headings} \textbf{Defining Features of Longitudinal Data
 Analysis.} The hallmark characteristic of longitudinal data analysis is
@@
 
+\hypertarget{general-guidelines-for-this-quarto-template}{%
+\section{General guidelines for this Quarto
+template}\label{general-guidelines-for-this-quarto-template}}
+
+This template shows how to use PLOS template from
+\url{https://plos.org/resources/writing-center/}. Each journal have a
+submission guideline page, please refer to it.
+
+\begin{itemize}
+\tightlist
+\item
+  \href{https://journals.plos.org/plosbiology/s/submission-guidelines}{PLOS
+  Biology}
+\item
+  \href{https://journals.plos.org/climate/s/submission-guidelines}{PLOS
+  Climate}
+\item
+  \href{https://journals.plos.org/digitalhealth/s/submission-guidelines}{PLOS
+  Digital Health}
+\item
+  \href{https://journals.plos.org/ploscompbiol/s/submission-guidelines}{PLOS
+  Computational Biology}
+\item
+  \href{https://journals.plos.org/plosgenetics/s/submission-guidelines}{PLOS
+  Genetics}
+\item
+  \href{https://journals.plos.org/globalpublichealth/s/submission-guidelines}{PLOS
+  Global Public Health}
+\item
+  \href{https://journals.plos.org/plosmedicine/s/submission-guidelines}{PLOS
+  Medicine}
+\item
+  \href{https://journals.plos.org/plosntds/s/submission-guidelines}{PLOS
+  Neglected Tropical Diseases}
+\item
+  \href{https://journals.plos.org/plosone/s/submission-guidelines}{PLOS
+  ONE}
+\item
+  \href{https://journals.plos.org/plospathogens/s/submission-guidelines}{PLOS
+  Pathogens}
+\item
+  \href{https://journals.plos.org/sustainabilitytransformation/s/submission-guidelines}{PLOS
+  Sustainability and Transformation}
+\item
+  \href{https://journals.plos.org/water/s/submission-guidelines}{PLOS
+  Water}
+\end{itemize}
+
+This template file contains some guidelines and recommandation initially
+given in \texttt{plos\_latex\_template.tex} that can be found in
+\url{https://github.com/quarto-journals/plos/blob/main/style-guide/plos_latex_template.tex}
+
+\hypertarget{metadata}{%
+\subsection{Metadata}\label{metadata}}
+
+\hypertarget{about-journal-id-field}{%
+\subsubsection{About journal id field}\label{about-journal-id-field}}
+
+This is an identifier for the target journal. It can be derived from
+https://plos.org/resources/writing-center/ following submission
+guidelines link, the identifier is the part of the URL after
+\texttt{https://journals.plos.org/\textless{}id\textgreater{}/s/submission-guidelines}
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule\noalign{}
+Journal & id \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+PLOS Biology & plosbiology \\
+PLOS Climate & climate \\
+PLOS Digital Health & digitalhealth \\
+PLOS Computational Biology & ploscompbiol \\
+PLOS Genetics & plosgenetics \\
+PLOS Global Public Health & globalpublichealth \\
+PLOS Medicine & plosmedicine \\
+PLOS Neglected Tropical Diseases & plosntds \\
+PLOS ONE & plosone \\
+PLOS Pathogens & plospathogens \\
+PLOS Sustainability and Transformation & sustainabilitytransformation \\
+PLOS Water & water \\
+\end{longtable}
+
+Example :
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{format}\KeywordTok{:}
+\AttributeTok{  }\FunctionTok{plos{-}pdf}\KeywordTok{:}
+\AttributeTok{    }\FunctionTok{journal}\KeywordTok{:}
+\AttributeTok{      }\FunctionTok{id}\KeywordTok{:}\AttributeTok{ water}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{once-your-paper-is-accepted-for-publication}{%
+\subsection{Once your paper is accepted for
+publication,}\label{once-your-paper-is-accepted-for-publication}}
+
+Do not include track change in LaTeX file and leave only the final text
+of your manuscript. PLOS recommends the use of latexdiff to track
+changes during review, as this will help to maintain a clean tex file.
+Visit \url{https://www.ctan.org/pkg/latexdiff?lang=en} for info or
+contact us at \href{mailto:latex@plos.org}{\nolinkurl{latex@plos.org}}.
+
+\emph{This should not be a problem using Quarto but still a
+recommandation from the journal}
+
+There are no restrictions on package use within the LaTeX files except
+that no packages listed in the template may be deleted.
+
+Please do not include colors or graphics in the text. Color can be used
+to apply background shading to table cells only.
+
+The manuscript LaTeX source should be contained within a single file (do
+not use\texttt{\textbackslash{}input},
+\texttt{\textbackslash{}externaldocument}, or similar commands).
+
+Please contact \href{mailto:latex@plos.org}{\nolinkurl{latex@plos.org}}
+with any questions submission guidelines. For anything Quarto related,
+please open an issue in \url{https://github.com/quarto-journals/plos}.
+If this is related to the LaTeX template, this could also be a good idea
+to contact PLOS directly.
+
+\hypertarget{figures-and-tables}{%
+\subsection{Figures and Tables}\label{figures-and-tables}}
+
+Please include tables/figure captions directly after the paragraph where
+they are first cited in the text.
+
+\hypertarget{figures}{%
+\subsubsection{Figures}\label{figures}}
+
+However, do not include graphics in your manuscript
+
+\begin{itemize}
+\tightlist
+\item
+  Figures should be uploaded separately from your manuscript file.
+\item
+  Figures generated using LaTeX should be extracted and removed from the
+  PDF before submission.
+\item
+  Figures containing multiple panels/subfigures must be combined into
+  one image file before submission.
+\end{itemize}
+
+\textbf{This means that, depending on how you create your figure, a
+manual post processing will be required.}
+
+For figure citations, please use ``Fig'' instead of ``Figure''. This has
+been made the default in this Quarto format:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{crossref}\KeywordTok{:}
+\AttributeTok{  }\FunctionTok{fig{-}title}\KeywordTok{:}\AttributeTok{ Fig }
+\end{Highlighting}
+\end{Shaded}
+
+Also, place figure captions after the first paragraph in which they are
+cited.
+
+See PLOS figure guidelines at
+\url{https://journals.plos.org/plosone/s/figures} and in your specific
+journal guideline.
+
+\hypertarget{tables}{%
+\subsubsection{Tables}\label{tables}}
+
+Tables should be cell-based and may not contain:
+
+\begin{itemize}
+\tightlist
+\item
+  spacing/line breaks within cells to alter layout or alignment
+\item
+  do not nest tabular environments (no tabular environments within
+  tabular environments)
+\item
+  no graphics or colored text (cell background color/shading OK)
+\end{itemize}
+
+See PLOS table guidelines at
+\url{http://journals.plos.org/plosone/s/tables} and in your specific
+journal guideline.
+
+For tables that exceed the width of the text column, use the adjustwidth
+environment as illustrated in the example table in text below. If you
+are in this case, you'll either need to manually post process the
+\texttt{.tex} file and recreate the PDF, or you need to include LaTeX
+tables directly.
+
+Also, place tables after the first paragraph in which they are cited.
+
+\hypertarget{equations-math-symbols-subscripts-and-superscripts}{%
+\subsection{Equations, math symbols, subscripts, and
+superscripts}\label{equations-math-symbols-subscripts-and-superscripts}}
+
+Below are a few tips to help format your equations and other special
+characters according to our specifications. For more tips to help reduce
+the possibility of formatting errors during conversion, please see our
+LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+
+\begin{itemize}
+\item
+  For inline equations, please be sure to include all portions of an
+  equation in the math environment. For example, \texttt{x\$\^{}2\$} is
+  incorrect; this should be formatted as \(x^2\) (or \(\mathrm{x}^2\) if
+  the romanized font is desired).
+\item
+  Do not include text that is not math in the math environment. For
+  example, \texttt{CO2} should be written as
+  \texttt{CO\textbackslash{}textsubscript\{2\}} giving
+  CO\textsubscript{2} instead of \texttt{CO\$\_2\$}.
+\item
+  Please add line breaks to long display equations when possible in
+  order to fit size of the column.
+\item
+  For inline equations, please do not include punctuation (commas, etc)
+  within the math environment unless this is part of the equation.
+\item
+  When adding superscript or subscripts outside of brackets/braces,
+  please group using \texttt{\{\}}. For example, change
+  \texttt{"{[}U(D,E,\textbackslash{}gamma){]}\^{}2"} to
+  \texttt{"\{{[}U(D,E,\textbackslash{}gamma){]}\}\^{}2"}.\\
+\item
+  Do not use \texttt{\textbackslash{}cal} for caligraphic font. Instead,
+  use \texttt{\textbackslash{}mathcal\{\}}
+\end{itemize}
+
+\hypertarget{title-and-headings}{%
+\subsection{Title and headings}\label{title-and-headings}}
+
+Please use ``sentence case'' for title and headings (capitalize only the
+first word in a title (or heading), the first word in a subtitle (or
+subheading), and any proper nouns).
+
+PLOS does not support heading levels beyond the 3rd, meaning no 4th
+level headings. Header 4 levels \texttt{\#\#\#\#} is used for the
+\emph{Supporting information} section
+
+\hypertarget{abstract-and-author-summary}{%
+\subsection{Abstract and author
+summary}\label{abstract-and-author-summary}}
+
+Abstract must be kept below 300 words.
+
+Author Summary must be kept between 150 and 200 words and first person
+must be used.
+
+For PLOS ONE, author summary won't be included as it is not valid for
+submission.
+
+\hypertarget{supplementary-information-syntax}{%
+\subsection{Supplementary information
+syntax}\label{supplementary-information-syntax}}
+
+Use this markdown syntax to create the supplementary information block
+with a custom block of class \texttt{.supp}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{::: \{.supp\}}
+\FunctionTok{\#\# SI TYPE \{\#id\}}
+
+\NormalTok{First paragraph is a title sentence that will be bold. (required)}
+
+\NormalTok{Optionnaly, add descriptive text after the title of the}
+\NormalTok{item. No third paragraph is allowed}
+\NormalTok{:::}
+\end{Highlighting}
+\end{Shaded}
+
+They need to be referenced in text using \texttt{nameref} by using this
+syntax \texttt{{[}id{]}(.nameref)} where \texttt{ìd} will be the id used
+on the header.
+
+\hypertarget{references}{%
+\subsection{References}\label{references}}
+
+Within Quarto, \texttt{natbib} will be used with \texttt{plos2015.bst},
+which expect numeric style citation. Use brackets for references, e.g
+\texttt{{[}@ref{]}}.
+
+\hypertarget{quarto-features-limitation}{%
+\subsection{Quarto features
+limitation}\label{quarto-features-limitation}}
+
+Some features are not working with this format in PDF:
+
+\begin{itemize}
+\tightlist
+\item
+  Callouts
+\item
+  Code highlighting customization (border left, background color)
+\end{itemize}
+
 \begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
 
-\hypertarget{references}{%
-\subsubsection{References}\label{references}}
-
-\renewcommand{\bibsection}{}
-\bibliography{references.bib,bibliography.bib}
+\begin{quote}
+\textbf{Following content of this document is from the LaTeX template
+content to demo journal style.}
+\end{quote}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\hypertarget{introduction-1}{%
+\section{Introduction}\label{introduction-1}}
+
+Lorem ipsum dolor sit~\citep{bib1} amet, consectetur adipiscing elit.
+Curabitur eget porta erat. Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam
+Eq~\ref{eq-schemeP} sapien tellus, commodo id velit id, eleifend
+volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel,
+pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor
+faucibus id.~\citep{bib2} Nunc convallis sodales ante, ut ullamcorper
+est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar,
+volutpat risus.
+
+\begin{equation}\protect\hypertarget{eq-schemeP}{}{
+\begin{aligned}
+\mathrm{P_Y} = \underbrace{H(Y_n) - H(Y_n|\mathbf{V}^{Y}_{n})}_{S_Y} + \underbrace{H(Y_n|\mathbf{V}^{Y}_{n})- H(Y_n|\mathbf{V}^{X,Y}_{n})}_{T_{X\rightarrow Y}}
+\end{aligned}
+}\label{eq-schemeP}\end{equation}
+
+\hypertarget{materials-and-methods}{%
+\section{Materials and methods}\label{materials-and-methods}}
+
+\hypertarget{etiam-eget-sapien-nibh}{%
+\subsection{Etiam eget sapien nibh}\label{etiam-eget-sapien-nibh}}
+
+Nulla mi mi, Fig~\ref{fig1} venenatis sed ipsum varius, volutpat euismod
+diam. Proin rutrum vel massa non gravida. Quisque tempor sem et
+dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing
+elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae
+diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat
+porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim
+semper, aliquam massa id, cursus neque. Praesent faucibus semper libero.
+
+% Place figure captions after the first paragraph in which they are cited.
+\begin{figure}[!h]
+\caption{{\bf Bold the figure title.}
+Figure caption text here, please use this space for the figure panel descriptions instead of using subfigure commands. A: Lorem ipsum dolor sit amet. B: Consectetur adipiscing elit.}
+\label{fig1}
+\end{figure}
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+Results and Discussion can be combined.
+
+Nulla mi mi, venenatis sed ipsum varius, Table~\ref{table1} volutpat
+euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et
+dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing
+elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae
+diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat
+porttitor lectus cursus, \nameref{s1-video} vel sagittis arcu lobortis.
+Aliquam in enim semper, aliquam massa id, cursus neque. Praesent
+faucibus semper libero.
+
+% Place tables after the first paragraph in which they are cited.
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
+\centering
+\caption{
+{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
+\begin{tabular}{|l+l|l|l|l|l|l|l|}
+\hline
+\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
+$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
+$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
+$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
+\end{tabular}
+\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl.
+Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
+\end{flushleft}
+\label{table1}
+\end{adjustwidth}
+\end{table}
+
+\hypertarget{lorem-and-ipsum-nunc-blandit-a-tortor}{%
+\subsection{\texorpdfstring{\textbf{LOREM}~and \textbf{IPSUM}~nunc
+blandit a
+tortor}{LOREM~and IPSUM~nunc blandit a tortor}}\label{lorem-and-ipsum-nunc-blandit-a-tortor}}
+
+\hypertarget{rd-level-heading}{%
+\subsubsection{3rd level heading}\label{rd-level-heading}}
+
+Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget
+sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod
+ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.
+Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed
+velit. Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur
+adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed
+suscipit iaculis mi at feugiat.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  react
+\item
+  diffuse free particles
+\item
+  increment time by dt and go to 1
+\end{enumerate}
+
+\hypertarget{sed-ac-quam-id-nisi-malesuada-congue}{%
+\subsection{Sed ac quam id nisi malesuada
+congue}\label{sed-ac-quam-id-nisi-malesuada-congue}}
+
+Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam. Proin
+rutrum vel massa non gravida. Quisque tempor sem et dignissim rutrum.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi at justo
+vitae nulla elementum commodo eu id massa. In vitae diam ac augue semper
+tincidunt eu ut eros. Fusce fringilla erat porttitor lectus cursus, vel
+sagittis arcu lobortis. Aliquam in enim semper, aliquam massa id, cursus
+neque. Praesent faucibus semper libero.
+
+\begin{itemize}
+\item
+  First bulleted item.
+\item
+  Second bulleted item.
+\item
+  Third bulleted item.
+\end{itemize}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+Nulla mi mi, venenatis sed ipsum varius,see Table~\ref{table1} volutpat
+euismod diam. Proin rutrum vel massa non gravida. Quisque tempor sem et
+dignissim rutrum. Lorem ipsum dolor sit amet, consectetur adipiscing
+elit. Morbi at justo vitae nulla elementum commodo eu id massa. In vitae
+diam ac augue semper tincidunt eu ut eros. Fusce fringilla erat
+porttitor lectus cursus, vel sagittis arcu lobortis. Aliquam in enim
+semper, aliquam massa id, cursus neque. Praesent faucibus semper
+libero~\citep{bib3}.
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices
+gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas
+ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus
+consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat
+eget, ullamcorper sed velit.
+
+Sed non aliquet felis. Lorem ipsum dolor sit amet, consectetur
+adipiscing elit. Mauris commodo justo ac dui pretium imperdiet. Sed
+suscipit iaculis mi at feugiat. Ut neque ipsum, luctus id lacus ut,
+laoreet scelerisque urna. Phasellus venenatis, tortor nec vestibulum
+mattis, massa tortor interdum felis, nec pellentesque metus tortor nec
+nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed. Nam
+condimentum sem eget mollis euismod. Nullam dui urna, gravida venenatis
+dui et, tincidunt sodales ex. Nunc est dui, sodales sed mauris nec,
+auctor sagittis leo. Aliquam tincidunt, ex in facilisis elementum,
+libero lectus luctus est, non vulputate nisl augue at dolor. For more
+information, see \nameref{s1-appendix}.
+
+\hypertarget{supporting-information}{%
+\section{Supporting information}\label{supporting-information}}
+
+\paragraph*{S1 Fig.}
+\label{s1-fig}
+{\textbf{Bold the title sentence.}} Add descriptive text after the title
+of the item (optional).
+
+\paragraph*{S2 Fig.}
+\label{s2-fig}
+{\textbf{Lorem ipsum.}} Maecenas convallis mauris sit amet sem ultrices
+gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas
+ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus
+consectetur pellentesque.
+
+\paragraph*{S1 File.}
+\label{s1-file}
+{\textbf{Lorem ipsum.}}
+
+\paragraph*{S1 Video.}
+\label{s1-video}
+{\textbf{Lorem ipsum.}} Maecenas convallis mauris sit amet sem ultrices
+gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas
+ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus
+consectetur pellentesque.
+
+\paragraph*{S1 Appendix.}
+\label{s1-appendix}
+{\textbf{Lorem ipsum.}} Maecenas convallis mauris sit amet sem ultrices
+gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas
+ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus
+consectetur pellentesque.
+
+\paragraph*{S1 Table.}
+\label{s1-table}
+{\textbf{Lorem ipsum.}} Maecenas convallis mauris sit amet sem ultrices
+gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas
+ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus
+consectetur pellentesque.
+
+\hypertarget{acknowledgments}{%
+\section{Acknowledgments}\label{acknowledgments}}
+
+Cras egestas velit mauris, eu mollis turpis pellentesque sit amet.
+Interdum et malesuada fames ac ante ipsum primis in faucibus. Nam id
+pretium nisi. Sed ac quam id nisi malesuada congue. Sed interdum aliquet
+augue, at pellentesque quam rhoncus vitae.
+
+zzzzz
 
 \nolinenumbers
+  \bibliography{references.bib}
 

### 2023-04-21 test/test.tex
cli commit 7

--- 2023-04-21 test.tex
@@
 \hypersetup{
-  pdftitle={Longitudinal Analysis Manuscript: Working Draft},
-  pdfauthor={Samuel W. Hawes, PhD.; Andrew K. Littlefield, PhD.; Daniel A. Lopez; Kenneth J. Sher; Wesley K. Thompson; Additional Co-authors1; Additional Co-authors2},
+  pdftitle={PLOS template},
+  pdfauthor={Name1 Surname; Name2 Surname; Name3 Surname; Name4 Surname; Name5 Surname; Name6 Surname; Name7 Surname; with the Lorem Ipsum Consortium},
   hidelinks,
@@
 {\Large
-\textbf\newline{Longitudinal Analysis Manuscript: Working
-Draft} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{PLOS
+template} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 % Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-Samuel W. Hawes, PhD.\textsuperscript{1*}, Andrew K. Littlefield,
-PhD.\textsuperscript{2}, Daniel A. Lopez\textsuperscript{3}, Kenneth J.
-Sher\textsuperscript{4}, Wesley K.
-Thompson\textsuperscript{5}, Additional
-Co-authors1\textsuperscript{1}, Additional
-Co-authors2\textsuperscript{1}
+Name1 Surname\textsuperscript{1,2\Yinyang}, Name2
+Surname\textsuperscript{2\Yinyang}, Name3
+Surname\textsuperscript{2,3}, Name4 Surname\textsuperscript{2}, Name5
+Surname\textsuperscript{2\dag}, Name6 Surname\textsuperscript{2}, Name7
+Surname\textsuperscript{1,2,3*}, with the Lorem Ipsum
+Consortium\textsuperscript{\textpilcrow}
 \\
 \bigskip
-\textbf{1} Center for Children and Families, Florida International
-University, Miami, FL, United States, \\ \textbf{2} Psychological
-Sciences, Texas Tech University, Lubbock, TX, United
-States, \\ \textbf{3} Department of Public Health, University of
-Rochester Medical Center, Rochester, NY, United
-States, \\ \textbf{4} Psychological Sciences, University of
-Missouri, Columbia, MO, United States, \\ \textbf{5} Population
-Neuroscience and Genetics Center, Laureate Institute for Brain
-Research, Tulsa, OK, United States, 
+\textbf{1} Affiliation Dept, Institution
+Name, City, State, Country, \\ \textbf{2} Affiliation Center,
+Institution Name, City, State, Country, \\ \textbf{3} Affiliation
+Program, Institution Name, City, State, Country, 
 \bigskip
@@
 % Group/Consortium Author Note
-\textpilcrow Biostatistics Working Group membership list can be found in
-the Acknowledgments sections
+\textpilcrow Membership list can be found in the Acknowledgments
+sections
 
 % Use the asterisk to denote corresponding authorship and provide email address in note below.
-* shawes@fiu.edu
+* correspondingauthor@institute.edu
 
@@
 \section*{Abstract}
-The Adolescent Brain Cognitive Development (ABCD) Study presents a
-unique opportunity for researchers to investigate developmental
-processes in a large, diverse cohort of children and adolescents. Given
-the complex nature of the longitudinal data collected in the ABCD Study,
-researcher are likely to encounter a myriad of methodological and
-analytic considerations and concerns. This review provides a
-comprehensive examination of key issues and techniques related to
-longitudinal data analysis, specifically focusing on the ABCD Study. The
-text discusses model assumptions, common violations (e.g., independent
-and identically distributed residuals, \emph{heterogeneous}
-\emph{variability}) and their implications for valid inference. The
-importance of appropriately modeling covariance structures,
-understandings trade-offs between model fit and parsimony, and
-challenges related to sample size, attrition, missing data are
-highlighted. Consideration is given to the importance of selecting
-appropriate statistical models to account for correlations in repeated
-measurements and the assumptions underlying these models. The review
-also differentiates between linear and non-linear models in the context
-of continuous and discrete data, emphasizing various distributional
-assumptions and the necessity of choosing appropriate models and
-statistical methods. By addressing these complexities, the review seeks
-to equip researchers with the necessary knowledge and tools to make
-informed decisions as they navigate effectively analyzing and
-interpreting data available in the ABCD Study.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
+porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
+dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
+velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
+finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
+diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
+ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
+pulvinar, volutpat risus.
 
 \section*{Author summary}
-The Adolescent Brain Cognitive Development (ABCD) Study offers valuable
-insights into adolescent brain development. Researchers working with
-this data face methodological and analytic concerns, including sample
-size, attrition, missing data, and measurement. This review addresses
-complexities in analyzing longitudinal data, with a focus on statistical
-models accounting for correlations in repeated measurements and their
-assumptions. We emphasize the importance of choosing appropriate models
-and analytic functions based on the data and the specific scientific
-question to be addressed. This concise resource aims to help researchers
-navigate longitudinal data analysis, make informed methodological
-decisions, and enhance the accuracy and reliability of their findings
-when working with data from the ABCD Study.
-
-\linenumbers\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-\label{sec:headings} The Adolescent Brain Cognitive Development (ABCD)
-Study® is the largest long-term investigation of neurodevelopment and
-child health in the United States. Conceived and initiated by the
-National Institutes of Health (NIH), this landmark prospective
-longitudinal study aims to transform our understanding of the genetic
-and environmental influences on brain development and their roles in
-behavioral and health outcomes in adolescents \citep{volkow2018}. At its
-heart, the study is designed to chart the course of human development
-across multiple, interacting domains from late childhood to early
-adulthood and to identify factors that lead to both positive and
-negative developmental outcomes. Central to achieving these goals is the
-ABCD Study's® commitment to an open science framework designed to
-facilitate access to and sharing of scientific knowledge by espousing
-practices that increase openness, integrity, and reproducibility of
-scientific research (e.g., public data releases). In this sense, the
-ABCD Study® is a collaboration with the larger research community, with
-the rich longitudinal nature of the ABCD Study dataset allowing
-researchers to perform a variety of analyses of both methodological and
-substantive interest. Together, this presents a unique opportunity to
-significantly advance our understanding of how a multitude of
-biopsychosocial processes emerge and unfold across critical periods of
-development.
-
-{[}section still be developed\ldots{]}
-
-\hypertarget{the-abcd-study-data}{%
-\subsection{The ABCD Study® Data}\label{the-abcd-study-data}}
-
-Participants enrolled in the ABCD Study include a large cohort of youth
-(n=11880) aged 9-10 years at baseline and their parents/guardians. The
-study sample was recruited from household populations in defined
-catchment areas for each of the 21 study sites across the United States
-(information regarding funding agencies, recruitment sites,
-investigators, and project organization can be obtained at the ABCD
-Study website). The ABCD Study is collecting longitudinal data on a rich
-variety of outcomes that will enable the construction of
-realistically-complex etiological models by incorporating factors from
-many domains simultaneously. Each new wave of data collection provides
-the building blocks for conducting probing longitudinal analyses that
-allow us to characterize normative development, identify variables that
-presage deviations from prototypic development, and assess a range of
-outcomes associated with variables of interest. This data includes a
-neurocognitive battery \citep{luciana2018a, thompson2019}, mental and
-physical health assessments \citep{barch2018}, measures of culture and
-environment \citep{zucker2018}, substance use {[}add citation{]},
-biospecimens \citep{uban2018}, structural and functional brain imaging
-\citep{casey2018, hagler2019}, geolocation-based environmental exposure
-data, wearables, and mobile technology \citep{bagot2018}, and whole
-genome genotyping \citep{loughnan2020}. Many of these measures are
-collected at in-person annual visits, with brain imaging collected at
-baseline and every other year going forward. A limited number of
-assessments are collected in semi-annual telephone interviews between
-in-person visits. Data are publicly released on an annual basis through
-the NIMH Data Archive. By necessity, the study's earliest data releases
-were cross-sectional (i.e., the baseline data), however, the most recent
-public data release (NDA Release 4.0) contains data collected across
-three annual assessments, including two imaging assessments (baseline
-and year 2 follow-up visits).
-
-\hypertarget{organization-of-current-manuscript}{%
-\subsection{Organization of current
-manuscript}\label{organization-of-current-manuscript}}
-
-The rich longitudinal nature of the ABCD Study dataset will allow
-researchers to perform analyses of both methodological and substantive
-interest. This report describes methods for longitudinal analyses of
-ABCD Study data that can address its fundamental scientific aims, as
-well as challenges inherent in a large population-based long-term study
-of adolescents. The manuscript is organized as follows:
-
-{[}section still be developed\ldots{]}
-
-\hypertarget{part-i-longitudinal-research-basic-concepts-and-considerations}{%
-\section{Part I: Longitudinal Research: Basic Concepts and
-Considerations}\label{part-i-longitudinal-research-basic-concepts-and-considerations}}
-
-\label{sec:headings} There are several important concepts to consider
-when conducting longitudinal analyses in a developmental context. These
-include different ways of thinking about developmental course, whether
-certain periods of development are relatively sensitive or insensitive
-to various types of insults or stressors, whether some time periods or
-situations inhibit the expression of individual differences due to
-extreme environmental pressures, and whether the same behavior
-manifested at different times represent the same phenomenon or different
-ones. Further, in the case of developmentally focused longitudinal
-research, each new measurement occasion not only provides a more
-extended portrait of the child's life course (and not just characterize
-growth during this period but also assesses the durability/chronicity of
-prior effects/consequences) but also brings with it greater
-methodological opportunities to exploit the statistical properties of
-longitudinal data in the furtherance of critical scientific questions.
-That is, we can ask more nuanced questions and make stronger inferences
-as our number of time-ordered observations grows, assuming we have
-assessed the ``right'' variables and the timings of our observations
-comport with the temporal dynamics of the mechanisms of interest.
-Appreciation of these and other issues can help to guide the analysis
-and interpretation of data and aid translation to clinical and public
-health applications.
-
-\textbf{Vulnerable periods}. Development normatively progresses from
-less mature to more mature levels of functioning. However, unique epochs
-and experiences can alter the course of this idealized form of
-development. Consider research that shows cannabis use during
-adolescence is associated with later psychosis to a greater degree than
-cannabis use initiated later in development {[}add citation{]}; or,
-similarly, experimental research on rodents that shows rodent brains to
-be especially sensitive to the neurotoxic effects of alcohol on brain
-structure and learning early in development (corresponding to early
-adolescence in humans){[}add citation{]}. These examples highlight the
-importance of considering the role of vulnerable periods -- temporal
-windows of rapid brain development or remodeling during which the
-effects of environmental stimuli (e.g.~cannabis exposure) on the
-developing brain may be particularly pronounced-- when trying to
-establish an accurate understanding of the association between exposures
-and outcomes.
-
-\textbf{Developmental disturbances}. Whereas vulnerable periods heighten
-neurobiological susceptibility to environmental influences, at other
-times environmental pressures will tend to suppress stability and
-disrupt the orderly stochastic process of normative development (e.g.,
-xxx-xxx). This situation reflects a developmental disturbance in that
-the normal course of development is ``disturbed'' for some time by some
-time-limited process. In such cases, we might find that prediction of
-behavior in the period of the disturbance is reduced and/or, similarly,
-the behavior exhibited during the disturbance might have less predictive
-power with respect to distal outcomes compared to the behavior exhibited
-before and following the disrupted period. That is, once the
-environmental stimuli are removed (or the individual is removed from the
-environment), individual differences are again more freely expressed and
-the autoregressive effects increase to levels similar to those before
-entering the environment.
-
-\textbf{Developmental snares and cascade effects}. Normative development
-can also be upended by experiences (e.g., drug use) that, through
-various mechanisms, disrupt the normal flow of development wherein each
-stage establishes a platform for the next. For instance, substance use
-could lead to association with deviant peers, precluding opportunities
-for learning various adaptive skills and prosocial behaviors, in effect,
-creating a ``snare'' that retards psychosocial development. Relatedly,
-the consequences of these types of events can cascade (e.g., school
-dropout, involvement in the criminal justice system) so that the effects
-of the snare are amplified. Although conceptually distinct from
-vulnerable periods, both of these types of developmental considerations
-highlight the importance of viewing behavior in the context of
-development and the importance of attempting to determine how various
-developmental pathways unfold.
-
-\textbf{Distinguishing developmental change from experience effects}.
-One can often observe systematic changes over time in a variable of
-interest and assume this change is attributable to development. To this
-point, cognitive abilities (e.g, verbal ability, problem-solving)
-normatively grow earlier in development and often decline in late life
-(e.g., memory, speed of processing). However, the observed patterns of
-growth and decline often differ between cross-sectional vs.~longitudinal
-effects \citep{salthouse2014} where subjects gain increasing experience
-with the assessment with each successive measurement occasion. Such
-experience effects on cognitive functioning have been demonstrated in
-adolescent longitudinal samples similar to ABCD \citep{sullivan2017} and
-highlight the need to consider these effects and address them
-analytically. In the case of performance-based measures {[}e.g., matrix
-reasoning related to neurocognitive functioning; see
-\citet{salthouse2014}{]}, this can be due to ``learning'' the task from
-previous test administrations (e.g., someone taking the test a second
-time performs better than they did the first time simply as a function
-of having taken it before). Even in the case of non-performance-based
-measures (e.g., levels of depression), where one cannot easily make the
-argument that one has acquired some task-specific skill through
-learning, it has been observed that respondents tend to endorse lower
-levels on subsequent assessments
-\citetext{\citealp[e.g.,][]{beck1961}; \citealp[see][]{french2010}} and
-this phenomenon has been well documented in research on structured
-diagnostic interviews \citep{robins1985}. While it is typically assumed
-that individuals are rescinding or telling us less information on
-follow-up interviews, there is reason to suspect that in some cases the
-initial assessment may be artefactually elevated
-\citep[see][]{shrout2018a}. Some designs (specifically, accelerated
-longitudinal designs) are especially well suited for discovering these
-effects and modeling them. While ABCD was not designed as an accelerated
-longitudinal design, the variability in age at the time of baseline
-recruitment (9 years, 0 months to 10 years, 11 months) allows some
-measures, collected every year, to be conceptualized as an accelerated
-longitudinal design. Moreover, it is possible that in later waves,
-patterns of longitudinal missing data will allow some analyses to assess
-the confounded effects of age and the number of prior assessments.
-However, ABCD is fundamentally a single-cohort, longitudinal design,
-where a number of prior assessments and age are highly confounded, and
-for, perhaps, most analyses, the possible influence of experience
-effects needs to be kept in mind.
-
-\hypertarget{part-ii-longitudinal-data-interpretation-issues-pitfalls-assumption}{%
-\section{Part II Longitudinal Data: Interpretation / Issues / Pitfalls
-\&
-Assumption}\label{part-ii-longitudinal-data-interpretation-issues-pitfalls-assumption}}
-
-\label{sec:headings} \textbf{Defining Features of Longitudinal Data
-Analysis.} The hallmark characteristic of longitudinal data analysis is
-its application to repeated assessments of the same assessment targets
-(e.g., individuals, families) across time. While the primary reason for
-collecting longitudinal data is in pursuit of addressing scientific
-questions, from a methodological perspective, having multiple
-observations over time allows researchers to identify potentially
-problematic observations when highly improbable longitudinal patterns
-are observed. That is, we can ask more nuanced questions and make
-stronger inferences as our number of time-ordered observations grows
-assuming we have assessed the ``right'' variables and the timings of our
-observations comport with the temporal dynamics of the mechanisms of
-interest .
-
-\hypertarget{modeling-data-across-two-time-points-versus-three-or-more-time-points.}{%
-\subsection{Modeling Data Across Two Time Points versus Three or More
-Time
-Points.}\label{modeling-data-across-two-time-points-versus-three-or-more-time-points.}}
-
-Although the clear leap to the realm of longitudinal data involves going
-from one assessment to two or more assessments, there are also notable
-distinctions in designs based on two-assessment points versus three or
-more measurement occasions. Just as cross-sectional data can be
-informative in some situations, two waves of data can be beneficial in
-contexts such as when experimental manipulation is involved (e.g.,
-pre/post tests), or if the central goal is prediction (e.g., trying to
-predict scores on Variable A at time T as a function of prior scores on
-Variable A and Variable B at time T-1). At the same time, data based on
-two assessments are inherently limited on multiple fronts. As
-\citep{rogosa1982} noted approximately forty years ago, ``Two waves of
-data are better than one, but maybe not much better''. These sentiments
-are reflected in more contemporary recommendations regarding
-best-practice guidelines for prospective data, which increasingly
-emphasize the benefits of additional measurement occasions for model
-identification and accurate parameter estimation. It is also consistent
-with research recommending that developmental studies include three or
-more assessment points, given it is impossible for data based on
-two-time points to determine the shape of development (since linear,
-straight line change is the only possible form, given two assessments;
-see \citep{duncan2009}). Research designs that include three or more
-time points allow for increasingly nuanced analyses that more adequately
-tease apart sources of variation and covariation among the repeated
-assessments \citep{king2018}-- a key aspect of inferential research. To
-illustrate, developmental theories are typically interested in
-understanding patterns of within-individual change over time (discussed
-in further detail, below); however, two data points provide meager
-information on change at the person level. This point is further
-underscored in a recent review of statistical models commonly touted as
-distinguishing within-individual vs between-individual sources of
-variance in which the study authors concluded ``\ldots{} researchers are
-limited when attempting to differentiate these sources of variation in
-psychological phenomenon when using two waves of data'' and perhaps more
-concerning, ``\ldots the models discussed here do not offer a feasible
-way to overcome these inherent limitations''
-\citet{littlefield20210603}. It is important to note, however, that
-despite the current focus on two-wave designs versus three or more
-assessment waves, garnering three assessment points is not a panacea for
-longitudinal modeling. Indeed, several contemporary longitudinal models
-designed to isolate within-individual variability {[}e.g., the Latent
-Curve Model with Structured Residuals; \citet{curran2014a}{]} require at
-least four assessments to parameterize fully and, more generally,
-increasingly accurate parameter estimates are obtained as more
-assessment occasions are used \citep{duncan2009}.
-
-\hypertarget{types-of-stability-and-change}{%
-\subsection{Types of stability and
-change}\label{types-of-stability-and-change}}
-
-If one were to try to sum up what development in a living organism is
-exactly, one could plausibly argue it's the characterization of
-stability and change as the organism traverses the life course. There
-are a few different ways to think of stability (and change). Consider we
-measure the height of all youth in a 6th-grade class, once in the fall
-at the beginning of the school year and once again in the spring at the
-end of the school year. A common first step may be to compare the
-class's average height values obtained at these two different
-measurement occasions. This comparison of the average scores for the
-same group of individuals at multiple time points is referred to as
-``mean-level'' stability as it provides information about continuity and
-change in the group level of an outcome of interest (e.g., height) over
-time. Another type of stability involves calculating the correlation
-between the values obtained at different time points (e.g., `height in
-the fall' with `height in the spring'). This type of ``rank-order''
-stability evaluates between-individual change by focusing on the degree
-to which individuals retain their relative placement in a group across
-time. Consider, someone who is the shortest person in their class in 6th
-grade may grow considerably over the school year (i.e., exhibit mean
-level change), but remain the shortest person among their classmates.
-That is, the individual is manifesting a type of rank-order stability.
-Both types of stability and change are important. Mean-level change in
-certain traits might help to explain why, in general, individuals are
-particularly vulnerable to social influences at some ages more than
-others; rank order change might help to quantify the extent to which
-certain characteristics of the individual are more trait-like. For
-example, in some areas of development, there is considerable mean-level
-change that occurs over time (e.g., changes in Big 5 personality
-traits), but relatively high rank-order stability. Despite the useful
-information afforded by examining mean-level and rank-order change,
-these approaches are limited in that they provide little information
-about patterns of ``within-individual'' change and, in turn, can result
-in fundamental misinterpretations about substantial or meaningful
-changes in an outcome of interest.
-
-There is growing recognition that statistical models commonly applied to
-longitudinal data often fail to comport with the developmental theory
-they are being used to assess (e.g., Curran, Lee, Howard, Lane, \&
-MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021. Specifically,
-developmental studies typically involve the use of prospective data to
-inform theories that are concerned with clear within-person (i.e.,
-intraindividual) processes (e.g., how phenotypes change or remain stable
-within individuals over time) \citep[e.g., see][]{curran2011}. Despite
-this, methods generally unsuited for disaggregating between- and
-within-person effects (e.g., cross-lagged panel models {[}CLPM{]})
-remain common within various extant literatures. As a result, experts
-increasingly caution about the need to xxxxxxxx {[}add citation{]}.
-Fortunately, there exists a range of models that have been proposed to
-tease apart between- and within-person sources of variance across time
-\citep[see][]{littlefield20210603, orth2021}. Most of these contemporary
-alternatives incorporate time-specific latent variables to capture
-between-person sources of variance and model within-person deviations
-around an individual's mean (or trait) level across time
-\citetext{\citealp[e.g.,
-RI-CLPM,][]{hamaker2015}; \citealp[LCM-SR,][]{curran2014a}}. It is
-important to note however that these models require multiple assessments
-waves (e.g., four or more to fully specify the LCM-SR), additional
-expertise to overcome issues with model convergence, and appreciation of
-modeling assumptions when attempting to adjudicate among potential
-models in each research context \citep[see][for further
-discussion]{littlefield20210603}.
-
-\hypertarget{model-assumptions}{%
-\subsection{Model Assumptions}\label{model-assumptions}}
-
-Many statistical models assume certain characteristics about the data to
-which they are being applied. As an example, common assumptions of
-parametric statistical models include normality, linearity, and equality
-of variances. These assumptions must be carefully considered before
-conducting analysis so that valid inferences can be made from the data;
-that is, violation of a model's assumptions can substantively alter the
-interpretation of results. Similarly, statistical models employed in the
-analyses of longitudinal data often entail a range of assumptions that
-must be closely inspected. One central issue for repeated measurements
-on an individual is how to account for the correlated nature of the
-data; another common feature of longitudinal data is heterogeneous
-variability; that is, the variance of the response changes over the
-duration of the study. Traditional techniques, such as a standard
-regression or ANOVA model, assume residuals are independent and thus are
-inappropriate for designs that assess (for example) the same individuals
-across time. That is, given the residuals are no longer independent, the
-standard errors from the models are biased and can produce misleading
-inferential results. Although there are formal tests of independence for
-time series data (e.g., the Durbin-Watson statistic; Durbin \& Watson,
-1950), more commonly independence is assumed to be violated in study
-designs with repeated assessments. Therefore, an initial question to be
-addressed by a researcher analyzing prospective data is how to best
-model the covariance structure of said data.
-
-\hypertarget{covariance-structures}{%
-\subsection{Covariance Structures}\label{covariance-structures}}
-
-Statistical models for longitudinal data include two main components to
-account for assumptions that are commonly violated when working with
-repeated measures data: a model for the covariance among repeated
-measures (both the correlations among pairs of repeated measures on an
-individual and the variability of the responses on different occasions),
-coupled with a model for the mean response and its dependence on
-covariates (eg, treatment group in the context of clinical trials). This
-allows for the specification of a range of so-called covariance
-structures, each with its own set of tradeoffs between model fit and
-parsimony \citep[e.g., see][]{kincaid2005}.
-
-\hypertarget{accounting-for-correlated-data}{%
-\subsection{Accounting for Correlated
-Data}\label{accounting-for-correlated-data}}
-
-As an example, one alternative structure that attempts to handle the
-reality that correlations between repeated assessments tend to diminish
-across time is the autoregressive design. As the name implies, the
-structure assumes a subsequent measurement occasion (e.g., assessment at
-Wave 2) is regressed onto (that is, is predicted by) a prior measurement
-occasion (e.g., assessment at Wave 1). The most common type of
-autoregressive design is the AR(1), where assessments at time T + 1 are
-regressed on assessments at Time T. Identical to compound symmetry, this
-model assumes the variances are homogenous across time. Diverting from
-compound symmetry, this model assumes the correlations between repeated
-assessments decline exponentially across time rather than remaining
-constant. For example, per the AR(1) structure, if the correlation
-between Time 1 and Time 2 data is thought to be .5, then the correlation
-between Time 1 and Time 3 data would be assumed to be .5\emph{.5 = .25,
-and the correlation between Time 1 and Time 4 data would be assumed to
-be .5}.5*.5 = .125. As with compound symmetry, the basic AR(1) model is
-parsimonious in that it only requires two parameters (the variance of
-the assessments and the autoregressive coefficient). Notably, the
-assumption of constant autoregressive relations between assessments is
-often relaxed in commonly employed designs that use autoregressive
-modeling (e.g., cross-lagged panel models {[}CLPM{]}). These designs
-still typically assume an AR(1) process (e.g., it is sufficient to
-regress the Time 3 assessment onto the Time 2 assessment and is not
-necessary to also regress the Time 3 assessment onto the Time 1
-assessment, which would result in an AR(2) process). However, the
-magnitude of these relations is often allowed to differ across different
-AR(1) pairs of assessment (e.g., the relation between Time 1 and Time 2
-can be different from the relation between Time 2 and Time 3). These
-more commonly employed models also often relax the assumption of equal
-variances of the repeated assessments. Although the AR(1) structure may
-involve a more realistic set of assumptions compared to compound
-symmetry, in that the AR(1) model allows for diminishing correlations
-across time, the basic AR(1) model, as well as autoregressive models
-more generally, can also suffer from several limitations in contexts
-that are common in prospective designs. In particular, recent work
-demonstrates that if a construct being assessed prospectively across
-time is trait-like in nature, then autoregressive relations fail to
-adequately account for this trait-like structure, with the downstream
-consequence that estimates derived from models based on AR structures
-(such as the CLPM) can be misleading and fail to adequately demarcate
-between- vs.~within-person sources of variance \citep{hamaker2015}.
-
-\hypertarget{linear-vs-non-linear-models}{%
-\subsection{Linear vs non-linear
-models}\label{linear-vs-non-linear-models}}
-
-Identification of optimal statistical models and appropriate
-mathematical functions requires an understanding of the type of data
-being used. Repeated assessments can be based on either continuous or
-discrete measures. Examples of discrete measures include repeated
-assessments of binary variables (e.g., past 12-month alcohol use
-disorder status measured across ten years), ordinal variables (e.g., a
-single item measuring the level of agreement to a statement on a
-three-point scale including the categories of ``disagree'', ``neutral'',
-and ``agree'' in an ecological momentary assessment study that involves
-multiple daily assessments), and count variables (e.g., number of
-cigarettes smoked per day across a daily diary study). In many ways, the
-distributional assumptions of indicators used in longitudinal designs
-mirror the decision points and considerations when delineating across
-different types of discrete outcome variables, a topic that spans entire
-textbooks \citep[e.g., see][]{lenz2016}. For example, the Mplus manual
-\citep{muthen2017} includes examples of a) censored and
-censored-inflated models, b) linear growth models for binary or ordinal
-variables, c) linear growth models for a count outcome assuming a
-Poisson model, d) linear growth models for a count outcome assuming a
-zero-inflated Poisson model and e) discrete- and continuous-time
-survival analysis for a binary outcome. Beyond these highlighted
-examples, other distributions (e.g., negative binomial) can be assumed
-for the indicators when modeling longitudinal data. These models can
-account for issues that can occur when working with discrete outcomes,
-including overdispersion (when the variance is higher than would be
-expected based on a given distribution) and zero-inflation {[}when more
-zeros occur than is expected based on a given distribution; see
-\citet{lenz2016}{]}. Models involving zero-inflation parameters are
-referred to as two-part models, given one part of the model predicts the
-zero-inflation whereas the other part of the model predicts outcomes
-consistent with a given distribution {[}e.g., Poisson distribution; see
-\citet{farewell2017}, for a review of two-part models for longitudinal
-data{]}. Although there exist several alternative models for discrete
-indicators, some more recent models that have been proposed for
-prospective data are only feasible in cases where indicators are assumed
-to be continuous rather than discrete {[}e.g., LCM-SR;
-\citet{curran2014a}{]}. Given the sheer breadth of issues relevant to
-determining better models for discrete outcomes, it is not uncommon for
-texts on longitudinal data analysis to only cover models and approaches
-that assume continuous indicators \citep[e.g.,][]{little2013}. However,
-some textbooks on categorical data analysis provide more detailed
-coverage of the myriad issues and modeling choices to consider when
-working with discrete outcomes {[}e.g., \citet{lenz2016}, Chapter 11 for
-matched pair/two-assessment designs; Chapter 12 for marginal and
-transitional models for repeated designs, such as generalized estimating
-equations, and Chapter 13 for random effects models for discrete
-outcomes{]}.
-
-\hypertarget{missing-dataattrition}{%
-\subsection{Missing Data/Attrition}\label{missing-dataattrition}}
-
-As recently reviewed by Littlefield (in press), investigators of
-prospective data are confronted with study attrition (i.e., participants
-may not provide data at a given wave of assessment) and thus approaches
-are needed to confront the issue of missing data. Three models of
-missingness are typically considered in the literature
-\citep[see][]{little1989}. These three models are data: a) missing
-completely at random (MCAR), b) missing at random (MAR), and c) missing
-not at random (MNAR). Data that are MCAR means missing data is a random
-sample of all the types of participants (e.g., males) in a given
-dataset. MAR suggests conditionally missing at random
-\citep[see][]{graham2009}. That is, MAR implies missingness is
-completely random (i.e., does not hinge on some unmeasured variables)
-once missingness has been adjusted by all available variables in a
-dataset (e.g., biological sex). Data that are MNAR are missing as a
-function of unobserved variables. \citet{graham2009} provides an
-excellent and easy-to-digest overview of further details involving
-missing data considerations.
-
-Multiple approaches have been posited to handle missing data. Before the
-advent of more contemporary approaches, common methods included several
-ad hoc procedures. These include eliminating the data of participants
-with missing data (e.g., listwise or pairwise deletion) or using mean
-imputation (i.e., replacing the missing value with the mean score of the
-sample that did participate). However, these methods are not recommended
-because they can contribute to biased parameter estimates and research
-conclusions \citep[see][]{graham2009}. More specifically, the last
-observation carried forward (LOCF) is a common approach to imputing
-missing data. LOCF replaces a participant's missing values after dropout
-with the last available measurement \citep{molnar2008}. This approach
-assumes stability (i.e., a given participant's score is not anticipated
-to increase or decline after study dropout) and that the data are MCA R.
-However, as described by \citet{molnar2008}, it is common for treatment
-groups to show higher attrition compared to control groups in studies of
-dementia drugs. Given that dementia worsens over time, using LOCF biases
-the results in favor of the treatment group \citep[see][for more
-details]{molnar2008}.
-
-More modern approaches, such as using maximum likelihood or multiple
-imputation to estimate missing data, are thought to avoid some of the
-biases of older approaches \citep[see][]{enders2010, graham2009}.
-\citet{graham2009} noted several ``myths'' regarding missing data. For
-example, Graham notes many assume the data must be minimally MAR to
-permit estimating procedures (such as maximum likelihood or multiple
-imputation) compared to other, more traditional approaches (e.g., using
-only complete case data). Violations of MAR impact both traditional and
-more modern data estimation procedures, though as noted by Graham,
-violations of MAR tend to have a greater effect on older methods. Graham
-thus suggests that estimating missing data is a better approach compared
-to the older procedures in most circumstances, regardless of the model
-of missingness {[}i.e., MCAR, MAR, MNAR; see \citet{graham2009}{]}.
-
-Attrition from a longitudinal panel study such as ABCD is inevitable and
-represents a threat to the validity of longitudinal analyses and
-cross-sectional analyses conducted at later time points, especially
-since attrition can only be expected to grow over time. While, to date,
-attrition in ABCD has been minimal (some cite here), it remains an
-important focus for longitudinal analysis and its significance is likely
-to only grow as the cohort ages. Ideally, one tries to minimize
-attrition through good retention practices from the outset via
-strategies designed to maintain engagement in the project
-\citep{cotter2005, hill2016, watson2018}. However, even the
-best-executed studies need to anticipate growing attrition over the
-length of the study and implement analytic strategies designed to
-provide the most valid inferences. Perhaps the most key concern when
-dealing with data that is missing due to attrition is determining the
-degree of bias in retained variables that is a consequence of attrition.
-Assuming that the data are not missing completely at random, attention
-to the nature of the missingness and employing techniques designed to
-mitigate attrition-related biases need to be considered in all
-longitudinal analyses. Several different approaches can be considered
-and employed depending upon the nature of the intended analyses, the
-degree of missingness, and data available to help estimate missing and
-unobserved values.
-
-\hypertarget{quantifying-effect-sizes-longitudinally}{%
-\subsection{Quantifying effect sizes
-longitudinally}\label{quantifying-effect-sizes-longitudinally}}
-
-Given longitudinal data involve different sources of variance,
-quantifying effect sizes longitudinally is a more difficult task
-compared to deriving such estimates from cross-sectional data. Effect
-size can be defined as, ``a population parameter (estimated in a sample)
-encapsulating the practical or clinical importance of a phenomenon under
-study.'' \citep{kraemer2014}. Common effect size metrics include r
-(i.e., the standardized covariance, or correlation, between two
-variables) and Cohen's d \citep{cohen1988}. Adjustments to common effect
-size calculations, such as Cohen's d, are required even when only two
-time points are considered \citep[e.g., see][]{morris2002}.
-\citet{wang2019a} note there are multiple approaches to obtaining
-standardized within-person effects, and that commonly suggested
-approaches (e.g., global standardization) can be problematic
-\citep[see][for more details]{wang2019a}. Thus, obtaining effect size
-metrics based on standardized estimates that are relatively simple in
-cross-sectional data (such as r) becomes more complex in the context of
-prospective data. \citet{feingold2009} noted that equations for effects
-sizes used in studies involving growth modeling analysis (e.g., latent
-growth curve modeling) were not mathematically equivalent, and the
-effect sizes were not in the same metric as effect sizes from
-traditional analysis \citep[see][for more details]{feingold2009}. Given
-this issue, there have been various proposals for adjusting effect size
-measures in repeated assessments. \citet{feingold2019} reviews the
-approach for effect size metrics for analyses based on growth modeling,
-including when considering linear and non-linear (i.e., quadratic)
-growth factors. \citet{morris2002} review various equations for effect
-size calculations relevant to when combining estimates in meta-analysis
-with repeated measures and independent-groups designs. Other approaches
-to quantifying effect sizes longitudinally may be based on standardized
-estimates from models that more optimally disentangle between- and
-within-person sources of variance (as reviewed above). As an example,
-within a RI-CLPM framework, standardized estimates between random
-intercepts (i.e., the correlation between two random intercepts for two
-different constructs assessed repeatedly) could be used to index the
-between-person relation, whereas standardized estimates among the
-structured residuals could be used as informing the effect sizes of
-within-person relations.
-
-\hypertarget{general-guidelines-for-this-quarto-template}{%
+Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
+porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
+dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
+velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
+finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
+diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
+ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
+pulvinar, volutpat risus.
+
+\linenumbers\hypertarget{general-guidelines-for-this-quarto-template}{%
 \section{General guidelines for this Quarto
@@
 
-\hypertarget{introduction-1}{%
-\section{Introduction}\label{introduction-1}}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
-zzzzz
-
 \nolinenumbers
-  \bibliography{references.bib}
+  \bibliography{bibliography.bib}
 

### 2023-04-21 test.qmd
cli commit 8

--- 2023-04-20 manuscript.qmd
@@
 ---
+title: "Longitudinal Analysis Manuscript: Working Draft"
+subtitle: Longitudinal Analysis using data from the ABCD Study 
+authors:
+  - name: Samuel W. Hawes
+    department: Center for Children and Families
+    affiliation: Florida International University
+    location: Miami, FL
+    email: shawes@fiu.edu
+  - name: Andrew K. Littlefield
+    department: Psychological Sciences
+    affiliation: Texas Tech University
+    location: Lubbock, TX
+    email: andrew.littlefield@ttu.edu
+  - name: Daniel A. Lopez
+    department: Department of Public Health
+    affiliation: University of Rochester Medical Center
+    location: Rochester, NY
+    email: Daniel_Lopez2@URMC.Rochester.edu
+  - name: Wesley K. Thompson
+    department: Center for Population Neuroscience and Genetics
+    affiliation: Laureate Institute for Brain Research
+    location: Tulsa, OK
+    email: wes.stat@gmail.com
+  - name: Kenneth J. Sher
+    department: Psychological Sciences
+    affiliation: University of Missouri
+    location: Columbia, MO
+    email: SherK@missouri.edu
+  - name: Additional Co-authors
+    department: xxxxxx
+    affiliation: University of xxxx
+    location: xxxxxxx
+    email: xxxxxx
 
-title: Manuscript Working Draft
+abstract: |
+ The Adolescent Brain Cognitive Development (ABCD) Study presents a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of children and adolescents.
+Given the complex nature of the longitudinal data collected in the ABCD Study, researcher are likely to encounter a myriad of methodological and analytic considerations and concerns.
+This review provides a comprehensive examination of key issues and techniques related to longitudinal data analysis, specifically focusing on the ABCD Study.
+The text discusses model assumptions, common violations (e.g., independent and identically distributed residuals, *heterogeneous* *variability*) and their implications for valid inference.
+The importance of appropriately modeling covariance structures, understandings trade-offs between model fit and parsimony, and challenges related to sample size, attrition, missing data are highlighted.
+Consideration is given to the importance of selecting appropriate statistical models to account for correlations in repeated measurements and the assumptions underlying these models.
+The review also differentiates between linear and non-linear models in the context of continuous and discrete data, emphasizing various distributional assumptions and the necessity of choosing appropriate models and statistical methods.
+By addressing these complexities, the review seeks to equip researchers with the necessary knowledge and tools to make informed decisions as they navigate effectively analyzing and interpreting data available in the ABCD Study.
 
-format: 
-  pdf: 
-    geometry: 
-      - top=20mm
-      - left=20mm
-  docx: default
-prefer-html: true
+author-summary: |
+  The Adolescent Brain Cognitive Development (ABCD) Study offers valuable insights into adolescent brain development.
+Researchers working with this data face methodological and analytic concerns, including sample size, attrition, missing data, and measurement.
+This review addresses complexities in analyzing longitudinal data, with a focus on statistical models accounting for correlations in repeated measurements and their assumptions.
+We emphasize the importance of choosing appropriate models and analytic functions based on the data and the specific scientific question to be addressed.
+This concise resource aims  to help researchers navigate longitudinal data analysis, make informed methodological decisions, and enhance the accuracy and reliability of their findings when working with data from the ABCD Study.
 
+    keywords:
+    *ABCD Study*,
+    *longitudinal analyses*,
+    *development*
+
+#format:
+#  pdf:
+#    geometry:
+#      - top=30mm
+#      - left=20mm
+#   docx: default
+# prefer-html: true
+
+output: 
+  rticles::arxiv_article
+
+bibliography: "references.bib" #references.bib
+biblio-style: unsrt
 ---
@@
 As an example, within a RI-CLPM framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.
-
----
-
-### References
-
-::: {#refs}
-:::

### 2023-04-21 test.qmd
cli commit 10

@@
     email: Daniel_Lopez2@URMC.Rochester.edu
-  - name: Wesley K. Thompson
-    department: Center for Population Neuroscience and Genetics
-    affiliation: Laureate Institute for Brain Research
-    location: Tulsa, OK
-    email: wes.stat@gmail.com
   - name: Kenneth J. Sher
@@
     email: xxxxxx
+  - name: Wesley K. Thompson
+    department: Center for Population Neuroscience and Genetics
+    affiliation: Laureate Institute for Brain Research
+    location: Tulsa, OK
+    email: wes.stat@gmail.com
 
@@
 To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level.
-This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield20210603.
+This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield2021.
 It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling.
@@
 As a result, experts increasingly caution about the need to xxxxxxxx [add citation].
-Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield20210603; @orth2021].
+Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021].
 Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a].
-It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield20210603, for further discussion].
+It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].
 
@@
 That is, given the residuals are no longer independent, the standard errors from the models are biased and can produce misleading inferential results.
-Although there are formal tests of independence for time series data (e.g., the Durbin-Watson statistic; Durbin & Watson, 1950), more commonly independence is assumed to be violated in study designs with repeated assessments.
+Although there are formal tests of independence for time series data (e.g., the Durbin-Watson statistic [@durbin1950]), more commonly independence is assumed to be violated in study designs with repeated assessments.
 Therefore, an initial question to be addressed by a researcher analyzing prospective data is how to best model the covariance structure of said data.
@@
 As an example, within a RI-CLPM framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.
+
+# Part III Section in Progress: TBD
+\label{sec:headings}
+**Subsection TBD.**
+
+# Part IV Discussion
+\label{sec:headings}
+**Subsection TBD.**

### 2023-04-21 test.tex
cli commit 10

--- 2023-04-21 manuscript.tex
@@
 \PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
-
-\PassOptionsToPackage{table}{xcolor}
-
 \documentclass[
-  10pt,
   letterpaper,
-]{article}
+  DIV=11,
+  numbers=noendperiod]{scrartcl}
 
@@
 \usepackage{xcolor}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
@@
 \makeatother
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
-
-% marvosym package for additional characters
-\usepackage{marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-% Using natbib instead
-% \usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+% load packages
+\usepackage{geometry}
+\usepackage{xcolor}
+\usepackage{eso-pic}
+\usepackage{fancyhdr}
+\usepackage{sectsty}
+\usepackage{fontspec}
+\usepackage{titlesec}
+
+%% Set page size with a wider right margin
+\geometry{a4paper, total={170mm,257mm}, left=20mm, top=20mm, bottom=20mm, right=50mm}
+
+%% Let's define some colours
+\definecolor{light}{HTML}{E6E6FA}
+\definecolor{highlight}{HTML}{800080}
+\definecolor{dark}{HTML}{330033}
+
+%% Let's add the border on the right hand side 
+\AddToShipoutPicture{% 
+    \AtPageLowerLeft{% 
+        \put(\LenToUnit{\dimexpr\paperwidth-3cm},0){% 
+            \color{light}\rule{3cm}{\LenToUnit\paperheight}%
+          }%
+     }%
+     % logo
+    \AtPageLowerLeft{% start the bar at the bottom right of the page
+        \put(\LenToUnit{\dimexpr\paperwidth-2.25cm},27.2cm){% move it to the top right
+            \color{light}\includegraphics[width=1.5cm]{_extensions/nrennie/PrettyPDF/logo.png}
+          }%
+     }%
 }
 
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in 
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr}
-\usepackage{epstopdf}
-%\pagestyle{myheadings}
-\pagestyle{fancy}
+%% Style the page number
+\fancypagestyle{mystyle}{
   \fancyhf{}
-%\setlength{\headheight}{27.023pt}
-%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\headrulewidth}{0pt}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\today}
-% Remove comment for double spacing
-% \usepackage{setspace}
-% \doublespacing
+  \renewcommand\headrulewidth{0pt}
+  \fancyfoot[R]{\thepage}
+  \fancyfootoffset{3.5cm}
+}
+\setlength{\footskip}{20pt}
+
+%% style the chapter/section fonts
+\chapterfont{\color{dark}\fontsize{20}{16.8}\selectfont}
+\sectionfont{\color{dark}\fontsize{20}{16.8}\selectfont}
+\subsectionfont{\color{dark}\fontsize{14}{16.8}\selectfont}
+\titleformat{\subsection}
+  {\sffamily\Large\bfseries}{\thesection}{1em}{}[{\titlerule[0.8pt]}]
+
+% left align title
+\makeatletter
+\renewcommand{\maketitle}{\bgroup\setlength{\parindent}{0pt}
+\begin{flushleft}
+  {\sffamily\huge\textbf{\MakeUppercase{\@title}}} \vspace{0.3cm} \newline
+  {\Large {\@subtitle}} \newline
+  \@author
+\end{flushleft}\egroup
+}
+\makeatother
+
+%% Use some custom fonts
+\setsansfont{Ubuntu}[
+    Path=_extensions/nrennie/PrettyPDF/Ubuntu/,
+    Scale=0.9,
+    Extension = .ttf,
+    UprightFont=*-Regular,
+    BoldFont=*-Bold,
+    ItalicFont=*-Italic,
+    ]
+
+\setmainfont{Ubuntu}[
+    Path=_extensions/nrennie/PrettyPDF/Ubuntu/,
+    Scale=0.9,
+    Extension = .ttf,
+    UprightFont=*-Regular,
+    BoldFont=*-Bold,
+    ItalicFont=*-Italic,
+    ]
+\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 \makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\@ifundefined{codebgcolor}{\definecolor{codebgcolor}{named}{light}}
+\makeatother
+\makeatletter
 \makeatother
@@
 \fi
-\usepackage[numbers,square,comma]{natbib}
-\bibliographystyle{plos2015}
 \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
@@
   pdftitle={Longitudinal Analysis Manuscript: Working Draft},
-  pdfauthor={Samuel W. Hawes; with the Lorem Ipsum Consortium},
-  hidelinks,
+  colorlinks=true,
+  linkcolor={highlight},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={highlight},
   pdfcreator={LaTeX via pandoc}}
 
+\title{Longitudinal Analysis Manuscript: Working Draft}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{Longitudinal Analysis using data from the ABCD Study}
+\author{Samuel W. Hawes \and Andrew K. Littlefield \and Daniel A.
+Lopez \and Kenneth J. Sher \and Additional Co-authors \and Wesley K.
+Thompson}
+\date{}
+
 \begin{document}
-\vspace*{0.2in}
-
-% Title must be 250 characters or less.
-\begin{flushleft}
-{\Large
-\textbf\newline{Longitudinal Analysis Manuscript: Working
-Draft} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
-}
-\newline
-\\
-% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
-Samuel W. Hawes\textsuperscript{1*}, with the Lorem Ipsum
-Consortium\textsuperscript{\textpilcrow}
-\\
-\bigskip
-\textbf{1} Center for Children and Families, Florida International
-University, Miami, Florida, United States, 
-\bigskip
-
-% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
-% 
-% Remove or comment out the author notes below if they aren't used.
-%
-% Primary Equal Contribution Note
-\Yinyang These authors contributed equally to this work.
-
-% Additional Equal Contribution Note
-% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
-%\ddag These authors also contributed equally to this work.
-
-% Current address notes
-\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
-% \textcurrency b Insert second current address 
-% \textcurrency c Insert third current address
-
-% Deceased author note
-\dag Deceased
-
-% Group/Consortium Author Note
-\textpilcrow Biostatistics Working Group membership list can be found in
-the Acknowledgments sections
-
-% Use the asterisk to denote corresponding authorship and provide email address in note below.
-* shawes@fiu.edu
-
-\end{flushleft}
-
-\section*{Abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
-porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
-dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
-velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
-finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
-diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
-ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
-pulvinar, volutpat risus.
-
-\section*{Author summary}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
-porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
-dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
-velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
-finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
-diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
-ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
-pulvinar, volutpat risus.
-
-\linenumbersxxxxx
-
-\#title: ``Longitudinal Analaysis Manuscript- Working Draft'' \# format:
-\# pdf: \# toc: true \# number-sections: true \# colorlinks: true \#
-geometry: \# - top=20mm \# - left=20mm \# docx: default \# prefer-html:
-true
-
-xxxxx
+\maketitle
+\begin{abstract}
+The Adolescent Brain Cognitive Development (ABCD) Study presents a
+unique opportunity for researchers to investigate developmental
+processes in a large, diverse cohort of children and adolescents. Given
+the complex nature of the longitudinal data collected in the ABCD Study,
+researcher are likely to encounter a myriad of methodological and
+analytic considerations and concerns. This review provides a
+comprehensive examination of key issues and techniques related to
+longitudinal data analysis, specifically focusing on the ABCD Study. The
+text discusses model assumptions, common violations (e.g., independent
+and identically distributed residuals, \emph{heterogeneous}
+\emph{variability}) and their implications for valid inference. The
+importance of appropriately modeling covariance structures,
+understandings trade-offs between model fit and parsimony, and
+challenges related to sample size, attrition, missing data are
+highlighted. Consideration is given to the importance of selecting
+appropriate statistical models to account for correlations in repeated
+measurements and the assumptions underlying these models. The review
+also differentiates between linear and non-linear models in the context
+of continuous and discrete data, emphasizing various distributional
+assumptions and the necessity of choosing appropriate models and
+statistical methods. By addressing these complexities, the review seeks
+to equip researchers with the necessary knowledge and tools to make
+informed decisions as they navigate effectively analyzing and
+interpreting data available in the ABCD Study.
+\end{abstract}
+\pagestyle{mystyle}
+
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, breakable, colback={codebgcolor}, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-label\{sec:headings\} The Adolescent Brain Cognitive Development (ABCD)
+\label{sec:headings} The Adolescent Brain Cognitive Development (ABCD)
 Study® is the largest long-term investigation of neurodevelopment and
@@
 and environmental influences on brain development and their roles in
-behavioral and health outcomes in adolescents \citep{volkow2018}. At its
-heart, the study is designed to chart the course of human development
-across multiple, interacting domains from late childhood to early
-adulthood and to identify factors that lead to both positive and
+behavioral and health outcomes in adolescents (Volkow et al. 2018). At
+its heart, the study is designed to chart the course of human
+development across multiple, interacting domains from late childhood to
+early adulthood and to identify factors that lead to both positive and
 negative developmental outcomes. Central to achieving these goals is the
@@
 outcomes associated with variables of interest. This data includes a
-neurocognitive battery \citep{luciana2018a, thompson2019}, mental and
-physical health assessments \citep{barch2018}, measures of culture and
-environment \citep{zucker2018}, substance use {[}add citation{]},
-biospecimens \citep{uban2018}, structural and functional brain imaging
-\citep{casey2018, hagler2019}, geolocation-based environmental exposure
-data, wearables, and mobile technology \citep{bagot2018}, and whole
-genome genotyping \citep{loughnan2020}. Many of these measures are
-collected at in-person annual visits, with brain imaging collected at
-baseline and every other year going forward. A limited number of
-assessments are collected in semi-annual telephone interviews between
-in-person visits. Data are publicly released on an annual basis through
-the NIMH Data Archive. By necessity, the study's earliest data releases
-were cross-sectional (i.e., the baseline data), however, the most recent
-public data release (NDA Release 4.0) contains data collected across
-three annual assessments, including two imaging assessments (baseline
-and year 2 follow-up visits).
+neurocognitive battery (Luciana et al. 2018; Thompson et al. 2019),
+mental and physical health assessments (Barch et al. 2018), measures of
+culture and environment (Zucker et al. 2018), substance use {[}add
+citation{]}, biospecimens (Uban et al. 2018), structural and functional
+brain imaging (Casey et al. 2018; Hagler et al. 2019), geolocation-based
+environmental exposure data, wearables, and mobile technology (Bagot et
+al. 2018), and whole genome genotyping (Loughnan et al. 2020). Many of
+these measures are collected at in-person annual visits, with brain
+imaging collected at baseline and every other year going forward. A
+limited number of assessments are collected in semi-annual telephone
+interviews between in-person visits. Data are publicly released on an
+annual basis through the NIMH Data Archive. By necessity, the study's
+earliest data releases were cross-sectional (i.e., the baseline data),
+however, the most recent public data release (NDA Release 4.0) contains
+data collected across three annual assessments, including two imaging
+assessments (baseline and year 2 follow-up visits).
 
@@
 growth and decline often differ between cross-sectional vs.~longitudinal
-effects \citep{salthouse2014} where subjects gain increasing experience
-with the assessment with each successive measurement occasion. Such
+effects (Salthouse 2014) where subjects gain increasing experience with
+the assessment with each successive measurement occasion. Such
 experience effects on cognitive functioning have been demonstrated in
-adolescent longitudinal samples similar to ABCD \citep{sullivan2017} and
-highlight the need to consider these effects and address them
+adolescent longitudinal samples similar to ABCD (Sullivan et al. 2017)
+and highlight the need to consider these effects and address them
 analytically. In the case of performance-based measures {[}e.g., matrix
-reasoning related to neurocognitive functioning; see
-\citet{salthouse2014}{]}, this can be due to ``learning'' the task from
-previous test administrations (e.g., someone taking the test a second
-time performs better than they did the first time simply as a function
-of having taken it before). Even in the case of non-performance-based
-measures (e.g., levels of depression), where one cannot easily make the
-argument that one has acquired some task-specific skill through
-learning, it has been observed that respondents tend to endorse lower
-levels on subsequent assessments
-\citetext{\citealp[e.g.,][]{beck1961}; \citealp[see][]{french2010}} and
+reasoning related to neurocognitive functioning; see Salthouse
+(2014){]}, this can be due to ``learning'' the task from previous test
+administrations (e.g., someone taking the test a second time performs
+better than they did the first time simply as a function of having taken
+it before). Even in the case of non-performance-based measures (e.g.,
+levels of depression), where one cannot easily make the argument that
+one has acquired some task-specific skill through learning, it has been
+observed that respondents tend to endorse lower levels on subsequent
+assessments (e.g., Beck et al. 1961; see French and Sutton 2010) and
 this phenomenon has been well documented in research on structured
-diagnostic interviews \citep{robins1985}. While it is typically assumed
-that individuals are rescinding or telling us less information on
-follow-up interviews, there is reason to suspect that in some cases the
-initial assessment may be artefactually elevated
-\citep[see][]{shrout2018a}. Some designs (specifically, accelerated
-longitudinal designs) are especially well suited for discovering these
-effects and modeling them. While ABCD was not designed as an accelerated
-longitudinal design, the variability in age at the time of baseline
-recruitment (9 years, 0 months to 10 years, 11 months) allows some
-measures, collected every year, to be conceptualized as an accelerated
-longitudinal design. Moreover, it is possible that in later waves,
-patterns of longitudinal missing data will allow some analyses to assess
-the confounded effects of age and the number of prior assessments.
-However, ABCD is fundamentally a single-cohort, longitudinal design,
-where a number of prior assessments and age are highly confounded, and
-for, perhaps, most analyses, the possible influence of experience
-effects needs to be kept in mind.
+diagnostic interviews (Robins 1985). While it is typically assumed that
+individuals are rescinding or telling us less information on follow-up
+interviews, there is reason to suspect that in some cases the initial
+assessment may be artefactually elevated (see Shrout et al. 2018). Some
+designs (specifically, accelerated longitudinal designs) are especially
+well suited for discovering these effects and modeling them. While ABCD
+was not designed as an accelerated longitudinal design, the variability
+in age at the time of baseline recruitment (9 years, 0 months to 10
+years, 11 months) allows some measures, collected every year, to be
+conceptualized as an accelerated longitudinal design. Moreover, it is
+possible that in later waves, patterns of longitudinal missing data will
+allow some analyses to assess the confounded effects of age and the
+number of prior assessments. However, ABCD is fundamentally a
+single-cohort, longitudinal design, where a number of prior assessments
+and age are highly confounded, and for, perhaps, most analyses, the
+possible influence of experience effects needs to be kept in mind.
 
@@
 Variable A and Variable B at time T-1). At the same time, data based on
-two assessments are inherently limited on multiple fronts. As
-\citep{rogosa1982} noted approximately forty years ago, ``Two waves of
-data are better than one, but maybe not much better''. These sentiments
-are reflected in more contemporary recommendations regarding
+two assessments are inherently limited on multiple fronts. As (Rogosa,
+Brandt, and Zimowski 1982) noted approximately forty years ago, ``Two
+waves of data are better than one, but maybe not much better''. These
+sentiments are reflected in more contemporary recommendations regarding
 best-practice guidelines for prospective data, which increasingly
@@
 straight line change is the only possible form, given two assessments;
-see \citep{duncan2009}). Research designs that include three or more
-time points allow for increasingly nuanced analyses that more adequately
-tease apart sources of variation and covariation among the repeated
-assessments \citep{king2018}-- a key aspect of inferential research. To
-illustrate, developmental theories are typically interested in
-understanding patterns of within-individual change over time (discussed
-in further detail, below); however, two data points provide meager
-information on change at the person level. This point is further
+see (Duncan and Duncan 2009)). Research designs that include three or
+more time points allow for increasingly nuanced analyses that more
+adequately tease apart sources of variation and covariation among the
+repeated assessments (King et al. 2018)-- a key aspect of inferential
+research. To illustrate, developmental theories are typically interested
+in understanding patterns of within-individual change over time
+(discussed in further detail, below); however, two data points provide
+meager information on change at the person level. This point is further
 underscored in a recent review of statistical models commonly touted as
@@
 concerning, ``\ldots the models discussed here do not offer a feasible
-way to overcome these inherent limitations''
-\citet{littlefield20210603}. It is important to note, however, that
-despite the current focus on two-wave designs versus three or more
-assessment waves, garnering three assessment points is not a panacea for
-longitudinal modeling. Indeed, several contemporary longitudinal models
-designed to isolate within-individual variability {[}e.g., the Latent
-Curve Model with Structured Residuals; \citet{curran2014a}{]} require at
-least four assessments to parameterize fully and, more generally,
-increasingly accurate parameter estimates are obtained as more
-assessment occasions are used \citep{duncan2009}.
+way to overcome these inherent limitations'' Littlefield et al. (2021).
+It is important to note, however, that despite the current focus on
+two-wave designs versus three or more assessment waves, garnering three
+assessment points is not a panacea for longitudinal modeling. Indeed,
+several contemporary longitudinal models designed to isolate
+within-individual variability {[}e.g., the Latent Curve Model with
+Structured Residuals; Curran et al. (2014){]} require at least four
+assessments to parameterize fully and, more generally, increasingly
+accurate parameter estimates are obtained as more assessment occasions
+are used (Duncan and Duncan 2009).
 
@@
 intraindividual) processes (e.g., how phenotypes change or remain stable
-within individuals over time) \citep[e.g., see][]{curran2011}. Despite
+within individuals over time) (e.g., see Curran and Bauer 2011). Despite
 this, methods generally unsuited for disaggregating between- and
@@
 tease apart between- and within-person sources of variance across time
-\citep[see][]{littlefield20210603, orth2021}. Most of these contemporary
-alternatives incorporate time-specific latent variables to capture
-between-person sources of variance and model within-person deviations
-around an individual's mean (or trait) level across time
-\citetext{\citealp[e.g.,
-RI-CLPM,][]{hamaker2015}; \citealp[LCM-SR,][]{curran2014a}}. It is
-important to note however that these models require multiple assessments
-waves (e.g., four or more to fully specify the LCM-SR), additional
-expertise to overcome issues with model convergence, and appreciation of
-modeling assumptions when attempting to adjudicate among potential
-models in each research context \citep[see][for further
-discussion]{littlefield20210603}.
+(see Littlefield et al. 2021; Orth et al. 2021). Most of these
+contemporary alternatives incorporate time-specific latent variables to
+capture between-person sources of variance and model within-person
+deviations around an individual's mean (or trait) level across time
+(e.g., RI-CLPM, Hamaker, Kuiper, and Grasman 2015; LCM-SR, Curran et al.
+2014). It is important to note however that these models require
+multiple assessments waves (e.g., four or more to fully specify the
+LCM-SR), additional expertise to overcome issues with model convergence,
+and appreciation of modeling assumptions when attempting to adjudicate
+among potential models in each research context (see Littlefield et al.
+2021, for further discussion).
 
@@
 inferential results. Although there are formal tests of independence for
-time series data (e.g., the Durbin-Watson statistic; Durbin \& Watson,
-1950), more commonly independence is assumed to be violated in study
+time series data (e.g., the Durbin-Watson statistic (Durbin and Watson
+1950)), more commonly independence is assumed to be violated in study
 designs with repeated assessments. Therefore, an initial question to be
@@
 structures, each with its own set of tradeoffs between model fit and
-parsimony \citep[e.g., see][]{kincaid2005}.
+parsimony (e.g., see Kincaid 2005).
 
@@
 (such as the CLPM) can be misleading and fail to adequately demarcate
-between- vs.~within-person sources of variance \citep{hamaker2015}.
+between- vs.~within-person sources of variance (Hamaker, Kuiper, and
+Grasman 2015).
 
@@
 different types of discrete outcome variables, a topic that spans entire
-textbooks \citep[e.g., see][]{lenz2016}. For example, the Mplus manual
-\citep{muthen2017} includes examples of a) censored and
-censored-inflated models, b) linear growth models for binary or ordinal
-variables, c) linear growth models for a count outcome assuming a
-Poisson model, d) linear growth models for a count outcome assuming a
-zero-inflated Poisson model and e) discrete- and continuous-time
-survival analysis for a binary outcome. Beyond these highlighted
-examples, other distributions (e.g., negative binomial) can be assumed
-for the indicators when modeling longitudinal data. These models can
-account for issues that can occur when working with discrete outcomes,
-including overdispersion (when the variance is higher than would be
-expected based on a given distribution) and zero-inflation {[}when more
-zeros occur than is expected based on a given distribution; see
-\citet{lenz2016}{]}. Models involving zero-inflation parameters are
-referred to as two-part models, given one part of the model predicts the
-zero-inflation whereas the other part of the model predicts outcomes
-consistent with a given distribution {[}e.g., Poisson distribution; see
-\citet{farewell2017}, for a review of two-part models for longitudinal
-data{]}. Although there exist several alternative models for discrete
-indicators, some more recent models that have been proposed for
-prospective data are only feasible in cases where indicators are assumed
-to be continuous rather than discrete {[}e.g., LCM-SR;
-\citet{curran2014a}{]}. Given the sheer breadth of issues relevant to
-determining better models for discrete outcomes, it is not uncommon for
-texts on longitudinal data analysis to only cover models and approaches
-that assume continuous indicators \citep[e.g.,][]{little2013}. However,
-some textbooks on categorical data analysis provide more detailed
-coverage of the myriad issues and modeling choices to consider when
-working with discrete outcomes {[}e.g., \citet{lenz2016}, Chapter 11 for
+textbooks (e.g., see Lenz 2016). For example, the Mplus manual (Muthén
+2017) includes examples of a) censored and censored-inflated models, b)
+linear growth models for binary or ordinal variables, c) linear growth
+models for a count outcome assuming a Poisson model, d) linear growth
+models for a count outcome assuming a zero-inflated Poisson model and e)
+discrete- and continuous-time survival analysis for a binary outcome.
+Beyond these highlighted examples, other distributions (e.g., negative
+binomial) can be assumed for the indicators when modeling longitudinal
+data. These models can account for issues that can occur when working
+with discrete outcomes, including overdispersion (when the variance is
+higher than would be expected based on a given distribution) and
+zero-inflation {[}when more zeros occur than is expected based on a
+given distribution; see Lenz (2016){]}. Models involving zero-inflation
+parameters are referred to as two-part models, given one part of the
+model predicts the zero-inflation whereas the other part of the model
+predicts outcomes consistent with a given distribution {[}e.g., Poisson
+distribution; see Farewell et al. (2017), for a review of two-part
+models for longitudinal data{]}. Although there exist several
+alternative models for discrete indicators, some more recent models that
+have been proposed for prospective data are only feasible in cases where
+indicators are assumed to be continuous rather than discrete {[}e.g.,
+LCM-SR; Curran et al. (2014){]}. Given the sheer breadth of issues
+relevant to determining better models for discrete outcomes, it is not
+uncommon for texts on longitudinal data analysis to only cover models
+and approaches that assume continuous indicators (e.g., T. D. Little
+2013). However, some textbooks on categorical data analysis provide more
+detailed coverage of the myriad issues and modeling choices to consider
+when working with discrete outcomes {[}e.g., Lenz (2016), Chapter 11 for
 matched pair/two-assessment designs; Chapter 12 for marginal and
@@
 are needed to confront the issue of missing data. Three models of
-missingness are typically considered in the literature
-\citep[see][]{little1989}. These three models are data: a) missing
-completely at random (MCAR), b) missing at random (MAR), and c) missing
-not at random (MNAR). Data that are MCAR means missing data is a random
-sample of all the types of participants (e.g., males) in a given
-dataset. MAR suggests conditionally missing at random
-\citep[see][]{graham2009}. That is, MAR implies missingness is
-completely random (i.e., does not hinge on some unmeasured variables)
-once missingness has been adjusted by all available variables in a
-dataset (e.g., biological sex). Data that are MNAR are missing as a
-function of unobserved variables. \citet{graham2009} provides an
-excellent and easy-to-digest overview of further details involving
-missing data considerations.
+missingness are typically considered in the literature (see R. J. Little
+and Rubin 1989). These three models are data: a) missing completely at
+random (MCAR), b) missing at random (MAR), and c) missing not at random
+(MNAR). Data that are MCAR means missing data is a random sample of all
+the types of participants (e.g., males) in a given dataset. MAR suggests
+conditionally missing at random (see Graham 2009). That is, MAR implies
+missingness is completely random (i.e., does not hinge on some
+unmeasured variables) once missingness has been adjusted by all
+available variables in a dataset (e.g., biological sex). Data that are
+MNAR are missing as a function of unobserved variables. Graham (2009)
+provides an excellent and easy-to-digest overview of further details
+involving missing data considerations.
 
@@
 because they can contribute to biased parameter estimates and research
-conclusions \citep[see][]{graham2009}. More specifically, the last
-observation carried forward (LOCF) is a common approach to imputing
-missing data. LOCF replaces a participant's missing values after dropout
-with the last available measurement \citep{molnar2008}. This approach
-assumes stability (i.e., a given participant's score is not anticipated
-to increase or decline after study dropout) and that the data are MCA R.
-However, as described by \citet{molnar2008}, it is common for treatment
-groups to show higher attrition compared to control groups in studies of
-dementia drugs. Given that dementia worsens over time, using LOCF biases
-the results in favor of the treatment group \citep[see][for more
-details]{molnar2008}.
+conclusions (see Graham 2009). More specifically, the last observation
+carried forward (LOCF) is a common approach to imputing missing data.
+LOCF replaces a participant's missing values after dropout with the last
+available measurement (Molnar, Hutton, and Fergusson 2008). This
+approach assumes stability (i.e., a given participant's score is not
+anticipated to increase or decline after study dropout) and that the
+data are MCA R. However, as described by Molnar, Hutton, and Fergusson
+(2008), it is common for treatment groups to show higher attrition
+compared to control groups in studies of dementia drugs. Given that
+dementia worsens over time, using LOCF biases the results in favor of
+the treatment group (see Molnar, Hutton, and Fergusson 2008, for more
+details).
 
@@
 imputation to estimate missing data, are thought to avoid some of the
-biases of older approaches \citep[see][]{enders2010, graham2009}.
-\citet{graham2009} noted several ``myths'' regarding missing data. For
-example, Graham notes many assume the data must be minimally MAR to
-permit estimating procedures (such as maximum likelihood or multiple
-imputation) compared to other, more traditional approaches (e.g., using
-only complete case data). Violations of MAR impact both traditional and
-more modern data estimation procedures, though as noted by Graham,
-violations of MAR tend to have a greater effect on older methods. Graham
-thus suggests that estimating missing data is a better approach compared
-to the older procedures in most circumstances, regardless of the model
-of missingness {[}i.e., MCAR, MAR, MNAR; see \citet{graham2009}{]}.
+biases of older approaches (see Enders 2010; Graham 2009). Graham (2009)
+noted several ``myths'' regarding missing data. For example, Graham
+notes many assume the data must be minimally MAR to permit estimating
+procedures (such as maximum likelihood or multiple imputation) compared
+to other, more traditional approaches (e.g., using only complete case
+data). Violations of MAR impact both traditional and more modern data
+estimation procedures, though as noted by Graham, violations of MAR tend
+to have a greater effect on older methods. Graham thus suggests that
+estimating missing data is a better approach compared to the older
+procedures in most circumstances, regardless of the model of missingness
+{[}i.e., MCAR, MAR, MNAR; see Graham (2009){]}.
 
@@
 attrition through good retention practices from the outset via
-strategies designed to maintain engagement in the project
-\citep{cotter2005, hill2016, watson2018}. However, even the
+strategies designed to maintain engagement in the project (Cotter et al.
+2005; Hill et al. 2016; Watson et al. 2018). However, even the
 best-executed studies need to anticipate growing attrition over the
@@
 encapsulating the practical or clinical importance of a phenomenon under
-study.'' \citep{kraemer2014}. Common effect size metrics include r
-(i.e., the standardized covariance, or correlation, between two
-variables) and Cohen's d \citep{cohen1988}. Adjustments to common effect
-size calculations, such as Cohen's d, are required even when only two
-time points are considered \citep[e.g., see][]{morris2002}.
-\citet{wang2019a} note there are multiple approaches to obtaining
-standardized within-person effects, and that commonly suggested
-approaches (e.g., global standardization) can be problematic
-\citep[see][for more details]{wang2019a}. Thus, obtaining effect size
-metrics based on standardized estimates that are relatively simple in
-cross-sectional data (such as r) becomes more complex in the context of
-prospective data. \citet{feingold2009} noted that equations for effects
-sizes used in studies involving growth modeling analysis (e.g., latent
-growth curve modeling) were not mathematically equivalent, and the
-effect sizes were not in the same metric as effect sizes from
-traditional analysis \citep[see][for more details]{feingold2009}. Given
-this issue, there have been various proposals for adjusting effect size
-measures in repeated assessments. \citet{feingold2019} reviews the
-approach for effect size metrics for analyses based on growth modeling,
-including when considering linear and non-linear (i.e., quadratic)
-growth factors. \citet{morris2002} review various equations for effect
-size calculations relevant to when combining estimates in meta-analysis
-with repeated measures and independent-groups designs. Other approaches
-to quantifying effect sizes longitudinally may be based on standardized
-estimates from models that more optimally disentangle between- and
-within-person sources of variance (as reviewed above). As an example,
-within a RI-CLPM framework, standardized estimates between random
-intercepts (i.e., the correlation between two random intercepts for two
-different constructs assessed repeatedly) could be used to index the
-between-person relation, whereas standardized estimates among the
-structured residuals could be used as informing the effect sizes of
-within-person relations.
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\hypertarget{references}{%
-\subsubsection{References}\label{references}}
-
-\renewcommand{\bibsection}{}
-\bibliography{references.bib,bibliography.bib}
-
-\nolinenumbers
+study.'' (Kraemer 2014). Common effect size metrics include r (i.e., the
+standardized covariance, or correlation, between two variables) and
+Cohen's d (Cohen 1988). Adjustments to common effect size calculations,
+such as Cohen's d, are required even when only two time points are
+considered (e.g., see Morris and DeShon 2002). Wang et al. (2019) note
+there are multiple approaches to obtaining standardized within-person
+effects, and that commonly suggested approaches (e.g., global
+standardization) can be problematic (see Wang et al. 2019, for more
+details). Thus, obtaining effect size metrics based on standardized
+estimates that are relatively simple in cross-sectional data (such as r)
+becomes more complex in the context of prospective data. Feingold (2009)
+noted that equations for effects sizes used in studies involving growth
+modeling analysis (e.g., latent growth curve modeling) were not
+mathematically equivalent, and the effect sizes were not in the same
+metric as effect sizes from traditional analysis (see Feingold 2009, for
+more details). Given this issue, there have been various proposals for
+adjusting effect size measures in repeated assessments. Feingold (2019)
+reviews the approach for effect size metrics for analyses based on
+growth modeling, including when considering linear and non-linear (i.e.,
+quadratic) growth factors. Morris and DeShon (2002) review various
+equations for effect size calculations relevant to when combining
+estimates in meta-analysis with repeated measures and independent-groups
+designs. Other approaches to quantifying effect sizes longitudinally may
+be based on standardized estimates from models that more optimally
+disentangle between- and within-person sources of variance (as reviewed
+above). As an example, within a RI-CLPM framework, standardized
+estimates between random intercepts (i.e., the correlation between two
+random intercepts for two different constructs assessed repeatedly)
+could be used to index the between-person relation, whereas standardized
+estimates among the structured residuals could be used as informing the
+effect sizes of within-person relations.
+
+\hypertarget{part-iii-section-in-progress-tbd}{%
+\section{Part III Section in Progress:
+TBD}\label{part-iii-section-in-progress-tbd}}
+
+\label{sec:headings} \textbf{Subsection TBD.}
+
+\hypertarget{part-iv-discussion}{%
+\section{Part IV Discussion}\label{part-iv-discussion}}
+
+\label{sec:headings} \textbf{Subsection TBD.}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-bagot2018}{}}%
+Bagot, K. S., S. A. Matthews, M. Mason, Lindsay M. Squeglia, J. Fowler,
+K. Gray, M. Herting, A. May, Ian Colrain, and J. Godino. 2018.
+{``Current, Future and Potential Use of Mobile and Wearable Technologies
+and Social Media Data in the {ABCD} Study to Increase Understanding of
+Contributors to Child Health.''} \emph{Developmental Cognitive
+Neuroscience} 32: 121--29.
+
+\leavevmode\vadjust pre{\hypertarget{ref-barch2018}{}}%
+Barch, Deanna M., Matthew D. Albaugh, Shelli Avenevoli, Linda Chang,
+Duncan B. Clark, Meyer D. Glantz, James J. Hudziak, Terry L. Jernigan,
+Susan F. Tapert, and Debbie Yurgelun-Todd. 2018. {``Demographic,
+Physical and Mental Health Assessments in the Adolescent Brain and
+Cognitive Development Study: {Rationale} and Description.''}
+\emph{Developmental Cognitive Neuroscience} 32: 55--66.
+
+\leavevmode\vadjust pre{\hypertarget{ref-beck1961}{}}%
+Beck, Aaron T., Calvin H. Ward, Mock Mendelson, Jeremiah Mock, and John
+Erbaugh. 1961. {``An Inventory for Measuring Depression.''}
+\emph{Archives of General Psychiatry} 4 (6): 561--71.
+
+\leavevmode\vadjust pre{\hypertarget{ref-casey2018}{}}%
+Casey, B. J., Tariq Cannonier, May I. Conley, Alexandra O. Cohen, Deanna
+M. Barch, Mary M. Heitzeg, Mary E. Soules, et al. 2018. {``The
+{Adolescent} {Brain} {Cognitive} {Development} ({ABCD}) Study: {Imaging}
+Acquisition Across 21 Sites.''} \emph{Developmental Cognitive
+Neuroscience}, The {Adolescent} {Brain} {Cognitive} {Development}
+({ABCD}) {Consortium}: {Rationale}, {Aims}, and {Assessment} {Strategy},
+32 (August): 43--54. \url{https://doi.org/10.1016/j.dcn.2018.03.001}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cohen1988}{}}%
+Cohen, Jacob. 1988. {``Statistical Power.''} \emph{Analysis for the
+Behavioral Sciences}, 273--406.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cotter2005}{}}%
+Cotter, Robert B., Jeffrey D. Burke, Magda Stouthamer-Loeber, and Rolf
+Loeber. 2005. {``Contacting Participants for Follow-up: How Much Effort
+Is Required to Retain Participants in Longitudinal Studies?''}
+\emph{Evaluation and Program Planning} 28 (1): 15--21.
+
+\leavevmode\vadjust pre{\hypertarget{ref-curran2011}{}}%
+Curran, Patrick J., and Daniel J. Bauer. 2011. {``The Disaggregation of
+Within-Person and Between-Person Effects in Longitudinal Models of
+Change.''} \emph{Annual Review of Psychology} 62: 583--619.
+
+\leavevmode\vadjust pre{\hypertarget{ref-curran2014a}{}}%
+Curran, Patrick J., Andrea L. Howard, Sierra Bainter, Stephanie T. Lane,
+and James S. McGinley. 2014. {``The {Separation} of {Between}-Person and
+{Within}-Person {Components} of {Individual} {Change} {Over} {Time}: {A}
+{Latent} {Curve} {Model} with {Structured} {Residuals}.''} \emph{J
+Consult Clin Psychol} 82 (5): 879--94.
+\url{https://doi.org/10.1037/a0035297}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-duncan2009}{}}%
+Duncan, Terry E., and Susan C. Duncan. 2009. {``The {ABC}'s of {LGM}:
+{An} {Introductory} {Guide} to {Latent} {Variable} {Growth} {Curve}
+{Modeling}.''} \emph{Social and Personality Psychology Compass} 3 (6):
+979--91. \url{https://doi.org/10.1111/j.1751-9004.2009.00224.x}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-durbin1950}{}}%
+Durbin, James, and Geoffrey S. Watson. 1950. {``Testing for Serial
+Correlation in Least Squares Regression: {I}.''} \emph{Biometrika} 37
+(3/4): 409--28.
+
+\leavevmode\vadjust pre{\hypertarget{ref-enders2010}{}}%
+Enders, Craig K. 2010. \emph{Applied {Missing} {Data} {Analysis}}.
+Guilford Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-farewell2017}{}}%
+Farewell, V. T., D. L. Long, B. D. M. Tom, S. Yiu, and L. Su. 2017.
+{``Two-{Part} and {Related} {Regression} {Models} for {Longitudinal}
+{Data}.''} \emph{Annual Review of Statistics and Its Application} 4 (1):
+283--315.
+\url{https://doi.org/10.1146/annurev-statistics-060116-054131}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-feingold2009}{}}%
+Feingold, Alan. 2009. {``Effect Sizes for Growth-Modeling Analysis for
+Controlled Clinical Trials in the Same Metric as for Classical
+Analysis.''} \emph{Psychological Methods} 14 (1): 43.
+
+\leavevmode\vadjust pre{\hypertarget{ref-feingold2019}{}}%
+---------. 2019. {``Time-Varying Effect Sizes for Quadratic Growth
+Models in Multilevel and Latent Growth Modeling.''} \emph{Structural
+Equation Modeling: A Multidisciplinary Journal} 26 (3): 418--29.
+
+\leavevmode\vadjust pre{\hypertarget{ref-french2010}{}}%
+French, David P., and Stephen Sutton. 2010. {``Reactivity of Measurement
+in Health Psychology: How Much of a Problem Is It? {What} Can Be Done
+about It?''} \emph{British Journal of Health Psychology} 15 (3):
+453--68.
+
+\leavevmode\vadjust pre{\hypertarget{ref-graham2009}{}}%
+Graham, John W. 2009. {``Missing {Data} {Analysis}: {Making} {It} {Work}
+in the {Real} {World}.''} \emph{Annual Review of Psychology} 60 (1):
+549--76. \url{https://doi.org/10.1146/annurev.psych.58.110405.085530}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hagler2019}{}}%
+Hagler, Donald J., SeanN. Hatton, M. Daniela Cornejo, Carolina Makowski,
+Damien A. Fair, Anthony Steven Dick, Matthew T. Sutherland, et al. 2019.
+{``Image Processing and Analysis Methods for the {Adolescent} {Brain}
+{Cognitive} {Development} {Study}.''} \emph{NeuroImage} 202 (November):
+116091. \url{https://doi.org/10.1016/j.neuroimage.2019.116091}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hamaker2015}{}}%
+Hamaker, Ellen L., Rebecca M. Kuiper, and Raoul P. P. P. Grasman. 2015.
+{``A Critique of the Cross-Lagged Panel Model.''} \emph{Psychological
+Methods} 20 (1): 102--16. \url{https://doi.org/10.1037/a0038889}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hill2016}{}}%
+Hill, Karl G., Danielle Woodward, Tiffany Woelfel, J. David Hawkins, and
+Sara Green. 2016. {``Planning for Long-Term Follow-up: {Strategies}
+Learned from Longitudinal Studies.''} \emph{Prevention Science} 17 (7):
+806--18.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kincaid2005}{}}%
+Kincaid, C. 2005. {``Guidelines for Selecting the Covariance Structure
+in Mixed Model Analysis, Paper 198-30 in {Proceedings} of the
+{Thirtieth} {Annual} {SAS} {Users} {Group} {Conference}.''} \emph{Inc.,
+Cary, North Carolina}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-king2018}{}}%
+King, Kevin M., Andrew K. Littlefield, Connor J. McCabe, Kathryn L.
+Mills, John Flournoy, and Laurie Chassin. 2018. {``Longitudinal Modeling
+in Developmental Neuroimaging Research: {Common} Challenges, and
+Solutions from Developmental Psychology.''} \emph{Developmental
+Cognitive Neuroscience}, Methodological {Challenges} in {Developmental}
+{Neuroimaging}: {Contemporary} {Approaches} and {Solutions}, 33
+(October): 54--72. \url{https://doi.org/10.1016/j.dcn.2017.11.009}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kraemer2014}{}}%
+Kraemer, Helena Chmura. 2014. {``Effect Size.''} \emph{The Encyclopedia
+of Clinical Psychology}, 1--3.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lenz2016}{}}%
+Lenz, Sylvia Tamara. 2016. {``Alan {Agresti} (2013): {Categorical} Data
+Analysis.''} \emph{Statistical Papers} 57 (3): 849.
+
+\leavevmode\vadjust pre{\hypertarget{ref-little1989}{}}%
+Little, Roderick J., and Donald B. Rubin. 1989. {``The {Analysis} of
+{Social} {Science} {Data} with {Missing} {Values}.''} \emph{Sociological
+Methods \& Research} 18 (2-3): 292--326.
+\url{https://doi.org/10.1177/0049124189018002004}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-little2013}{}}%
+Little, Todd D. 2013. \emph{The {Oxford} {Handbook} of {Quantitative}
+{Methods}, {Vol}. 2: {Statistical} {Analysis}}. Oxford University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-littlefield2021}{}}%
+Littlefield, Andrew K., Kevin M. King, Samuel F. Acuff, Katherine T.
+Foster, James G. Murphy, and Katie Witkiewitz. 2021. {``Limitations of
+Cross-Lagged Panel Models in Addiction Research and Alternative Models:
+{An} Empirical Example Using Project {MATCH}.''} \emph{Psychology of
+Addictive Behaviors}. \url{https://doi.org/10.1037/adb0000750}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-loughnan2020}{}}%
+Loughnan, Robert J., Clare E. Palmer, Wesley K. Thompson, Anders M.
+Dale, Terry L. Jernigan, and Chun Chieh Fan. 2020. {``Polygenic Score of
+Intelligence Is More Predictive of Crystallized Than Fluid Performance
+Among Children.''} \emph{bioRxiv}, 637512.
+
+\leavevmode\vadjust pre{\hypertarget{ref-luciana2018a}{}}%
+Luciana, M., J. M. Bjork, B. J. Nagel, D. M. Barch, R. Gonzalez, S. J.
+Nixon, and M. T. Banich. 2018. {``Adolescent Neurocognitive Development
+and Impacts of Substance Use: {Overview} of the Adolescent Brain
+Cognitive Development ({ABCD}) Baseline Neurocognition Battery.''}
+\emph{Developmental Cognitive Neuroscience} 32: 67--79.
+
+\leavevmode\vadjust pre{\hypertarget{ref-molnar2008}{}}%
+Molnar, Frank J., Brian Hutton, and Dean Fergusson. 2008. {``Does
+Analysis Using {`Last Observation Carried Forward'} Introduce Bias in
+Dementia Research?''} \emph{Cmaj} 179 (8): 751--53.
+
+\leavevmode\vadjust pre{\hypertarget{ref-morris2002}{}}%
+Morris, Scott B., and Richard P. DeShon. 2002. {``Combining Effect Size
+Estimates in Meta-Analysis with Repeated Measures and Independent-Groups
+Designs.''} \emph{Psychological Methods} 7 (1): 105.
+
+\leavevmode\vadjust pre{\hypertarget{ref-muthen2017}{}}%
+Muthén, L. K. 2017. {``Mplus User's Guide. {Los} {Angeles}: {Muthén} \&
+{Muthén}; 1998.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-orth2021}{}}%
+Orth, Ulrich, D. Angus Clark, M. Brent Donnellan, and Richard W. Robins.
+2021. {``Testing Prospective Effects in Longitudinal Research:
+{Comparing} Seven Competing Cross-Lagged Models.''} \emph{Journal of
+Personality and Social Psychology} 120 (4): 1013.
+
+\leavevmode\vadjust pre{\hypertarget{ref-robins1985}{}}%
+Robins, Lee. 1985. {``Epidemiology: {Reflections} on {Testing} the
+{Validity} of {Psychiatric} {Interviews} {\textbar} {JAMA} {Psychiatry}
+{\textbar} {JAMA} {Network}.''}
+\url{https://jamanetwork.com/journals/jamapsychiatry/article-abstract/493658}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rogosa1982}{}}%
+Rogosa, David, David Brandt, and Michele Zimowski. 1982. {``A Growth
+Curve Approach to the Measurement of Change.''} \emph{Psychological
+Bulletin} 92 (3): 726.
+
+\leavevmode\vadjust pre{\hypertarget{ref-salthouse2014}{}}%
+Salthouse, Timothy A. 2014. {``Why Are There Different Age Relations in
+Cross-Sectional and Longitudinal Comparisons of Cognitive
+Functioning?''} \emph{Current Directions in Psychological Science} 23
+(4): 252--56.
+
+\leavevmode\vadjust pre{\hypertarget{ref-shrout2018a}{}}%
+Shrout, Patrick E., Gertraud Stadler, Sean P. Lane, M. Joy McClure,
+Grace L. Jackson, Frederick D. Clavél, Masumi Iida, Marci E. J. Gleason,
+Joy H. Xu, and Niall Bolger. 2018. {``Initial Elevation Bias in
+Subjective Reports.''} \emph{PNAS} 115 (1): E15--23.
+\url{https://doi.org/10.1073/pnas.1712277115}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sullivan2017}{}}%
+Sullivan, Edith V., Ty Brumback, Susan F. Tapert, Devin Prouty, Rosemary
+Fama, Wesley K. Thompson, Sandra A. Brown, Kevin Cummins, Ian M.
+Colrain, and Fiona C. Baker. 2017. {``Effects of Prior Testing Lasting a
+Full Year in {NCANDA} Adolescents: Contributions from Age, Sex,
+Socioeconomic Status, Ethnicity, Site, Family History of Alcohol or Drug
+Abuse, and Baseline Performance.''} \emph{Developmental Cognitive
+Neuroscience} 24: 72--83.
+
+\leavevmode\vadjust pre{\hypertarget{ref-thompson2019}{}}%
+Thompson, Wesley K., Deanna M. Barch, James M. Bjork, Raul Gonzalez,
+Bonnie J. Nagel, Sara Jo Nixon, and Monica Luciana. 2019. {``The
+Structure of Cognition in 9 and 10 Year-Old Children and Associations
+with Problem Behaviors: {Findings} from the {ABCD} Study's Baseline
+Neurocognitive Battery.''} \emph{Developmental Cognitive Neuroscience}
+36: 100606.
+
+\leavevmode\vadjust pre{\hypertarget{ref-uban2018}{}}%
+Uban, Kristina A., Megan K. Horton, Joanna Jacobus, Charles Heyser,
+Wesley K. Thompson, Susan F. Tapert, Pamela A. F. Madden, and Elizabeth
+R. Sowell. 2018. {``Biospecimens and the {ABCD} Study: {Rationale},
+Methods of Collection, Measurement and Early Data.''}
+\emph{Developmental Cognitive Neuroscience}, The {Adolescent} {Brain}
+{Cognitive} {Development} ({ABCD}) {Consortium}: {Rationale}, {Aims},
+and {Assessment} {Strategy}, 32 (August): 97--106.
+\url{https://doi.org/10.1016/j.dcn.2018.03.005}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-volkow2018}{}}%
+Volkow, Nora D., George F. Koob, Robert T. Croyle, Diana W. Bianchi,
+Joshua A. Gordon, Walter J. Koroshetz, Eliseo J. Pérez-Stable, et al.
+2018. {``The Conception of the {ABCD} Study: {From} Substance Use to a
+Broad {NIH} Collaboration.''} \emph{Developmental Cognitive
+Neuroscience}, The {Adolescent} {Brain} {Cognitive} {Development}
+({ABCD}) {Consortium}: {Rationale}, {Aims}, and {Assessment} {Strategy},
+32 (August): 4--7. \url{https://doi.org/10.1016/j.dcn.2017.10.002}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wang2019a}{}}%
+Wang, Lijuan, Qian Zhang, Scott E. Maxwell, and C. S. Bergeman. 2019.
+{``On Standardizing Within-Person Effects: {Potential} Problems of
+Global Standardization.''} \emph{Multivariate Behavioral Research} 54
+(3): 382--403.
+
+\leavevmode\vadjust pre{\hypertarget{ref-watson2018}{}}%
+Watson, Nicole, Eva Leissou, Heidi Guyer, and Mark Wooden. 2018. {``Best
+{Practices} for {Panel} {Maintenance} and {Retention}.''} In
+\emph{Advances in {Comparative} {Survey} {Methods}}, 597--622. John
+Wiley \& Sons, Ltd. \url{https://doi.org/10.1002/9781118884997.ch29}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-zucker2018}{}}%
+Zucker, Robert A., Raul Gonzalez, Sarah W. Feldstein Ewing, Martin P.
+Paulus, Judith Arroyo, Andrew Fuligni, Amanda Sheffield Morris, Mariana
+Sanchez, and Thomas Wills. 2018. {``Assessment of Culture and
+Environment in the {Adolescent} {Brain} and {Cognitive} {Development}
+{Study}: {Rationale}, Description of Measures, and Early Data.''}
+\emph{Developmental Cognitive Neuroscience} 32: 107--20.
+
+\end{CSLReferences}
 

### 2023-04-21 2_HowTos_ResidualizedChangeScores.qmd
cli 12

--- 2023-04-19 2_Tutorials_ResidualizedChangeScores.qmd
@@
 
-## Overview
-Residualized change scores are a statistical method used to assess the degree of change in a variable while controlling for its initial level.
-The approach involves calculating the residualized change scores by regressing post-treatment scores on pre-treatment scores and using the resulting residuals as the measure of change.
-This estimated change score adjusts for baseline scores while ignoring any prior group assignments/differences.
-The residualized change score is often included in subsequent analysis, such as an examination of intervention effects in pre-test/post-test designs [@Kisbu-Sakarya2013].
-This method is useful when there is a high correlation between the pre-treatment and post-treatment scores, which can make it difficult to determine the true degree of change.
-Research studies have shown that this method can provide more accurate and reliable results compared to other methods of measuring change, such as raw change scores or gain scores (Kenny, 1979; Rogosa & Willett, 1985).
-
-Residualized change scores are a method used to analyze change in a variable over time, while accounting for the initial level of that variable.
-By using residualized change scores, we can examine the extent to which changes in one variable are associated with changes in another variable, controlling for the initial level of the variables.
-
-![Residualized Change Score 1_Tutorial](./figures/Residualized Change Score.png)
-
-xxxxxx
-### When to use Residualized Change Score Models?
-You should use residualized change scores in the following scenario:
-1.	You want to know xxxxx
-2.	Your variable xxxxx
-3.	You have xxxxx
-
-### Getting Started with Residualized Change Score Models
-In this tutorial, we will introduce the concept of residualized change scores and guide you through a simple example using a small dataset. By the end of this tutorial, you will be able to:
-
-1. Understand the basic concepts of residualized change scores.
-2. Calculate residualized change scores using example data.
-3. Interpret the results of the residualized change scores analysis.
-
-## Basic Example
-For this tutorial, we will use a simple dataset containing the following information for a group of individuals:
-
-- Time point (T1, T2)
-- Job satisfaction (independent variable)
-- Life satisfaction (dependent variable)
-
-| Individual | Time Point | Job Satisfaction | Life Satisfaction |
-| ---------- | ---------- | ---------------- | ----------------- |
-| A          | T1         | 7                | 6                 |
-| A          | T2         | 8                | 7                 |
-| B          | T1         | 6                | 5                 |
-| B          | T2         | 7                | 6                 |
-
-### Model Specification and Estimation
-
-To calculate residualized change scores, we will follow these steps:
-
-1. Calculate the raw change scores for each variable by subtracting the T1 value from the T2 value.
-2.
-Fit a regression model with the change score of the dependent variable (life satisfaction) as the outcome and the change score of the independent variable (job satisfaction) and the initial level of the dependent variable (T1 life satisfaction) as predictors.
-3. Extract the residuals from the regression model. These residuals represent the residualized change scores.
-
-Using the example dataset, we can calculate the residualized change scores for life satisfaction as follows:
-
-1. Calculate raw change scores:
-
-   | Individual | Job Satisfaction Change | Life Satisfaction Change |
-   | ---------- | ----------------------- | ------------------------ |
-   | A          | 1                       | 1                        |
-   | B          | 1                       | 1                        |
-
-2. Fit the regression model:
-
-   Life Satisfaction Change = b0 + b1 * Job Satisfaction Change + b2 * T1 Life Satisfaction
-
-3. Extract residuals:
-
-   | Individual | Residualized Change Score |
-   | ---------- | ------------------------- |
-   | A          | e1                        |
-   | B          | e2                        |
-
-### Interpreting the Results
-
-Residualized change scores represent the change in the dependent variable (life satisfaction) after accounting for the initial level of that variable and the change in the independent variable (job satisfaction).
-By examining the association between the residualized change scores of two variables, we can gain insights into how changes in one variable are related to changes in another variable, controlling for initial levels.
+Residualized Change Scores How tos test page

### 2023-04-21 3_HowTos_LinearMixedModels.qmd
cli 12

---
title: "Linear Mixed Models"
author: "Biostatistics Working Group"
---

Linear Mixed Models test page

### 2023-04-21 3_HowTos_LinearMixedModels.qmd
cli 15

@@
 
-Linear Mixed Models test page
+# ex1.interceptonly {.unnumbered}
+
+<style>
+div.blue pre { background-color:lightblue; }
+div.blue pre.r { background-color:lightgrey; }
+
+.nav-tabs > li.active:nth-child(1) > a {
+  background-color: ##964B00;
+}
+
+.nav-tabs > li > a {
+  background-color: ##FFFF00;
+  color: blue;
+}
+
+main-container {
+   max-width: unset;
+}
+.descriptives {
+   overflow: auto; /*do not use `resize`*/
+}
+
+.tableby {
+   overflow: auto;
+}
+
+.descriptives {
+   overflow: auto;
+   resize: both;
+}
+
+table, th, td {
+  padding-top: 0px !important;
+  padding-bottom: 0px !important;
+  vertical-align: middle;
+}
+</style>
+
+::: column-body-outset
+## Linear Mixed Model with a Random-Intercept (LMM:ri)
+<p>
+The **`LMM:ri`** is similar to traditional (fixed-effect) linear regression extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value
+
+In this example, we will use the **`LMM:ri`** to analyze trajectories of scores on the externalizing subscale of the child behavior checklist (CBCL) obtained across three measurement occasions in a sample of youth taking part in the ABCD Study.
+Our primary aim is to characterize stability and change in CBCL externalizing scores across assessments, while accounting for observations that are clustered within youth over time.
+To do so, we will use the **`LMM:ri`** to simultaneously model an overall sample mean trajectory (fixed effect) and subject-specific (random) effects that vary randomly about the sample mean trajectory.
+</p>
+
+:::
+:::
+
+###############################
+
+::: {.column-body-outset}
+## Preliminary Setup
+::: {.panel-tabset}
+### Load Libraries {.tabset .tabset-fade .tabset-pills}
+example-text before code block
+<div class = "blue">
+```{r pckg-install}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+
+## Install necessary packages (if not already installed)
+if (!("lme4" %in% installed.packages())) install.packages("lme4")
+if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
+if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
+if (!("arrow" %in% installed.packages())) install.packages("arrow")
+if (!("afex" %in% installed.packages())) install.packages("afex")
+if (!("janitor" %in% installed.packages())) install.packages("janitor")
+if (!("skimr" %in% installed.packages())) install.packages("skimr")
+if (!("sdamr" %in% installed.packages())) install.packages("sdamr")
+if (!("formatR" %in% installed.packages())) install.packages("formatR")
+if (!("report" %in% installed.packages())) install.packages("report")
+if (!("easystats" %in% installed.packages())) install.packages("easystats")
+if (!("emmeans" %in% installed.packages())) install.packages("emmeans")
+if (!("poorman" %in% installed.packages())) install.packages("poorman")
+if (!("parameters" %in% installed.packages())) install.packages("parameters")
+if (!("modelbased" %in% installed.packages())) install.packages("modelbased")
+if (!("DT" %in% installed.packages())) install.packages("DT")
+if (!("data.table" %in% installed.packages())) install.packages("data.table")
+if (!("arsenal" %in% installed.packages())) install.packages("arsenal")
+if (!("kableExtra" %in% installed.packages())) install.packages("kableExtra")
+if (!("equatiomatic" %in% installed.packages())) install.packages("equatiomatic")
+if (!("gtsummary" %in% installed.packages())) install.packages("gtsummary")
+```
+</div>
+example-text after code block
+
+### Load Packages
+<div class = "blue">
+```{r pckg-load}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+#| class.source: fold-hide
+
+## Load packages
+library(lme4)
+library(lmerTest)
+library(tidyverse)
+library(arrow)
+library(afex)
+library(janitor)
+library(skimr)
+library(sdamr)
+library(formatR)
+library(report)
+library(easystats)
+library(emmeans)
+library(poorman)
+library(parameters)
+library(modelbased)
+library(DT)
+library(data.table)
+library(arsenal)
+library(kableExtra)
+library(equatiomatic)
+library(gtsummary)
+```
+</div>
+
+### Config Options
+<div class = "blue">
+```{r config}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+
+knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
+                      comment=NA, cache=T, code_folding=T,
+                      R.options=list(width=220), fig.align='center', 
+                      out.width='75%', fig.asp=.75)
+```
+</div>
+:::
+:::
+
+::: {.column-body-outset}
+## Descriptives Overview
+::: {.panel-tabset}
+### Data Table {.tabset .tabset-fade .tabset-pills}
+<div class = "blue">
+```{r data-table}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+
+df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
+#df_long_sub<- read_csv("/Users/shawes/Desktop/df_long_sub.csv")
+
+## Create a viewable 'datatable' of the primary dataframe (df)
+datatable(head(df_long, 50, rownames = FALSE),extensions = 'AutoFill','ColReorder', options = list(autoFill = TRUE,colReorder = TRUE,
+  columnDefs = list(list(className = 'dt-center', targets = 10)),
+  order = list(list(3, 'asc'), list(4, 'desc')),
+  pageLength = 10,
+  lengthMenu = c(10, 15, 20),
+  initComplete = JS(
+    "function(settings, json) {",
+    "$(this.api().table().header()).css({'background-color': '#808080', 'color': '#fff'});",
+    "}")
+))
+
+## Not run: 
+# Specify table size by pixels
+#kable(cbind(df_long, df_long), "html") %>%
+#    kable_styling() %>%
+#    scroll_box(width = "100%", height = "500px")
+```
+</div>
+
+### Descriptives
+<div class = "blue">
+```{r descriptives}
+#| messages: FALSE
+#| echo: TRUE
+#| warning: FALSE
+#| results: "asis"
+
+## Create a descriptives table of study variables by measurement occasion 
+descriptives_1  <- tableby.control(test=FALSE, total=FALSE,
+                               numeric.test="kwt", cat.test="chisq",
+                               numeric.stats=c("N", "meansd", "median", "range" 
+                                               ), #"Nmiss2"
+                               cat.stats=c("countpct"), #"Nmiss2"
+                               stats.labels=list(N='Count', meansd="Mean (SD)", median
+                                                 ='Median', range='Min - Max'
+                                                 )) #, Nmiss2 ='Missing'
+
+my_cont_labels <- list(
+  age = "Age",
+  vg_total = "Weekly # of Video Gaming Hrs",
+  cbcl_extern = "CBCL Externalizing Scale"
+)
+
+tab_descriptives_1 <- tableby(event ~ age + vg_total + 
+                            cbcl_extern,
+                            data=df_long, control=descriptives_1)
+
+#summary(tab_descriptives_1, labelTranslations = my_cont_labels , text=TRUE, title = #"Continuous Outcomes", term.name = TRUE)
+
+# Push table object through kable and kable_styling
+tab_descriptives_1 %>%
+  summary(text=TRUE, digits.pct=1, digits=1) %>%
+  kable(caption = "Continuous Outcomes") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, html_font = "Cambria",
+                font_size = 15,
+                position = "center", fixed_thead = T) %>%
+  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
+  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
+  column_spec(2, width = "20em", border_right = T) %>%
+  column_spec(3, width = "20em", border_right = T) %>%
+  footnote(general = "Here is a general comments of the table. ") %>%
+  scroll_box(width = "75%", height = "500px")
+```
+</div>
+
+<div class = "blue">
+```{r catvar-table}
+#| messages: FALSE
+#| warning: FALSE
+#| results: "asis"
+
+## Create a descriptives table of study variables by measurement occasion 
+descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
+                               numeric.test="kwt", cat.test="chisq",
+                               numeric.stats=c("N", "meansd", "median", "range"
+                                               ), # "Nmiss2"
+                               cat.stats=c("countpct"), # "Nmiss2"
+                               stats.labels=list(N='Count', meansd="Mean (SD)", median
+                                                 ='Median', range='Min - Max'
+                                                 )) # , Nmiss2 ='Missing'
+
+my_cat_labels  <- list(
+  event = "Year",
+  sex = "Sex",
+  mature_vg = "Mature Video Games"
+  )
+
+tab_descriptives_2 <- tableby(event ~ sex + mature_vg, 
+                            data=df_long, control=descriptives_2)
+
+#summary(tab_descriptives_2, labelTranslations = my_cat_labels , text=TRUE, title = #"Categorical Outcomes", term.name = TRUE)
+
+# Push table object through kable and kable_styling
+tab_descriptives_2 %>%
+  summary(text=TRUE, digits.pct=1, digits=1) %>%
+  kable(caption = "Categorical Outcomes") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size = 15,
+                position = "center", fixed_thead = T) %>%
+  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
+  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
+  column_spec(2, width = "20em", border_right = T) %>%
+  #column_spec(3, width = "20em", border_right = T) %>%
+  footnote(general = "Here is a general comments of the table. ") %>%
+  scroll_box(width = "75%", height = "500px")
+
+```
+</div>
+:::
+:::
+
+::: {.column-body-outset}
+## Results 
+::: {.panel-tabset}
+### Build Model {.tabset .tabset-fade .tabset-pills}
+<div class = "blue">
+```{r build-model}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+
+## Linear Mixed Model with a random intercept (LMM-ri)
+random_intercepts <- lmer(cbcl_extern ~ 1 + event + sex + (1|ids), data = df_long, REML=T)
+
+print(random_intercepts)
+```
+</div>
+
+> The code snippet above tells R to run a linear mixed model using lmer
+> from the lme4 library. The left side of the "\~" symbol specifies the
+> dependent variable; the right side specifies days as the independent
+> variable(s). The code (1 \| Subject) specifies a random intercept for
+> each subject.
+
+### Model Output
+<div class = "blue">
+```{r model-output}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+#| code-summary: testing
+
+## Output and reports extending from the LMM-ri analyses
+summary(random_intercepts)
+confint(random_intercepts, level = 0.95, method = "Wald")
+report(random_intercepts)
+report_performance(random_intercepts)
+report_statistics(random_intercepts)
+random <- estimate_grouplevel(random_intercepts)
+```
+</div>
+
+### Model Report
+<div class = "blue">
+```{r mod-report}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+
+#df_long %>%
+#  report() %>%
+#  summary()
+```
+</div>
+:::
+:::
+
+::: {.column-body-outset}
+## Visualizations 
+::: {.panel-tabset}
+### Extract Params {.tabset .tabset-fade .tabset-pills}
+<div class = "blue">
+```{r extrct-FE}
+#| messages: FALSE
+#| warning: FALSE
+
+## Obtain LMM-ri model parameters to plot/graph of results
+model_intercept <- as.numeric(fixef(random_intercepts)[1])
+model_slope <- as.numeric(fixef(random_intercepts)[2])
+```
+</div>
+
+> This code allows for extracting the fixed effects estimates for the
+> model intercept and slope.
+
+<div class = "blue">
+```{r extrct-RE}
+#| class.source: 'fold-hide'
+#| message: FALSE
+#| warning: FALSE
+#| echo: FALSE
+
+intercepts <- rep(coef(random_intercepts)$ids[,1], each = 4)
+```
+</div>
+
+> Extract the individual participant intercepts (random effects) for this model and add
+> it to the data frame for plotting
+
+### Diagnostic Plots
+<div class = "blue">
+```{r diag-plots}
+#| messages: FALSE
+#| warning: FALSE
+#| echo: FALSE
+#| column: screen-inset-shaded
+#| layout-nrow: 2
+
+## xxxxx plot
+random_intercepts.std <- resid(random_intercepts)/sd(resid(random_intercepts))
+plot(random_intercepts.std, ylab="Standardized Residuals")
+
+## Outlier Tests. 
+#car::outlierTest(random_intercepts)
+car::infIndexPlot(random_intercepts)
+
+## qqplot
+ggplot(as.data.frame(random_intercepts.std), aes(sample = random_intercepts.std)) +
+  geom_qq() +
+  geom_qq_line()
+
+## Predicted vs Actual plot
+predicted <- estimate_expectation(random_intercepts, data = "grid")
+plot(predicted)
+
+## Estimated means plot
+means <- estimate_means(random_intercepts)
+plot(means)
+
+```
+</div>
+
+> Build diagnostic plots. Observations with a Bonferroni p \< .05 might
+> be considered as outliers and might need further investigation.
+
+### Bar Plot
+```{r barplot}
+#| messages: FALSE
+#| warning: FALSE
+#| echo: TRUE
+
+# Basic ggplot2 barplot with mean
+p<-df_long %>%
+    filter(!is.na(mature_vg)) %>% # filter on non-missing values  
+    ggplot(aes(x = mature_vg, y = cbcl_extern, fill = sex)) +
+    geom_col(position = "dodge",
+           stat = "summary",
+           fun = "mean",
+           #fill = "#AA4A44",
+           #color = "#0099f9"
+           ) +
+    #scale_fill_brewer(palette = "Set1") +
+    scale_fill_manual(values = c("#3db5ff", "#0099f9")) +
+    #scale_fill_gradient(high = "#B00B69", low = "#0e0e63") +
+    #geom_text(aes(label = cu_traits), position = position_dodge(0.9), 
+              #vjust = 2, size = 4, color = "#ffffff") +
+    labs(title = "Behavior Problems by Video Game Groups", subtitle = "Simple bar chart",
+         caption = "Simple caption", x = "DBD Group", y = "CBCL Externalizing") +
+              #coord_flip() +
+    #geom_hline(yintercept = mean(df$cbcl_agg), linetype = "dashed", size = 1) +
+    theme_minimal() +
+    theme(
+          plot.title = element_text(color = "#0099f9", size = 20, hjust = 0.5),
+          plot.subtitle = element_text(face = "bold", hjust = 1),
+          plot.caption = element_text(face = "italic", hjust = 0),
+          axis.title.x = element_text(color = "#0099f9", size = 15, face = "bold"),
+          axis.title.y = element_text(size = 15, face = "italic")
+  )
+p
+```
+
+### Random Intercept LMM Plot/Graph
+<div class = "blue">
+```{r lmmri-graph}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+#| label: my-figure
+#| fig-asp: 0.61
+#| fig-cap: |
+#|   This is a caption for my figure, 
+#|   using YAML formatting, etc.#|
+
+#random_intercepts <- lmer(Externalizing ~ 1 + Event + DBD + Sex + (1|ids), data = df, #REML=T)
+
+## LMM-ri plot of estimated random intercepts 
+model <- lmer(cbcl_extern ~ mature_vg + (1 | sex), data = df_long)
+preds <- estimate_relation(model, include_random = TRUE)
+
+## Adding fixed effect trajectory to LMM-ri plot 
+fixed_pred <- estimate_relation(model) # This time, include_random is FALSE (default)
+plot(preds, ribbon = list(alpha = 0)) + # Previous plot
+  geom_ribbon(data = fixed_pred, aes(x = mature_vg, ymin = CI_low, ymax = CI_high), alpha = 0.4) +
+  geom_line(data = fixed_pred, aes(x = mature_vg, y = Predicted), size = 2)
+
+######
+# ggplot(df_long_sub, aes(x = event, y = cbcl_agg, color=sex, shape=sex)) + 
+# geom_abline(slope = model_slope, intercept = model_intercept, linetype = "solid", color = "red", linewidth = 1) + 
+# geom_point(color = "grey70") + 
+# geom_smooth(method=lm, se=FALSE, fullrange=TRUE)+
+# scale_shape() + 
+# geom_segment(aes(x = event, xend = event, y = cbcl_agg, yend = fitted(cbcl_agg)), color = "grey70") + 
+# scale_y_continuous(expand = c(0, 0), breaks = c(0, 20, 40, 60, 80, 100), limits = c(0, 100)) + 
+# scale_x_continuous(expand = c(0, 0), breaks = c(1, 2, 3, 4), limits = c(0, 4)) +
+# theme(panel.background = element_blank(),         
+#         panel.grid.major = element_blank(),
+#         panel.grid.minor = element_blank(),
+#         panel.border = element_rect(colour = "black", fill = NA),
+#         legend.position = "none",
+#         axis.text = element_text(size = 14), 
+#         axis.title = element_text(size = 14)) +
+#         labs (x = "Event", y= "CBCL Aggression")
+
+#Save the plot
+#ggsave("random_intercept.png", units = "in", width = 9, height = 6, dpi = 300)
+
+```
+</div>
+:::
+:::
+
+<p class="text-center" style="background-color: gray">Figure x: The overall group-mean (fixed effects) trajectory is shown in blue.
+The faded lines represent each individual youth's estimated trajectory.
+An examination of this figure shows happiness scores to be increasing across measurement occasions, however, there appears to be substantial variability in the youth's initial happiness scores.</p>
+
+::: column-body-outset
+#### Interpretation
+The estimated correlation between x and y was "r cor(x,y)`.
+There are "r nrow(my_data)` individuals.
+
+From the fixed effects section of the model summary, we can conclude
+that there is strong evidence that RT increased significantly over time
+(i.e., \# of years). On average for each additional year, RT increased
+by 10.46 (b = 10.46, SE = 0.80, p \< .001).We are 95% confident that the
+average increase was between 8.89 and 12.04.
+
+In the above, we estimate that the average intercept across all
+participants is 251.4. Results from the random effects section below
+show that the variance of the intercept for Subject is 1378.2. Taking
+the square root, the standard deviation of the intercept is thus 37.1.
+
+We can calculate the 95% coverage interval as 251.4 ± 1.96\*37.1. The
+lower bound of the 95% coverage interval is thus 178.7 and the upper
+bound is 324.1. We therefore estimated that 95% of the participants have
+an intercept between 178.7 and 324.1. This means that 95% of the
+participants have a reaction time between 178.7 and 324.1 at Day 0. This
+is not to be confused with the 95% confidence interval of the intercept.
+The 95% confidence interval is (232.3, 270.5), and this indicates that
+we are 95% confident that the average intercept is somewhere between
+232.3 and 270.5.
+
+In this model, we have accounted for the repeated measures design
+(observations nested within individuals) by including a random intercept
+for each participants. Each individual has his/her own intercept. The
+effect of time(Year) on RT is assumed to be the same across individuals.
+This assumption can be relaxed by fitting a random slope model.
+
+#### Mock Write-up
+
+A random slope model is used to test if sleep deprivation affects
+reaction time. To account for the repeated measures design, a random
+intercept was specified for participants. The random slope for days of
+sleep deprivation was included in the model to allow the effect of sleep
+deprivation to vary across participants. Results are shown in Table 1.
+Using a significant level of 0.05, results indicate that sleep
+deprivation significantly increased reaction time. On average, each
+additional day of sleep deprivation increased reaction time by 10.47ms
+(b = 10.47, 95% CI = \[7.44, 13.50\], p \< .001). Model fit comparison
+between model with and without random slope for sleep deprivation shows
+that the effect of sleep deprivation varied across participants, χ2(2)=
+42.14, p \< .001. The 95% coverage interval for the random slope of
+sleep deprivation is (-1.14, 22.07), indicating that the effect of sleep
+deprivation was between -1.14 and 22.07 for 95% of the participants.
+:::
+
+## References
+
+[@brown2021] <!-- [@Singmann2019]
+[@sarafian]
+[#@https://m-clark.github.io/]
+[#@Fitzmaurice]
+[#@mspeekenbrink.github.io]
+[#https://cran.r-project.org/web/packages/lme4/lme4.pdf] -->

### 2023-04-22 3_HowTos_LinearMixedModels.qmd
added link to pdf version of manuscript

@@
 
-# ex1.interceptonly {.unnumbered}
-
-<style>
-div.blue pre { background-color:lightblue; }
-div.blue pre.r { background-color:lightgrey; }
-
-.nav-tabs > li.active:nth-child(1) > a {
-  background-color: ##964B00;
-}
-
-.nav-tabs > li > a {
-  background-color: ##FFFF00;
-  color: blue;
-}
-
-main-container {
-   max-width: unset;
-}
-.descriptives {
-   overflow: auto; /*do not use `resize`*/
-}
-
-.tableby {
-   overflow: auto;
-}
-
-.descriptives {
-   overflow: auto;
-   resize: both;
-}
-
-table, th, td {
-  padding-top: 0px !important;
-  padding-bottom: 0px !important;
-  vertical-align: middle;
-}
-</style>
-
-::: column-body-outset
 ## Linear Mixed Model with a Random-Intercept (LMM:ri)
 <p>
-The **`LMM:ri`** is similar to traditional (fixed-effect) linear regression extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value
+The LMM:ri is similar to traditional (fixed-effect) linear regression extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value zzzzt
 
-In this example, we will use the **`LMM:ri`** to analyze trajectories of scores on the externalizing subscale of the child behavior checklist (CBCL) obtained across three measurement occasions in a sample of youth taking part in the ABCD Study.
+In this example, we will use the LMM:ri to analyze trajectories of scores on the externalizing subscale of the child behavior checklist (CBCL) obtained across three measurement occasions in a sample of youth taking part in the ABCD Study.
 Our primary aim is to characterize stability and change in CBCL externalizing scores across assessments, while accounting for observations that are clustered within youth over time.
-To do so, we will use the **`LMM:ri`** to simultaneously model an overall sample mean trajectory (fixed effect) and subject-specific (random) effects that vary randomly about the sample mean trajectory.
+To do so, we will use the LMM:ri to simultaneously model an overall sample mean trajectory (fixed effect) and subject-specific (random) effects that vary randomly about the sample mean trajectory.
 </p>
 
-:::
-:::
-
-###############################
-
-::: {.column-body-outset}
 ## Preliminary Setup
@@
 :::
-:::
 
-::: {.column-body-outset}
 ## Descriptives Overview
@@
 :::
-:::
 
-::: {.column-body-outset}
 ## Results 
@@
 :::
-:::
 
-::: {.column-body-outset}
 ## Visualizations 
@@
 :::
-:::
 
@@
 
-::: column-body-outset
 #### Interpretation

### 2023-04-22 TailwindTest.qmd
added link to pdf version of manuscript

---
title: "Tailwind Test Doc"
editor: visual
---

ssss

<script src="https://cdn.tailwindcss.com"></script>

sssss

xxxx

:::{.text-xl .text-blue .rounded-lg .py-4 .px-2 .bg-gradient--to-r .from-green-700 .to-blue-900 .shadow-xl}

Place Content here

:::

xxxxx

.bg-gradient--to-r

.from-green-700

.to-blue-900

.shadow-xl

![](images/image-1130704809.png)

### 2023-04-22 TailwindTest.qmd
testing tailwind cdn

@@
 editor: visual
+
+format:
+  html:
+    css: styles.css
+
 ---
+<!-- tailwind cdn temp link -->
+<script src = "https://cdn.tailwindcss.com"></script>
+<!---->
 
-ssss
+<div class="bg-gradient-to-r from-indigo-500 from-10% via-purple-500 via-30% to-pink-500 to-90% ...">
+  <!-- ... -->
+  xxxxxx
+</div>
 
-<script src="https://cdn.tailwindcss.com"></script>
+xxxxxx
 
-sssss
-
-xxxx
+<div class="bg-gradient-to-r from-indigo-500 from-10% via-purple-500 via-30% to-pink-500 to-90% ...">
+  <!-- ... -->
+  include content
+</div>
 
@@
 
+<p .bg-sky-900>
+
+include content
+
+</p>
+
 xxxxx
+
+.bg-sky-900
 
@@
 .shadow-xl
-
-![](images/image-1130704809.png)

### 2023-04-22 tailwind.config.js

@@ -1,0 +1,11 @@
+/** @type {import('tailwindcss').Config} */
+module.exports = {
+  content: [],
+  theme: {
+    extend: {},
+  },
+  plugins: [
+    require('@tailwindcss/forms'),
+    require('@tailwindcss/typography'),
+  ]
+}

### 2023-04-23 TailwindTest2.qmd
landing page update

---
title: "TailwindTest2"
format:
  html:
    css: styles.css
    theme: cosmo
    fontsize: 1.1em
    linestretch: 1.7
---

<!-- tailwind cdn temp link -->
<script src = "https://cdn.tailwindcss.com"></script>
<!---->

<div class="bg-gradient-to-r from-indigo-500 from-10% via-purple-500 via-30% to-pink-500 to-90% ...">
This is github pages test site for the longitudinal analysis project. The information included within the site is for testing purposes only and contents may be inaccurate.
</div>

::: {.grid}

::: {.g-col-4}
This column takes 1/3 of the page
:::

::: {.g-col-8}
This column takes 2/3 of the page
![lights](./figures/lights.png)
:::

:::

zzzzzzzzz

---
#title: "Temporary Landing Page for Longitudinal Project"
# format: 
#   html:
#     page-layout: full
#     theme:
#       light: cosmos #flatly
#       dark: darkly
#     css: styles.css
#     smooth-scroll: true
#     linestretch: 1.7

---
<!-- tailwind cdn temp link -->
<script src = "https://cdn.tailwindcss.com"></script>
<!---->

<div class="bg-slate-400">

<div class="text-3xl text-center bg-gradient-to-r from-blue-700 from-10% via-blue-500 via-30% to-blue-300 to-90%">
This is the temporary landing page for the Longitudinal Analysis Project. <br> The site is currently under construction and used for testing purposes only. <br> Contents may be inaccurate.
</div>

<br>

::: {.grid}

::: {.g-col-2}
:::

::: {.g-col-8}
![](./figures/lights.png)
:::

::: {.g-col-2}
:::

:::

</div>

### 2023-04-23 index.qmd
landing page update

--- 2023-04-19 index.qmd
@@
 ---
-title: "swhawes.github.io"
+#title: "Temporary Landing Page for Longitudinal Project"
+#format:
+#  html:
+#    css: styles.css
+#    theme: cosmo
+#    fontsize: 1.1em
+    #linestretch: 1.7
 ---
 
-This is a Quarto website.
+<div class="text-3xl text-center">
+This is the temporary landing page for the Longitudinal Analysis Project. <br> The site is currently under construction and used for testing purposes only. <br> Contents may be inaccurate.
+</div>
 
-To learn more about Quarto websites visit <https://quarto.org/docs/websites>.
+<br>
+
+::: {.grid}
+
+::: {.g-col-2}
+:::
+
+::: {.g-col-8}
+![](./figures/lights.png)
+:::
+
+::: {.g-col-2}
+:::
+
+:::
+
+</div>

### 2023-04-23 tailwind.config.js

@@ -7,5 +7,6 @@ module.exports = {
   plugins: [
     require('@tailwindcss/forms'),
     require('@tailwindcss/typography'),
+    require('@tailwindcss/aspect-ratio'),
   ]
 }

### 2023-04-23 1b_Difference Scores simple regression_HowTo.qmd
hide 3b how-to for mixed models temporarily

---
format: 
  #html:
  #  toc: true
  #  code-fold: true
  docx:
    toc: true
prefer-html: true
---

# Simple Regression {.unnumbered}

## Overview

<p>

This example will examine whether average change in CBCL externalizing scores from Basline (T0) to the 1-Year follow-up (T1) differs significantly between boys and girls taking part in the ABCD Study. This analysis is conducted in two primary steps: 1) computing a difference score (DiffScore = CBCL EXT T2 - CBCL EXT T1); 2) conducting a simple regression on the difference score. Our primary aim is to determine whether group membership (boys, girls) predict the average difference value in CBCL externalizing scores from T1 to T2.

</p>

## Preliminary Setup
::: {.panel-tabset}
### Install Packages {.tabset .tabset-fade .tabset-pills}
<div class = "blue">

> **This code installs the r packages necessary for this example, if they are not already installed**

```{r}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: FALSE

#rm(list = ls())

# Install necessary packages (if not already installed)
if (!("lme4" %in% installed.packages())) install.packages("lme4")
if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("arrow" %in% installed.packages())) install.packages("arrow")
if (!("arsenal" %in% installed.packages())) install.packages("arsenal")
if (!("kableExtra" %in% installed.packages())) install.packages("kableExtra")
if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
if (!("rstatix" %in% installed.packages())) install.packages("rstatix")
```
</div>

### Load Packages
<div class = "blue">

**This code loads the r libraries necessary for this example**

```{r pckg-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| class.source: fold-hide

#rm(list = ls())

#Load packages
library(lme4)
library(lmerTest)
library(tidyverse)
library(arrow)
library(arsenal)
library(kableExtra)
library(rstatix)
library(ggpubr)
```

</div>

### Config Options
<div class = "blue">

**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                                  comment=NA, cache=T, code_folding=T,
                                  R.options=list(width=220), fig.align='center', 
                                  out.width='75%', fig.asp=.75)
```
</div>
:::

## Descriptives Overview
::: {.panel-tabset}
### Read and View Data {.tabset .tabset-fade .tabset-pills}
<div class = "blue">

**This code reads in and shows the data to be used in the current example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Read data
df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
df_wide <- read_csv("/Users/shawes/Desktop/data/df_wide.csv")
str(df_wide)
# set types
df_long$ids <- as.factor(df_long$ids)
df_long$event <- as.factor(df_long$event)
df_long$site_id <- as.factor(df_long$site_id)
df_long$fam_id <- as.factor(df_long$fam_id)
df_long$sex <- as.factor(df_long$sex)
df_long$age <- as.numeric(df_long$age)
df_long$mature_vg <- as.numeric(df_long$mature_vg)
df_long$cbcl_extern <- as.numeric(df_long$cbcl_extern)
df_long$vg_total <- as.numeric(df_long$vg_total)

```

</div>

:::

### Continuous Outcomes

::: blue
**This code creates a table of descriptive information for continuous
outcomes**

```{r contvar-table}
#| messages: FALSE
#| echo: TRUE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_1 <- tableby.control(test=FALSE, total=FALSE,
numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range" 
), #"Nmiss2"
cat.stats=c("countpct"), #"Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median
='Median', range='Min - Max'
))  # , Nmiss2 ='Missing'

my_cont_labels <- list(
  age = "Age",
  vg_total = "Weekly # of Video Gaming Hrs",
  cbcl_extern = "CBCL Externalizing Scale"
)

tab_descriptives_1 <- tableby(event ~ age + vg_total + 
                                      cbcl_extern,
                                      data=df_long, control=tab_descriptives_1)

#summary(tab_descriptives_1, labelTranslations = my_cont_labels , text=TRUE, title = #"Continuous Outcomes", term.name = TRUE)

# Push table object through kable and kable_styling
tab_descriptives_1 %>%
                summary(text=TRUE, digits.pct=1, digits=1) %>%
            kable(caption = "Continuous Outcomes") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, html_font = "Cambria",
                font_size = 15,
                position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
            column_spec(3, width = "20em", border_right = T) %>%
            footnote(general = "Here is a general comments of the table. ") %>%
            scroll_box(width = "75%", height = "500px")
```   
:::

### Categorical Outcomes

::: blue
**This code creates a descriptives table for categorical outcomes**

```{r catvar-table}
#| class.source: 'fold-hide'
#| messages: FALSE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"
), # "Nmiss2"
cat.stats=c("countpct"), # "Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median
='Median', range='Min - Max'
)) # , Nmiss2 ='Missing'

my_cat_labels  <- list(
  event = "Year",
  sex = "Sex",
  mature_vg = "Mature Video Games"
  )

tab_descriptives_2 <- tableby(event ~ sex + mature_vg, data=df_long, control=tab_descriptives_2)

#summary(tab_descriptives_2, labelTranslations = my_cat_labels , text=TRUE, title = #"Categorical Outcomes", term.name = TRUE)

# Push table object through kable and kable_styling
tab_descriptives_2 %>%
            summary(text=TRUE, digits.pct=1, digits=1) %>%
            kable(caption = "Categorical Outcomes") %>%
            kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size = 15,
                position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
            column_spec(2, width = "20em", border_right = T) %>%
  #column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")

```
:::
:::

## Results

<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue

The code snippet below tells R to compute a difference score by subtracting each participant's Externalizing score at T2 from their Externalizing score at T1. 
Next, a simple regression analyses is conducted to examine whether a grouping variable (participant sex) significantly predicts the difference score value (indicating significant group differences in the average difference score). The simple regression function (lm) is provided by the r 'stats' package.

xxx
First, compute the difference ($\Delta$) between a score on some measure ($x$) assessed at baseline ($x_{t1}$) and follow-up ($x_{t2}$). The result of this formula ($\Delta$=$x_{t2}$ - $x_{t1}$) can be included as the outcome variable in a regression model that analyzes the role of the grouping variable (e.g., 0 = boy; 1 = girl) on changes in scores on the measure across follow-ups.
xxx

**STEP 1: Compute Difference Score**
```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Compute difference score based on CBCL Externalizing subscale scores at baseline (t1) and 1-Year Follow-up (t2)
df_wide$diffscore <- (df_wide$cbcl_extern_Year_1 - df_wide$cbcl_extern_Baseline)

```

```{r model-checking}
# Compute statistical summaries for the difference score variable
summary(df_wide$diffscore)

```

This summary of the difference score variable indicates xxxxx.

```{r}
hist(df_long$vg_total)

## Summary statistics
summary<-df_long %>%
group_by(event) %>%
get_summary_stats(cbcl_extern, type = "mean_sd")
data.frame(summary)

```

This histogram indicates xxxxxx. The summary command shows xxxxxx.

```{r shapiro-wilks}
### Shapiro-Wilk test and normality (Q-Q) plot (visualization of correlation between a given sample and the normal distribution)
shapiro.test(df_wide$cbcl_extern_Baseline[0:5000])

```

This Shapiro-wilkes test shows xxxxxx.

```{r qqplot}
qqplot <- ggqqplot(df_wide$cbcl_extern_Baseline,
    ylab = "Externalizing Difference Score", xlab = FALSE,
    ggtheme = theme_minimal()
)

suppressWarnings(print(qqplot)) 
```

This qqplot shows xxxxxx.

**STEP 2: Conduct simple regression on Difference Score**
```{r build-model-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

#Simple linear regression is used to examine the relationship between a continuous predictor variable and the average difference score.

#Regression model
result <- lm(df_wide$diffscore ~ vg_total_Baseline, data = df_wide)
#summarize the reults
summary(result)
```

The output from our model provides: i. a parameter estimate; ii. standard error; iii. p-value. In our example p = .001 which is less than .05 indicating significant differences in predicted cbcl externalizing scores between boys and girls in the ABCD Study. An examination of the regression coeefficient for sex (female = 0; male = 1) is b = .xx, indicating the difference in cbcl externalizing scores from Time 1 to Time 2 is significantly greater in boys taking part in the ABCD Study relative to girls. Then we conclude that participant sex (does/does not) predict change in cbcl externalizing scores from t1 to t2.

### Model Plots

::: blue
```{r scatterplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

#Scatterplot to visualize relationship between a continuous predictor and a difference score
scatterplot <- ggplot (df_wide, aes(x = diffscore, y = vg_total_Baseline)) + geom_point(size = 3) + geom_smooth(method = lm, se = F) +
xlab("Difference Score (x)") +
ylab("Screentime hours (y) Baseline")

suppressWarnings(print(scatterplot))
```

Examination of this scatterplot indicates xxxxx.
:::

> This output shows xxxxxxxx. Briefly walk through each metric
:::

## Wrapping Up
<div class = "blue">

<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue

A simple regression analysis was conducted to examine whether participant's sex predicted change in cbcl externalzing scores measured across two timepoints. Findings showed a significant positive effect such that boys demonstrated increases in cbcl externalzing scores relative to girls (b = .xx; SE = .xx; p = .xx)

:::
:::

### 2023-04-23 2_HowTos_ResidualizedChangeScores.qmd
cli 15
hide 3b how-to for mixed models temporarily

@@
 
-Residualized Change Scores How tos test page
+# Mulitple Regression {.unnumbered}
+
+## Overview
+
+<p>
+insert text
+</p>
+
+## Preliminary Setup
+::: {.panel-tabset}
+### Install Packages {.tabset .tabset-fade .tabset-pills}
+<div class = "blue">
+
+> **This code installs the r packages necessary for this example, if they are not already installed**
+
+```{r}
+#| class.source: 'fold-hide'
+#| message: FALSE
+#| warning: FALSE
+#| echo: FALSE
+
+#rm(list = ls())
+
+# Install necessary packages (if not already installed)
+if (!("rstatix" %in% installed.packages())) install.packages("rstatix")
+if (!("lme4" %in% installed.packages())) install.packages("lme4")
+if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
+if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
+if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
+if (!("emmeans" %in% installed.packages())) install.packages("emmeans")
+```
+</div>
+
+### Load Packages
+<div class = "blue">
+
+> **This code loads the r libraries necessary for this example**
+
+```{r pckg-load}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+#| class.source: fold-hide
+
+#rm(list = ls())
+
+#Load packages
+library(rstatix)
+library(lme4)
+library(lmerTest)
+library(tidyverse)
+library(ggpubr)
+library(emmeans)
+library(arrow)
+library(arsenal)
+library(kableExtra)
+library(rstatix)
+
+```
+
+</div>
+
+### Config Options
+<div class = "blue">
+
+**This code configures knitr code chunk options**
+
+```{r config}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+
+knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
+                      comment=NA, cache=T, code_folding=T,
+                      R.options=list(width=220), fig.align='center',
+                      out.width='75%', fig.asp=.75)
+```
+</div>
+:::
+
+## Descriptives Overview
+::: {.panel-tabset}
+### Read and View Data {.tabset .tabset-fade .tabset-pills}
+<div class = "blue">
+
+**This code reads in and shows the data to be used in the current example**
+
+```{r read-data}
+#| echo: TRUE
+#| messages: FALSE
+#| warning: FALSE
+
+## Read data
+df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
+df_wide <- read_csv("/Users/shawes/Desktop/data/df_wide.csv")
+
+# set types (if necessary)
+df_long$ids <- as.factor(df_long$ids)
+df_long$event <- as.factor(df_long$event)
+df_long$site_id <- as.factor(df_long$site_id)
+df_long$fam_id <- as.factor(df_long$fam_id)
+df_long$sex <- as.factor(df_long$sex)
+df_long$age <- as.numeric(df_long$age)
+df_long$mature_vg <- as.numeric(df_long$mature_vg)
+df_long$cbcl_extern <- as.numeric(df_long$cbcl_extern)
+df_long$vg_total <- as.numeric(df_long$vg_total)
+```
+
+</div>
+
+:::
+
+### Continuous Outcomes
+
+::: blue
+**This code creates a table of descriptive information for continuous
+outcomes**
+
+```{r contvar-table}
+#| messages: FALSE
+#| echo: TRUE
+#| warning: FALSE
+#| results: "asis"
+
+## Create a descriptives table of study variables by measurement occasion 
+tab_descriptives_1 <- tableby.control(test=FALSE, total=FALSE, numeric.test="kwt", cat.test="chisq",
+numeric.stats=c("N", "meansd", "median", "range"), #"Nmiss2"
+cat.stats=c("countpct"), #"Nmiss2"
+stats.labels=list(N='Count', meansd="Mean (SD)", median ='Median', 
+                  range='Min - Max'))  #, Nmiss2 ='Missing'
+my_cont_labels <- list(age = "Age", vg_total = "Weekly # of Video Gaming Hrs",
+            cbcl_extern = "CBCL Externalizing Scale"
+)
+
+tab_descriptives_1 <- tableby(event ~ age + vg_total + 
+                                      cbcl_extern,
+                                      data=df_long, control=tab_descriptives_1)
+
+# Push table object through kable and kable_styling
+tab_descriptives_1 %>%
+  summary(text=TRUE, digits.pct=1, digits=1) %>%
+  kable(caption = "Continuous Outcomes") %>%
+  kable_styling(bootstrap_options = "striped", 
+                full_width = FALSE, html_font = "Cambria",
+                font_size = 15, position = "center", fixed_thead = T) %>%
+  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>% 
+  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
+  column_spec(2, width = "20em", border_right = T) %>%
+  column_spec(3, width = "20em", border_right = T) %>%
+  footnote(general = "Here is a general comments of the table. ") %>%
+  scroll_box(width = "75%", height = "500px")
+```  
+:::
+
+### Categorical Outcomes
+
+::: blue
+**This code creates a descriptives table for categorical outcomes**
+
+```{r catvar-table}
+#| class.source: 'fold-hide'
+#| messages: FALSE
+#| warning: FALSE
+#| results: "asis"
+
+## Create a descriptives table of study variables by measurement occasion 
+tab_descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
+numeric.test="kwt", cat.test="chisq",
+numeric.stats=c("N", "meansd", "median", "range"
+), # "Nmiss2"
+cat.stats=c("countpct"), # "Nmiss2"
+stats.labels=list(N='Count', meansd="Mean (SD)", median
+='Median', range='Min - Max'
+)) # , Nmiss2 ='Missing'
+
+my_cat_labels  <- list(
+  event = "Year",
+  sex = "Sex",
+  mature_vg = "Mature Video Games"
+  )
+
+tab_descriptives_2 <- tableby(event ~ sex + mature_vg, data=df_long, control=tab_descriptives_2)
+
+# Push table object through kable and kable_styling
+tab_descriptives_2 %>%
+            summary(text=TRUE, digits.pct=1, digits=1) %>%
+            kable(caption = "Categorical Outcomes") %>%
+            kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size =
+                            15, position = "center", fixed_thead = T) %>%
+  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
+  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
+  column_spec(2, width = "20em", border_right = T) %>%
+  #column_spec(3, width = "20em", border_right = T) %>%
+  footnote(general = "Here is a general comments of the table. ") %>%
+  scroll_box(width = "75%", height = "500px")
+
+```
+:::
+:::
+
+## Results
+
+<div class = "blue">
+
+::: panel-tabset
+### Build Model {.tabset .tabset-fade .tabset-pills}
+
+::: blue
+
+The code snippet below tells R to compute xxxxx by xxxxxxx.
+Next, xxxxx is conducted to examine whether xxxxxx does xxxxxxx (indicating xxxxx).
+The xxxx function (xxx) is provided by the r 'xxx' package.
+
+**STEP 1: Compute xxxxxxxx**
+```{r build-model-1}
+#| class.source: 'fold-hide'
+#| message: FALSE
+#| warning: FALSE
+#| echo: TRUE
+
+#A multiple regression approach can be used to examine effects of a continuous covariate  when controlling for autoregressive effects (t1->t2)
+
+#Multiple regression model
+result = lm(cbcl_extern_Year_1 ~ cbcl_extern_Baseline + vg_total_Baseline, data = df_wide)
+str(df_wide)
+# fit ML model including mean structure to make comparable with FIML fit below
+# (means are always included with FIML model fits)
+# sem.mod.fit <- sem(sem.mod, data=tpb.df, meanstructure=TRUE)
+
+# fit again including missing data also
+#sem.mod.fit.fiml <- sem(sem.mod, data=tpb.df, missing="ML")
+
+```
+
+```{r model-checking}
+#Diagnostic Plots
+
+summary(result)
+
+# residuals vs. fitted values plot.
+resid_plot<- plot(result, add.smooth = FALSE, which = 1)
+
+# normal probability plot
+norm_plot<- plot(result, which = 2)
+
+# scale location plot
+scale_plot<- plot(result, add.smooth = FALSE, which = 3)
+
+# boxplot
+#boxplot <- ggplot(df_wide, aes(x = Externalizing_T1, y = Externalizing_T0, colour = Sex)) + geom_point()
+
+#suppressWarnings(print(resid_plot))
+#suppressWarnings(print(norm_plot))
+#suppressWarnings(print(scale_plot))
+#suppressWarnings(print(boxplot))
+
+```
+
+This summary of xxxx indicates xxxxx.
+
+The output from our model provides: i. xxxx; ii. xxxxx; iii. xxxx. In our example xxxxx indicates xxxxx. An examination of xxxxx is xxxx, indicating xxxxxx. Then we conclude that xxxxxx.
+
+```{r}
+# Boxplots
+
+df_wide %>%
+            group_by(sex) %>%  
+            summarise(mean_grade = mean(cbcl_extern_Baseline),
+            sd_grade = sd(cbcl_extern_Baseline),
+            mean_exam = mean(cbcl_extern_Year_1),
+            sd_exam = sd(cbcl_extern_Year_1))
+
+# Boxplot of CBCL Externalizing scores at baseline by sex
+boxplot_cbclExt_T1 <- boxplot(cbcl_extern_Baseline ~ sex,data = df_wide,
+main = "CBCL Ext Baseline Score by sex",
+xlab = "Sex",ylab = "Baseline CBCL Externalizing Score",
+col = "red",border = "black")
+
+# Boxplot of CBCL Externalizing scores at 1-Year follow-up by sex
+boxplot_cbclExt_T2 <- boxplot(cbcl_extern_Year_1 ~ sex, data = df_wide,
+main = "CBCL Ext 1-Year Follow-up Score by sex",
+xlab = "Sex",ylab = "1-Year Follow-up CBCL Externalizing Score",
+col = "red",border = "black")
+
+suppressWarnings(print(boxplot_cbclExt_T1))
+suppressWarnings(print(boxplot_cbclExt_T2))
+```
+
+These boxplots indicates xxxxxx.
+
+## Wrapping Up
+<div class = "blue">
+
+::: panel-tabset
+### Write-up {.tabset .tabset-fade .tabset-pills}
+
+::: blue
+
+A xxxxxx analysis was conducted to examine xxxxx. Findings showed xxxxxx
+
+:::
+:::

### 2023-04-23 3_Tutorials_LinearMixedModels.qmd
tutorials update 1
hide 3b how-to for mixed models temporarily

@@
 author: "Biostatistics Working Group"
+execute:
+  warning: false
 ---
@@
 
-[+add primary diagram/figure]
+![Linear Mixed Models](./figures/Linear Mixed Models.png)
 
@@
 You should use LMMs in the following scenario:
+
 1.	You want to know xxxxx

### 2023-04-23 5_Tutorials_MarginalModels.qmd
tutorials update 1
hide 3b how-to for mixed models temporarily

@@
 author: "Biostatistics Working Group"
+format:
+  html:
+    mermaid:
+      theme: forest
 ---
+
+xxxxxxxxx
+
+```{mermaid}
+graph LR
+A(( )) -- Intercept --> B(( ))
+B -- Slope --> C(( ))
+C -- Quadratic --> D(( ))
+
+B -- Y1 --> E[square]
+C -- Y2 --> F[square]
+D -- Y3 --> G[square]
+```
+
+xxxxxxxx
 
@@
 You should use xxxxxxx in the following scenario:
+
 1.	You want to know xxxxx

### 2023-04-23 ListofReferences.qmd
hide 3b how-to for mixed models temporarily

---
title: "List of Reference Material"
---

GLM @sec-glm
Non-independence @Non-independence

## GLM (Generalized Linear Model): 
A flexible generalization of ordinary linear regression that allows for response variables that have error distribution models other than a normal distribution.{#sec-glm}

## Non-independence: 
The situation where observations are not statistically independent, often arising in longitudinal or clustered data.

3.	Sample average of differences: The mean of the differences between paired observations, typically used in paired-sample designs to assess the average change between two time points or conditions.

4.	Fixed effects: Effects in a statistical model that are constant across all observations, representing the average relationship between predictors and the outcome variable.

5.	Random effects: Effects in a statistical model that vary across different levels of a grouping factor (e.g., subjects, clusters), capturing the variability within each level.

6.	Hierarchical data structures: Data structures where observations are nested within higher-order units (e.g., students within schools, employees within companies), resulting in dependency among observations.

7.	Within-subject correlations: Correlations between repeated measurements on the same subject or unit, often present in longitudinal or clustered data.

8.	Model fit indices: Quantitative measures used to assess how well a statistical model fits the data, such as R-squared, AIC, or BIC.

9.	Nonparametric: Statistical methods that do not rely on assumptions about the underlying distribution of the data.

10.	Resampling permutation method: A nonparametric method for hypothesis testing that involves randomly rearranging the observed data to create new samples and assess the probability of the observed test statistic under the null hypothesis.

11.	Clustered data: Data where observations are grouped or clustered within higher-order units, resulting in dependency among observations within each cluster.

12.	Population-averaged effects: Effects in a statistical model that represent the average relationship between predictors and the outcome variable across the entire population, rather than for specific levels of a grouping factor.

13.	Identity link: A link function in GLMs that relates the linear predictor to the expected value of the outcome variable without any transformation (E[Y] = Xβ).

14.	Logit link: A link function in GLMs used for binary outcomes, relating the linear predictor to the log-odds of the outcome variable (log(E[Y]/(1-E[Y])) = Xβ).

15.	Log link: A link function in GLMs used for count outcomes, relating the linear predictor to the natural logarithm of the expected value of the outcome variable (log(E[Y]) = Xβ).

16.	Logistic link: Synonymous with the logit link.

17.	Binomial distribution: A probability distribution that describes the number of successes in a fixed number of Bernoulli trials with the same probability of success.

18.	Exchangeable correlation structure: A working correlation structure in GEE models, assuming constant correlation between any pair of repeated measurements within subjects.

19.	Autoregressive correlation structure: A working correlation structure in GEE models, assuming correlation between repeated measurements decreases as the time lag between measurements increases.

20.	Unstructured correlation structure: A working correlation structure in GEE models, allowing for different correlations between each pair of repeated measurements within subjects.

21.	Sandwich estimator: A method for computing robust standard errors in GEE models, providing valid inference even if the working correlation structure is misspecified.

22.	Quasi-likelihood under the Independence model Criterion (QIC): A model selection criterion for GEE models, analogous to AIC, used to compare models with different working correlation structures.

23.	Odds ratios: A measure of association for binary outcomes, representing the ratio of the odds of an event occurring in one group to the odds in another group.

24.	Rate ratios: A measure of association for count outcomes, representing the ratio of the rates of an event occurring in one group to the rates in another group.

25.	Maximum likelihood estimation: A statistical method for estimating model parameters by finding the values that maximize the likelihood of the observed data given the model.

26.	Unbalanced data: Data where the number of observations or the number of measurements per subject or group is unequal across subjects or groups.

27.	Unequally spaced data: Data where the time intervals between repeated measurements or the spacing of observations are not equal.

28.	Maximum likelihood (ML): A method of estimation that involves finding the parameter values that maximize the likelihood of the observed data given a statistical model.

29.	Restricted maximum likelihood (REML): A modification of maximum likelihood estimation that is less biased for estimating variance components in mixed-effects models, particularly when the sample size is small.

30.	Penalized quasi-likelihood (PQL): An approximate method for fitting generalized linear mixed models, particularly for non-normal outcomes, which involves a penalized likelihood function and iteratively weighted least squares.

31.	Akaike Information Criterion (AIC): A model selection criterion that balances goodness-of-fit and model complexity, calculated as AIC = -2 * log-likelihood + 2 * k, where k is the number of parameters. Lower AIC values indicate better model fit.

32.	Bayesian Information Criterion (BIC): A model selection criterion that balances goodness-of-fit and model complexity, calculated as BIC = -2 * log-likelihood + k * log(n), where k is the number of parameters and n is the sample size. Lower BIC values indicate better model fit.

### 2023-04-23 _10_Autoregressive Crosslagged Panel Models_HowTo.qmd
hide 3b how-to for mixed models temporarily

# ARCL Models

**text below is notes only**
---
###  EXAMPLE 5.3: Autoregressive Cross-lag (ARCL) model, with Continuous Indicators
***
The ARCL model does xxxxxxx.

    EXAMPLE 5.3 CODE 
    --- 
    TITLE: this is an example of an Autoregressive Cross-lag (ARCL) model. This model extends on the CLPM presented in example 5.3 by including autoregressive (t1 -> t2) effects in the model, with continuous indicators
    DATA: FILE IS example-data.dat; 
    VARIABLE: NAMES ARE x1-x2 y1-y2; 
    MODEL: y2 on x1;
           x2 on y1;
           x1 with y1;
           x2 with y2;
           y2 on y1;
           x2 on x1;

![EXAMPLE 5.3 Diagram. this is an example of a cross-lag panel model including contemporaneous effects, with Continuous Indicators.](/Users/shawes/Desktop/outdoor house.png)

<br>

#### **EXAMPLE 5.3 CODE COMMENTARY**
    TITLE: this is an example of an Autoregressive Cross-lag (ARCL) model. This model extends on the CLPM presented in example 5.3 by including autoregressive (t1 -> t2) effects in the model, with continuous indicators
>   *The TITLE command is used to provide a title for the analysis. The title is printed in the output just before the Summary of Analysis.*
<br>

    DATA: FILE IS example-data.dat;
>   *The DATA command is used to provide information about the data set to be analyzed. The FILE option is used to specify the name of the file that contains the data to be analyzed, ex5.1.dat. Because the data set is in free format, the default, a FORMAT statement is not required.*
<br>

    VARIABLE: NAMES ARE x1-x2 y1-y2;
>   *The VARIABLE command is used to provide information about the variables in the data set to be analyzed. The NAMES option is used to assign names to the variables in the data set. The data set in this example contains six variables: y1, y2, y3, y4, y5, y6. Note that the hyphen can be used as a convenience feature in order to generate a list of names.*
<br>

    MODEL: y2 on x1;
           x2 on y1;
           x1 with y1;
           x2 with y2;
           y2 on y1;
           x2 on x1;
>   *The MODEL command is used to describe the model to be estimated. Here the two BY statements specify that f1 is measured by y1, y2, and y3, and f2 is measured by y4, y5, and y6. The metric of the factors is set automatically by the program by fixing the first factor loading in each BY statement to 1. This option can be overridden. The intercepts and residual variances of the factor indicators are estimated and the residuals are not correlated as the default. The variances of the factors are estimated as the default. The factors are correlated as the default because they are independent (exogenous) variables. The default estimator for this type of analysis is maximum likelihood. The ESTIMATOR option of the ANALYSIS command can be used to select a different estimator.*

***

The difference between this example and Example 5.1 is that the factor indicators are binary or ordered categorical (ordinal) variables instead of continuous variables.  The CATEGORICAL option is used to specify which dependent variables are treated as binary or ordered categorical (ordinal) variables in the model and its estimation.  In the example above, all six factor indicators are binary or ordered categorical variables.  The program determines the number of categories for each factor indicator.  The default estimator for this type of analysis is a robust weighted least squares estimator (Muthén, 1984; Muthén, du Toit, & Spisic, 1997).  With this estimator, probit regressions for the factor indicators regressed on the factors are estimated.  The ESTIMATOR option of the ANALYSIS command can be used to select a different estimator.  An explanation of the other commands can be found in Example 5.1.

With maximum likelihood estimation, logistic regressions for the factor indicators regressed on the factors are estimated using a numerical integration algorithm.  This is shown in Example 5.5.  Note that numerical integration becomes increasingly more computationally demanding as the number of factors and the sample size increase.

## **APPENDIX CHAPTER 5 EXAMPLES: AUTOREGRESSIVE AND CROSSLAGED MODELS**

Using the xxxxx package, we evaluated the performance of the ARCL applied to longitudinal ABCD Study dataset to provide a practical example (all code available at [githublink]). For this example, we will consider the temporal association between youth report scores on the xxx (a measure of xxxxxxx) and yyy (a measure of xxxxxxx). First, we constructed a basic CLPM to simultaneously estimate the lagged association between xxxxxx (xt1) and subsequent yyyyy (yt2), as well as the lagged association between yyyyy (yt1) and subsequent xxxxx (xt2) (see single-headed arrow in Figure x, section xx). Results from this model reveal xxxxxxx (see Table x). However, the basic CLPM does not tell us anything about xxxx and assumes xxxxx.

### **CHAPTER 5 TOC**
* 5.1: Basic Crosslagged Panel Model
* 5.2: Crosslagged Panel Model w/contemporaneous effects
* 5.3: Autoregressive Crosslagged Model

### EXAMPLE 5.1: Basic Cross-lag Panel Model (CLPM) with Continuous Indicators
***
The basic CLPM model does xxxxxxx.

    EXAMPLE 5.1 CODE 
    --- 
    TITLE: this is an example of a basic cross-lag panel model (only cross-lag parameters are modeled) with continuous indicators
    DATA: FILE IS example-data.dat; 
    VARIABLE: NAMES ARE x1-x2 y1-y2; 
    MODEL: y2 on x1;
           x2 on y1;

![EXAMPLE 5.1 Diagram. This is an example of a basic cross-lag panel model (only cross-lag parameters are modeled) with continuous indicators.](/Users/shawes/Desktop/outdoor house.png)

<br>

#### **EXAMPLE 5.1 CODE COMMENTARY**
    TITLE: this is an example of a CFA with continuous factor indicators
>   *The TITLE command is used to provide a title for the analysis. The title is printed in the output just before the Summary of Analysis.*
<br>

    DATA: FILE IS example-data.dat;
>   *The DATA command is used to provide information about the data set to be analyzed. The FILE option is used to specify the name of the file that contains the data to be analyzed, ex5.1.dat. Because the data set is in free format, the default, a FORMAT statement is not required.*
<br>

    VARIABLE: NAMES ARE x1-x2 y1-y2; 
>   *The VARIABLE command is used to provide information about the variables in the data set to be analyzed. The NAMES option is used to assign names to the variables in the data set. The data set in this example contains four variables: x1, x2, y1, y2.

<br>

    MODEL: y2 on x1;
           x2 on y1;

>   *The MODEL command is used to describe the model to be estimated. Here the two BY statements specify that f1 is measured by y1, y2, and y3, and f2 is measured by y4, y5, and y6. The metric of the factors is set automatically by the program by fixing the first factor loading in each BY statement to 1. This option can be overridden. The intercepts and residual variances of the factor indicators are estimated and the residuals are not correlated as the default. The variances of the factors are estimated as the default. The factors are correlated as the default because they are independent (exogenous) variables. The default estimator for this type of analysis is maximum likelihood. The ESTIMATOR option of the ANALYSIS command can be used to select a different estimator.*

<br>

### EXAMPLE 5.2: Cross-lag Panel Model (CLPM), including contemporaneous effects, with Continuous Indicators
***
The CLPM model with included contemporaneous effects does xxxxxxx.

    EXAMPLE 5.2 CODE 
    --- 
    TITLE: this is an example of a cross-lag panel model including contemporaneous effects, with Continuous Indicators
    DATA: FILE IS example-data.dat; 
    VARIABLE: NAMES ARE x1-x2 y1-y2; 
    MODEL: y2 on x1;
           x2 on y1;
           x1 with y1;
           x2 with y2;

![EXAMPLE 5.2 Diagram. this is an example of a cross-lag panel model including contemporaneous effects, with Continuous Indicators.](/Users/shawes/Desktop/outdoor house.png)

<br>

#### **EXAMPLE 5.2 CODE COMMENTARY**
    TITLE: this is an example of a cross-lag panel model including contemporaneous effects, with Continuous Indicators
>   *The TITLE command is used to provide a title for the analysis. The title is printed in the output just before the Summary of Analysis.*
<br>

    DATA: FILE IS example-data.dat;
>   *The DATA command is used to provide information about the data set to be analyzed. The FILE option is used to specify the name of the file that contains the data to be analyzed, ex5.1.dat. Because the data set is in free format, the default, a FORMAT statement is not required.*
<br>

    VARIABLE: NAMES ARE x1-x2 y1-y2;
>   *The VARIABLE command is used to provide information about the variables in the data set to be analyzed. The NAMES option is used to assign names to the variables in the data set. The data set in this example contains six variables: y1, y2, y3, y4, y5, y6. Note that the hyphen can be used as a convenience feature in order to generate a list of names.*
<br>

    MODEL: y2 on x1;
           x2 on y1;
           x1 with y1;
           x2 with y2;
>   *The MODEL command is used to describe the model to be estimated. Here the two BY statements specify that f1 is measured by y1, y2, and y3, and f2 is measured by y4, y5, and y6. The metric of the factors is set automatically by the program by fixing the first factor loading in each BY statement to 1. This option can be overridden. The intercepts and residual variances of the factor indicators are estimated and the residuals are not correlated as the default. The variances of the factors are estimated as the default. The factors are correlated as the default because they are independent (exogenous) variables. The default estimator for this type of analysis is maximum likelihood. The ESTIMATOR option of the ANALYSIS command can be used to select a different estimator.*

### 2023-04-23 _3b_HowTos_Linear Mixed Models.qmd
hide 3b how-to for mixed models temporarily

---
title: "Linear Mixed Models: Random Intercept and Slope"
author: "Biostatistics Working Group"
---

# LMM: Intercept & Slope {.unnumbered}

## Overview

<p>

In this example, we will use the **`LMM:ri`** to analyze trajectories of scores on the externalizing subscale of the child behavior checklist (CBCL) obtained across three measurement occasions in a sample of youth taking part in the ABCD Study. Our primary aim is to characterize stability and change in CBCL externalizing scores across assessments, while accounting for observations that are clustered within youth over time. To do so, we will use the **`LMM:ri`** to simultaneously model an overall sample mean trajectory (fixed effect) and subject-specific (random) effects that vary randomly about the sample mean trajectory. 
</p>

## Preliminary Setup
::: {.panel-tabset}
### Install Packages {.tabset .tabset-fade .tabset-pills}
<div class = "blue">

> **This code installs the r packages necessary for this example, if they are not already installed**

```{r}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: FALSE

#rm(list = ls())

# Install necessary packages (if not already installed)
if (!("lme4" %in% installed.packages())) install.packages("lme4")
if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("afex" %in% installed.packages())) install.packages("afex")
if (!("sdamr" %in% installed.packages())) install.packages("sdamr")
```
</div>

### Load Packages
<div class = "blue">

> **This code loads the r libraries necessary for this example**

```{r pckg-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| class.source: fold-hide

#rm(list = ls())

#Load packages
#library(lme4)
#library(lmerTest)
#library(tidyverse)
#library(afex)
#library(janitor)
#library(skimr)
#library(sdamr)

library(lme4)
library(lmerTest)
library(tidyverse)
library(arrow)
library(afex)
library(janitor)
library(skimr)
library(sdamr)
library(formatR)
library(report)
library(easystats)
library(emmeans)
library(poorman)
library(parameters)
library(modelbased)
library(DT)
library(data.table)
library(arsenal)
library(kableExtra)
library(equatiomatic)
library(gtsummary)
```

</div>

### Config Options
<div class = "blue">

**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                        comment=NA, cache=T, code_folding=T,
                        R.options=list(width=220), fig.align='center',
                        out.width='75%', fig.asp=.75)
```
</div>
:::

## Descriptives Overview
::: {.panel-tabset}
### Read and View Data {.tabset .tabset-fade .tabset-pills}
<div class = "blue">

**This code reads in and shows the data to be used in the current example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Read data
df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
#df_wide <- read_csv("/path/to/datafile")

# set types (if necessary)
#df_long$x <- as.factor(df_long$x)
#df_long$y <- as.numeric(df_long$y)
```

</div>

### Descriptives
<div class = "blue">
```{r descriptives}
#| messages: FALSE
#| echo: TRUE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
descriptives_1  <- tableby.control(test=FALSE, total=FALSE,
                               numeric.test="kwt", cat.test="chisq",
                               numeric.stats=c("N", "meansd", "median", "range" 
                                               ), #"Nmiss2"
                               cat.stats=c("countpct"), #"Nmiss2"
                               stats.labels=list(N='Count', meansd="Mean (SD)", median
                                                 ='Median', range='Min - Max'
                                                 )) #, Nmiss2 ='Missing'

my_cont_labels <- list(
  age = "Age",
  vg_total = "Weekly # of Video Gaming Hrs",
  cbcl_extern = "CBCL Externalizing Scale"
)

tab_descriptives_1 <- tableby(event ~ age + vg_total + 
                            cbcl_extern,
                            data=df_long, control=descriptives_1)

#summary(tab_descriptives_1, labelTranslations = my_cont_labels , text=TRUE, title = #"Continuous Outcomes", term.name = TRUE)

# Push table object through kable and kable_styling
tab_descriptives_1 %>%
  summary(text=TRUE, digits.pct=1, digits=1) %>%
  kable(caption = "Continuous Outcomes") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, html_font = "Cambria",
                font_size = 15,
                position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")
```
</div>

<div class = "blue">
```{r catvar-table}
#| messages: FALSE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
                               numeric.test="kwt", cat.test="chisq",
                               numeric.stats=c("N", "meansd", "median", "range"
                                               ), # "Nmiss2"
                               cat.stats=c("countpct"), # "Nmiss2"
                               stats.labels=list(N='Count', meansd="Mean (SD)", median
                                                 ='Median', range='Min - Max'
                                                 )) # , Nmiss2 ='Missing'

my_cat_labels  <- list(
  event = "Year",
  sex = "Sex",
  mature_vg = "Mature Video Games"
  )

tab_descriptives_2 <- tableby(event ~ sex + mature_vg, 
                            data=df_long, control=descriptives_2)

#summary(tab_descriptives_2, labelTranslations = my_cat_labels , text=TRUE, title = #"Categorical Outcomes", term.name = TRUE)

# Push table object through kable and kable_styling
tab_descriptives_2 %>%
  summary(text=TRUE, digits.pct=1, digits=1) %>%
  kable(caption = "Categorical Outcomes") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size = 15,
                position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  #column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")

```
</div>
:::

## Results

<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue

The code snippet below tells R to compute xxxxx by xxxxxxx. Next, xxxxx is conducted to examine whether xxxxxx does xxxxxxx (indicating xxxxx). The xxxx function (xxx) is provided by the r 'xxx' package.
The code above is very similar to the code for the random intercept model, except that now we use the code (1 + Days \| Subject) to specify a random intercept for each Subject, and a random slope of Days for each Subject. This allows both the intercept and the slope of days to vary across participants.

**STEP 1: Compute xxxxxxxx**
```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

## Linear Mixed Model with a random intercept (LMM-ri)
#really still an intercept only model
random_slopes <- lmer(cbcl_extern ~ 1 + event + (1|ids), data = df_long, REML=F)
print(random_slopes)

```

```{r model-output}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

summary(random_slopes)

## Output and reports extending from the LMM-ri analyses
summary(random_slopes)
confint(random_slopes, level = 0.95, method = "Wald")
report(random_slopes)
report_performance(random_slopes)
report_statistics(random_slopes)
random <- estimate_grouplevel(random_slopes)

```
</div>

The summary function is used to print out the results from the random intercept model and

**STEP 2: Conduct xxxxxxx**
```{r build-model-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

## Obtain LMM-ri model parameters to plot/graph of results
## Fixed-Effects
model_intercept <- rep(coef(random_slopes)$ids[,1], each = 4)
slopes2 <- rep(coef(random_slopes)$ids[,2], each = 4)
#This code allows for extracting the fixed effects estimates for the model intercept and slope.

```

The output from our model provides: i. a parameter estimate; ii. standard error; iii. p-value. In our example p = .001 which is less than .05 indicating significant differences in predicted cbcl externalizing scores between boys and girls in the ABCD Study. An examination of the regression coeefficient for sex (female = 0; male = 1) is b = .xx, indicating the difference in cbcl externalizing scores from Time 1 to Time 2 is significantly greater in boys taking part in the ABCD Study relative to girls. Then we conclude that participant sex (does/does not) predict change in cbcl externalizing scores from t1 to t2.

From the fixed effects section of the model summary, we can conclude that there is strong evidence that number of days of sleep deprivation significantly increased reaction time under a significance level of 0.05. On average, for each additional day of sleep deprivation, reaction time increased by 10.47ms (b = 10.47, SE = 1.55, p \< .001).

We are 95% confident that the average increase was between 7.44 and 13.50.

From the fixed effects section, we estimated the average intercept and average effect of days of sleep deprivation (slopes of Days). We also estimated the corresponding 95% confidence intervals. Using the results from the random effects section, we can calculate the coverage intervals for both the intercept and the slope.

The variance of the intercept and for the slope of Days is 612.1 and 35.1 respectively. Taking their square root, the corresponding standard deviations are 24.7 and 5.9. The 95% coverage interval of the intercept is 251.4 ± 1.96\*24.7. The lower bound is 203.0 and the upper bound is 299.8. It is therefore estimated that 95% of the participants had a reaction time between 203.0 and 299.8 at Day 0.

The 95% coverage interval of the slope of Days is 10.5 ± 1.96\*5.9. The lower bound is -1.1 and the upper bound is 22.1. It is therefore estimated that for 95% of the participants, the effect of each additional day of sleep deprivation was between -1.1 and 22.1. This is not to be confused with the 95% confidence interval (7.4 and 13.5), which means that we are 95% confident that average effect (across all participants) of each additional day of sleep deprivation on reaction time was between 7.4 and 13.5. The model now estimates a variance of the random slopes effects, as well as a correlation between the random intercept and slope effects. :::

```{r}
#| message: FALSE
#| warning: FALSE

ggplot(df_long, aes(x = event, y = cbcl_extern)) + 
  geom_abline(model_slope = slopes2, model_intercept = intercepts2, 
              linetype = "solid", color = "black", linewidth = 1) + 
  # geom_abline(mapping = aes(slopes2 = model_slope, intercepts2 = model_intercept, linetype = PID), 
  #             linetype = "dashed", color = "grey70", linewidth = .4) +
  #geom_point(aes(shape = ids), size = 3.25, color = "grey70") + 
  #scale_shape_manual() + 
  # geom_segment(aes(x = event, xend = event, 
  #                  y = cbcl1, yend = fitted(model_slope)), 
  #              color = "grey70") +
  scale_y_continuous(expand = c(0, 0), breaks = c(0, 100, 200, 300, 400, 500), 
                     limits = c(0, 500)) +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9), 
                     limits = c(-0.5, 10.5)) +
  theme(panel.background = element_blank(),         
          panel.grid.major = element_blank(),
            panel.grid.minor = element_blank(),
            panel.border = element_rect(colour = "black", fill = NA),
          legend.position = "none", 
          axis.text = element_text(size = 14),
          axis.title = element_text(size = 14)) +
    labs (x = "event", y = "cbcl_extern")

#Save the figure
 ggsave("random_slopes.png", units = "in", width = 9, height = 6, dpi = 300)
```

###   Model Plots

:::blue

```{r diag-plots}
#| messages: FALSE
#| warning: FALSE
#| echo: FALSE
#| column: screen-inset-shaded
#| layout-nrow: 2

## xxxxx plot
random_intercepts.std <- resid(model_intercept)/sd(resid(model_intercept))
plot (random_intercepts.std, ylab="Standardized Residuals")

## Outlier Tests. 
#car::outlierTest(random_intercepts)
car::infIndexPlot(random_intercepts)

## qqplot
ggplot(as.data.frame(random_intercepts.std), aes(sample = random_intercepts.std)) +
  geom_qq() +
  geom_qq_line()

## Predicted vs Actual plot
predicted <- estimate_expectation(random_intercepts, data = "grid")
plot(predicted)

## Estimated means plot
means <- estimate_means(random_intercepts)
plot(means)

```

#Build diagnostic plots. Observations with a Bonferroni p \< .05 might
#be considered as outliers and might need further investigation.
Examination of this scatterplot indicates xxxxx.

```{r barplot}
#| messages: FALSE
#| warning: FALSE
#| echo: TRUE

# Basic ggplot2 barplot with mean
p <-df_long %>%
    filter(!is.na(DBD)) %>% # filter on non-missing values  
    ggplot(aes(x = MatureVideoGames, y = Externalizing, fill = Sex)) +
    geom_col(position = "dodge",
           stat = "summary",
           fun = "mean",
           #fill = "#AA4A44",
           #color = "#0099f9"
           ) +
    #scale_fill_brewer(palette = "Set1") +
    scale_fill_manual(values = c("#3db5ff", "#0099f9")) +
    #scale_fill_gradient(high = "#B00B69", low = "#0e0e63") +
    #geom_text(aes(label = cu_traits), position = position_dodge(0.9), 
              #vjust = 2, size = 4, color = "#ffffff") +
    labs(title = "Behavior Problems by Video Game Groups", subtitle = "Simple bar chart",
         caption = "Simple caption", x = "DBD Group", y = "CBCL Externalizing") +
              #coord_flip() +
    #geom_hline(yintercept = mean(df$cbcl_agg), linetype = "dashed", size = 1) +
    theme_minimal() +
    theme(
          plot.title = element_text(color = "#0099f9", size = 20, hjust = 0.5),
          plot.subtitle = element_text(face = "bold", hjust = 1),
          plot.caption = element_text(face = "italic", hjust = 0),
            axis.title.x = element_text(color = "#0099f9", size = 15, face = "bold"),
            axis.title.y = element_text(size = 15, face = "italic")
    )
   p 
    ```

 Examination of this scatterplot indicates xxxxx.

```{r lmmri-graph}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| label: my-figure
#| fig-asp: 0.61

#Random Intercept LMM Plot/Graph
#random_intercepts <- lmer(Externalizing ~ 1 + Event + DBD + Sex + (1|ids), data = df, #REML=T)

##  LMM-ri plot of estimated random intercepts 
model <- lmer(cbcl_extern ~ mature_vg + (1 | sex), data = df_long)
preds <- estimate_relation(model, include_random = TRUE)

## Adding fixed effect trajectory to LMM-ri plot 
fixed_pred <- estimate_relation(model) # This time, include_random is FALSE (default)
 plot(preds, ribbon = list(alpha = 0)) + # Previous plot
  geom_ribbon(data = fixed_pred, aes(x = mature_vg, ymin = CI_low, ymax = CI_high), alpha = 0.4) +
  geom_line(data = fixed_pred, aes(x = mature_vg, y = Predicted), size = 2)

# ggplot(df_long_sub, aes(x = event, y = cbcl_agg, color=sex, shape=sex)) + 
# geom_abline(slope = model_slope, intercept = model_intercept, linetype = "solid", color = "red", linewidth = 1) + 
# geom_point(color = "grey70") + 
# geom_smooth(method=lm, se=FALSE, fullrange=TRUE)+
# scale_shape() + 
# geom_segment(aes(x = event, xend = event, y = cbcl_agg, yend = fitted(cbcl_agg)), color = "grey70") + 
# scale_y_continuous(expand = c(0, 0), breaks = c(0, 20, 40, 60, 80, 100), limits = c(0, 100)) + 
# scale_x_continuous(expand = c(0, 0), breaks = c(1, 2, 3, 4), limits = c(0, 4)) +
# theme(panel.background = element_blank(),         
#         panel.grid.major = element_blank(),
#         panel.grid.minor = element_blank(),
#         panel.border = element_rect(colour = "black", fill = NA),
#         legend.position = "none",
#         axis.text = element_text(size = 14), 
#         axis.title = element_text(size = 14)) +
#         labs (x = "Event", y= "CBCL Aggression")

#Save the plot
#ggsave("random_intercept.png", units = "in", width = 9, height = 6, dpi = 300)

```

<p class="text-center" style="background-color: gray">Figure x: The overall group-mean (fixed effects) trajectory is shown in blue. The faded lines represent each individual youth's estimated trajectory. An examination of this figure shows happiness scores to be increasing across measurement occasions, however, there appears to be substantial variability in the youth's initial happiness scores.</p>

Examination of this barplot indicates xxxxx.

## Wrapping Up
<div class = "blue">

<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue

A random slope model is used to test if sleep deprivation affects reaction time. To account for the repeated measures design, a random intercept was specified for participants. The random slope for days of sleep deprivation was included in the model to allow the effect of sleep deprivation to vary across participants. Results are shown in Table 1. Using a significant level of 0.05, results indicate that sleep deprivation significantly increased reaction time. On average, each additional day of sleep deprivation increased reaction time by 10.47ms (b = 10.47, 95% CI = \[7.44, 13.50\], p \< .001). Model fit comparison between model with and without random slope for sleep deprivation shows that the effect of sleep deprivation varied across participants, χ2(2)= 42.14, p \< .001. The 95% coverage interval for the random slope of sleep deprivation is (-1.14, 22.07), indicating that the effect of sleep deprivation was between -1.14 and 22.07 for 95% of the participants.
:::

## References

[@brown2021] <!-- [@Singmann2019]
[@sarafian]
[#@https://m-clark.github.io/]
[#@Fitzmaurice]
[#@mspeekenbrink.github.io]
[#https://cran.r-project.org/web/packages/lme4/lme4.pdf] -->

### 2023-04-23 index.qmd
hide 3b how-to for mixed models temporarily

@@
 ---
+number-sections: true
 #title: "Temporary Landing Page for Longitudinal Project"

### 2023-04-23 reference.qmd
hide 3b how-to for mixed models temporarily

---
title: "Reference Material: A Brief Overview"
---

[Section in progress...]

### 2023-04-24 explanations.qmd
divio docs section's update

---
title: "Explanations: A Brief Overview"
---

The "Explanation" section focuses on providing context, background information, and a deeper understanding of the concepts, methodologies, or processes related to a specific analyses or topic. Unlike tutorials and how-to guides, which are goal-oriented and problem-oriented respectively, explanations aim to clarify and enhance the user's knowledge. The explanation section aims to create informative, engaging, and accessible content that helps users build a solid foundation of understanding and knowledge about the subject matter.

The explanation section should:

1. Be informative: Explanations should offer valuable insights and information, helping users develop a deeper understanding of the topic. This includes discussing the rationale behind certain choices, exploring alternative approaches, and explaining the advantages and disadvantages of various options.

2. Address "why" questions: Explanations should aim to answer the "why" questions users may have, delving into the reasoning, motivations, or principles that underlie a specific concept or approach.

3. Be clear and concise: While explanations may go into greater depth than tutorials or how-to guides, they should still be presented in a clear and concise manner. This means using plain language, avoiding jargon, and providing examples to help users grasp the concepts more easily.

4. Be well-structured: Explanations should be organized in a logical and coherent manner, making it easy for users to follow the flow of information and understand the relationships between different concepts.

5. Support other documentation types: Explanations should complement and support other types of documentation, such as tutorials and how-to guides, by providing the necessary context and background information. This may involve cross-referencing other resources or linking to related documentation.

6. Engage the audience: Explanations should be engaging and interesting, capturing the user's attention and encouraging them to explore the topic further. This may involve using analogies, storytelling, or real-world examples to illustrate the concepts and make them more relatable.

For additional information/resources, see [https://documentation.divio.com/]

### 2023-04-24 howto.qmd
divio docs section's update

@@
 
-[Section in progress...]
+The "how-to" section focuses on providing practical solutions to specific methods or problems users might encounter.
+Unlike tutorials, which aim to teach a concept, how-to guides concentrate on the efficient execution of specicif analyses without going into in-depth background explanations.
+By following these principles, the how-to section aims to create practical, efficient, and user-friendly guides that help users quickly identify solutions and accomplish their goals.
+
+The how-to section should:
+
+1.
+Be problem-oriented: How-to guides should address specific problems or analyses users may face, offering clear and concise solutions.
+Each guide should focus on a single issue, making it easy for users to find and apply the relevant information.
+
+2.
+Offer step-by-step instructions: Similar to tutorials, how-to guides should provide a clear sequence of steps that users can follow to solve their problem or complete their analysis.
+Each step of the instructions should be explained in a straightforward manner.
+
+3.
+Be concise: How-to guides should be to the point, offering just enough information to help users accomplish their goal.
+This means avoiding unnecessary explanations or background information, as users are typically looking for a quick solution.
+
+4.
+Be actionable: The content in the how-to section should be practical, with a focus on actionable advice that users can apply immediately.
+This may involve providing code snippets, configuration settings, or other specific instructions that users can implement in their projects.
+
+5. Include examples: How-to guides should include relevant examples that demonstrate the solution in action, allowing users to see how the instructions can be applied in a real-world context.
+
+6.
+Cross-reference other resources: How-to guides should link to other relevant documentation, such as explanations or reference material, for users who may require additional background information or a deeper understanding of the topic.
+
+For additional information/resources, see [https://documentation.divio.com/]

### 2023-04-24 reference.qmd
divio docs section's update

@@
 
-[Section in progress...]
+The "Reference" section focuses on providing accurate, concise, and structured information about specific components, functions, or features of a method, analysis, or framework.
+Unlike tutorials, how-to guides, and explanations, which aim to teach, solve problems, and provide context, respectively, reference materials serve as a quick lookup resource for users to find the exact information they need.
+By following these principles, the reference section aims to create a reliable, user-friendly, and easily searchable resource that provides users with the specific information they need to effectively use and understand the method, analysis, or framework.
+
+The reference section should:
+
+1.
+Be comprehensive: Reference materials should cover all relevant aspects of the method, analysis, or framework, including components, functions, features, and configurations.
+The goal is to provide users with a complete and authoritative source of information.
+
+2.
+Be accurate: The information in the reference section should be precise and up-to-date, ensuring that users can rely on it for their projects.
+This may involve regular updates and checks to maintain accuracy.
+
+3.
+Be concise: Reference materials should be brief and to the point, providing just enough information for users to understand and use the specific analysis or method.
+This means avoiding unnecessary explanations, examples, or background information.
+
+4.
+Follow a consistent format: The reference section should use a consistent format and structure, making it easy for users to locate and access the information they need.
+This may involve using templates, standardized headings, or other organizational tools.
+
+5.
+Be easily searchable: Reference materials should be designed with searchability in mind, enabling users to quickly find the information they need.
+This may involve using clear and descriptive titles, headings, and keywords, as well as implementing search functionality or indexing.
+
+6.
+Cross-reference other resources: The reference section should link to other relevant documentation, such as tutorials, how-to guides, or explanations, for users who may require additional information or a deeper understanding of the topic.
+
+For additional information/resources, see [https://documentation.divio.com/]

### 2023-04-24 tutorials.qmd
cli commit 3
divio docs section's update

@@
 
-[Section in progress...]
+The primary goal of the tutorial section is to help users learn and understand a specific topic or method by providing an introductory and hands-on, step-by-step guide.
+This section aims to create engaging, accessible, and effective entry learning experiences that enable users to develop their skills and understanding of the subject matter.
+
+The tutorial section should:
+
+1. Be goal-oriented: Tutorials should focus on a specific task the user wants to accomplish. By concentrating on a particular outcome, the tutorial becomes more engaging and relevant to the user.
+
+2.
+Provide clear steps: Tutorials should offer step-by-step instructions, guiding users through the process of achieving their goal.
+Each step should be clearly explained and easy to follow, making sure users can successfully complete the tutorial.
+
+3.
+Use examples: Tutorials should include real-world examples and use cases that are relatable and meaningful to the user.
+This helps users better understand the concepts being taught and how they can apply them in their own projects.
+
+4.
+Be practical: Tutorials should emphasize hands-on learning, encouraging users to actively engage with the content and practice the skills they are learning.
+This may involve providing code samples, interactive exercises, or other activities that help users develop their understanding.
+
+5.
+Encourage exploration: Tutorials should encourage users to explore and experiment with the concepts they are learning.
+This may involve providing opportunities for users to adapt the tutorial to their own needs or offering suggestions for further experimentation.
+
+For additional information/resources, see [https://documentation.divio.com/]

### 2023-04-25 ListofReferences.qmd
css updates

@@
 
-GLM @sec-glm
-Non-independence @Non-independence
-
-## GLM (Generalized Linear Model): 
+## Generalized Linear Model (GLM): 
 A flexible generalization of ordinary linear regression that allows for response variables that have error distribution models other than a normal distribution.{#sec-glm}
@@
 
-3.	Sample average of differences: The mean of the differences between paired observations, typically used in paired-sample designs to assess the average change between two time points or conditions.
+## Sample average of differences: 
+The mean of the differences between paired observations, typically used in paired-sample designs to assess the average change between two time points or conditions.
 
-4.	Fixed effects: Effects in a statistical model that are constant across all observations, representing the average relationship between predictors and the outcome variable.
+## Fixed effects: 
+Effects in a statistical model that are constant across all observations, representing the average relationship between predictors and the outcome variable.
 
-5.	Random effects: Effects in a statistical model that vary across different levels of a grouping factor (e.g., subjects, clusters), capturing the variability within each level.
+## Random effects: 
+Effects in a statistical model that vary across different levels of a grouping factor (e.g., subjects, clusters), capturing the variability within each level.
 
-6.	Hierarchical data structures: Data structures where observations are nested within higher-order units (e.g., students within schools, employees within companies), resulting in dependency among observations.
+## Hierarchical data structures: 
+Data structures where observations are nested within higher-order units (e.g., students within schools, employees within companies), resulting in dependency among observations.
 
-7.	Within-subject correlations: Correlations between repeated measurements on the same subject or unit, often present in longitudinal or clustered data.
+## Within-subject correlations: 
+Correlations between repeated measurements on the same subject or unit, often present in longitudinal or clustered data.
 
-8.	Model fit indices: Quantitative measures used to assess how well a statistical model fits the data, such as R-squared, AIC, or BIC.
+## Model fit indices: 
+Quantitative measures used to assess how well a statistical model fits the data, such as R-squared, AIC, or BIC.
 
-9.	Nonparametric: Statistical methods that do not rely on assumptions about the underlying distribution of the data.
+## Nonparametric: 
+Statistical methods that do not rely on assumptions about the underlying distribution of the data.
 
-10.	Resampling permutation method: A nonparametric method for hypothesis testing that involves randomly rearranging the observed data to create new samples and assess the probability of the observed test statistic under the null hypothesis.
+## Resampling permutation method: 
+A nonparametric method for hypothesis testing that involves randomly rearranging the observed data to create new samples and assess the probability of the observed test statistic under the null hypothesis.
 
-11.	Clustered data: Data where observations are grouped or clustered within higher-order units, resulting in dependency among observations within each cluster.
+## Clustered data: 
+Data where observations are grouped or clustered within higher-order units, resulting in dependency among observations within each cluster.
 
-12.	Population-averaged effects: Effects in a statistical model that represent the average relationship between predictors and the outcome variable across the entire population, rather than for specific levels of a grouping factor.
+## Population-averaged effects: 
+Effects in a statistical model that represent the average relationship between predictors and the outcome variable across the entire population, rather than for specific levels of a grouping factor.
 
-13.	Identity link: A link function in GLMs that relates the linear predictor to the expected value of the outcome variable without any transformation (E[Y] = Xβ).
+## Identity link: 
+A link function in GLMs that relates the linear predictor to the expected value of the outcome variable without any transformation (E[Y] = Xβ).
 
-14.	Logit link: A link function in GLMs used for binary outcomes, relating the linear predictor to the log-odds of the outcome variable (log(E[Y]/(1-E[Y])) = Xβ).
+## Logit link: 
+A link function in GLMs used for binary outcomes, relating the linear predictor to the log-odds of the outcome variable (log(E[Y]/(1-E[Y])) = Xβ).
 
-15.	Log link: A link function in GLMs used for count outcomes, relating the linear predictor to the natural logarithm of the expected value of the outcome variable (log(E[Y]) = Xβ).
+## Log link: 
+A link function in GLMs used for count outcomes, relating the linear predictor to the natural logarithm of the expected value of the outcome variable (log(E[Y]) = Xβ).
 
-16.	Logistic link: Synonymous with the logit link.
+## Logistic link: 
+Synonymous with the logit link.
 
-17.	Binomial distribution: A probability distribution that describes the number of successes in a fixed number of Bernoulli trials with the same probability of success.
+## Binomial distribution: 
+A probability distribution that describes the number of successes in a fixed number of Bernoulli trials with the same probability of success.
 
-18.	Exchangeable correlation structure: A working correlation structure in GEE models, assuming constant correlation between any pair of repeated measurements within subjects.
+## Exchangeable correlation structure: 
+A working correlation structure in GEE models, assuming constant correlation between any pair of repeated measurements within subjects.
 
-19.	Autoregressive correlation structure: A working correlation structure in GEE models, assuming correlation between repeated measurements decreases as the time lag between measurements increases.
+## Autoregressive correlation structure: 
+A working correlation structure in GEE models, assuming correlation between repeated measurements decreases as the time lag between measurements increases.
 
-20.	Unstructured correlation structure: A working correlation structure in GEE models, allowing for different correlations between each pair of repeated measurements within subjects.
+## Unstructured correlation structure: 
+A working correlation structure in GEE models, allowing for different correlations between each pair of repeated measurements within subjects.
 
-21.	Sandwich estimator: A method for computing robust standard errors in GEE models, providing valid inference even if the working correlation structure is misspecified.
+## Sandwich estimator: 
+A method for computing robust standard errors in GEE models, providing valid inference even if the working correlation structure is misspecified.
 
-22.	Quasi-likelihood under the Independence model Criterion (QIC): A model selection criterion for GEE models, analogous to AIC, used to compare models with different working correlation structures.
+## Quasi-likelihood under the Independence model Criterion (QIC): 
+A model selection criterion for GEE models, analogous to AIC, used to compare models with different working correlation structures.
 
-23.	Odds ratios: A measure of association for binary outcomes, representing the ratio of the odds of an event occurring in one group to the odds in another group.
+## Odds ratios: 
+A measure of association for binary outcomes, representing the ratio of the odds of an event occurring in one group to the odds in another group.
 
-24.	Rate ratios: A measure of association for count outcomes, representing the ratio of the rates of an event occurring in one group to the rates in another group.
+## Rate ratios: 
+A measure of association for count outcomes, representing the ratio of the rates of an event occurring in one group to the rates in another group.
 
-25.	Maximum likelihood estimation: A statistical method for estimating model parameters by finding the values that maximize the likelihood of the observed data given the model.
+## Maximum likelihood estimation: 
+A statistical method for estimating model parameters by finding the values that maximize the likelihood of the observed data given the model.
 
-26.	Unbalanced data: Data where the number of observations or the number of measurements per subject or group is unequal across subjects or groups.
+## Unbalanced data: 
+Data where the number of observations or the number of measurements per subject or group is unequal across subjects or groups.
 
-27.	Unequally spaced data: Data where the time intervals between repeated measurements or the spacing of observations are not equal.
+## Unequally spaced data: 
+Data where the time intervals between repeated measurements or the spacing of observations are not equal.
 
-28.	Maximum likelihood (ML): A method of estimation that involves finding the parameter values that maximize the likelihood of the observed data given a statistical model.
+## Maximum likelihood (ML): 
+A method of estimation that involves finding the parameter values that maximize the likelihood of the observed data given a statistical model.
 
-29.	Restricted maximum likelihood (REML): A modification of maximum likelihood estimation that is less biased for estimating variance components in mixed-effects models, particularly when the sample size is small.
+## Restricted maximum likelihood (REML): 
+A modification of maximum likelihood estimation that is less biased for estimating variance components in mixed-effects models, particularly when the sample size is small.
 
-30.	Penalized quasi-likelihood (PQL): An approximate method for fitting generalized linear mixed models, particularly for non-normal outcomes, which involves a penalized likelihood function and iteratively weighted least squares.
+## Penalized quasi-likelihood (PQL): 
+An approximate method for fitting generalized linear mixed models, particularly for non-normal outcomes, which involves a penalized likelihood function and iteratively weighted least squares.
 
-31.	Akaike Information Criterion (AIC): A model selection criterion that balances goodness-of-fit and model complexity, calculated as AIC = -2 * log-likelihood + 2 * k, where k is the number of parameters.
+## Akaike Information Criterion (AIC): 
+A model selection criterion that balances goodness-of-fit and model complexity, calculated as AIC = -2 * log-likelihood + 2 * k, where k is the number of parameters.
 Lower AIC values indicate better model fit.
 
-32.	Bayesian Information Criterion (BIC): A model selection criterion that balances goodness-of-fit and model complexity, calculated as BIC = -2 * log-likelihood + k * log(n), where k is the number of parameters and n is the sample size.
+## Bayesian Information Criterion (BIC): 
+A model selection criterion that balances goodness-of-fit and model complexity, calculated as BIC = -2 * log-likelihood + k * log(n), where k is the number of parameters and n is the sample size.
 Lower BIC values indicate better model fit.

### 2023-04-25 explanations.qmd
css updates

@@
 title: "Explanations: A Brief Overview"
+number-sections: true
+format:
+ html:
+   css: styles.css
+   theme: cosmo
+   fontsize: 1.1em
+   linestretch: 1.5
 ---
 
-The "Explanation" section focuses on providing context, background information, and a deeper understanding of the concepts, methodologies, or processes related to a specific analyses or topic.
+## What are Explanations?
+
+::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.o; font-family: Georgia, serif;"}
+[**The primary goal**]{style="color: goldenrod; font-size: 1.2em; font-family: Verdana, sans-serif;"} 
+of the explanation section is to provide context, background information, and a deeper understanding of the concepts, methodologies, or processes related to a specific analysis or topic.
 Unlike tutorials and how-to guides, which are goal-oriented and problem-oriented respectively, explanations aim to clarify and enhance the user's knowledge.
 The explanation section aims to create informative, engaging, and accessible content that helps users build a solid foundation of understanding and knowledge about the subject matter.
+:::
+
+::: {style="text-align: justify; line-height: normal; color: black; border: solid black; background-color: antiquewhite; padding: 70px; font-family: Georgia, serif;"}
 
@@
 1.
-Be informative: Explanations should offer valuable insights and information, helping users develop a deeper understanding of the topic.
+**Be informative**: Explanations should offer valuable insights and information, helping users develop a deeper understanding of the topic.
 This includes discussing the rationale behind certain choices, exploring alternative approaches, and explaining the advantages and disadvantages of various options.
 
-2. Address "why" questions: Explanations should aim to answer the "why" questions users may have, delving into the reasoning, motivations, or principles that underlie a specific concept or approach.
+2.
+**Address "why" questions**: Explanations should aim to answer the "why" questions users may have, delving into the reasoning, motivations, or principles that underlie a specific concept or approach.
 
 3.
-Be clear and concise: While explanations may go into greater depth than tutorials or how-to guides, they should still be presented in a clear and concise manner.
+**Be clear and concise**: While explanations may go into greater depth than tutorials or how-to guides, they should still be presented in a clear and concise manner.
 This means using plain language, avoiding jargon, and providing examples to help users grasp the concepts more easily.
@@
 4.
-Be well-structured: Explanations should be organized in a logical and coherent manner, making it easy for users to follow the flow of information and understand the relationships between different concepts.
+**Be well-structured**: Explanations should be organized in a logical and coherent manner, making it easy for users to follow the flow of information and understand the relationships between different concepts.
 
 5.
-Support other documentation types: Explanations should complement and support other types of documentation, such as tutorials and how-to guides, by providing the necessary context and background information.
+**Support other documentation types**: Explanations should complement and support other types of documentation, such as tutorials and how-to guides, by providing the necessary context and background information.
 This may involve cross-referencing other resources or linking to related documentation.
@@
 6.
-Engage the audience: Explanations should be engaging and interesting, capturing the user's attention and encouraging them to explore the topic further.
+**Engage the audience**: Explanations should be engaging and interesting, capturing the user's attention and encouraging them to explore the topic further.
 This may involve using analogies, storytelling, or real-world examples to illustrate the concepts and make them more relatable.
 
-For additional information/resources, see [https://documentation.divio.com/]
+::: {style="text-align: center"}
+For additional information, see [https://documentation.divio.com/]
+:::
+
+:::

### 2023-04-25 howto.qmd
css updates

--- 2023-04-20 howto.qmd
@@
 title: "How-To's: A Brief Overview"
+number-sections: true
+format:
+ html:
+   css: styles.css
+   theme: cosmo
+   fontsize: 1.1em
+   linestretch: 1.5
 ---
 
-[Section in progress...]
+## What are How-To's?
+::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.o; font-family: Georgia, serif;"}
+[**The primary goal**]{style="color: goldenrod; font-size: 1.2em; font-family: Verdana, sans-serif;"} 
+of the "how-to" section is on providing practical solutions to specific methods or analytic problems users might encounter.
+Unlike tutorials, which aim to teach a concept, how-to guides concentrate on the efficient execution of specific analyses without going into in-depth background explanations.
+By following these principles, the how-to section aims to create practical, efficient, and user-friendly guides that help users quickly identify solutions and accomplish their goals.
+:::
+
+::: {style="text-align: justify; line-height: normal; color: black; border: solid black; background-color: antiquewhite; padding: 70px; font-family: Georgia, serif;"}
+
+The how-to section should:
+
+1.
+**Be problem-oriented**: How-to guides should address specific problems or analyses users may face, offering clear and concise solutions.
+Each guide should focus on a single issue, making it easy for users to find and apply the relevant information.
+
+2.
+**Offer step-by-step instructions**: Similar to tutorials, how-to guides should provide a clear sequence of steps that users can follow to solve their problem or complete their analysis.
+Each step of the instructions should be explained in a straightforward manner.
+
+3.
+**Be concise**: How-to guides should be to the point, offering just enough information to help users accomplish their goal.
+This means avoiding unnecessary explanations or background information, as users are typically looking for a quick solution.
+
+4.
+**Be actionable**: The content in the how-to section should be practical, with a focus on actionable advice that users can apply immediately.
+This may involve providing code snippets, configuration settings, or other specific instructions that users can implement in their projects.
+
+5. **Include examples**: How-to guides should include relevant examples that demonstrate the solution in action, allowing users to see how the instructions can be applied in a real-world context.
+
+6.
+**Cross-reference other resources**: How-to guides should link to other relevant documentation, such as explanations or reference material, for users who may require additional background information or a deeper understanding of the topic.
+
+::: {style="text-align: center"}
+For additional information, see [https://documentation.divio.com/]
+:::
+
+:::

### 2023-04-25 index.qmd
css updates

--- 2023-04-23 index.qmd
@@
 ---
-#title: "Temporary Landing Page for Longitudinal Project"
-#format:
-#  html:
-#    css: styles.css
-#    theme: cosmo
-#    fontsize: 1.1em
-    #linestretch: 1.7
+number-sections: true
+#title: "Longitudinal Analysis"
+format:
+  html:
+    css: styles.css
+    theme: cosmo
+    fontsize: 1.3em
+    linestretch: 1.7
 ---
 
-<div class="text-3xl text-center">
+::: {style="text-align: justify; text-align: center; line-height: normal; color: white; border: solid black; background-color: darkgoldenrod; padding: 70px; font-family: Georgia, serif;"}
+
 This is the temporary landing page for the Longitudinal Analysis Project. <br> The site is currently under construction and used for testing purposes only. <br> Contents may be inaccurate.
-</div>
+
+:::
 

### 2023-04-25 reference.qmd
css updates

--- 2023-04-23 reference.qmd
@@
 ---
-title: "Reference Material: A Brief Overview"
+title: "References: A Brief Overview"
+number-sections: true
+format:
+ html:
+   css: styles.css
+   theme: cosmo
+   fontsize: 1.1em
+   linestretch: 1.5
 ---
 
-[Section in progress...]
+## What are References?
+
+::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.o; font-family: Georgia, serif;"}
+[**The primary goal**]{style="color: goldenrod; font-size: 1.2em; font-family: Verdana, sans-serif;"} 
+of the "Reference" section is on providing accurate, concise, and structured information about specific components, functions, or features of a method, analysis, or framework.
+Unlike tutorials, how-to guides, and explanations, which aim to teach, solve problems, and provide context, respectively, reference materials serve as a quick lookup resource for users to find the exact information they need.
+By following these principles, the reference section aims to create a reliable, user-friendly, and easily searchable resource that provides users with the specific information they need to effectively use and understand the method, analysis, or framework.
+:::
+
+::: {style="text-align: justify; line-height: normal; color: black; border: solid black; background-color: antiquewhite; padding: 70px; font-family: Georgia, serif;"}
+
+The reference section should:
+
+1.
+**Be comprehensive**: Reference materials should cover all relevant aspects of the method, analysis, or framework, including components, functions, features, and configurations.
+The goal is to provide users with a complete and authoritative source of information.
+
+2.
+**Be accurate**: The information in the reference section should be precise and up-to-date, ensuring that users can rely on it for their projects.
+This may involve regular updates and checks to maintain accuracy.
+
+3.
+**Be concise**: Reference materials should be brief and to the point, providing just enough information for users to understand and use the specific analysis or method.
+This means avoiding unnecessary explanations, examples, or background information.
+
+4.
+**Follow a consistent format**: The reference section should use a consistent format and structure, making it easy for users to locate and access the information they need.
+This may involve using templates, standardized headings, or other organizational tools.
+
+5.
+**Be easily searchable**: Reference materials should be designed with searchability in mind, enabling users to quickly find the information they need.
+This may involve using clear and descriptive titles, headings, and keywords, as well as implementing search functionality or indexing.
+
+6.
+**Cross-reference other resources**: The reference section should link to other relevant documentation, such as tutorials, how-to guides, or explanations, for users who may require additional information or a deeper understanding of the topic.
+
+::: {style="text-align: center"}
+For additional information, see [https://documentation.divio.com/]
+:::
+
+:::

### 2023-04-25 temp.qmd
css updates

---
title: "Tutorials: A Brief Overview"
number-sections: true
format:
 html:
   css: styles.css
   theme: cosmo
   fontsize: 1.1em
   linestretch: 1.7
---

This is the TUTORIAL main page example.

## Main Section

::: {style="color: red; border: solid black; background-color: grey;"}
This is the [middle section]{style="background-color: yellow; color: blue;"} 
:::

::: {.grid}

::: {.g-col-2}
:::

::: {.g-col-8}

::: {style="color: red; border: solid black; background-color: green;"}
Testing [Sample Test!]{style="background-color: yellow; color: blue;"}
:::

:::

::: {.g-col-2}
:::

:::

The primary goal of the tutorial section is to help users learn and understand a specific topic or method by providing an introductory and hands-on, step-by-step guide. This section aims to create engaging, accessible, and effective entry learning experiences that enable users to develop their skills and understanding of the subject matter.

<br>

::: {.grid}

::: {.g-col-2}
:::

::: {.g-col-8}
![](./figures/lights.png)
:::

::: {.g-col-2}
:::

:::

### 2023-04-25 tutorials.qmd
css updates

--- 2023-04-20 tutorials.qmd
@@
 title: "Tutorials: A Brief Overview"
+number-sections: true
+format:
+ html:
+   css: styles.css
+   theme: cosmo
+   fontsize: 1.1em
+   linestretch: 1.5
 ---
 
-[Section in progress...]
+## What are Tutorials?
+
+::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.o; font-family: Georgia, serif;"}
+[**The primary goal**]{style="color: goldenrod; font-size: 1.2em; font-family: Verdana, sans-serif;"} 
+of the tutorial section is to help users learn and understand a specific topic or method by providing an introductory and hands-on, step-by-step guide.
+This section aims to create engaging, accessible, and effective entry learning experiences that enable users to develop their skills and understanding of the subject matter.
+:::
+
+::: {style="text-align: justify; line-height: normal; color: black; border: solid black; background-color: antiquewhite; padding: 70px; font-family: Georgia, serif;"}
+
+The tutorial section should:
+
+1. **Be goal-oriented**: Tutorials should focus on a specific task the user wants to accomplish. By concentrating on a particular outcome, the tutorial becomes more engaging and relevant to the user.
+
+2.
+**Provide clear steps**: Tutorials should offer step-by-step instructions, guiding users through the process of achieving their goal.
+Each step should be clearly explained and easy to follow, making sure users can successfully complete the tutorial.
+
+3.
+**Use examples**: Tutorials should include real-world examples and use cases that are relatable and meaningful to the user.
+This helps users better understand the concepts being taught and how they can apply them in their own projects.
+
+4.
+**Be practical**: Tutorials should emphasize hands-on learning, encouraging users to actively engage with the content and practice the skills they are learning.
+This may involve providing code samples, interactive exercises, or other activities that help users develop their understanding.
+
+5.
+**Encourage exploration**: Tutorials should encourage users to explore and experiment with the concepts they are learning.
+This may involve providing opportunities for users to adapt the tutorial to their own needs or offering suggestions for further experimentation.
+
+::: {style="text-align: center"}
+For additional information, see [https://documentation.divio.com/]
+:::
+
+:::

### 2023-04-25 1a_HowTos_DifferenceScores_PairedTtests.qmd
yaml section-nesting updates for how-tos

---
title: "Difference Scores: Paired Samples T-test"
author: "Biostatistics Working Group"
---

# Paired Samples T-test {.unnumbered}

## Overview

<p>

This example examines the question whether participants reported
different mean levels of externalizing problems on the CBCL when
measured at Baseline (T0) compared to when measured at the 1-Year
follow-up (T1). This analysis is conducted in two primary steps: 1)
computing a difference score (DiffScore = CBCL EXT T2 - CBCL EXT T1); 2)
conducting a paired samples t-test on the difference score. Our primary
aim is to determine whether average CBCL externalizing scores of the
participants change from assessement 1 (T1) to assessement 2 (T2), while
accounting for observations that are clustered within youth over time.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

#rm(list = ls())

if (!("lme4" %in% installed.packages())) install.packages("lme4")
if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("arrow" %in% installed.packages())) install.packages("arrow")
if (!("arsenal" %in% installed.packages())) install.packages("arsenal")
if (!("kableExtra" %in% installed.packages())) install.packages("kableExtra")
if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
if (!("rstatix" %in% installed.packages())) install.packages("rstatix")

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| class.source: fold-hide

#rm(list = ls())

library(lme4)
library(lmerTest)
library(tidyverse)
library(arrow)
library(arsenal)
library(kableExtra)
library(rstatix)
library(ggpubr)
```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                                  comment=NA, cache=T, code_folding=T,
                                  R.options=list(width=220), fig.align='center', 
                                  out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Read data
df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
df_wide <- read_csv("/Users/shawes/Desktop/data/df_wide.csv")

# set types
df_long$ids <- as.factor(df_long$ids)
df_long$event <- as.factor(df_long$event)
df_long$site_id <- as.factor(df_long$site_id)
df_long$fam_id <- as.factor(df_long$fam_id)
df_long$sex <- as.factor(df_long$sex)
df_long$age <- as.numeric(df_long$age)
df_long$mature_vg <- as.numeric(df_long$mature_vg)
df_long$cbcl_extern <- as.numeric(df_long$cbcl_extern)
df_long$vg_total <- as.numeric(df_long$vg_total)
```
:::

### Continuous Outcomes

::: blue
**This code creates a table of descriptive information for continuous
outcomes**

```{r contvar-table}
#| messages: FALSE
#| echo: TRUE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_1 <- tableby.control(test=FALSE, total=FALSE, numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"), #"Nmiss2"
cat.stats=c("countpct"), #"Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median ='Median', 
                  range='Min - Max'))  #, Nmiss2 ='Missing'
my_cont_labels <- list(age = "Age", vg_total = "Weekly # of Video Gaming Hrs",
            cbcl_extern = "CBCL Externalizing Scale"
)

tab_descriptives_1 <- tableby(event ~ age + vg_total + 
                                      cbcl_extern,
                                      data=df_long, control=tab_descriptives_1)

# Push table object through kable and kable_styling
tab_descriptives_1 %>%
  summary(text=TRUE, digits.pct=1, digits=1) %>%
  kable(caption = "Continuous Outcomes") %>%
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, html_font = "Cambria",
                font_size = 15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>% 
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")
```
:::

### Categorical Outcomes

::: blue
**This code creates a descriptives table for categorical outcomes**

```{r catvar-table}
#| class.source: 'fold-hide'
#| messages: FALSE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"
), # "Nmiss2"
cat.stats=c("countpct"), # "Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median
='Median', range='Min - Max'
)) # , Nmiss2 ='Missing'

my_cat_labels  <- list(
  event = "Year",
  sex = "Sex",
  mature_vg = "Mature Video Games"
  )

tab_descriptives_2 <- tableby(event ~ sex + mature_vg, data=df_long, control=tab_descriptives_2)

# Push table object through kable and kable_styling
tab_descriptives_2 %>%
            summary(text=TRUE, digits.pct=1, digits=1) %>%
            kable(caption = "Categorical Outcomes") %>%
            kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size =
                            15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  #column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")

```
:::
:::

## Results

<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

The code snippet below tells R to compute a difference score by
subtracting each participant's Externalizing score at T2 from their
Externalizing score at T1. Next, a one-sample t-test is computed to
examine whether the average difference score is different than zero
(indicating change).

**STEP 1: Compute Difference Score**

```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Compute difference score based on CBCL Externalizing subscale scores at baseline (t1) and 1-Year Follow-up (t2)
df_wide$diffscore <- df_wide$cbcl_extern_Year_1 - df_wide$cbcl_extern_Baseline
```

```{r model-checking}
# Compute statistical summaries for the difference score variable
summary(df_wide$diffscore)

```

This summary of the difference score variable indicates xxxxx.

```{r}
hist(df_long$vg_total)

## Summary statistics
summary<-df_long %>%
group_by(event) %>%
get_summary_stats(cbcl_extern, type = "mean_sd")
data.frame(summary)
```

This histogram indicates xxxxxx. The summary command shows xxxxxx.

```{r shapiro-wilks}
### Shapiro-Wilk test and normality (Q-Q) plot (visualization of correlation between a given sample and the normal distribution)
shapiro.test(df_wide$cbcl_extern_Baseline[0:5000])
```

This Shapiro-Wilkes test shows xxxxxx.

```{r qqplot}
qqplot <- ggqqplot(df_wide$cbcl_extern_Baseline,
    ylab = "Externalizing Difference Score", xlab = FALSE,
    ggtheme = theme_minimal()
)

suppressWarnings(print(qqplot)) 
```

This qqplot shows xxxxxx.

**STEP 2: Conduct t-test on Difference Score**

```{r build-model-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# One-sample t-test to determine whether the average difference score is significantly different than 0.
result <- t.test(df_wide$cbcl_extern_Baseline, mu = 0, alternative = "two.sided")
# Print the results
result
```
:::

The output from our model provides: i. a t-statistic ii. degrees of
freedom under iii. p-value (\< .05 then we reject the null hypothesis)
iv. a mean estimate v. 95% confidence intervals

In our example p = .002 which is less than .05 then we reject the null
hypothesis and conclude that the mean difference score µ is different
from zero. So we can conclude that there𝐷 is a difference between cbcl
externalizing scores from t1 to t2. This output also provides the mean
for cbcl externalizing scores (4.45),as well as the correpsonding
confidence intervals

### Model Plots

::: blue
```{r scatterplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

# Scatterplot to visualize relationship between t1 & t2 data used to create difference score
scatterplot <- ggplot(df_wide, aes(x = cbcl_extern_Baseline, y = cbcl_extern_Year_1)) +
    geom_point(size = 3) +
    geom_smooth(method = lm, se = F) +
    xlab("CBCL Externalizing (x) Baseline") +
    ylab("CBCL Externalizing (x) 1-Year Follow-up")

suppressWarnings(print(scatterplot))
```

Examination of this scatterplot indicates xxxxx.

```{r boxplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

### boxplot
diffscore_boxplot <- ggboxplot(df_wide$cbcl_extern_Baseline,
    ylab = "Externalizing Difference Score", xlab = FALSE,
    ggtheme = theme_minimal()
)

suppressWarnings(print(diffscore_boxplot))
```

Examination of this boxplot indicates xxxxx.
:::

> This output shows xxxxxxxx. Briefly walk through each metric
:::

## Wrapping Up

<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue
A paired samples t test was conducted to compare differences in
participants cbcl externalzing scores measured across two timepoints.
Findings indicated that there was a difference between cbcl externalzing
scores obtained at T1 compared to cbcl externalzing scores obtained at
T2, t(179) = 3.10, p = .002, Cohen's d = 0.23. Youth exhibited an
increase of cbcl externalzing scores at t2 ( M = 5.67, SD = 1.24)
compared to t1 ( M = 5.83, SD = 1.21).
:::
:::

### 2023-04-25 3b_HowTos_LinearMixedModels.qmd
yaml section-nesting updates for how-tos

--- 2023-04-23 _3b_HowTos_Linear Mixed Models.qmd
@@
 
-ggplot(df_long, aes(x = event, y = cbcl_extern)) + 
-  geom_abline(model_slope = slopes2, model_intercept = intercepts2, 
-              linetype = "solid", color = "black", linewidth = 1) + 
-  # geom_abline(mapping = aes(slopes2 = model_slope, intercepts2 = model_intercept, linetype = PID), 
-  #             linetype = "dashed", color = "grey70", linewidth = .4) +
-  #geom_point(aes(shape = ids), size = 3.25, color = "grey70") + 
-  #scale_shape_manual() + 
-  # geom_segment(aes(x = event, xend = event, 
-  #                  y = cbcl1, yend = fitted(model_slope)), 
-  #              color = "grey70") +
-  scale_y_continuous(expand = c(0, 0), breaks = c(0, 100, 200, 300, 400, 500), 
-                     limits = c(0, 500)) +
-  scale_x_continuous(expand = c(0, 0), breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9), 
-                     limits = c(-0.5, 10.5)) +
-  theme(panel.background = element_blank(),         
-          panel.grid.major = element_blank(),
-            panel.grid.minor = element_blank(),
-            panel.border = element_rect(colour = "black", fill = NA),
-          legend.position = "none", 
-          axis.text = element_text(size = 14),
-          axis.title = element_text(size = 14)) +
-    labs (x = "event", y = "cbcl_extern")
-
-#Save the figure
- ggsave("random_slopes.png", units = "in", width = 9, height = 6, dpi = 300)
+# # ! THIS PLOT RUNS & PRINTS TO SCREEN. I am temporarily commenting it out due to a side-effect of the 
+# #"intercepts2 variable"
+# ggplot(df_long, aes(x = event, y = cbcl_extern)) + 
+#   geom_abline(model_slope = slopes2, model_intercept = intercepts2, 
+#               linetype = "solid", color = "black", linewidth = 1) + 
+#   # geom_abline(mapping = aes(slopes2 = model_slope, intercepts2 = model_intercept, linetype = PID), 
+#   #             linetype = "dashed", color = "grey70", linewidth = .4) +
+#   #geom_point(aes(shape = ids), size = 3.25, color = "grey70") + 
+#   #scale_shape_manual() + 
+#   # geom_segment(aes(x = event, xend = event, 
+#   #                  y = cbcl1, yend = fitted(model_slope)), 
+#   #              color = "grey70") +
+#   scale_y_continuous(expand = c(0, 0), breaks = c(0, 100, 200, 300, 400, 500), 
+#                      limits = c(0, 500)) +
+#   scale_x_continuous(expand = c(0, 0), breaks = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9), 
+#                      limits = c(-0.5, 10.5)) +
+#   theme(panel.background = element_blank(),         
+#           panel.grid.major = element_blank(),
+#             panel.grid.minor = element_blank(),
+#             panel.border = element_rect(colour = "black", fill = NA),
+#           legend.position = "none", 
+#           axis.text = element_text(size = 14),
+#           axis.title = element_text(size = 14)) +
+#     labs (x = "event", y = "cbcl_extern") 
+#   
+# #Save the figure
+#  ggsave("random_slopes.png", units = "in", width = 9, height = 6, dpi = 300)
 ```
@@
 ```{r diag-plots}
-#| messages: FALSE
-#| warning: FALSE
-#| echo: FALSE
-#| column: screen-inset-shaded
-#| layout-nrow: 2
-
-## xxxxx plot
-random_intercepts.std <- resid(model_intercept)/sd(resid(model_intercept))
-plot (random_intercepts.std, ylab="Standardized Residuals")
-
-## Outlier Tests. 
-#car::outlierTest(random_intercepts)
-car::infIndexPlot(random_intercepts)
-
-## qqplot
-ggplot(as.data.frame(random_intercepts.std), aes(sample = random_intercepts.std)) +
-  geom_qq() +
-  geom_qq_line()
-
-## Predicted vs Actual plot
-predicted <- estimate_expectation(random_intercepts, data = "grid")
-plot(predicted)
-
-## Estimated means plot
-means <- estimate_means(random_intercepts)
-plot(means)
+# #| messages: FALSE
+# #| warning: FALSE
+# #| echo: FALSE
+# #| column: screen-inset-shaded
+# #| layout-nrow: 2
+#     
+# ## xxxxx plot
+# random_intercepts.std <- resid(model_intercept)/sd(resid(model_intercept))
+# plot (random_intercepts.std, ylab="Standardized Residuals")
+# 
+# ## Outlier Tests. 
+# #car::outlierTest(random_intercepts)
+# car::infIndexPlot(random_intercepts)
+# 
+# ## qqplot
+# ggplot(as.data.frame(random_intercepts.std), aes(sample = random_intercepts.std)) +
+#   geom_qq() +
+#   geom_qq_line()
+# 
+# ## Predicted vs Actual plot
+# predicted <- estimate_expectation(random_intercepts, data = "grid")
+# plot(predicted)
+# 
+# ## Estimated means plot
+# means <- estimate_means(random_intercepts)
+# plot(means)
 
@@
 p <-df_long %>%
-    filter(!is.na(DBD)) %>% # filter on non-missing values  
-    ggplot(aes(x = MatureVideoGames, y = Externalizing, fill = Sex)) +
+    filter(!is.na(event)) %>% # filter on non-missing values  
+    ggplot(aes(x = mature_vg, y = cbcl_extern, fill = sex)) +
     geom_col(position = "dodge",
@@
     )
-   p 
 ```

### 2023-04-25 explanations.qmd
yaml section-nesting updates for how-tos

@@
 title: "Explanations: A Brief Overview"
-number-sections: true
+number-sections: false
 format:
@@
 
-## What are Explanations?
+## ## <i class="bi bi-book"></i> What are Explanations?
 

### 2023-04-25 howto.qmd
yaml section-nesting updates for how-tos

@@
 title: "How-To's: A Brief Overview"
-number-sections: true
+number-sections: false
 format:
@@
 
-## What are How-To's?
+## ## <i class="bi bi-book"></i>  What are How-To's?
+
 ::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.o; font-family: Georgia, serif;"}

### 2023-04-25 manuscript.qmd
yaml section-nesting updates for how-tos

--- 2023-04-20 manuscript.qmd
@@
 ---
+title: "Longitudinal Analysis Manuscript: Working Draft"
+subtitle: Longitudinal Analysis using data from the ABCD Study 
+authors:
+  - name: Samuel W. Hawes
+    department: Center for Children and Families
+    affiliation: Florida International University
+    location: Miami, FL
+    email: shawes@fiu.edu
+  - name: Andrew K. Littlefield
+    department: Psychological Sciences
+    affiliation: Texas Tech University
+    location: Lubbock, TX
+    email: andrew.littlefield@ttu.edu
+  - name: Daniel A. Lopez
+    department: Department of Public Health
+    affiliation: University of Rochester Medical Center
+    location: Rochester, NY
+    email: Daniel_Lopez2@URMC.Rochester.edu
+  - name: Kenneth J. Sher
+    department: Psychological Sciences
+    affiliation: University of Missouri
+    location: Columbia, MO
+    email: SherK@missouri.edu
+  - name: Additional Co-authors
+    department: xxxxxx
+    affiliation: University of xxxx
+    location: xxxxxxx
+    email: xxxxxx
+  - name: Wesley K. Thompson
+    department: Center for Population Neuroscience and Genetics
+    affiliation: Laureate Institute for Brain Research
+    location: Tulsa, OK
+    email: wes.stat@gmail.com
 
-title: Manuscript Working Draft
+abstract: |
+ The Adolescent Brain Cognitive Development (ABCD) Study presents a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of children and adolescents.
+Given the complex nature of the longitudinal data collected in the ABCD Study, researcher are likely to encounter a myriad of methodological and analytic considerations and concerns.
+This review provides a comprehensive examination of key issues and techniques related to longitudinal data analysis, specifically focusing on the ABCD Study.
+The text discusses model assumptions, common violations (e.g., independent and identically distributed residuals, *heterogeneous* *variability*) and their implications for valid inference.
+The importance of appropriately modeling covariance structures, understandings trade-offs between model fit and parsimony, and challenges related to sample size, attrition, missing data are highlighted.
+Consideration is given to the importance of selecting appropriate statistical models to account for correlations in repeated measurements and the assumptions underlying these models.
+The review also differentiates between linear and non-linear models in the context of continuous and discrete data, emphasizing various distributional assumptions and the necessity of choosing appropriate models and statistical methods.
+By addressing these complexities, the review seeks to equip researchers with the necessary knowledge and tools to make informed decisions as they navigate effectively analyzing and interpreting data available in the ABCD Study.
+
+author-summary: |
+  The Adolescent Brain Cognitive Development (ABCD) Study offers valuable insights into adolescent brain development.
+Researchers working with this data face methodological and analytic concerns, including sample size, attrition, missing data, and measurement.
+This review addresses complexities in analyzing longitudinal data, with a focus on statistical models accounting for correlations in repeated measurements and their assumptions.
+We emphasize the importance of choosing appropriate models and analytic functions based on the data and the specific scientific question to be addressed.
+This concise resource aims  to help researchers navigate longitudinal data analysis, make informed methodological decisions, and enhance the accuracy and reliability of their findings when working with data from the ABCD Study.
+
+    keywords:
+    *ABCD Study*,
+    *longitudinal analyses*,
+    *development*
 
 format: 
+  html:
+    grid:
+      sidebar-width: 200px
+      margin-width: 300px
   pdf: 
     geometry:
-      - top=20mm
+      - top=30mm
       - left=20mm
-  docx: default
-prefer-html: true
+    #fontfamily: libertinus
+    #fontsize: 20pt
+format-links: [html, pdf]
 
+  #format:
+#  pdf:
+#    geometry:
+#      - top=30mm
+#      - left=20mm
+#   docx: default
+# prefer-html: true
+
+output: 
+  rticles::arxiv_article
+
+bibliography: "references.bib" #references.bib
+biblio-style: unsrt
 ---
@@
 
-# Part II Longitudinal Data: Interpretation / Issues / Pitfalls & Assumption
+# Part II: Longitudinal Data: Interpretation / Issues / Pitfalls & Assumption
 \label{sec:headings}
@@
 
----
+# Part III: Additional Section(s) in Progress: TBD]
+\label{sec:headings}
+**Subsection TBD.**
 
-### References
-
-::: {#refs}
-:::
+# Discussion
+\label{sec:headings}
+**Subsection TBD.**

### 2023-04-25 reference.qmd
yaml section-nesting updates for how-tos

@@
 title: "References: A Brief Overview"
-number-sections: true
+number-sections: false
 format:
@@
 
-## What are References?
+## ## <i class="bi bi-book"></i> What are References?
 

### 2023-04-26 manuscript.qmd
misc. updates

@@
 title: "Longitudinal Analysis Manuscript: Working Draft"
+number-sections: true
+number-depth: 3
 subtitle: Longitudinal Analysis using data from the ABCD Study 
@@
   html:
+    toc: true
+    toc-depth: 1
+    toc-expand: 2
+    toc-title: Table of Contents
     grid:
-      sidebar-width: 200px
+      sidebar-width: 300px
       margin-width: 300px
@@
 
-# Part I: Longitudinal Research: Basic Concepts and Considerations
+# Part I: Longitudinal Research
 \label{sec:headings}
+
+## Basic Concepts and Considerations
 There are several important concepts to consider when conducting longitudinal analyses in a developmental context.
@@
 
-# Part II: Longitudinal Data: Interpretation / Issues / Pitfalls & Assumption
+# Part II: Longitudinal Data
 \label{sec:headings}
+
+## Interpretation / Issues / Pitfalls & Assumption
 **Defining Features of Longitudinal Data Analysis.**
@@
 
-## Modeling Data Across Two Time Points versus Three or More Time Points.
+### Modeling Data Across Two Time Points versus Three or More Time Points.
 Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions.
@@
 
-## Types of stability and change
+### Types of stability and change
 If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it’s the characterization of stability and change as the organism traverses the life course.
@@
 
-## Model Assumptions
+### Model Assumptions
 Many statistical models assume certain characteristics about the data to which they are being applied.
@@
 
-## Covariance Structures
+### Covariance Structures
 Statistical models for longitudinal data include two main components to account for assumptions that are commonly violated when working with repeated measures data: a model for the covariance among repeated measures (both the correlations among pairs of repeated measures on an individual and the variability of the responses on different occasions), coupled with a model for the mean response and its dependence on covariates (eg, treatment group in the context of clinical trials).
@@
 
-## Accounting for Correlated Data
+### Accounting for Correlated Data
 As an example, one alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design.
@@
 
-## Linear vs non-linear models
+### Linear vs non-linear models
 Identification of optimal statistical models and appropriate mathematical functions requires an understanding of the type of data being used.
@@
 
-## Missing Data/Attrition
+### Missing Data/Attrition
 As recently reviewed by Littlefield (in press), investigators of prospective data are confronted with study attrition (i.e., participants may not provide data at a given wave of assessment) and thus approaches are needed to confront the issue of missing data.
@@
 
-## Quantifying effect sizes longitudinally
+### Quantifying effect sizes longitudinally
 Given longitudinal data involve different sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data.
@@
 
-# Part III: Additional Section(s) in Progress: TBD]
+# Part III: Additional Section(s)
 \label{sec:headings}
-**Subsection TBD.**
+## In Progress: [TBD]
 
-# Discussion
+### Subsection TBD.
+
+# Part IV: Discussion
 \label{sec:headings}
-**Subsection TBD.**
+
+### Subsection TBD.
+
+# Part V: References

### 2023-04-26 temp.qmd
yaml section-nesting updates for how-tos
misc. updates

@@
 ## Main Section
+
+```{ojs}
+sunburst = {
+  const root = partition(flareData);
+
+  root.each(d => d.current = d);
+
+  const svg = d3.create("svg")
+      .attr("viewBox", [0, 0, width, width])
+      .style("font", "15px sans-serif");
+
+  const g = svg.append("g")
+      .attr("transform", `translate(${width / 2},${width / 2})`);
+
+  const path = g.append("g")
+    .selectAll("path")
+    .data(root.descendants().slice(1))
+    .join("path")
+      .attr("fill", d => { while (d.depth > 1) d = d.parent; return color(d.data.name); })
+      .attr("fill-opacity", d => arcVisible(d.current) ? (d.children ? 0.6 : 0.4) : 0)
+      .attr("d", d => arc(d.current));
+
+  path.filter(d => d.children)
+      .style("cursor", "pointer")
+      .on("click", clicked);
+
+  path.append("title")
+      .text(d => `${d.ancestors().map(d => d.data.name).reverse().join("/")}\n${format(d.value)}`);
+
+  const label = g.append("g")
+      .attr("pointer-events", "none")
+      .attr("text-anchor", "middle")
+      .style("user-select", "none")
+    .selectAll("text")
+    .data(root.descendants().slice(1))
+    .join("text")
+      .attr("dy", "0.35em")
+      .attr("fill-opacity", d => +labelVisible(d.current))
+      .attr("transform", d => labelTransform(d.current))
+      .text(d => d.data.name);
+
+  const parent = g.append("circle")
+      .datum(root)
+      .attr("r", radius)
+      .attr("fill", "none")
+      .attr("pointer-events", "all")
+      .on("click", clicked);
+
+  function clicked(event, p) {
+    parent.datum(p.parent || root);
+
+    root.each(d => d.target = {
+      x0: Math.max(0, Math.min(1, (d.x0 - p.x0) / (p.x1 - p.x0))) * 2 * Math.PI,
+      x1: Math.max(0, Math.min(1, (d.x1 - p.x0) / (p.x1 - p.x0))) * 2 * Math.PI,
+      y0: Math.max(0, d.y0 - p.depth),
+      y1: Math.max(0, d.y1 - p.depth)
+    });
+
+    const t = g.transition().duration(750);
+
+    // Transition the data on all arcs, even the ones that aren’t visible,
+    // so that if this transition is interrupted, entering arcs will start
+    // the next transition from the desired position.
+    path.transition(t)
+        .tween("data", d => {
+          const i = d3.interpolate(d.current, d.target);
+          return t => d.current = i(t);
+        })
+      .filter(function(d) {
+        return +this.getAttribute("fill-opacity") || arcVisible(d.target);
+      })
+        .attr("fill-opacity", d => arcVisible(d.target) ? (d.children ? 0.6 : 0.4) : 0)
+        .attrTween("d", d => () => arc(d.current));
+
+    label.filter(function(d) {
+        return +this.getAttribute("fill-opacity") || labelVisible(d.target);
+      }).transition(t)
+        .attr("fill-opacity", d => +labelVisible(d.target))
+        .attrTween("transform", d => () => labelTransform(d.current));
+  }
+
+  function arcVisible(d) {
+    return d.y1 <= 3 && d.y0 >= 1 && d.x1 > d.x0;
+  }
+
+  function labelVisible(d) {
+    return d.y1 <= 3 && d.y0 >= 1 && (d.y1 - d.y0) * (d.x1 - d.x0) > 0.03;
+  }
+
+  function labelTransform(d) {
+    const x = (d.x0 + d.x1) / 2 * 180 / Math.PI;
+    const y = (d.y0 + d.y1) / 2 * radius;
+    return `rotate(${x - 90}) translate(${y},0) rotate(${x < 180 ? 0 : 180})`;
+  }
+
+  return svg.node();
+}
+```
 

### 2023-04-26 tutorials.qmd
brief update

@@
 title: "Tutorials: A Brief Overview"
-number-sections: true
+number-sections: false
 format:
@@
    fontsize: 1.1em
-   linestretch: 1.5
 ---
 
-## What are Tutorials?
+xxxxx
+```{ojs}
+sunburst = {
+  const root = partition(flareData);
 
-::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.o; font-family: Georgia, serif;"}
+  root.each(d => d.current = d);
+
+  const svg = d3.create("svg")
+      .attr("viewBox", [0, 0, width, width])
+      .style("font", "15px sans-serif");
+
+  const g = svg.append("g")
+      .attr("transform", `translate(${width / 2},${width / 2})`);
+
+  const path = g.append("g")
+    .selectAll("path")
+    .data(root.descendants().slice(1))
+    .join("path")
+      .attr("fill", d => { while (d.depth > 1) d = d.parent; return color(d.data.name); })
+      .attr("fill-opacity", d => arcVisible(d.current) ? (d.children ? 0.6 : 0.4) : 0)
+      .attr("d", d => arc(d.current));
+
+  path.filter(d => d.children)
+      .style("cursor", "pointer")
+      .on("click", clicked);
+
+  path.append("title")
+      .text(d => `${d.ancestors().map(d => d.data.name).reverse().join("/")}\n${format(d.value)}`);
+
+  const label = g.append("g")
+      .attr("pointer-events", "none")
+      .attr("text-anchor", "middle")
+      .style("user-select", "none")
+    .selectAll("text")
+    .data(root.descendants().slice(1))
+    .join("text")
+      .attr("dy", "0.35em")
+      .attr("fill-opacity", d => +labelVisible(d.current))
+      .attr("transform", d => labelTransform(d.current))
+      .text(d => d.data.name);
+
+  const parent = g.append("circle")
+      .datum(root)
+      .attr("r", radius)
+      .attr("fill", "none")
+      .attr("pointer-events", "all")
+      .on("click", clicked);
+
+  function clicked(event, p) {
+    parent.datum(p.parent || root);
+
+    root.each(d => d.target = {
+      x0: Math.max(0, Math.min(1, (d.x0 - p.x0) / (p.x1 - p.x0))) * 2 * Math.PI,
+      x1: Math.max(0, Math.min(1, (d.x1 - p.x0) / (p.x1 - p.x0))) * 2 * Math.PI,
+      y0: Math.max(0, d.y0 - p.depth),
+      y1: Math.max(0, d.y1 - p.depth)
+    });
+
+    const t = g.transition().duration(750);
+
+    // Transition the data on all arcs, even the ones that aren’t visible,
+    // so that if this transition is interrupted, entering arcs will start
+    // the next transition from the desired position.
+    path.transition(t)
+        .tween("data", d => {
+          const i = d3.interpolate(d.current, d.target);
+          return t => d.current = i(t);
+        })
+      .filter(function(d) {
+        return +this.getAttribute("fill-opacity") || arcVisible(d.target);
+      })
+        .attr("fill-opacity", d => arcVisible(d.target) ? (d.children ? 0.6 : 0.4) : 0)
+        .attrTween("d", d => () => arc(d.current));
+
+    label.filter(function(d) {
+        return +this.getAttribute("fill-opacity") || labelVisible(d.target);
+      }).transition(t)
+        .attr("fill-opacity", d => +labelVisible(d.target))
+        .attrTween("transform", d => () => labelTransform(d.current));
+  }
+
+  function arcVisible(d) {
+    return d.y1 <= 3 && d.y0 >= 1 && d.x1 > d.x0;
+  }
+
+  function labelVisible(d) {
+    return d.y1 <= 3 && d.y0 >= 1 && (d.y1 - d.y0) * (d.x1 - d.x0) > 0.03;
+  }
+
+  function labelTransform(d) {
+    const x = (d.x0 + d.x1) / 2 * 180 / Math.PI;
+    const y = (d.y0 + d.y1) / 2 * radius;
+    return `rotate(${x - 90}) translate(${y},0) rotate(${x < 180 ? 0 : 180})`;
+  }
+
+  return svg.node();
+}
+```
+
+xxxx
+
+## <i class="bi bi-book"></i> What are Tutorials?
+
+::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.0; font-family: Georgia, serif;"}
 [**The primary goal**]{style="color: goldenrod; font-size: 1.2em; font-family: Verdana, sans-serif;"} 

### 2023-04-26 documentation.qmd
update to documentation buttons

---
title: Documentation
author: Biostatistics Working Group
---

::: {.grid}
::: {.g-col-6}
::: {style="background-color: gray;"}
[**Tutorials**](./tutorials.qmd){.btn .btn-primary .btn-outline-warning .btn-lg .btn-block .btn role="link" data-toggle="tooltip" title="Link to the Tutorials documentation"}
:::
:::
::: {.g-col-6}
::: {style="background-color: gray;"}
[How-Tos](./howto.qmd){.btn .btn-primary  .btn-outline-warning .btn-lg .btn-block .btn role="button" data-toggle="tooltip" title="Link to the How-Tos documentation"}
:::
:::
:::

<br>

::: {.grid}
::: {.g-col-6}
::: {style="background-color: gray;"}
[Explanations](./explanations.qmd){.btn .btn-outline-warning .btn-lg .btn role="button"  data-toggle="tooltip" title="Link to the Explanations documentation"}
:::
:::
::: {.g-col-6}
::: {style="background-color: gray;"}
[Reference](./reference.qmd){.btn .btn-outline-warning .btn-lg .btn role="button" data-toggle="tooltip" title="Link to the Reference documentation"}
:::
:::
:::

<br>

### 2023-04-26 tutorials.qmd
update to documentation buttons

--- 2023-04-25 tutorials.qmd
@@
    fontsize: 1.1em
-   linestretch: 1.5
 ---
 
-## What are Tutorials?
+## <i class="bi bi-book"></i> What are Tutorials?
 
-::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.o; font-family: Georgia, serif;"}
+::: {style="text-align: justify; font-size: 1.2em; line-stretch: 1.0; font-family: Georgia, serif;"}
 [**The primary goal**]{style="color: goldenrod; font-size: 1.2em; font-family: Verdana, sans-serif;"} 

### 2023-04-27 index.qmd
landing page updates

--- 2023-04-23 index.qmd
@@
 ---
-#title: "Temporary Landing Page for Longitudinal Project"
-#format:
-#  html:
-#    css: styles.css
-#    theme: cosmo
-#    fontsize: 1.1em
-    #linestretch: 1.7
+#title: "Longitudinal Analysis"
+format:
+  html:
+    page-layout: full
+    grid:
+      sidebar-width: 0px
+      body-width: 2000px
+      margin-width: 0px
+      gutter-width: 1.5rem
 ---
 
-<div class="text-3xl text-center">
-This is the temporary landing page for the Longitudinal Analysis Project. <br> The site is currently under construction and used for testing purposes only. <br> Contents may be inaccurate.
-</div>
-
-<br>
-
-::: {.grid}
-
-::: {.g-col-2}
-:::
-
-::: {.g-col-8}
-![](./figures/lights.png)
-:::
-
-::: {.g-col-2}
-:::
-
-:::
-
-</div>
+![](./figures/Figlanding4.svg)

### 2023-04-27 temp.qmd
landing page updates

--- 2023-04-25 temp.qmd
@@
 title: "Tutorials: A Brief Overview"
-number-sections: true
 format:
   html:
-   css: styles.css
-   theme: cosmo
-   fontsize: 1.1em
-   linestretch: 1.7
+    page-layout: full
+    grid:
+      sidebar-width: 0px
+      body-width: 2000px
+      margin-width: 0px
+      gutter-width: 1.5rem
+#    css: styles.css
+#    theme: cosmo
+#    fontsize: 1.1em
+#    linestretch: 1.7
+
 ---
 
-This is the TUTORIAL main page example.
+![FigmaLandingTest2](./figures/Figlanding3.svg)
 
-## Main Section
+<div class="page-layout-custom">
+  ![FigmaLandingTest2](./figures/Figlanding2.svg)
+</div>
+
+xxxxx
+It is important to match a theory of change with design and statistical model to assess change. (p. 4)
+Models with two time points cannot provide estimates of within-person change.
+“Stability” over time can reflect many things, only some of which mean a lack of change. (p. 14)
+It is helpful to describe not only average growth (e.g. intercept and slope means), but variances in the growth parameters. (p. 25)
+
+Models that focus exclusively on mean-level change sometimes called Marginal Models
+Those that emphasize individual change often do so through inclusion of Random Effects
+often called “growth models”
+
+Longitudinal Psychometrics
+
+xxxxxx
 
@@
 :::
+
+::: {style="text-align: justify; text-align: center; line-height: normal; color: white; border: solid black; background-color: darkgoldenrod; padding: 70px; font-family: Georgia, serif;"}
+
+This is the temporary landing page for the Longitudinal Analysis Project. <br> The site is currently under construction and used for testing purposes only. <br> Contents may be inaccurate.
+
+:::
+
+<br>
+
+::: {.grid}
+
+::: {.g-col-2}
+:::
+
+::: {.g-col-8}
+![](./figures/lights.png)
+:::
+
+::: {.g-col-2}
+:::
+
+:::
+
+</div>

### 2023-04-28 manuscript.qmd
landing page updates
landing page updates 3

@@
 # Introduction
-\label{sec:headings}
-The Adolescent Brain Cognitive Development (ABCD) Study® is the largest long-term investigation of neurodevelopment and child health in the United States.
+
+\label{sec:headings} The Adolescent Brain Cognitive Development (ABCD) Study® is the largest long-term investigation of neurodevelopment and child health in the United States.
 Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018].
 At its heart, the study is designed to chart the course of human development across multiple, interacting domains from late childhood to early adulthood and to identify factors that lead to both positive and negative developmental outcomes.
-Central to achieving these goals is the ABCD Study’s® commitment to an open science framework designed to facilitate access to and sharing of scientific knowledge by espousing practices that increase openness, integrity, and reproducibility of scientific research (e.g., public data releases).
+Central to achieving these goals is the ABCD Study's® commitment to an open science framework designed to facilitate access to and sharing of scientific knowledge by espousing practices that increase openness, integrity, and reproducibility of scientific research (e.g., public data releases).
 In this sense, the ABCD Study® is a collaboration with the larger research community, with the rich longitudinal nature of the ABCD Study dataset allowing researchers to perform a variety of analyses of both methodological and substantive interest.
@@
 
-[section still be developed...]
+\[section still be developed...\]
 
 ## The ABCD Study® Data
+
 Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians.
@@
 Each new wave of data collection provides the building blocks for conducting probing longitudinal analyses that allow us to characterize normative development, identify variables that presage deviations from prototypic development, and assess a range of outcomes associated with variables of interest.
-This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use [add citation], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020].
+This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use \[add citation\], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020].
 Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward.
@@
 Data are publicly released on an annual basis through the NIMH Data Archive.
-By necessity, the study’s earliest data releases were cross-sectional (i.e., the baseline data), however, the most recent public data release (NDA Release 4.0) contains data collected across three annual assessments, including two imaging assessments (baseline and year 2 follow-up visits).
+By necessity, the study's earliest data releases were cross-sectional (i.e., the baseline data), however, the most recent public data release (NDA Release 4.0) contains data collected across three annual assessments, including two imaging assessments (baseline and year 2 follow-up visits).
 
 ## Organization of current manuscript
+
 The rich longitudinal nature of the ABCD Study dataset will allow researchers to perform analyses of both methodological and substantive interest.
@@
 
-[section still be developed...]
+\[section still be developed...\]
 
 # Part I: Longitudinal Research
+
 \label{sec:headings}
@@
 ## Basic Concepts and Considerations
+
 There are several important concepts to consider when conducting longitudinal analyses in a developmental context.
 These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represent the same phenomenon or different ones.
-Further, in the case of developmentally focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child’s life course (and not just characterize growth during this period but also assesses the durability/chronicity of prior effects/consequences)  but also brings with it greater methodological opportunities to exploit the statistical properties of longitudinal data in the furtherance of critical scientific questions.
-That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows, assuming we have assessed the “right” variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest.
+Further, in the case of developmentally focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child's life course (and not just characterize growth during this period but also assesses the durability/chronicity of prior effects/consequences) but also brings with it greater methodological opportunities to exploit the statistical properties of longitudinal data in the furtherance of critical scientific questions.
+That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows, assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest.
 Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.
@@
 However, unique epochs and experiences can alter the course of this idealized form of development.
-Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [add citation]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans)[add citation].
-These examples highlight the importance of considering the role of vulnerable periods – temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli (e.g.
-cannabis exposure) on the developing brain may be particularly pronounced– when trying to establish an accurate understanding of the association between exposures and outcomes.
+Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development \[add citation\]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans)\[add citation\].
+These examples highlight the importance of considering the role of vulnerable periods -- temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli (e.g.
+cannabis exposure) on the developing brain may be particularly pronounced-- when trying to establish an accurate understanding of the association between exposures and outcomes.
 
@@
 Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times environmental pressures will tend to suppress stability and disrupt the orderly stochastic process of normative development (e.g., xxx-xxx).
-This situation reflects a developmental disturbance in that the normal course of development is “disturbed” for some time by some time-limited process.
+This situation reflects a developmental disturbance in that the normal course of development is "disturbed" for some time by some time-limited process.
 In such cases, we might find that prediction of behavior in the period of the disturbance is reduced and/or, similarly, the behavior exhibited during the disturbance might have less predictive power with respect to distal outcomes compared to the behavior exhibited before and following the disrupted period.
@@
 Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next.
-For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect, creating a “snare” that retards psychosocial development.
+For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect, creating a "snare" that retards psychosocial development.
 Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified.
@@
 Such experience effects on cognitive functioning have been demonstrated in adolescent longitudinal samples similar to ABCD [@sullivan2017] and highlight the need to consider these effects and address them analytically.
-In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to “learning” the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before).
+In the case of performance-based measures [e.g., matrix reasoning related to neurocognitive functioning; see @salthouse2014], this can be due to "learning" the task from previous test administrations (e.g., someone taking the test a second time performs better than they did the first time simply as a function of having taken it before).
 Even in the case of non-performance-based measures (e.g., levels of depression), where one cannot easily make the argument that one has acquired some task-specific skill through learning, it has been observed that respondents tend to endorse lower levels on subsequent assessments [e.g., @beck1961; see @french2010] and this phenomenon has been well documented in research on structured diagnostic interviews [@robins1985].
@@
 # Part II: Longitudinal Data
+
 \label{sec:headings}
@@
 ## Interpretation / Issues / Pitfalls & Assumption
-**Defining Features of Longitudinal Data Analysis.**
-The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
+
+**Defining Features of Longitudinal Data Analysis.** The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
 While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed.
-That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the “right” variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .
+That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .
+
+### \[may add section here about the "structure" of longitudinal data we're mostly focused on-- repeated measures of panel data...\]
+
+"traditional approaches": marginal models
+
+random effects: Growth Curves- SEM (lgcm) vs ME (lmm)
+
+(between- and within-person levels?)
 
 ### Modeling Data Across Two Time Points versus Three or More Time Points.
+
 Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions.
@@
 At the same time, data based on two assessments are inherently limited on multiple fronts.
-As [@rogosa1982] noted approximately forty years ago, “Two waves of data are better than one, but maybe not much better”.
+As [@rogosa1982] noted approximately forty years ago, "Two waves of data are better than one, but maybe not much better".
 These sentiments are reflected in more contemporary recommendations regarding best-practice guidelines for prospective data, which increasingly emphasize the benefits of additional measurement occasions for model identification and accurate parameter estimation.
 It is also consistent with research recommending that developmental studies include three or more assessment points, given it is impossible for data based on two-time points to determine the shape of development (since linear, straight line change is the only possible form, given two assessments; see [@duncan2009]).
-Research designs that include three or more time points allow for increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]– a key aspect of inferential research.
+Research designs that include three or more time points allow for increasingly nuanced analyses that more adequately tease apart sources of variation and covariation among the repeated assessments [@king2018]-- a key aspect of inferential research.
 To illustrate, developmental theories are typically interested in understanding patterns of within-individual change over time (discussed in further detail, below); however, two data points provide meager information on change at the person level.
-This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded “… researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data” and perhaps more concerning, “…the models discussed here do not offer a feasible way to overcome these inherent limitations” @littlefield2021.
+This point is further underscored in a recent review of statistical models commonly touted as distinguishing within-individual vs between-individual sources of variance in which the study authors concluded "...
+researchers are limited when attempting to differentiate these sources of variation in psychological phenomenon when using two waves of data" and perhaps more concerning, "...the models discussed here do not offer a feasible way to overcome these inherent limitations" @littlefield2021.
 It is important to note, however, that despite the current focus on two-wave designs versus three or more assessment waves, garnering three assessment points is not a panacea for longitudinal modeling.
@@
 ### Types of stability and change
-If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it’s the characterization of stability and change as the organism traverses the life course.
+
+If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it's the characterization of stability and change as the organism traverses the life course.
 There are a few different ways to think of stability (and change).
 Consider we measure the height of all youth in a 6th-grade class, once in the fall at the beginning of the school year and once again in the spring at the end of the school year.
-A common first step may be to compare the class’s average height values obtained at these two different measurement occasions.
-This comparison of the average scores for the same group of individuals at multiple time points is referred to as “mean-level” stability as it provides information about continuity and change in the group level of an outcome of interest (e.g., height) over time.
-Another type of stability involves calculating the correlation between the values obtained at different time points (e.g., ‘height in the fall’ with ‘height in the spring’).
-This type of “rank-order” stability evaluates between-individual change by focusing on the degree to which individuals retain their relative placement in a group across time.
+A common first step may be to compare the class's average height values obtained at these two different measurement occasions.
+This comparison of the average scores for the same group of individuals at multiple time points is referred to as "mean-level" stability as it provides information about continuity and change in the group level of an outcome of interest (e.g., height) over time.
+Another type of stability involves calculating the correlation between the values obtained at different time points (e.g., 'height in the fall' with 'height in the spring').
+This type of "rank-order" stability evaluates between-individual change by focusing on the degree to which individuals retain their relative placement in a group across time.
 Consider, someone who is the shortest person in their class in 6th grade may grow considerably over the school year (i.e., exhibit mean level change), but remain the shortest person among their classmates.
@@
 For example, in some areas of development, there is considerable mean-level change that occurs over time (e.g., changes in Big 5 personality traits), but relatively high rank-order stability.
-Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of “within-individual” change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest.
+Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of "within-individual" change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest.
 
@@
 Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021].
-Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a].
+Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual's mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a].
 It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].
@@
 ### Model Assumptions
+
 Many statistical models assume certain characteristics about the data to which they are being applied.
 As an example, common assumptions of parametric statistical models include normality, linearity, and equality of variances.
-These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model’s assumptions can substantively alter the interpretation of results.
+These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model's assumptions can substantively alter the interpretation of results.
 Similarly, statistical models employed in the analyses of longitudinal data often entail a range of assumptions that must be closely inspected.
@@
 ### Covariance Structures
+
 Statistical models for longitudinal data include two main components to account for assumptions that are commonly violated when working with repeated measures data: a model for the covariance among repeated measures (both the correlations among pairs of repeated measures on an individual and the variability of the responses on different occasions), coupled with a model for the mean response and its dependence on covariates (eg, treatment group in the context of clinical trials).
@@
 ### Accounting for Correlated Data
+
 As an example, one alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design.
@@
 Diverting from compound symmetry, this model assumes the correlations between repeated assessments decline exponentially across time rather than remaining constant.
-For example, per the AR(1) structure, if the correlation between Time 1 and Time 2 data is thought to be .5, then the correlation between Time 1 and Time 3 data would be assumed to be .5*.5 = .25, and the correlation between Time 1 and Time 4 data would be assumed to be .5*.5*.5 = .125.
+For example, per the AR(1) structure, if the correlation between Time 1 and Time 2 data is thought to be .5, then the correlation between Time 1 and Time 3 data would be assumed to be .5*.5 = .25, and the correlation between Time 1 and Time 4 data would be assumed to be .5*.5\*.5 = .125.
 As with compound symmetry, the basic AR(1) model is parsimonious in that it only requires two parameters (the variance of the assessments and the autoregressive coefficient).
-Notably, the assumption of constant autoregressive relations between assessments is often relaxed in commonly employed designs that use autoregressive modeling (e.g., cross-lagged panel models [CLPM]).
+Notably, the assumption of constant autoregressive relations between assessments is often relaxed in commonly employed designs that use autoregressive modeling (e.g., cross-lagged panel models \[CLPM\]).
 These designs still typically assume an AR(1) process (e.g., it is sufficient to regress the Time 3 assessment onto the Time 2 assessment and is not necessary to also regress the Time 3 assessment onto the Time 1 assessment, which would result in an AR(2) process).
@@
 ### Linear vs non-linear models
+
 Identification of optimal statistical models and appropriate mathematical functions requires an understanding of the type of data being used.
 Repeated assessments can be based on either continuous or discrete measures.
-Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., a single item measuring the level of agreement to a statement on a three-point scale including the categories of “disagree”, “neutral”, and “agree” in an ecological momentary assessment study that involves multiple daily assessments), and count variables (e.g., number of cigarettes smoked per day across a daily diary study).
+Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., a single item measuring the level of agreement to a statement on a three-point scale including the categories of "disagree", "neutral", and "agree" in an ecological momentary assessment study that involves multiple daily assessments), and count variables (e.g., number of cigarettes smoked per day across a daily diary study).
 In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016].
@@
 
+### Marginal vs Conditional Models (?)
+
+\[section still under development...\]
+
 ### Missing Data/Attrition
+
 As recently reviewed by Littlefield (in press), investigators of prospective data are confronted with study attrition (i.e., participants may not provide data at a given wave of assessment) and thus approaches are needed to confront the issue of missing data.
@@
 More specifically, the last observation carried forward (LOCF) is a common approach to imputing missing data.
-LOCF replaces a participant’s missing values after dropout with the last available measurement [@molnar2008].
+LOCF replaces a participant's missing values after dropout with the last available measurement [@molnar2008].
 This approach assumes stability (i.e., a given participant's score is not anticipated to increase or decline after study dropout) and that the data are MCA R.
@@
 More modern approaches, such as using maximum likelihood or multiple imputation to estimate missing data, are thought to avoid some of the biases of older approaches [see @enders2010; @graham2009].
-@graham2009 noted several “myths” regarding missing data.
+@graham2009 noted several "myths" regarding missing data.
 For example, Graham notes many assume the data must be minimally MAR to permit estimating procedures (such as maximum likelihood or multiple imputation) compared to other, more traditional approaches (e.g., using only complete case data).
@@
 ### Quantifying effect sizes longitudinally
+
 Given longitudinal data involve different sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data.
-Effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” [@kraemer2014].
-Common effect size metrics include r (i.e., the standardized covariance, or correlation, between two variables) and Cohen’s d [@cohen1988].
-Adjustments to common effect size calculations, such as Cohen’s d, are required even when only two time points are considered [e.g., see @morris2002].
+Effect size can be defined as, "a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study." [@kraemer2014].
+Common effect size metrics include r (i.e., the standardized covariance, or correlation, between two variables) and Cohen's d [@cohen1988].
+Adjustments to common effect size calculations, such as Cohen's d, are required even when only two time points are considered [e.g., see @morris2002].
 @wang2019a note there are multiple approaches to obtaining standardized within-person effects, and that commonly suggested approaches (e.g., global standardization) can be problematic [see @wang2019a, for more details].
@@
 # Part III: Additional Section(s)
-\label{sec:headings}
-## In Progress: [TBD]
+
+\label{sec:headings} \## In Progress: \[TBD\]
+
+Section focused on different ways of considering/grouping/deciding on which specific analysis to choose (see curran&bauer slides)
 
@@
 # Part IV: Discussion
+
 \label{sec:headings}

### 2023-04-28 temp.qmd
landing page updates 3

@@
 
-![FigmaLandingTest2](./figures/Figlanding3.svg)
+![FigmaLandingTest2](./figures/Figlanding5.svg)
+(remove marginal models from "examples" address (vs random-effects) in the manuscript)
+
+- section: "Traditional Linear Models"
+          contents:
+            - 1_Tutorials_DifferenceScores.qmd
+            - 2_Tutorials_ResidualizedChangeScores.qmd
+        - section: "Traditional Non-linear Models"
+          contents:
+            - 4_Tutorials_SignedRankTest.qmd
+            - 6_Tutorials_GeneralizedEstimatingEquations.qmd (population average effects)
+        - section: "Modern GLM Extensions(?)"
+            - 7_Tutorials_LinearMixedModels.qmd  
+            - 8_Tutorials_GeneralizedLinearMixedEffectsModels.qmd
+        - section: "SEM Approaches"
+          contents:
+            - 8_Tutorials_AutoregressiveCrosslaggedPanelModels.qmd
+            - 9_Tutorials_LatentChangeScoresModels.qmd
+            - 10_Tutorials_LatentGrowthCurveModels.qmd
+            - 11_Tutorials_MultivariateLatentGrowthCurveModels.qmd
+            - 12_Tutorials_LatentTransitionAnalysis.qmd
+            - 13_Tutorials_GrowthMixtureModels.qmd
+            - 14_Tutorials_StateTraitModels.qmd
+        - section: "Advanced SEM"
+          contents:
+            - 15_Tutorials_RandomInterceptCrosslaggedPanelModels.qmd
+            - 16_Tutorials_LatentCurveModelsStructuredResiduals.qmd
+
+GEE, on the other hand, is fitting a marginal model. These model population-averages. You're modeling the expectation conditional only on your fixed design matrix.
+GLMM is fitting a mixed-effects model; Thus the the estimates are subject-specific.
+What differs between GEE and GLMM is the target of inference: population-average or subject-specific.
+
+This is in contrast to mixed effect models as explained above which condition on both the fixed design matrix and the random effects.
+So with the marginal model above you're saying, "forget about the difference among classrooms, I just want the population (school-wise) rate of failure and its association with gender." You fit the model and get an odds ratio that is the population-averaged odds ratio of failure associated with gender.
+
+Molenberghs, Verbeke 2005 has an entire chapter on marginal vs. random effects models.
+
+xxxxxx
+
+2. Models for Longitudinal Data
+Since both the GLMM and WGEE are extensions of the GLM, we start with a brief overview of the latter.
+
+2.1 Generalized Linear Models (GLM)
+
+LMM vs LGCM?
+For modeling growth, researchers typically employ some type of random effects model such that a mean growth trajectory is estimated for all observations in the data but a unique growth curve is estimated for each individual in the data as well (Curran & Bauer, 2011).
+These types of models are generally referred to as subject-specific models (Zeger, Liang, & Albert, 1988), which are more commonly known in psychology as growth curve models.
+Growth models have many aliases but can be broadly grouped into two different classes of methods: the latent-curve (LC) approach, which treats the repeated measures as multivariate (also known as the “wide” data format) and tends to be fit with general structural equation modeling (SEM) software (Meredith & Tisak, 1990; Tucker, 1958; Willett & Sayer, 1994), and the mixed-effect (ME) approach, which treats the repeated measures as univariate (also known as the “long” data format) and is generally fitted with regression software (Bryk & Raudenbush, 1987; Laird & Ware, 1982; Rao 1965).Footnote1
+
+Over the past 20 years, methodological research has shown that the LC and ME approaches are actually nuanced twists on the same idea; they have been shown to converge, and to be mathematically equivalent in many cases (e.g., Bauer, 2003; Curran, 2003; Ledermann & Kenny, 2017; Mehta & Neale, 2005).
 

### 2023-04-28 index.qmd
landing page updates 3
landing page updates 4
landing page updates 6

@@
   html:
+    css: styles.css
+    theme: cosmo
     page-layout: full
     grid:
-      sidebar-width: 0px
-      body-width: 2000px
+      sidebar-width: 200px
+      body-width: 1500px
       margin-width: 0px
@@
 
-![](./figures/Figlanding4.svg)
+![](./figures/Figlanding5.svg)

### 2023-04-28 manuscript.qmd
landing page updates 8

@@
     email: SherK@missouri.edu
+  - name: Erin L. Thompson
+    department: Center for Children and Families
+    affiliation: Florida International University
+    location: Miami, FL
+    email: erthomps@fiu.edu
+  - name: Raul Gonzalez
+    department: Center for Children and Families
+    affiliation: Florida International University
+    location: Miami, FL
+    email: gonzara@fiu.edu  
   - name: Additional Co-authors
@@
 
+![test](./figures/Mindmap1.png)
+
 # Part IV: Discussion

### 2023-04-29 figures/Code for LDA Structure Mermaid Diagram.qmd
updated LDA Struct image for manuscript

---
title: "Code for LDA Structure Mermaid Diagram"
editor: visual
---

```{mermaid}

%%| echo: false
%%| fig-width: 3
%%| fig-height: 3

graph TB
A[LDA Structures] 
A -.-> |Focus on if & when an event occurs| B[Time-to-Event]
B -.-> C[Continuous-Time <br/>Survival Analysis]
B -.-> D[Discrete-Time <br/>Survival Analysis]
A -->|Focus on Change & Growth| E[Repeated Measures]
E -.-> F[Time-Series <br/> Models]
E -.-> G[Intensive <br/> Longitudinal Data]
E --> H[Longitudinal Panel Data]
H --Traditional Models--> J[Mean-level change & Marginal models]
H --Contemporary  Methods--> I[Within-person change & Random-effects models]

J --> K[ANOVA <br/> Generalized Estimating Equations <br/> ... ]
I --> L[Mixed Models <br/> Growth Curves <br/> ...]

click A href "https://www.github.com" "This is a link"
click B href "https://www.github.com" "This is a link"
click C href "https://www.github.com" "This is a link"
click D href "https://www.github.com" "This is a link"

style A fill:blue,stroke:#333,stroke-width:4px,color:#fff
style E fill:blue,stroke:#333,stroke-width:4px,color:#fff
style B fill:floralwhite,stroke:black
style C fill:floralwhite,stroke:black
style D fill:floralwhite,stroke:black
style F fill:floralwhite,stroke:black
style G fill:floralwhite,stroke:black

style H fill:blue,stroke:#333,stroke-width:4px,color:#fff
style J fill:blue,stroke:#333,stroke-width:2px,color:#fff
style I fill:blue,stroke:#333,stroke-width:2px,color:#fff
style K fill:blue,stroke:#333,stroke-width:2px,color:#fff,text-align:center
style L fill:blue,stroke:#333,stroke-width:2px,color:#fff,text-align:center

```

### 2023-04-29 manuscript.qmd
updated LDA Struct image for manuscript

@@
   html:
+    page-layout: article
+    mermaid:
+      theme: forest
     toc: true
@@
     toc-title: Table of Contents
+    code-tools:
+      source: true
+      toggle: false
+      caption: none
     grid:
-      sidebar-width: 300px
-      margin-width: 300px
+      sidebar-width: 250px
+      margin-width: 250px
   pdf: 
@@
 
-  #format:
-#  pdf:
-#    geometry:
-#      - top=30mm
-#      - left=20mm
-#   docx: default
-# prefer-html: true
+execute:
+  echo: false
 
@@
 
-bibliography: "references.bib" #references.bib
+bibliography: "references.bib"
 biblio-style: unsrt
@@
 
+![LDA Structural Diagram](./figures/LDA Structure 2.png){#fig-LDAstruct  fig-alt="LDA Structure Diagram." width=600 height=400 fig-align="center"}
+
 **Defining Features of Longitudinal Data Analysis.** The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
@@
 That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .
-
-### \[may add section here about the "structure" of longitudinal data we're mostly focused on-- repeated measures of panel data...\]
 
@@
 It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].
+
+This is illustrated well by @fig-LDAstruct.
 
@@
 
+xxxxxx
+
+xxxxxx
+
 ![test](./figures/Mindmap1.png)
@@
 
+|                                           |                         |                          |                                      |                         |
+|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
+| Models                                    |     Form of Change      | Minimum \# of Timepoints | Variable-Centered vs Person-Centered | Time-varying Covariates |
+| Difference Scores                         |       mean-level        |            2+            |                  \-                  |          none           |
+| Change Scores                             |       rank-order        |            2+            |                  \-                  |          none           |
+| Signed-Rank Test                          |       rank-order        |            2+            |                  \-                  |          none           |
+| Generalized Estimating Equations          |       rank-order        |            2+            |                  \-                  |         allowed         |
+| Linear Mixed Models                       |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Generalized Linear Mixed-Effects Models   |   within and between    |            3+            |                                      |         allowed         |
+| Autoregressive Crosslagged Panel Models   |       rank-order        |            3+            |          Variable-Centered           |         allowed         |
+| Latent Change Scores Models               |       rank-order        |            2+            |          Variable-Centered           |         allowed         |
+| Latent Growth Curve Models                |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Multivariate Latent Growth Curve Model    |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Latent Transition Analysis                |   within and between    |            3+            |                Mixed                 |         allowed         |
+| Latent Class Growth Analysis              |   within and between    |            3+            |           Person-Centered            |         allowed         |
+| Growth Mixture Models                     |   within and between    |            3+            |                Mixed                 |         allowed         |
+| State-Trait Models                        | continuous and discrete |            3+            |          Variable-Centered           |         allowed         |
+| Random-Intercept Crosslagged Panel Models |   within and between    |            4+            |          Variable-Centered           |         allowed         |
+| Latent Curve Models Structured Residuals  |   within and between    |            4+            |          Variable-Centered           |         allowed         |
+|                                           |                         |                          |                                      |                         |
+
+: Models for the analysis of longitudinal data {.striped .hover .responsive-xl}
+
 # Part V: References

### 2023-04-29 temp.qmd
landing page updates 8
landing page updates 9
updated LDA Struct image for manuscript

@@
       gutter-width: 1.5rem
+    mermaid:
+      theme: forest
 #    css: styles.css
@@
 
-![FigmaLandingTest2](./figures/Figlanding5.svg)
-(remove marginal models from "examples" address (vs random-effects) in the manuscript)
+![LPA Structure](./figures/LDA Structure 3.svg)
 
-- section: "Traditional Linear Models"
+## Section III
+
+Section focused on different ways of considering/grouping/deciding on which specific analysis to choose (see curran&bauer slides)
+
+---
+
+Steps
+
+1. update manuscript text
+2. section order via yaml file
+3. call out boxes
+
+**
+2 call out blocks/boxes to emphasize main sections of parts II (e.g., vulnerable periods, experience effects) and III (e.g., # of time points, types of stability and change).
+Help to add some visuals to first sections of manuscript
+**
+
+xxxxxxxxxxx
+Traditional Models
+  -Linear
+    - Difference Scores
+    - Change Scores
+  -Non-Linear
+    - Signed-Rank test
+    - Generalized Estimating Equations
+
+Modern Approaches
+  -Linear
+    - Linear Mixed Models
+  -Non-Linear
+    - Generalized Linear Mixed Models
+
+SEM
+  - Variable-Centered
+    - Autoregressive Crosslagged Panel Model
+    - Latent Change Score
+    - Latent Growth Curve Models
+    - Multivariate Latent Growth Curve Models
+    - State-Trait Models
+  - Person-Centered
+    - Latent Transition Analysis
+    - Latent Class Growth Analysis
+  - Hybrid
+    - Growth Mixture Model
+
+Advanced SEM
+  - Random-Intercept Crosslagged Panel Model
+  - Latent Curve Model with Structured Residuals
+
+Longitudinal Twin Designs/Genetic Models
+
+Longitudinal Neuroimaging
+
+xxxxxxxxxx\
+
+Traditional Models
+  -Linear
+    - Difference Scores
+    - Change Scores
+  -Non-Linear
+    - Signed-Rank test
+    - Generalized Estimating Equations
+
+xxxxxxxxx
+    - title: "Tutorials"
+      contents:
+        - tutorials.qmd
+        - section: "Traditional Models"
           contents:
@@
             - 2_Tutorials_ResidualizedChangeScores.qmd
+            - 3_Tutorials_LinearMixedModels.qmd
         - section: "Traditional Non-linear Models"
@@
             - 4_Tutorials_SignedRankTest.qmd
-            - 6_Tutorials_GeneralizedEstimatingEquations.qmd (population average effects)
-        - section: "Modern GLM Extensions(?)"
-            - 7_Tutorials_LinearMixedModels.qmd  
-            - 8_Tutorials_GeneralizedLinearMixedEffectsModels.qmd
+            - 5_Tutorials_MarginalModels.qmd
+            - 6_Tutorials_GeneralizedEstimatingEquations.qmd
+            - 7_Tutorials_GeneralizedLinearMixedEffectsModels.qmd
         - section: "SEM Approaches"
@@
             - 16_Tutorials_LatentCurveModelsStructuredResiduals.qmd
+        - section: "Longitudinal Neuroimaging"
+          contents:
+            - blank.qmd
+        - section: "Twin Modeling & Genetics Designs"
+          contents:
+            - blank.qmd
+        - section: "Data Wrangling"
+          contents:
+            - blank.qmd
+    - title: "How-To"
+      contents:
+        - howto.qmd
+        - section: "Traditional Linear Models"
+          contents:
+            - section: "Difference Scores"
+              contents:
+                - 1a_HowTos_DifferenceScores_PairedTtests.qmd
+                - 1b_HowTos_DifferenceScores_SimpleRegression.qmd
+            - section: "Residualized Change Scores"
+              contents:
+                - 2_HowTos_ResidualizedChangeScores.qmd
+            - section: "Linear Mixed Models"
+              contents:
+                - 3a_HowTos_LinearMixedModels.qmd
+                - 3b_HowTos_LinearMixedModels.qmd
+        - section: "Traditional Non-linear Models"
+          contents:
+            - section: "Signed-Rank Test"
+              contents: 
+                - 4_HowTos_SignedRankTest.qmd
+            - section: "Marginal Models"
+              contents:
+                - 5_HowTos_MarginalModels.qmd
+            - section: "Generalized Estimating Equations"
+              contents:
+                - 6_HowTos_GeneralizedEstimatingEquations.qmd
+            - section: "Generalized Linear Mixed-Effects Models"
+              contents:
+                - 7_HowTos_GeneralizedLinearMixedEffectsModels.qmd
+        - section: "SEM Approaches"
+          contents:
+            - section: "Autoregressive Crosslagged Panel Models"
+              contents:
+                - 8_HowTos_AutoregressiveCrosslaggedPanelModels.qmd
+            - section: "Latent Change Scores Models"
+              contents:
+                - 9_HowTos_LatentChangeScoresModels.qmd
+            - section: "LatentGrowthCurveModels"
+              contents:
+                - 10_HowTos_LatentGrowthCurveModels.qmd
+            - section: "Multivariate Latent Growth Curve Models"
+              contents:
+                - 11_HowTos_MultivariateLatentGrowthCurveModels.qmd
+            - section: "Latent Transition Analysis"
+              contents:
+                - 12_HowTos_LatentTransitionAnalysis.qmd
+            - section: "Growth Mixture Models"
+              contents:
+                - 13_HowTos_GrowthMixtureModels.qmd
+            - section: "State-Trait Models"
+              contents:
+                - 14_HowTos_StateTraitModels.qmd
+        - section: "Advanced SEM"
+          contents:
+            - 15_HowTos_RandomInterceptCrosslaggedPanelModels.qmd
+            - 16_HowTos_LatentCurveModelsStructuredResiduals.qmd
+        - section: "Longitudinal Neuroimaging"
+          contents:
+            - blank.qmd
+        - section: "Twin Modeling & Genetics Designs"
+          contents:
+            - blank.qmd
+        - section: "Data Wrangling"
+          contents:
+            - blank.qmd
+        # navigation items
+    - title: "Explanations"
+      contents:
+        - explanations.qmd
+        - section: "Traditional Linear Models"
+          contents:
+            - 1_Explanations_DifferenceScores.qmd
+            - 2_Explanations_ResidualizedChangeScores.qmd
+            - 3_Explanations_LinearMixedModels.qmd
+        - section: "Traditional Non-linear Models"
+          contents:
+            - 4_Explanations_SignedRankTest.qmd
+            - 5_Explanations_MarginalModels.qmd
+            - 6_Explanations_GeneralizedEstimatingEquations.qmd
+            - 7_Explanations_GeneralizedLinearMixedEffectsModels.qmd
+        - section: "SEM Approaches"
+          contents:
+            - 8_Explanations_AutoregressiveCrosslaggedPanelModels.qmd
+            - 9_Explanations_LatentChangeScoresModels.qmd
+            - 10_Explanations_LatentGrowthCurveModels.qmd
+            - 11_Explanations_MultivariateLatentGrowthCurveModels.qmd
+            - 12_Explanations_LatentTransitionAnalysis.qmd
+            - 13_Explanations_GrowthMixtureModels.qmd
+            - 14_Explanations_StateTraitModels.qmd
+        - section: "Advanced SEM"
+          contents:
+            - 15_Explanations_RandomInterceptCrosslaggedPanelModels.qmd
+            - 16_Explanations_LatentCurveModelsStructuredResiduals.qmd
+        - section: "Longitudinal Neuroimaging"
+          contents:
+            - blank.qmd
+        - section: "Twin Modeling & Genetics Designs"
+          contents:
+            - blank.qmd
+        - section: "Data Wrangling"
+          contents:
+            - blank.qmd
+        # navigation items
 
-GEE, on the other hand, is fitting a marginal model. These model population-averages. You're modeling the expectation conditional only on your fixed design matrix.
+    - title: "Reference"
+      contents:
+        - reference.qmd
+        - section: "List of Reference Material"
+          contents:
+            - ListofReferences.qmd
+
+    - title: "Documentation"
+      contents:
+        - documentation.qmd
+
+xxxxxxxxxxx
+
+![test](./figures/Mindmap1.png)
+
+xxx
+
+| fruit  | price  |
+|--------|--------|
+| apple  | 2.05   |
+| pear   | 1.37   |
+| orange | 3.09   |
+
+: Fruit prices {.striped .hover}
+
+xxx
+
+xxx
+
+```{r}
+library(leaflet)
+leaflet() %>%
+  addTiles() %>%  # Add default OpenStreetMap map tiles
+  addMarkers(lng=-77.0168, lat=38.7814, popup="You are here - in Maryland!")
+
+```
+
+xxx
+
+```{mermaid}
+
+%%| fig-width: 6.5
+%%| fig-height: 6.5
+
+graph TB
+A[LDA Structures] 
+A -.-> |Focus on if & when an event occurs| B[Time-to-Event]
+B -.-> C[Continuous-Time <br/>Survival Analysis]
+B -.-> D[Discrete-Time <br/>Survival Analysis]
+A -->|Focus on Change & Growth| E[Repeated Measures]
+E -.-> F[Time-Series <br/> Models]
+E -.-> G[Intensive <br/> Longitudinal Data]
+E --> H[Longitudinal Panel Data]
+H --Traditional Models--> J[Mean-level change & Marginal models]
+H --Contemporary  Methods--> I[Within-person change & Random-effects models]
+
+J --> K[ANOVA <br/> Generalized Estimating Equations <br/> ... ]
+I --> L[Mixed Models <br/> Growth Curves <br/> ...]
+
+click A href "https://www.github.com" "This is a link"
+click B href "https://www.github.com" "This is a link"
+click C href "https://www.github.com" "This is a link"
+click D href "https://www.github.com" "This is a link"
+
+style A fill:blue,stroke:#333,stroke-width:4px,color:#fff
+style E fill:blue,stroke:#333,stroke-width:4px,color:#fff
+style B fill:floralwhite,stroke:black
+style C fill:floralwhite,stroke:black
+style D fill:floralwhite,stroke:black
+style F fill:floralwhite,stroke:black
+style G fill:floralwhite,stroke:black
+
+style H fill:blue,stroke:#333,stroke-width:4px,color:#fff
+style J fill:blue,stroke:#333,stroke-width:2px,color:#fff
+style I fill:blue,stroke:#333,stroke-width:2px,color:#fff
+style K fill:blue,stroke:#333,stroke-width:2px,color:#fff,text-align:center
+style L fill:blue,stroke:#333,stroke-width:2px,color:#fff,text-align:center
+
+```
+
+![FigmaLandingTest2](./figures/Figlanding5.svg) (remove marginal models from "examples" address (vs random-effects) in the manuscript)
+
+-   section: "Traditional Linear Models" contents: - 1_Tutorials_DifferenceScores.qmd - 2_Tutorials_ResidualizedChangeScores.qmd - section: "Traditional Non-linear Models" contents: - 4_Tutorials_SignedRankTest.qmd - 6_Tutorials_GeneralizedEstimatingEquations.qmd (population average effects) - section: "Modern GLM Extensions(?)" - 7_Tutorials_LinearMixedModels.qmd\
+    -   8_Tutorials_GeneralizedLinearMixedEffectsModels.qmd - section: "SEM Approaches" contents: - 8_Tutorials_AutoregressiveCrosslaggedPanelModels.qmd - 9_Tutorials_LatentChangeScoresModels.qmd - 10_Tutorials_LatentGrowthCurveModels.qmd - 11_Tutorials_MultivariateLatentGrowthCurveModels.qmd - 12_Tutorials_LatentTransitionAnalysis.qmd - 13_Tutorials_GrowthMixtureModels.qmd - 14_Tutorials_StateTraitModels.qmd - section: "Advanced SEM" contents: - 15_Tutorials_RandomInterceptCrosslaggedPanelModels.qmd - 16_Tutorials_LatentCurveModelsStructuredResiduals.qmd
+
+GEE, on the other hand, is fitting a marginal model.
+These model population-averages.
+You're modeling the expectation conditional only on your fixed design matrix.
 GLMM is fitting a mixed-effects model; Thus the the estimates are subject-specific.
@@
 
-2. Models for Longitudinal Data
-Since both the GLMM and WGEE are extensions of the GLM, we start with a brief overview of the latter.
+|                                           |                         |                          |                                      |                         |
+|:-----------------------------|:----------------------:|:---------------------:|:---------------------------:|:-----------------:|
+| Models                                    |     Form of Change      | Minimum \# of Timepoints | Variable-Centered vs Person-Centered | Time-varying Covariates |
+| Difference Scores                         |       mean-level        |            2+            |                  \-                  |          none           |
+| Change Scores                             |       rank-order        |            2+            |                  \-                  |          none           |
+| Signed-Rank Test                          |       rank-order        |            2+            |                  \-                  |          none           |
+| Generalized Estimating Equations          |       rank-order        |            2+            |                  \-                  |         allowed         |
+| Linear Mixed Models                       |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Generalized Linear Mixed-Effects Models   |   within and between    |            3+            |                                      |         allowed         |
+| Autoregressive Crosslagged Panel Models   |       rank-order        |            3+            |          Variable-Centered           |         allowed         |
+| Latent Change Scores Models               |       rank-order        |            2+            |          Variable-Centered           |         allowed         |
+| Latent Growth Curve Models                |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Multivariate Latent Growth Curve Model    |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Latent Transition Analysis                |   within and between    |            3+            |                Mixed                 |         allowed         |
+| Latent Class Growth Analysis              |   within and between    |            3+            |           Person-Centered            |         allowed         |
+| Growth Mixture Models                     |   within and between    |            3+            |                Mixed                 |         allowed         |
+| State-Trait Models                        | continuous and discrete |            3+            |          Variable-Centered           |         allowed         |
+| Random-Intercept Crosslagged Panel Models |   within and between    |            4+            |          Variable-Centered           |         allowed         |
+| Latent Curve Models Structured Residuals  |   within and between    |            4+            |          Variable-Centered           |         allowed         |
+|                                           |                         |                          |                                      |                         |
+: Models for the analysis of longitudinal data {.striped .hover .responsive-xl}
+
+xxxxxx
+
+2.  Models for Longitudinal Data Since both the GLMM and WGEE are extensions of the GLM, we start with a brief overview of the latter.
 
@@
 These types of models are generally referred to as subject-specific models (Zeger, Liang, & Albert, 1988), which are more commonly known in psychology as growth curve models.
-Growth models have many aliases but can be broadly grouped into two different classes of methods: the latent-curve (LC) approach, which treats the repeated measures as multivariate (also known as the “wide” data format) and tends to be fit with general structural equation modeling (SEM) software (Meredith & Tisak, 1990; Tucker, 1958; Willett & Sayer, 1994), and the mixed-effect (ME) approach, which treats the repeated measures as univariate (also known as the “long” data format) and is generally fitted with regression software (Bryk & Raudenbush, 1987; Laird & Ware, 1982; Rao 1965).Footnote1
+Growth models have many aliases but can be broadly grouped into two different classes of methods: the latent-curve (LC) approach, which treats the repeated measures as multivariate (also known as the "wide" data format) and tends to be fit with general structural equation modeling (SEM) software (Meredith & Tisak, 1990; Tucker, 1958; Willett & Sayer, 1994), and the mixed-effect (ME) approach, which treats the repeated measures as univariate (also known as the "long" data format) and is generally fitted with regression software (Bryk & Raudenbush, 1987; Laird & Ware, 1982; Rao 1965).Footnote1
 
@@
 
-<div class="page-layout-custom">
+::: page-layout-custom
 ![FigmaLandingTest2](./figures/Figlanding2.svg)
-</div>
+:::
+
+xxxxx It is important to match a theory of change with design and statistical model to assess change.
+(p.
+4) Models with two time points cannot provide estimates of within-person change.
+"Stability" over time can reflect many things, only some of which mean a lack of change.
+(p.
+14) It is helpful to describe not only average growth (e.g.
+intercept and slope means), but variances in the growth parameters.
+(p.
+25)
+
+Models that focus exclusively on mean-level change sometimes called Marginal Models Those that emphasize individual change often do so through inclusion of Random Effects often called "growth models"
 
 xxxxx
-It is important to match a theory of change with design and statistical model to assess change. (p. 4)
-Models with two time points cannot provide estimates of within-person change.
-“Stability” over time can reflect many things, only some of which mean a lack of change. (p. 14)
-It is helpful to describe not only average growth (e.g. intercept and slope means), but variances in the growth parameters. (p. 25)
 
-Models that focus exclusively on mean-level change sometimes called Marginal Models
-Those that emphasize individual change often do so through inclusion of Random Effects
-often called “growth models”
+```{mermaid}
+mindmap
+  root((mindmap))
+    Origins
+      Long history
+      ::icon(fa fa-book)
+      Popularisation
+        British popular psychology author Tony Buzan
+    Research
+      On effectiveness<br/>and features
+      On Automatic creation
+        Uses
+            Creative techniques
+            Strategic planning
+            Argument mapping
+    Tools
+      Pen and paper
+      Mermaid
+
+```
+
+xxxxx
 
@@
 
-::: {.grid}
-
-::: {.g-col-2}
+::: grid
+::: g-col-2
 :::
 
-::: {.g-col-8}
-
+::: g-col-8
 ::: {style="color: red; border: solid black; background-color: green;"}
@@
 :::
-
 :::
 
-::: {.g-col-2}
+::: g-col-2
 :::
-
 :::
@@
 
-::: {.grid}
-
-::: {.g-col-2}
+::: grid
+::: g-col-2
 :::
 
-::: {.g-col-8}
+::: g-col-8
 ![](./figures/lights.png)
@@
 
-::: {.g-col-2}
+::: g-col-2
 :::
-
 :::
@@
 ::: {style="text-align: justify; text-align: center; line-height: normal; color: white; border: solid black; background-color: darkgoldenrod; padding: 70px; font-family: Georgia, serif;"}
-
 This is the temporary landing page for the Longitudinal Analysis Project. <br> The site is currently under construction and used for testing purposes only. <br> Contents may be inaccurate.
-
 :::

### 2023-04-29 index.qmd
updated LDA Struct image for manuscript 2

--- 2023-04-27 index.qmd
@@
   html:
+    css: styles.css
+    theme: cosmo
     page-layout: full
     grid:
-      sidebar-width: 0px
+      sidebar-width: 250px
       body-width: 2000px
-      margin-width: 0px
+      margin-width: 100px
     gutter-width: 1.5rem
@@
 
-![](./figures/Figlanding4.svg)
+::: {.grid}
+
+::: {.g-col-12}
+![](./figures/Figlanding5.svg)
+:::
+
+:::

### 2023-04-30 index.qmd
updates to manucript figs

--- 2023-04-27 index.qmd
@@
   html:
+    css: styles.css
+    theme: cosmo
     page-layout: full
@@
       margin-width: 0px
-      gutter-width: 1.5rem
+    #gutter-width: 1.5rem
 ---
 
-![](./figures/Figlanding4.svg)
+::: {.column-screen}
+![](./figures/Figlanding5.svg)
+:::

### 2023-04-30 manuscript.qmd
updated LDA Struct image for manuscript 2
updates to manucript figs

@@
     grid:
-      sidebar-width: 250px
-      margin-width: 250px
+      sidebar-width: 200px
+      margin-width: 0px
   pdf: 
@@
       - left=20mm
+    keep_tex: yes
     #fontfamily: libertinus
     #fontsize: 20pt
+    colorlinks: true
 format-links: [html, pdf]
@@
   rticles::arxiv_article
-
 bibliography: "references.bib"
@@
 ## The ABCD Study® Data
-
 Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians.
@@
 \[section still be developed...\]
+
+::: {.callout-note  collapse="true"}
+## Tip with Title
+Note that there are five types of callouts, including:
+`xxxxx`, `xxxxxx`, `xxxxxx`.
+:::
 
@@
 
-xxxxxx
-
-xxxxxx
-
-![test](./figures/Mindmap1.png)
+![Longitudinal Models](./figures/Mindmap1.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width=600 height=400 fig-align="center"}
 
@@
 
-: Models for the analysis of longitudinal data {.striped .hover .responsive-xl}
+: LDA Models {#tbl-lda .striped .hover .responsive-xl}
+
+axasx
+
+See @tbl-lda.
+
+xxxxx
 

### 2023-04-30 temp.qmd
updates to manucript figs

@@
 ---
-title: "Tutorials: A Brief Overview"
+#title: "Tutorials: A Brief Overview"
 format:
@@
     grid:
+      #sidebar-width: 250px
       sidebar-width: 0px
@@
       margin-width: 0px
-      gutter-width: 1.5rem
-    mermaid:
-      theme: forest
-#    css: styles.css
+      #gutter-width: 1.5rem
+    #mermaid:
+      #theme: forest
+    css: styles.css
 #    theme: cosmo
@@
 
+<script src="https://cdn.tailwindcss.com"></script>
+
+```{html}
+<div class="alert alert-primary" role="alert">
+  This is a primary alert—check it out!
+</div>
+```
+
+xxxyyyzzz
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```       
+\begin{tabular}[t]{@{}l}
+This is where the reciepient goes \\ 
+Address 1 \\
+Address 2 \\
+Address 3 \\
+Address 3 \\
+\end{tabular}
+\hfill
+\begin{tabular}[t]{@{}l}
+This is where the sender goes \\  
+Person \\
+Department \\ 
+Address 1 \\ 
+Address 2 \\  
+Address 3 \\  
+Address 4 \\  
+Address 5 \\  
+Address 6 \\ 
+Telephone \\  
+Fax \\  
+Email \\
+\end{tabular}
+
+xxxyyyzzz
+xxxxx
+
+<div class="alert alert-primary" role="alert">
+  This is a primary alert with <a href="#" class="alert-link">an example link</a>. Give it a click if you like.
+</div>
+
+<span class="badge badge-pill badge-primary">Primary</span>
+
+<a href="#" class="badge badge-primary">Primary</a>
+
+<div class="card" style="width: 18rem;">
+  <img class="card-img-top" src="..." alt="Card image cap">
+  <div class="card-body">
+    <div class="card-title">Card title</div>
+    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
+    <a href="#" class="btn btn-primary">Go somewhere</a>
+  </div>
+</div>
+
+<div class="card">
+  <div class="card-body">
+    This is some text within a card body.
+  </div>
+</div>
+
+<div class="card" style="width: 18rem;">
+  <img class="card-img-top" src="..." alt="Card image cap">
+  <div class="card-body">
+    <h5 class="card-title">Card title</h5>
+    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
+  </div>
+  <ul class="list-group list-group-flush">
+    <li class="list-group-item">Cras justo odio</li>
+    <li class="list-group-item">Dapibus ac facilisis in</li>
+    <li class="list-group-item">Vestibulum at eros</li>
+  </ul>
+  <div class="card-body">
+    <a href="#" class="card-link">Card link</a>
+    <a href="#" class="card-link">Another link</a>
+  </div>
+</div>
+
+<div class="card">
+  <div class="card-header">
+    Featured
+  </div>
+  <div class="card-body">
+    <h5 class="card-title">Special title treatment</h5>
+    <p class="card-text">With supporting text below as a natural lead-in to additional content.</p>
+    <a href="#" class="btn btn-primary">Go somewhere</a>
+  </div>
+</div>
+
+<div class="card">
+  <h5 class="card-header">Featured</h5>
+  <div class="card-body">
+    <h5 class="card-title">Special title treatment</h5>
+    <p class="card-text">With supporting text below as a natural lead-in to additional content.</p>
+    <a href="#" class="btn btn-primary">Go somewhere</a>
+  </div>
+</div>
+
+<div class="card text-white bg-primary mb-3" style="max-width: 18rem;">
+  <div class="card-header">Header</div>
+  <div class="card-body">
+    <h5 class="card-title">Primary card title</h5>
+    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
+  </div>
+</div>
+
+<div class="card border-primary mb-3" style="max-width: 18rem;">
+  <div class="card-header">Header</div>
+  <div class="card-body text-primary">
+    <h5 class="card-title">Primary card title</h5>
+    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
+  </div>
+</div>
+
+<div class="card border-success mb-3" style="max-width: 18rem;">
+  <div class="card-header bg-transparent border-success">Header</div>
+  <div class="card-body text-success">
+    <h5 class="card-title">Success card title</h5>
+    <p class="card-text">Some quick example text to build on the card title and make up the bulk of the card's content.</p>
+  </div>
+  <div class="card-footer bg-transparent border-success">Footer</div>
+</div>
+<div class="alert alert-success" role="alert">
+  <h4 class="alert-heading">Well done!</h4>
+  <p>Aww yeah, you successfully read this important alert message. This example text is going to run a bit longer so that you can see how spacing within an alert works with this kind of content.</p>
+  <hr>
+  <p class="mb-0">Whenever you need to, be sure to use margin utilities to keep things nice and tidy.</p>
+</div>
+
+<dl class="row">
+  <dt class="col-sm-3">Description lists</dt>
+  <dd class="col-sm-9">A description list is perfect for defining terms.</dd>
+
+  <dt class="col-sm-3">Euismod</dt>
+  <dd class="col-sm-9">
+    <p>Vestibulum id ligula porta felis euismod semper eget lacinia odio sem nec elit.</p>
+    <p>Donec id elit non mi porta gravida at eget metus.</p>
+  </dd>
+
+  <dt class="col-sm-3">Malesuada porta</dt>
+  <dd class="col-sm-9">Etiam porta sem malesuada magna mollis euismod.</dd>
+
+  <dt class="col-sm-3 text-truncate">Truncated term is truncated</dt>
+  <dd class="col-sm-9">Fusce dapibus, tellus ac cursus commodo, tortor mauris condimentum nibh, ut fermentum massa justo sit amet risus.</dd>
+
+  <dt class="col-sm-3">Nesting</dt>
+  <dd class="col-sm-9">
+    <dl class="row">
+      <dt class="col-sm-4">Nested definition list</dt>
+      <dd class="col-sm-8">Aenean posuere, tortor sed cursus feugiat, nunc augue blandit nunc.</dd>
+    </dl>
+  </dd>
+</dl>
+
+<!-- Button trigger modal -->
+<button type="button" class="btn btn-primary" data-toggle="modal" data-target="#exampleModal">
+  Launch demo modal
+</button>
+
+<!-- Modal -->
+<div class="modal fade" id="exampleModal" tabindex="-1" role="dialog" aria-labelledby="exampleModalLabel" aria-hidden="true">
+  <div class="modal-dialog" role="document">
+    <div class="modal-content">
+      <div class="modal-header">
+        <h5 class="modal-title" id="exampleModalLabel">Modal title</h5>
+        <button type="button" class="close" data-dismiss="modal" aria-label="Close">
+          <span aria-hidden="true">&times;</span>
+        </button>
+      </div>
+      <div class="modal-body">
+        ...
+      </div>
+      <div class="modal-footer">
+        <button type="button" class="btn btn-secondary" data-dismiss="modal">Close</button>
+        <button type="button" class="btn btn-primary">Save changes</button>
+      </div>
+    </div>
+  </div>
+</div>
+
+<button type="button" class="btn btn-lg btn-danger" data-toggle="popover" title="Popover title" data-content="And here's some amazing content.
+It's very engaging.
+Right?">Click to toggle popover</button>
+
+<div class="w-25 p-3" style="background-color: #eee;">Width 25%</div>
+<div class="w-50 p-3" style="background-color: #eee;">Width 50%</div>
+<div class="w-75 p-3" style="background-color: #eee;">Width 75%</div>
+<div class="w-100 p-3" style="background-color: #eee;">Width 100%</div>
+
+<div style="height: 100px; background-color: rgba(255,0,0,0.1);">
+  <div class="h-25 d-inline-block" style="width: 120px; background-color: rgba(0,0,255,.1)">Height 25%</div>
+  <div class="h-50 d-inline-block" style="width: 120px; background-color: rgba(0,0,255,.1)">Height 50%</div>
+  <div class="h-75 d-inline-block" style="width: 120px; background-color: rgba(0,0,255,.1)">Height 75%</div>
+  <div class="h-100 d-inline-block" style="width: 120px; background-color: rgba(0,0,255,.1)">Height 100%</div>
+</div>
+
+<p class="text-justify">Ambitioni dedisse scripsisse iudicaretur.
+Cras mattis iudicium purus sit amet fermentum.
+Donec sed odio operae, eu vulputate felis rhoncus.
+Praeterea iter est quasdam res quas ex communi.
+At nos hinc posthac, sitientis piros Afros.
+Petierunt uti sibi concilium totius Galliae in diem certam indicere.
+Cras mattis iudicium purus sit amet fermentum.</p>
+
+<p class="text-left">Left aligned text on all viewport sizes.</p>
+<p class="text-center">Center aligned text on all viewport sizes.</p>
+<p class="text-right">Right aligned text on all viewport sizes.</p>
+
+<p class="text-sm-left">Left aligned text on viewports sized SM (small) or wider.</p>
+<p class="text-md-left">Left aligned text on viewports sized MD (medium) or wider.</p>
+<p class="text-lg-left">Left aligned text on viewports sized LG (large) or wider.</p>
+<p class="text-xl-left">Left aligned text on viewports sized XL (extra-large) or wider.</p>
+
+<p class="font-weight-bold">Bold text.</p>
+<p class="font-weight-normal">Normal weight text.</p>
+<p class="font-weight-light">Light weight text.</p>
+<p class="font-italic">Italic text.</p>
+
+<table class="table table-bordered">
+  <thead>
+    <tr>
+      <th scope="col">#</th>
+      <th scope="col">First</th>
+      <th scope="col">Last</th>
+      <th scope="col">Handle</th>
+    </tr>
+  </thead>
+  <tbody>
+    <tr>
+      <th scope="row">1</th>
+      <td>Mark</td>
+      <td>Otto</td>
+      <td>@mdo</td>
+    </tr>
+    <tr>
+      <th scope="row">2</th>
+      <td>Jacob</td>
+      <td>Thornton</td>
+      <td>@fat</td>
+    </tr>
+    <tr>
+      <th scope="row">3</th>
+      <td colspan="2">Larry the Bird</td>
+      <td>@twitter</td>
+    </tr>
+  </tbody>
+</table>
+
+<blockquote class="blockquote">
+  <p class="mb-0">Lorem ipsum dolor sit amet, consectetur adipiscing elit. Integer posuere erat a ante.</p>
+</blockquote>
+
+<table class="table">
+  <thead>
+    <tr>
+      <th scope="col">#</th>
+      <th scope="col">First</th>
+      <th scope="col">Last</th>
+      <th scope="col">Handle</th>
+    </tr>
+  </thead>
+  <tbody>
+    <tr>
+      <th scope="row">1</th>
+      <td>Mark</td>
+      <td>Otto</td>
+      <td>@mdo</td>
+    </tr>
+    <tr>
+      <th scope="row">2</th>
+      <td>Jacob</td>
+      <td>Thornton</td>
+      <td>@fat</td>
+    </tr>
+    <tr>
+      <th scope="row">3</th>
+      <td>Larry</td>
+      <td>the Bird</td>
+      <td>@twitter</td>
+    </tr>
+  </tbody>
+</table>
+
+<table class="hover:table-fixed">
+  <thead>
+    <tr>
+      <th>Song</th>
+      <th>Artist</th>
+      <th>Year</th>
+    </tr>
+  </thead>
+  <tbody>
+    <tr>
+      <td>The Sliding Mr. Bones (Next Stop, Pottersville)</td>
+      <td>Malcolm Lockyer</td>
+      <td>1961</td>
+    </tr>
+    <tr>
+      <td>Witchy Woman</td>
+      <td>The Eagles</td>
+      <td>1972</td>
+    </tr>
+    <tr>
+      <td>Shining Star</td>
+      <td>Earth, Wind, and Fire</td>
+      <td>1975</td>
+    </tr>
+  </tbody>
+</table>
+
+::: {.column-screen}
+![-](./figures/Figlanding5.svg)
+:::
+
+xxxxx
+
+file:///Users/shawes/git/swhawes.github.io/docs/manuscript.html
+xxxxx
+
+```{dot}
+digraph "Linux_kernel_diagram" {
+	fontname="Helvetica,Arial,sans-serif"
+	node [fontname="Helvetica,Arial,sans-serif"]
+	edge [fontname="Helvetica,Arial,sans-serif"]
+	graph [
+		newrank = true,
+		nodesep = 0.3,
+		ranksep = 0.2,
+		overlap = true,
+		splines = false,
+	]
+	node [
+		fixedsize = false,
+		fontsize = 24,
+		height = 1,
+		shape = box,
+		style = "filled,setlinewidth(5)",
+		width = 2.2
+	]
+	edge [
+		arrowhead = none,
+		arrowsize = 0.5,
+		labelfontname = "Ubuntu",
+		weight = 10,
+		style = "filled,setlinewidth(5)"
+	]
+	subgraph system {
+		node [color = "#e27dd6ff"]
+		edge [color = "#e27dd6ff"]
+		system_ [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			shape = point
+		]
+		system [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/System",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		system -> system_ [
+			arrowhead = "",
+			row = func];
+		SCI [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Syscalls",
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = "System calls",
+			row = usr,
+			shape = ellipse]
+		sysfs [
+			fillcolor = "#b2d3e4",
+			label = "proc & sysfs\nfile systems"]
+		SCI -> sysfs
+		DM [
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			fontsize = 20,
+			height = 0.8,
+			label = "Device\nModel",
+			shape = octagon,
+			width = 2]
+		sysfs -> DM
+		log_sys [
+			fillcolor = "#6a9ab1",
+			fontsize = 20,
+			label = "system run,\nmodules,\ngeneric\nHW access "]
+		DM -> log_sys
+		bus_drv [
+			fillcolor = "#71809b",
+			label = "bus drivers"]
+		log_sys -> bus_drv
+		buses [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			label = "buses:\nPCI, USB ...",
+			row = chip]
+		bus_drv -> buses
+	}
+	subgraph networking {
+		node [color = "#61c2c5"]
+		edge [color = "#61c2c5"]
+		networking_ [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			shape = point
+				width = 0]
+		networking [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Networking",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		networking -> networking_ [
+			arrowhead = "",
+			row = func]
+		sock [
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = Sockets,
+			row = usr,
+			shape = ellipse]
+		prot_fam [
+			fillcolor = "#b2d3e4",
+			label = "protocol\nfamilies"]
+		sock -> prot_fam
+		log_prot [
+			fillcolor = "#6a9ab1",
+			label = "protocols:\nTCP, UDP, IP"]
+		prot_fam -> log_prot
+		netif [
+			fillcolor = "#71809b",
+			fontsize = 20,
+			label = "network\ninterfaces\nand drivers"]
+		log_prot -> netif
+		net_hw [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			label = "network:\nEthernet, WiFi ...",
+			row = chip]
+		netif -> net_hw
+		NFS [
+			color = "#8383cc",
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			height = 0.8,
+			label = NFS,
+			shape = octagon,
+			width = 1.2]
+		NFS -> log_prot [weight = 0]
+	}
+	subgraph processing {
+		node [color = "#c46747"]
+		edge [color = "#c46747"]
+		processing_ [
+			fixedsize = true,
+			height = 0,
+			shape = point
+				style = invis,
+			width = 0]
+		processing [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Processing",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		processing -> processing_ [
+			arrowhead = "",
+			row = func]
+		proc [
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = Processes,
+			row = usr,
+			shape = ellipse]
+		Tasks [
+			fillcolor = "#b2d3e4"]
+		proc -> Tasks
+		sync [
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			fontsize = 20,
+			fontname = "Arial Narrow"
+			label = synchronization,
+			height = 0.7,
+			//width = 2,
+			shape = octagon]
+		Tasks -> sync
+		sched [
+			fillcolor = "#6a9ab1",
+			label = Scheduler]
+		sync -> sched
+		IRQ [
+			fillcolor = "#71809b",
+			fontsize = 20,
+			label = "interrupts\ncore,\nCPU arch"]
+		sched -> IRQ
+		CPU [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			row = chip]
+		IRQ -> CPU
+	}	// processing
+	subgraph mem {
+		node [
+			color = "#51bf5b",
+			height = 1
+		]
+		edge [color = "#51bf5b"]
+		MA [
+			color = "#51bf5b",
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = "memory\naccess",
+			row = usr,
+			height = 1,
+			shape = ellipse]
+		MA -> VM
+		mmap [
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			fontsize = 20,
+			height = 0.8,
+			label = "memory\nmapping",
+			shape = octagon,
+			width = 2]
+		mmap -> log_mem
+		log_mem -> PA
+		SW [
+			color = "#8383cc",
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			label = Swap,
+			height = 0.8,
+			shape = octagon,
+			width = 1.2]
+		mmap -> SW [weight = 1]
+		SW -> block [
+			color = "#8383cc", weight = 1]
+		PA [
+			fillcolor = "#71809b",
+			label = "Page\nAllocator"
+		]
+		PC -> PA [weight = 0 color="#51bf5b"]
+		RAM [
+			color = "#51bf5b",
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			label = "MMU, RAM",
+			height = 1,
+			row = chip]
+		PA -> RAM
+		memory -> memory_ [
+			arrowhead = "",
+			row = func]
+		VM -> mmap
+	}	// mem
+	subgraph storage {
+		node [color = "#8383cc"]
+		edge [color = "#8383cc"]
+		NFS;
+		storage_ [
+			shape = point,
+			fixedsize = true,
+			height = 0,
+			style = invis,
+			width = 0]
+		storage [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Storage",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		storage -> storage_ [
+			arrowhead = "",
+			row = func]
+		FS [
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = "files and\ndirectories",
+			row = usr,
+			shape = ellipse]
+		VFS [
+			fillcolor = "#b2d3e4",
+			label = "Virtual\nFile System"]
+		FS -> VFS
+		VFS -> mmap [weight = 0]
+		VFS -> NFS [weight = 0]
+		logFS [
+			fillcolor = "#6a9ab1",
+			fontsize = 20,
+			label = "logical\nfilesystems:\next3, xfs ..."]
+		VFS -> logFS
+		PC [
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			fontsize = 20,
+			height = 0.8,
+			label = "page\ncache",
+			shape = octagon,
+			width = 1.2]
+		VFS -> PC [weight = 0]
+		block [
+			fillcolor = "#71809b",
+			fontsize = 20,
+			label = "Block\ndevices\nand drivers"]
+		logFS -> block
+		SD [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			label = "storage devices:\nSCSI, NVMe ...",
+			row = chip]
+		block -> SD
+	}	// storge
+	subgraph HI {
+		node [color = "#cfbf57ff"]
+		edge [
+			color = "#cfbf57ff",
+			weight = 10
+		]
+		HI_ [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		HI [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Human_interfaces",
+			fillcolor = white,
+			fixedsize = true,
+			fontsize = 12,
+			height = 0.6,
+			label = "Forms\nof\nchange",
+			row = func,
+			width = 2]
+		HI -> HI_ [
+			arrowhead = "",
+			row = func]
+		char [
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = "char\ndevices",
+			row = usr,
+			shape = ellipse]
+		input [
+			fillcolor = "#b2d3e4",
+			label = "input\nsubsystem"]
+		char -> input
+		F7 [
+			fillcolor = "#6a9ab1",
+			label = "HI class\ndrivers"]
+		input -> F7
+		HID [
+			fillcolor = "#71809b",
+			fontsize = 20,
+			URL = "https://www.kernel.org/doc/html/latest/hid/",
+			label = "HI\nperipherals\ndrivers"]
+		F7 -> HID
+		display [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 19,
+			label = "keyboard, mouse,\ndisplay, audio",
+			row = chip]
+		HID -> display
+	} // HI
+	subgraph functions {
+		graph [rank = same]
+		edge [
+			style = invis,
+			weight = 1
+		]
+		system;
+		networking;
+		system -> processing [weight = 1]
+		storage -> networking [weight = 1]
+		memory [
+			color = "#51bf5b",
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Memory",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		memory -> storage [weight = 1]
+		processing -> memory [weight = 1]
+		functions_ [
+			fixedsize = true,
+			height = 0,
+			shape = point
+			style = invis,
+			width = 0]
+		functions_ -> HI -> system [weight = 1]
+		functions [
+			color = gray,
+			tooltip = "Columns represent main functionalities of the kernel",
+			URL = "http://www.makelinux.net/ldd3/chp-1-sect-2.shtml",
+			fillcolor = gray,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			style = dashed,
+			width = 1.6]
+		functions -> functions_ [
+			arrowhead = "",
+			color = gray,
+			style = "",
+			weight = ""]
+	}
+	subgraph interfaces {
+		graph [rank = same]
+		SCI;
+		sock;
+		FS;
+		proc;
+		char;
+		usr_ [
+			fixedsize = true,
+			height = 0,
+			shape = point
+				style = invis,
+			width = 0.5]
+		usr [
+			fillcolor = "#d9e7eeff",
+			fixedsize = true,
+			label = "user space\ninterfaces",
+			row = usr,
+			shape = ellipse,
+			style = "filled,setlinewidth(0)"]
+		MA;
+	}
+	{
+		edge [style = invis weight = 10 ]
+		system_;
+		SCI;
+		system_ -> SCI;
+		networking_;
+		sock;
+		networking_ -> sock;
+		storage_;
+		FS;
+		storage_ -> FS;
+		processing_;
+		proc;
+		processing_ -> proc;
+		HI_;
+		char;
+		HI_ -> char;
+		MA;
+		memory_ [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		memory_ -> MA;
+	}
+	subgraph virtual {
+		graph [rank = same]
+		sysfs;
+		prot_fam;
+		VFS;
+		Tasks;
+		input;
+		D0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		virt [
+			fillcolor = "#b2d3e4",
+			label = "virtual\nsubsystems",
+			URL = "https://en.wikipedia.org/wiki/Proxy_pattern",
+			tooltip = "proxy between standard user space interfaces and internal implementations",
+			style = "filled,setlinewidth(0)"]
+		VM [
+			color = "#51bf5b",
+			fillcolor = "#b2d3e4",
+			label = "Virtual\nmemory"]
+	}
+	subgraph bridges {
+		graph [rank = same]
+		bridges [
+			fillcolor = "#91b5c9",
+			shape = octagon,
+			tooltip = "bridges between uniform virtual interfaces and various implementations",
+			URL = "https://en.wikipedia.org/wiki/Bridge_pattern",
+			style = "filled,setlinewidth(0)"]
+		DM;
+		NFS;
+		mmap;
+		sync;
+		E0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		//PC
+	}
+	subgraph logical {
+		graph [rank = same]
+		log_sys;
+		log_prot;
+		logFS;
+		sched;
+		F7;
+		F0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		logical [
+			fillcolor = "#6a9ab1",
+			style = "filled,setlinewidth(0)"]
+		log_mem [
+			color = "#51bf5b",
+			fillcolor = "#6a9ab1",
+			label = "logical\nmemory"]
+		//SW
+	}
+	subgraph HWI {
+		graph [rank = same]
+		HWI [
+			fillcolor = "#71809b",
+			label = "hardware\ninterfaces",
+			style = "filled,setlinewidth(0)"]
+		bus_drv;
+		netif;
+		block;
+		//PA;
+		IRQ;
+		HID;
+		G0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+	}
+	subgraph HW {
+		graph [rank = same]
+		HW [
+			fillcolor = "#777777",
+			fontcolor = white,
+			label = "electronics,\nhardware",
+			row = chip,
+			style = "filled,setlinewidth(0)"]
+		buses;
+		net_hw;
+		SD;
+		CPU;
+		display;
+		H0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		RAM;
+	}
+	bottom [
+		label = "© 2007-2022 Costa Shulyupin http://www.MakeLinux.net/kernel/diagram",
+		URL = "http://www.MakeLinux.net/kernel/diagram",
+		shape = plaintext,
+		style = ""]
+	CPU -> bottom [style = invis]
+	layers [
+		fillcolor = lightgray,
+		tooltip = "Functionalities are divided to common layers. It is approximate division.",
+		height = 0.1,
+		style = "filled,setlinewidth(0)",
+		width = 0.5]
+	functions -> layers [style = invis ]
+	usr -> usr_ [
+		arrowhead = "",
+		color = "#d9e7eeff",
+		minlen = 2]
+	usr -> virt [
+		color = "#d9e7eeff"]
+	virt -> D0 [
+		arrowhead = "",
+		color = "#b2d3e4",
+		minlen = 2]
+	virt -> bridges [
+		color = "#b2d3e4"]
+	bridges -> E0 [
+		arrowhead = "",
+		color = "#91b5c9",
+		minlen = 2,
+		style = "filled,setlinewidth(6)",
+		weight = ""]
+	bridges -> logical [
+		color = "#91b5c9",
+		style = "filled,setlinewidth(6)"]
+	logical -> F0 [
+		arrowhead = "",
+		color = "#6a9ab1",
+		minlen = 2,
+		row = logical,
+		style = "filled,setlinewidth(6)",
+		weight = ""]
+	logical -> HWI [
+		color = "#6a9ab1",
+		row = logical,
+		style = "filled,setlinewidth(6)"]
+	HWI -> G0 [
+		arrowhead = "",
+		color = "#71809b",
+		minlen = 2,
+		row = HWI,
+		style = "filled,setlinewidth(6)",
+		weight = ""]
+	HWI -> HW [
+		color = "#71809b",
+		row = HWI,
+		style = "filled,setlinewidth(6)"]
+	HW -> H0 [
+		arrowhead = "",
+		color = "#777777",
+		minlen = 2,
+		row = chip,
+		style = "filled,setlinewidth(6)",
+		weight = ""]
+	layers -> usr [
+		arrowhead = "",
+		color = gray,
+		style = "filled,setlinewidth(1)"]
+	LKD [
+		fontsize = 40,
+		label = "Linux kernel diagram",
+		shape = plain,
+		style = ""]
+	LKD -> processing [style = invis]
+}
+
+```
+
+xxxxx
+
 ![LPA Structure](./figures/LDA Structure 3.svg)
@@
 Section focused on different ways of considering/grouping/deciding on which specific analysis to choose (see curran&bauer slides)
+
+```{dot}
+digraph "Linux_kernel_diagram" {
+	fontname="Helvetica,Arial,sans-serif"
+	node [fontname="Helvetica,Arial,sans-serif"]
+	edge [fontname="Helvetica,Arial,sans-serif"]
+	graph [
+		newrank = true,
+		nodesep = 0.3,
+		ranksep = 0.2,
+		overlap = true,
+		splines = false,
+	]
+	node [
+		fixedsize = false,
+		fontsize = 24,
+		height = 1,
+		shape = box,
+		style = "filled,setlinewidth(5)",
+		width = 2.2
+	]
+	edge [
+		arrowhead = none,
+		arrowsize = 0.5,
+		labelfontname = "Ubuntu",
+		weight = 10,
+		style = "filled,setlinewidth(5)"
+	]
+	subgraph system {
+		node [color = "#e27dd6ff"]
+		edge [color = "#e27dd6ff"]
+		system_ [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			shape = point
+		]
+		system [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/System",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		system -> system_ [
+			arrowhead = "",
+			row = func];
+		SCI [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Syscalls",
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = "System calls",
+			row = usr,
+			shape = ellipse]
+		sysfs [
+			fillcolor = "#b2d3e4",
+			label = "proc & sysfs\nfile systems"]
+		SCI -> sysfs
+		DM [
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			fontsize = 20,
+			height = 0.8,
+			label = "Device\nModel",
+			shape = octagon,
+			width = 2]
+		sysfs -> DM
+		log_sys [
+			fillcolor = "#6a9ab1",
+			fontsize = 20,
+			label = "system run,\nmodules,\ngeneric\nHW access "]
+		DM -> log_sys
+		bus_drv [
+			fillcolor = "#71809b",
+			label = "bus drivers"]
+		log_sys -> bus_drv
+		buses [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			label = "buses:\nPCI, USB ...",
+			row = chip]
+		bus_drv -> buses
+	}
+	subgraph networking {
+		node [color = "#61c2c5"]
+		edge [color = "#61c2c5"]
+		networking_ [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			shape = point
+				width = 0]
+		networking [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Networking",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		networking -> networking_ [
+			arrowhead = "",
+			row = func]
+		sock [
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = Sockets,
+			row = usr,
+			shape = ellipse]
+		prot_fam [
+			fillcolor = "#b2d3e4",
+			label = "protocol\nfamilies"]
+		sock -> prot_fam
+		log_prot [
+			fillcolor = "#6a9ab1",
+			label = "protocols:\nTCP, UDP, IP"]
+		prot_fam -> log_prot
+		netif [
+			fillcolor = "#71809b",
+			fontsize = 20,
+			label = "network\ninterfaces\nand drivers"]
+		log_prot -> netif
+		net_hw [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			label = "network:\nEthernet, WiFi ...",
+			row = chip]
+		netif -> net_hw
+		NFS [
+			color = "#8383cc",
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			height = 0.8,
+			label = NFS,
+			shape = octagon,
+			width = 1.2]
+		NFS -> log_prot [weight = 0]
+	}
+	subgraph processing {
+		node [color = "#c46747"]
+		edge [color = "#c46747"]
+		processing_ [
+			fixedsize = true,
+			height = 0,
+			shape = point
+				style = invis,
+			width = 0]
+		processing [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Processing",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		processing -> processing_ [
+			arrowhead = "",
+			row = func]
+		proc [
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = Processes,
+			row = usr,
+			shape = ellipse]
+		Tasks [
+			fillcolor = "#b2d3e4"]
+		proc -> Tasks
+		sync [
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			fontsize = 20,
+			fontname = "Arial Narrow"
+			label = synchronization,
+			height = 0.7,
+			//width = 2,
+			shape = octagon]
+		Tasks -> sync
+		sched [
+			fillcolor = "#6a9ab1",
+			label = Scheduler]
+		sync -> sched
+		IRQ [
+			fillcolor = "#71809b",
+			fontsize = 20,
+			label = "interrupts\ncore,\nCPU arch"]
+		sched -> IRQ
+		CPU [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			row = chip]
+		IRQ -> CPU
+	}	// processing
+	subgraph mem {
+		node [
+			color = "#51bf5b",
+			height = 1
+		]
+		edge [color = "#51bf5b"]
+		MA [
+			color = "#51bf5b",
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = "memory\naccess",
+			row = usr,
+			height = 1,
+			shape = ellipse]
+		MA -> VM
+		mmap [
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			fontsize = 20,
+			height = 0.8,
+			label = "memory\nmapping",
+			shape = octagon,
+			width = 2]
+		mmap -> log_mem
+		log_mem -> PA
+		SW [
+			color = "#8383cc",
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			label = Swap,
+			height = 0.8,
+			shape = octagon,
+			width = 1.2]
+		mmap -> SW [weight = 1]
+		SW -> block [
+			color = "#8383cc", weight = 1]
+		PA [
+			fillcolor = "#71809b",
+			label = "Page\nAllocator"
+		]
+		PC -> PA [weight = 0 color="#51bf5b"]
+		RAM [
+			color = "#51bf5b",
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			label = "MMU, RAM",
+			height = 1,
+			row = chip]
+		PA -> RAM
+		memory -> memory_ [
+			arrowhead = "",
+			row = func]
+		VM -> mmap
+	}	// mem
+	subgraph storage {
+		node [color = "#8383cc"]
+		edge [color = "#8383cc"]
+		NFS;
+		storage_ [
+			shape = point,
+			fixedsize = true,
+			height = 0,
+			style = invis,
+			width = 0]
+		storage [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Storage",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		storage -> storage_ [
+			arrowhead = "",
+			row = func]
+		FS [
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = "files and\ndirectories",
+			row = usr,
+			shape = ellipse]
+		VFS [
+			fillcolor = "#b2d3e4",
+			label = "Virtual\nFile System"]
+		FS -> VFS
+		VFS -> mmap [weight = 0]
+		VFS -> NFS [weight = 0]
+		logFS [
+			fillcolor = "#6a9ab1",
+			fontsize = 20,
+			label = "logical\nfilesystems:\next3, xfs ..."]
+		VFS -> logFS
+		PC [
+			fillcolor = "#91b5c9",
+			fixedsize = true,
+			fontsize = 20,
+			height = 0.8,
+			label = "page\ncache",
+			shape = octagon,
+			width = 1.2]
+		VFS -> PC [weight = 0]
+		block [
+			fillcolor = "#71809b",
+			fontsize = 20,
+			label = "Block\ndevices\nand drivers"]
+		logFS -> block
+		SD [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 20,
+			label = "storage devices:\nSCSI, NVMe ...",
+			row = chip]
+		block -> SD
+	}	// storge
+	subgraph HI {
+		node [color = "#cfbf57ff"]
+		edge [
+			color = "#cfbf57ff",
+			weight = 10
+		]
+		HI_ [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		HI [
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Human_interfaces",
+			fillcolor = white,
+			fixedsize = true,
+			fontsize = 12,
+			height = 0.6,
+			label = "human\ninterface",
+			row = func,
+			width = 2]
+		HI -> HI_ [
+			arrowhead = "",
+			row = func]
+		char [
+			fillcolor = "#d9e7ee",
+			fixedsize = true,
+			label = "char\ndevices",
+			row = usr,
+			shape = ellipse]
+		input [
+			fillcolor = "#b2d3e4",
+			label = "input\nsubsystem"]
+		char -> input
+		F7 [
+			fillcolor = "#6a9ab1",
+			label = "HI class\ndrivers"]
+		input -> F7
+		HID [
+			fillcolor = "#71809b",
+			fontsize = 20,
+			URL = "https://www.kernel.org/doc/html/latest/hid/",
+			label = "HI\nperipherals\ndrivers"]
+		F7 -> HID
+		display [
+			fillcolor = "#777777",
+			fontcolor = white,
+			fontsize = 19,
+			label = "keyboard, mouse,\ndisplay, audio",
+			row = chip]
+		HID -> display
+	} // HI
+	subgraph functions {
+		graph [rank = same]
+		edge [
+			style = invis,
+			weight = 1
+		]
+		system;
+		networking;
+		system -> processing [weight = 1]
+		storage -> networking [weight = 1]
+		memory [
+			color = "#51bf5b",
+			URL = "https://en.wikibooks.org/wiki/The_Linux_Kernel/Memory",
+			fillcolor = white,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			width = 2]
+		memory -> storage [weight = 1]
+		processing -> memory [weight = 1]
+		functions_ [
+			fixedsize = true,
+			height = 0,
+			shape = point
+			style = invis,
+			width = 0]
+		functions_ -> HI -> system [weight = 1]
+		functions [
+			color = gray,
+			tooltip = "Columns represent main functionalities of the kernel",
+			URL = "http://www.makelinux.net/ldd3/chp-1-sect-2.shtml",
+			fillcolor = gray,
+			fixedsize = true,
+			height = 0.6,
+			row = func,
+			style = dashed,
+			width = 1.6]
+		functions -> functions_ [
+			arrowhead = "",
+			color = gray,
+			style = "",
+			weight = ""]
+	}
+	subgraph interfaces {
+		graph [rank = same]
+		SCI;
+		sock;
+		FS;
+		proc;
+		char;
+		usr_ [
+			fixedsize = true,
+			height = 0,
+			shape = point
+				style = invis,
+			width = 0.5]
+		usr [
+			fillcolor = "#d9e7eeff",
+			fixedsize = true,
+			label = "user space\ninterfaces",
+			row = usr,
+			shape = ellipse,
+			style = "filled,setlinewidth(0)"]
+		MA;
+	}
+	{
+		edge [style = invis weight = 10 ]
+		system_;
+		SCI;
+		system_ -> SCI;
+		networking_;
+		sock;
+		networking_ -> sock;
+		storage_;
+		FS;
+		storage_ -> FS;
+		processing_;
+		proc;
+		processing_ -> proc;
+		HI_;
+		char;
+		HI_ -> char;
+		MA;
+		memory_ [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		memory_ -> MA;
+	}
+	subgraph virtual {
+		graph [rank = same]
+		sysfs;
+		prot_fam;
+		VFS;
+		Tasks;
+		input;
+		D0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		virt [
+			fillcolor = "#b2d3e4",
+			label = "virtual\nsubsystems",
+			URL = "https://en.wikipedia.org/wiki/Proxy_pattern",
+			tooltip = "proxy between standard user space interfaces and internal implementations",
+			style = "filled,setlinewidth(0)"]
+		VM [
+			color = "#51bf5b",
+			fillcolor = "#b2d3e4",
+			label = "Virtual\nmemory"]
+	}
+	subgraph bridges {
+		graph [rank = same]
+		bridges [
+			fillcolor = "#91b5c9",
+			shape = octagon,
+			tooltip = "bridges between uniform virtual interfaces and various implementations",
+			URL = "https://en.wikipedia.org/wiki/Bridge_pattern",
+			style = "filled,setlinewidth(0)"]
+		DM;
+		NFS;
+		mmap;
+		sync;
+		E0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		//PC
+	}
+	subgraph logical {
+		graph [rank = same]
+		log_sys;
+		log_prot;
+		logFS;
+		sched;
+		F7;
+		F0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		logical [
+			fillcolor = "#6a9ab1",
+			style = "filled,setlinewidth(0)"]
+		log_mem [
+			color = "#51bf5b",
+			fillcolor = "#6a9ab1",
+			label = "logical\nmemory"]
+		//SW
+	}
+	subgraph HWI {
+		graph [rank = same]
+		HWI [
+			fillcolor = "#71809b",
+			label = "hardware\ninterfaces",
+			style = "filled,setlinewidth(0)"]
+		bus_drv;
+		netif;
+		block;
+		//PA;
+		IRQ;
+		HID;
+		G0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+	}
+	subgraph HW {
+		graph [rank = same]
+		HW [
+			fillcolor = "#777777",
+			fontcolor = white,
+			label = "electronics,\nhardware",
+			row = chip,
+			style = "filled,setlinewidth(0)"]
+		buses;
+		net_hw;
+		SD;
+		CPU;
+		display;
+		H0 [
+			fixedsize = true,
+			height = 0,
+			shape = point,
+			style = invis,
+			width = 0]
+		RAM;
+	}
+	bottom [
+		label = "© 2007-2022 Costa Shulyupin http://www.MakeLinux.net/kernel/diagram",
+		URL = "http://www.MakeLinux.net/kernel/diagram",
+		shape = plaintext,
+		style = ""]
+	CPU -> bottom [style = invis]
+	layers [
+		fillcolor = lightgray,
+		tooltip = "Functionalities are divided to common layers. It is approximate division.",
+		height = 0.1,
+		style = "filled,setlinewidth(0)",
+		width = 0.5]
+	functions -> layers [style = invis ]
+	usr -> usr_ [
+		arrowhead = "",
+		color = "#d9e7eeff",
+		minlen = 2]
+	usr -> virt [
+		color = "#d9e7eeff"]
+	virt -> D0 [
+		arrowhead = "",
+		color = "#b2d3e4",
+		minlen = 2]
+	virt -> bridges [
+		color = "#b2d3e4"]
+	bridges -> E0 [
+		arrowhead = "",
+		color = "#91b5c9",
+		minlen = 2,
+		style = "filled,setlinewidth(6)",
+		weight = ""]
+	bridges -> logical [
+		color = "#91b5c9",
+		style = "filled,setlinewidth(6)"]
+	logical -> F0 [
+		arrowhead = "",
+		color = "#6a9ab1",
+		minlen = 2,
+		row = logical,
+		style = "filled,setlinewidth(6)",
+		weight = ""]
+	logical -> HWI [
+		color = "#6a9ab1",
+		row = logical,
+		style = "filled,setlinewidth(6)"]
+	HWI -> G0 [
+		arrowhead = "",
+		color = "#71809b",
+		minlen = 2,
+		row = HWI,
+		style = "filled,setlinewidth(6)",
+		weight = ""]
+	HWI -> HW [
+		color = "#71809b",
+		row = HWI,
+		style = "filled,setlinewidth(6)"]
+	HW -> H0 [
+		arrowhead = "",
+		color = "#777777",
+		minlen = 2,
+		row = chip,
+		style = "filled,setlinewidth(6)",
+		weight = ""]
+	layers -> usr [
+		arrowhead = "",
+		color = gray,
+		style = "filled,setlinewidth(1)"]
+	LKD [
+		fontsize = 40,
+		label = "Linux kernel diagram",
+		shape = plain,
+		style = ""]
+	LKD -> processing [style = invis]
+}
+
+```
 

### 2023-04-30 test.qmd
testing tailwind cdn
landing page update
updates to pdf, probs with manuscript render

@@
+######
+######
+######
+######
+
 ---
 title: "Longitudinal Analysis Manuscript: Working Draft"
+number-sections: true
+number-depth: 3
 subtitle: Longitudinal Analysis using data from the ABCD Study 
@@
 
-#format:
-#  pdf:
-#    geometry:
-#      - top=30mm
-#      - left=20mm
-#   docx: default
-# prefer-html: true
+format: 
+  html:
+    page-layout: article
+    mermaid:
+      theme: forest
+    toc: true
+    toc-depth: 1
+    toc-expand: 2
+    toc-title: Table of Contents
+    code-tools:
+      source: true
+      toggle: false
+      caption: none
+    grid:
+      sidebar-width: 200px
+      margin-width: 0px
+  pdf: 
+    geometry:
+      - top=30mm
+      - left=20mm
+    keep_tex: yes
+    #fontfamily: libertinus
+    #fontsize: 20pt
+    colorlinks: true
+format-links: [html, pdf]
+
+execute:
+  echo: false
 
@@
   rticles::arxiv_article
-
-bibliography: "references.bib" #references.bib
+bibliography: "references.bib"
 biblio-style: unsrt
@@
 
-[section still be developed...]
+\[section still be developed...\]
 
@@
 
-[section still be developed...]
+\[section still be developed...\]
 
-# Part I: Longitudinal Research: Basic Concepts and Considerations
+::: {.callout-note  collapse="true" icon=false}
+## Organization `callout`
+Note that there are five types of callouts, including:
+`xxxxx`, `xxxxxx`, `xxxxxx`.
+:::
+
+# Part I: Longitudinal Research
+
 \label{sec:headings}
+
+## Basic Concepts and Considerations
+
 There are several important concepts to consider when conducting longitudinal analyses in a developmental context.
 These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represent the same phenomenon or different ones.
-Further, in the case of developmentally focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child’s life course (and not just characterize growth during this period but also assesses the durability/chronicity of prior effects/consequences)  but also brings with it greater methodological opportunities to exploit the statistical properties of longitudinal data in the furtherance of critical scientific questions.
-That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows, assuming we have assessed the “right” variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest.
+Further, in the case of developmentally focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child's life course (and not just characterize growth during this period but also assesses the durability/chronicity of prior effects/consequences) but also brings with it greater methodological opportunities to exploit the statistical properties of longitudinal data in the furtherance of critical scientific questions.
+That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows, assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest.
 Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.
@@
 
-# Part II Longitudinal Data: Interpretation / Issues / Pitfalls & Assumption
+# Part II: Longitudinal Data
+
 \label{sec:headings}
-**Defining Features of Longitudinal Data Analysis.**
-The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
+
+## Interpretation / Issues / Pitfalls & Assumption
+
+![LDA Structural Diagram](./figures/LDA Structure 2.png){#fig-LDAstruct  fig-alt="LDA Structure Diagram." width=600 height=400 fig-align="center"}
+
+**Defining Features of Longitudinal Data Analysis.** The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
 While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed.
-That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the “right” variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .
+That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .
 
-## Modeling Data Across Two Time Points versus Three or More Time Points.
+"traditional approaches": marginal models
+
+random effects: Growth Curves- SEM (lgcm) vs ME (lmm)
+
+(between- and within-person levels?)
+
+### Modeling Data Across Two Time Points versus Three or More Time Points.
+
 Although the clear leap to the realm of longitudinal data involves going from one assessment to two or more assessments, there are also notable distinctions in designs based on two-assessment points versus three or more measurement occasions.
@@
 
-## Types of stability and change
-If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it’s the characterization of stability and change as the organism traverses the life course.
+### Types of stability and change
+
+If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it's the characterization of stability and change as the organism traverses the life course.
 There are a few different ways to think of stability (and change).
 Consider we measure the height of all youth in a 6th-grade class, once in the fall at the beginning of the school year and once again in the spring at the end of the school year.
-A common first step may be to compare the class’s average height values obtained at these two different measurement occasions.
-This comparison of the average scores for the same group of individuals at multiple time points is referred to as “mean-level” stability as it provides information about continuity and change in the group level of an outcome of interest (e.g., height) over time.
-Another type of stability involves calculating the correlation between the values obtained at different time points (e.g., ‘height in the fall’ with ‘height in the spring’).
-This type of “rank-order” stability evaluates between-individual change by focusing on the degree to which individuals retain their relative placement in a group across time.
+A common first step may be to compare the class's average height values obtained at these two different measurement occasions.
+This comparison of the average scores for the same group of individuals at multiple time points is referred to as "mean-level" stability as it provides information about continuity and change in the group level of an outcome of interest (e.g., height) over time.
+Another type of stability involves calculating the correlation between the values obtained at different time points (e.g., 'height in the fall' with 'height in the spring').
+This type of "rank-order" stability evaluates between-individual change by focusing on the degree to which individuals retain their relative placement in a group across time.
 Consider, someone who is the shortest person in their class in 6th grade may grow considerably over the school year (i.e., exhibit mean level change), but remain the shortest person among their classmates.
@@
 Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021].
-Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual’s mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a].
+Most of these contemporary alternatives incorporate time-specific latent variables to capture between-person sources of variance and model within-person deviations around an individual's mean (or trait) level across time [e.g., RI-CLPM, @hamaker2015; LCM-SR, @curran2014a].
 It is important to note however that these models require multiple assessments waves (e.g., four or more to fully specify the LCM-SR), additional expertise to overcome issues with model convergence, and appreciation of modeling assumptions when attempting to adjudicate among potential models in each research context [see @littlefield2021, for further discussion].
 
-## Model Assumptions
+This is illustrated well by @fig-LDAstruct.
+
+### Model Assumptions
+
 Many statistical models assume certain characteristics about the data to which they are being applied.
 As an example, common assumptions of parametric statistical models include normality, linearity, and equality of variances.
-These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model’s assumptions can substantively alter the interpretation of results.
+These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model's assumptions can substantively alter the interpretation of results.
 Similarly, statistical models employed in the analyses of longitudinal data often entail a range of assumptions that must be closely inspected.
@@
 
-## Covariance Structures
+### Covariance Structures
+
 Statistical models for longitudinal data include two main components to account for assumptions that are commonly violated when working with repeated measures data: a model for the covariance among repeated measures (both the correlations among pairs of repeated measures on an individual and the variability of the responses on different occasions), coupled with a model for the mean response and its dependence on covariates (eg, treatment group in the context of clinical trials).
@@
 
-## Accounting for Correlated Data
+### Accounting for Correlated Data
+
 As an example, one alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design.
@@
 
-## Linear vs non-linear models
+### Linear vs non-linear models
+
 Identification of optimal statistical models and appropriate mathematical functions requires an understanding of the type of data being used.
 Repeated assessments can be based on either continuous or discrete measures.
-Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., a single item measuring the level of agreement to a statement on a three-point scale including the categories of “disagree”, “neutral”, and “agree” in an ecological momentary assessment study that involves multiple daily assessments), and count variables (e.g., number of cigarettes smoked per day across a daily diary study).
+Examples of discrete measures include repeated assessments of binary variables (e.g., past 12-month alcohol use disorder status measured across ten years), ordinal variables (e.g., a single item measuring the level of agreement to a statement on a three-point scale including the categories of "disagree", "neutral", and "agree" in an ecological momentary assessment study that involves multiple daily assessments), and count variables (e.g., number of cigarettes smoked per day across a daily diary study).
 In many ways, the distributional assumptions of indicators used in longitudinal designs mirror the decision points and considerations when delineating across different types of discrete outcome variables, a topic that spans entire textbooks [e.g., see @lenz2016].
@@
 
-## Missing Data/Attrition
+### Marginal vs Conditional Models (?)
+
+\[section still under development...\]
+
+### Missing Data/Attrition
+
 As recently reviewed by Littlefield (in press), investigators of prospective data are confronted with study attrition (i.e., participants may not provide data at a given wave of assessment) and thus approaches are needed to confront the issue of missing data.
@@
 
-## Quantifying effect sizes longitudinally
+### Quantifying effect sizes longitudinally
+
 Given longitudinal data involve different sources of variance, quantifying effect sizes longitudinally is a more difficult task compared to deriving such estimates from cross-sectional data.
-Effect size can be defined as, “a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study.” [@kraemer2014].
-Common effect size metrics include r (i.e., the standardized covariance, or correlation, between two variables) and Cohen’s d [@cohen1988].
-Adjustments to common effect size calculations, such as Cohen’s d, are required even when only two time points are considered [e.g., see @morris2002].
+Effect size can be defined as, "a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study." [@kraemer2014].
+Common effect size metrics include r (i.e., the standardized covariance, or correlation, between two variables) and Cohen's d [@cohen1988].
+Adjustments to common effect size calculations, such as Cohen's d, are required even when only two time points are considered [e.g., see @morris2002].
 @wang2019a note there are multiple approaches to obtaining standardized within-person effects, and that commonly suggested approaches (e.g., global standardization) can be problematic [see @wang2019a, for more details].
@@
 
-# Part III Section in Progress: TBD
+[psychometrics as a new "Part" below, or facets of psychometrics (e.g., invariance, reliability, etc...) inserted above?]
+
+# Part III: Additional Section(s)
+
+\label{sec:headings} \## In Progress: \[TBD\]
+
+Section focused on different ways of considering/grouping/deciding on which specific analysis to choose (see curran&bauer slides)
+
+### Subsection TBD.
+
+![Longitudinal Models](./figures/Mindmap1.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width=600 height=400 fig-align="center"}
+
+# Part IV: Discussion
+
 \label{sec:headings}
-**Subsection TBD.**
 
-# Part IV Discussion
-\label{sec:headings}
-**Subsection TBD.**
+### Subsection TBD.
+
+|                                           |                         |                          |                                      |                         |
+|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
+| Models                                    |     Form of Change      | Minimum \# of Timepoints | Variable-Centered vs Person-Centered | Time-varying Covariates |
+| Difference Scores                         |       mean-level        |            2+            |                  \-                  |          none           |
+| Change Scores                             |       rank-order        |            2+            |                  \-                  |          none           |
+| Signed-Rank Test                          |       rank-order        |            2+            |                  \-                  |          none           |
+| Generalized Estimating Equations          |       rank-order        |            2+            |                  \-                  |         allowed         |
+| Linear Mixed Models                       |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Generalized Linear Mixed-Effects Models   |   within and between    |            3+            |                                      |         allowed         |
+| Autoregressive Crosslagged Panel Models   |       rank-order        |            3+            |          Variable-Centered           |         allowed         |
+| Latent Change Scores Models               |       rank-order        |            2+            |          Variable-Centered           |         allowed         |
+| Latent Growth Curve Models                |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Multivariate Latent Growth Curve Model    |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Latent Transition Analysis                |   within and between    |            3+            |                Mixed                 |         allowed         |
+| Latent Class Growth Analysis              |   within and between    |            3+            |           Person-Centered            |         allowed         |
+| Growth Mixture Models                     |   within and between    |            3+            |                Mixed                 |         allowed         |
+| State-Trait Models                        | continuous and discrete |            3+            |          Variable-Centered           |         allowed         |
+| Random-Intercept Crosslagged Panel Models |   within and between    |            4+            |          Variable-Centered           |         allowed         |
+| Latent Curve Models Structured Residuals  |   within and between    |            4+            |          Variable-Centered           |         allowed         |
+|                                           |                         |                          |                                      |                         |
+
+: LDA Models {#tbl-lda .striped .hover .responsive-xl}
+
+axasx
+
+See @tbl-lda.
+
+xxxxx
+
+# Part V: References

### 2023-05-05 _temp.qmd
updated section III of manuscript

--- 2023-04-30 temp.qmd
@@
 
-<script src="https://cdn.tailwindcss.com"></script>
+---
+\[psychometrics as a new "Part" below, or facets of psychometrics (e.g., invariance, reliability, etc...) inserted above? MEASUREMENT CONSIDERATIONS IN LONGITUDINAL RESEARCH?\]
+[[
+- SectionIIIa on Data Structures and Data Wrangling(?)
+- SectionIIIb on Psychometrics(?)
+- SectionIIIc on Methods/Analyses(?)
+]]
+
+Measurement?
+Data Wrangling?
+
+With the large and continually expanding body of research on statistical methods for longitudinal analyses, determining which longitudinal model to implement can be challenging.
+The aim of this section is to help researchers to navigate these many options in order to identify the statistical approach most appropriate to their unique research question.
+Notably, there are a myriad of viable ways one can go about grouping various types of longitudinal models for presentation.
+Common examples iclude grouping by linearity (linear vs nonlinear models; see Collins, 2006(?)), the number of measurement occassions (see King et al., 2018), and xxxxxx (see xxxx, xxx).
+The ordering/organzation we use below overlaps in a number of ways with these examples, and with Curran & Bauer (2019); however, it is important to note that in each case, the ordering/organization that is chosen is somewhat arbitrary and primarly intended to allow the reader to compare and contrast of various analytic approaches.
+In the following we will describe (briefly) and summarize the advantages/disadvantages for a variety longitudinal models, organized into the following x# `fuzzy` groupings: Traditional Models, Modern GLM Extensions, SEM, Advanced SEM, xxx, xxxx.
+We note that this is note an exhaustive review of each of these methods, but for more in-depth detail we do provide the reader relevant resources.
+As aptly summarized by Curran & Bauer (2019), "there are many exceptions, alternatives, nuances, “what ifs”, and “but couldn’t yous” that aren’t addressed here."
+
+##### Traditional Models
+
+Traditional methods for longitudinal analysis primarily focus on modeling mean-level change. 
+For example, is there a difference in average math scores obtained across multiple assessments between boys and girls?
+Longitudinal models that focus on mean-level change are also referred to as `marginal models` and examples of specific methods include `repeated measures ANOVA` (and `MANOVA`), `ANCOVA`, and `Generalized Estimating Equations` (GEE).
+These methods are commonly used when data is only available from 2 measurement occassions.
+For example, computing a `difference score` (e.g., math scores at t2 - math scores at t1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups.
+We also note the longitudinal `signed-rank test` which is a nonparametric alternative to the paired t-test that can be a useful tool for analyzing non-normal paired data.
+Another common approach, often used in pre-/post-design studies, is to use `residualized change score analysis` to assess the degree of change in a variable, while controlling for its initial level.
+For example, regressing post-treatment scores on pre-treatment scores and then using the resulting residuals as a measure of change that is adjusted for baseline scores (ignoring any prior group assignments/differences).
+Similar to difference scores, the residualized change score is often included in subsequent analysis, such as for evaluating intervention effects in pre-test/post-test designs [@Kisbu-Sakarya2013].
+
+Traditional longitudinal methods can still be useful in some contexts (e.g., < 3 measurement occasions), but overall their practical utility for answering questions about developmental processes is limited.
+Perhaps most notably, these models do not allow for characterizing patterns of within-person change, which is often an emphasis of longitudinal research.
+Moreover, even determining which of these procedures to use for comparing change over two time points across groups can be surprisingly complicated.
+A particularly vexing example is that of imbalanced baseline scores which can produce biased effects when using these methods.
+Indeed, it is well-documented that this can result in different conclusions between these approaches, which has led to many articles debating the merits of which procedures is most appropriate to use and when [see xxxxx for a review].
+Beyond these concerns, additional limitations include xxxxxx.
+Given these shortcomings and the complexity of the issues surrounding some of these methods, when possible, it is typically recommended that researchers make use of more modern approaches for analyzing longitudinal data (and preferably make use of data collected across 3 or more time points.
+
+xxxxxxxxx
+Many advantages of traditional approaches 
+ straightforward to estimate as standard path models 
+ obtain overall tests of fit and tests direct and indirect effects 
+ corresponds to some models of change posited by some theories
+
+ But also many disadvantages 
+ only considers change in a series of two time-point comparisons 
+ assumes all measures have perfect reliability 
+ can be biased if systematic growth exists in repeated measures 
+ poorly corresponds to most theoretical models of change
+
+zzzzzzzzzxxxxxxx
+
+Modern approaches to longitudinal data analysis have advanced beyond traditional methods by offering greater flexibility and a more in-depth understanding of within-person and between-person variability.
+`Generalized Estimating Equations (GEE)`, `Linear Mixed Models (LMM)`, `Generalized Linear Mixed Models (GLMM)`, and `Autoregressive Cross-Lagged Panel Models (ARCL)` are examples of such contemporary techniques.
+GEE is an extension of Generalized Linear Models.
+This model combines the generalized linear model for non-normal outcomes with repeated measures (marginal) model and is suitable for analyzing correlated longitudinal data and modeling population-averaged effects.
+LMMs, also known as multilevel or hierarchical linear models, facilitate the simultaneous analysis of within-person and between-person variability, making them ideal for nested data structures or repeated measures.
+GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data.
+Finally, ARCL models are uses to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects.
+
+The strengths of these modern methods lie in their ability to account for individual differences, within-person change, and time-varying predictors, thereby providing a more comprehensive understanding of complex relationships in longitudinal data.
+Despite these advantages, modern approaches may require more complex modeling assumptions and higher computational demands compared to traditional methods.
+Additionally, proper model specification and the interpretation of results can be more challenging, especially in cases of high multicollinearity or missing data.
+However, modern longitudinal analysis methods have generally surpassed traditional methods in addressing a wider range of research questions, accommodating diverse data structures, and elucidating the intricate dynamics of developmental processes.
+
+Overview—A GEE model is designed for analyzing the regression relationship between covariates and repeated responses, but not the correlation structure of the repeated responses.
+If the latter is of interest, a GEE is inappropriate and one should consider MER (“Mixed effects regression (MER)” section).
+In estimating the regression parameters, the correlation structure in a GEE is represented using a working, potentially incorrect model (see “Modeling correlation” section).
+Even when the working model is incorrect, however, the GEE approach yields unbiased parameter estimates.
+Traditionally, GEEs are intended for two-level hierarchical data (e.g., cases 1 and 2), but recent work [15] has allowed extensions to three levels (e.g., case 3).
+
+Generalized estimating equations (GEEs) are a statistical method used to analyze longitudinal or clustered data, while accounting for within-subject correlation.
+The goal GEEs are to make inferences about population-averaged effects (controlling for within-subject correlations), rather than individual subject-level effects.
+GEEs can handle a wide range of outcome distributions, including binary, count, and continuous data (Fitzmaurice et al., 2011; Hardin & Hilbe, 2012).
+
+GLMM.
+Unlike the WGEE, the GLMM accommodates correlated responses Yit by directly modeling their joint distribution.
+Since multivariate distributions are extremely complex except for the multivariate normal, latent variables bi are generally employed to model the correlated responses.
+Thus, although Yit is still modeled for each time point t, by including the random effect b, in the specification of the conditional distribution of Yit given bi (Xit and Zit), the GLMM in (9) allows the resulting Yit ‘s to be correlated (conditional on Xit and Zit only).
+This approach allows one to specify multivariate distributions using familiar univariate distributions such as the Bernoulli (for binary responses) and the Poisson (for count responses).
+
+zzzzzzzzzxxxxxxx
+##### Modern GLM Extensions
+**Generalized Linear Mixed Models** 
+**Linear Mixed Models**
+**Generalized Linear Mixed Models**
+**ARCL**
+
+`Generalized Linear Mixed Models`, `Linear Mixed Models`, `Generalized Linear Mixed Models`, `Autoregressive Crosslagged Panel Models`
+
+Two preferred methods for longitudinal data are generalized estimating equations model (GEE) [13] and mixed effects regression (MER) [14].
+Both allow time-invariant predictors that never change (e.g., gender, genotype) and time-varying predictors (e.g., age), and handle irregularly timed and missing data without the need for explicit imputation.
+
+GEE particularly popular in longitudinal health research because it
+provides robust inferences without making strong assumptions
+about individual differences
+ Particularly useful when working with discrete outcomes
+
+ Those that emphasize individual change often do so through
+inclusion of Random Effects
+ often called “growth models”
+
+ and individual-level (within person) change…
+ What do individual trajectories of vocabulary production look like?
+ How great are the individual differences in these trajectories?
+
+include bullet points to summarize?
+
+##### SEM
+Variable-centered. 
+**Latent Change Score** (add internal link to documentation)
+**Latent Growth Curve Model** (add internal link to documentation)
+**Multivariate Latent Growth Curve Model** (add internal link to documentation) 
+**State-Trait Model** (add internal link to documentation)
+
+ Those that emphasize individual change often do so through
+inclusion of Random Effects
+ often called “growth models”
+
+Within-person change Panel (3+waves)
+
+Within-person change Panel (3+waves) Quantitative Ind. Differences:
+mixed effects, latent curve
+ind. Differences:
+Latent class growth analysis,
+Semiparametric groups-based approach
+General growth mixture model
+
+Person-centered.
+**Latent Transition Analysis** (add internal link to documentation)
+**Latent Class Growth Analysis** (add internal link to documentation)
+
+Progression through
+stages/sequence
+Panel data Latent transition analysis
+
+Hybrid.
+**Growth Mixture Modeling** (add internal link to documentation)
+
+##### Advanced SEM
+**RI-CLPM** (add internal link to documentation)
+**LCM-SR** (add internal link to documentation)
+
+Examine
+relationships
+between timeadjacent changes
+
+Change + bidirectional Panel data Latent curve model with structured residuals
+Timing of predictor effects Intensive longitudinal
+Long-term panel
+Time-varying effects models
+
+##### (possibly others xxxx?)
+(longitudinal neuroimaging?)
+(longitudinal twins/genetics?)
+
+![Longitudinal Models/Data Structures](./figures/Mindmap1.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="600" height="400" fig-align="center"}
+
+### Brief section to wrap up (and highlight the availiblity of the online examples)
+
+xxxxxxxxxxxxxxxxxxxxx
+xxxxxxxxxxxxxxxxxxxxx
+xxxxxxxxxxxxxxxxxxxxx
+xxxxxxxxxxxxxxxxxxxxx
+
+Data structures are the building blocks for more sophisticated applications.
+They are designed by composing data elements into a logical unit representing an abstract data type that has relevance to the algorithm or application.
+An example of an abstract data type is a "customer name" that is composed of the character strings for "first name," "middle name" and "last name."
+
+It is not only important to use data structures, but it is also important to choose the proper data structure for each task.
+Choosing an ill-suited data structure could result in slow runtimes or unresponsive code.
+Five factors to consider when picking a data structure include the following:
+
+What kind of information will be stored?
+How will that information be used?
+Where should data persist, or be kept, after it is created?
+What is the best way to organize the data?
+What aspects of memory and storage reservation management should be considered?
+
+the choice of several types of longitudinal designs that researchers employ to investigate change over time. 
+Clarify the analysis approaches that are associated with each type of data structure and the questions they can answer
+
+Section focused on different ways of considering/grouping/deciding on which specific analysis to choose (see curran&bauer slides)
+
+data structure is not only used for organizing the data. It is also used for processing, retrieving, and storing data
+The data structure type used in a particular situation is determined by the type of operations that will be required or the kinds of algorithms that will be applied.
+
+types of data structures, all designed to arrange data to suit a specific purpose.
+Data structures make it easy for users to access and work with the data they need in appropriate ways.
+Most importantly, data structures frame the organization of information so that machines and humans can better understand it.
+
+applications that ingest, manipulate and produce information must understand how data should be organized to simplify processing.
+Data structures bring together the data elements in a logical way and facilitate the effective use, persistence and sharing of data.
+They provide a formal model that describes the way the data elements are organized.
+
+xxxxxxx
+This article provides an overview of the imaging procedures of the ABCD study, the basis for their selection and preliminary quality assurance and results that provide evidence for the feasibility and age-appropriateness of procedures and generalizability of findings to the existent literature.
+
+Retention is critical in longitudinal, multi-site studies of adolescents.
+
+•
+Differential retention can impact generalizability of study outcomes.
+
+•
+We review efforts utilized by the ABCD study staff to ensure high retention.
+
+Learning What To Do: Survey of Techniques
+
+Objectives  Distinguish between different longitudinal data structures 
+ Clarify the analysis approaches that are associated with each type of data structure and the questions they can answer
+
+Panel Data  Most often, what we mean by longitudinal data is panel data  Panel data is data collected over a relatively small number of time points on a relatively large number of units  Examples:  Biennial assessments of alcohol and substance use in adolescents and young adults from ages 14 to 30.
+ Monthly assessments of vocabulary production by infants/toddlers from 8 to 24 months of age  Other similar data structures arise from  Randomized clinical trials  Accelerated longitudinal designs
+
+Analysis Goals  A common goal when collecting panel data is to evaluate change over time  One can distinguish between mean-level change...
+ On average, how much more quickly does the vocabulary production of girls increase relative to the vocabulary production of boys?
+ and individual-level (within person) change...
+ What do individual trajectories of vocabulary production look like?
+ How great are the individual differences in these trajectories?
+
+traditional and contemporary approaches to modeling developmental change
+
+Theoretical Model of Change
+Temporal Design Used to Observe the Change Phenomenon
+Statistical Model of Change
+
+xxxxxxx
+
+Evaluate the measurement properties of the variable for invariance (e.g., configural, metric) before testing whether change has occurred.
+Analytical Issues 10.
+Be aware of potential violations in statistical assumptions inherent in longitudinal designs (e.g., correlated residuals, nonindependence).
+11.
+Describe how time is coded (e.g., polynomials, orthogonal polynomials) and why.
 
@@
 </div>
+
+|                                           |                         |                          |                                      |                         |
+|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
+| Models                                    |     Form of Change      | Minimum \# of Timepoints | Variable-Centered vs Person-Centered | Time-varying Covariates |
+| Difference Scores                         |       mean-level        |            2+            |                  \-                  |          none           |
+| Change Scores                             |       rank-order        |            2+            |                  \-                  |          none           |
+| Signed-Rank Test                          |       rank-order        |            2+            |                  \-                  |          none           |
+| Generalized Estimating Equations          |       rank-order        |            2+            |                  \-                  |         allowed         |
+| Linear Mixed Models                       |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Generalized Linear Mixed-Effects Models   |   within and between    |            3+            |                                      |         allowed         |
+| Autoregressive Crosslagged Panel Models   |       rank-order        |            3+            |          Variable-Centered           |         allowed         |
+| Latent Change Scores Models               |       rank-order        |            2+            |          Variable-Centered           |         allowed         |
+| Latent Growth Curve Models                |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Multivariate Latent Growth Curve Model    |   within and between    |            3+            |          Variable-Centered           |         allowed         |
+| Latent Transition Analysis                |   within and between    |            3+            |                Mixed                 |         allowed         |
+| Latent Class Growth Analysis              |   within and between    |            3+            |           Person-Centered            |         allowed         |
+| Growth Mixture Models                     |   within and between    |            3+            |                Mixed                 |         allowed         |
+| State-Trait Models                        | continuous and discrete |            3+            |          Variable-Centered           |         allowed         |
+| Random-Intercept Crosslagged Panel Models |   within and between    |            4+            |          Variable-Centered           |         allowed         |
+| Latent Curve Models Structured Residuals  |   within and between    |            4+            |          Variable-Centered           |         allowed         |
+|                                           |                         |                          |                                      |                         |
+
+: LDA Models {#tbl-lda .striped .hover .responsive-xl}
+
+axasx
+
+See @tbl-lda.

### 2023-05-05 manuscript.qmd
updated section III of manuscript

--- 2023-04-28 manuscript.qmd
@@
 ---
-title: "Longitudinal Analysis Manuscript: Working Draft"
+title: |
+  Longitudinal Analysis Manuscript: Working Draft
+subtitle: Longitudinal Analysis using data from the ABCD Study
+short-title: Longitudinal Analysis
+published: "Under review at *xxxxx Journal*"
+code-repo: "Access the code, data, and analysis at <http://swhawes.github.io/test/>"
 number-sections: true
 number-depth: 3
-subtitle: Longitudinal Analysis using data from the ABCD Study 
-authors:
-  - name: Samuel W. Hawes
+correspondence-prefix: "Correspondence concerning this article should be addressed to"
+author:
+  - name: Samuel Hawes
+    email: shawes@fiu.edu
+    title: Associate Research Professor
+    affiliations: 
+      #- id: ds9
+        name: Florida International University
+        department: Center for Children & Families
+        address: 11200 SW 8th St, AHC-IV
+        city: Miami
+        region: FL
+        country: USA
+        postal-code: 33199
+    attributes:
+      corresponding: true
+  - name: Andrew K. Littlefield
+    email: andrew.littlefield@ttu.edu
+    affiliations: 
+      #- id: ds9
+        name: Texas Tech University
+        department: Psychological Sciences
+        city: Lubbock
+        region: TX
+        country: USA
+  - name: Daniel A. Lopez
+    email: Daniel_Lopez2@URMC.Rochester.edu
+    affiliations:
+      #- id: terok
+        name: University of Rochester Medical Center
+        department: Department of Public Health
+        city: Rochester
+        region: NY
+        country: USA
+      #- ref: ds9
+  - name: Kenneth J. Sher
+    email: SherK@missouri.edu
+    affiliations: 
+      #- id: ds9
+        name: University of Missouri
+        department: Psychological Sciences
+        city: Columbia
+        region: MO
+        country: USA
+  - name: Erin L. Thompson
+    email: erthomps@fiu.edu
+    affiliations:
+      #- id: terok
+        name: Florida International University
         department: Center for Children and Families
-    affiliation: Florida International University
-    location: Miami, FL
-    email: shawes@fiu.edu
-  - name: Andrew K. Littlefield
-    department: Psychological Sciences
-    affiliation: Texas Tech University
-    location: Lubbock, TX
-    email: andrew.littlefield@ttu.edu
-  - name: Daniel A. Lopez
-    department: Department of Public Health
-    affiliation: University of Rochester Medical Center
-    location: Rochester, NY
-    email: Daniel_Lopez2@URMC.Rochester.edu
-  - name: Kenneth J. Sher
-    department: Psychological Sciences
-    affiliation: University of Missouri
-    location: Columbia, MO
-    email: SherK@missouri.edu
-  - name: Erin L. Thompson
+        city: Miami
+        region: FL
+        country: USA
+  - name: Raul Gonzalez
+    email: gonzara@fiu.edu
+    affiliations: 
+      #- id: ds9
+        name: Florida International University
         department: Center for Children and Families
-    affiliation: Florida International University
-    location: Miami, FL
-    email: erthomps@fiu.edu
-  - name: Raul Gonzalez
-    department: Center for Children and Families
-    affiliation: Florida International University
-    location: Miami, FL
-    email: gonzara@fiu.edu  
+        city: Miami
+        region: FL
+        country: USA
   - name: Additional Co-authors
-    department: xxxxxx
-    affiliation: University of xxxx
-    location: xxxxxxx
-    email: xxxxxx
+    email: xxxxx@fiu.edu
+    affiliations:
+      #- id: terok
+        name: xxxxxx University
+        department: xxxxxx Dept
+        city: xxxx
+        region: xxxxx
+        country: xxxxx
   - name: Wesley K. Thompson
+    email: wes.stat@gmail.com
+    affiliations:
+      #- id: terok
+        name: Laureate Institute for Brain Research
         department: Center for Population Neuroscience and Genetics
-    affiliation: Laureate Institute for Brain Research
-    location: Tulsa, OK
-    email: wes.stat@gmail.com
+        city: Tulsa
+        region: OK
+        country: USA
 
@@
 By addressing these complexities, the review seeks to equip researchers with the necessary knowledge and tools to make informed decisions as they navigate effectively analyzing and interpreting data available in the ABCD Study.
-
-author-summary: |
-  The Adolescent Brain Cognitive Development (ABCD) Study offers valuable insights into adolescent brain development.
-Researchers working with this data face methodological and analytic concerns, including sample size, attrition, missing data, and measurement.
-This review addresses complexities in analyzing longitudinal data, with a focus on statistical models accounting for correlations in repeated measurements and their assumptions.
-We emphasize the importance of choosing appropriate models and analytic functions based on the data and the specific scientific question to be addressed.
-This concise resource aims  to help researchers navigate longitudinal data analysis, make informed methodological decisions, and enhance the accuracy and reliability of their findings when working with data from the ABCD Study.
-
+# thanks: |
+#   Placeholder text generated at <https://vlad-saling.github.io/star-trek-ipsum/>.
+# additional-info: |
+#   We have no known conflict of interest to disclose.
 keywords:
-    *ABCD Study*,
-    *longitudinal analyses*,
-    *development*
-
+  - ABCD Study
+  - longitudinal analyses
+  - development
 format:
-  html:
-    toc: true
-    toc-depth: 1
-    toc-expand: 2
-    toc-title: Table of Contents
-    grid:
-      sidebar-width: 300px
-      margin-width: 300px
+  #hikmah-pdf: default
+  hikmah-manuscript-pdf: default
+  #html:
+    #page-layout: article
+    #toc: true
+    #toc-depth: 1
+    #toc-expand: 2
+    #toc-title: Table of Contents
+  #   code-tools:
+  #     source: true
+  #     toggle: false
+  #     caption: none
+  #   grid:
+  #     sidebar-width: 200px
+  #     margin-width: 0px
   pdf:
@@
       - left=20mm
+  #   keep_tex: yes
     #fontfamily: libertinus
     #fontsize: 20pt
-format-links: [html, pdf]
+#colorlinks: true
+#format-links: [html, pdf]
 
-  #format:
-#  pdf:
-#    geometry:
-#      - top=30mm
-#      - left=20mm
-#   docx: default
-# prefer-html: true
+execute:
+ echo: false
 
-output: 
-  rticles::arxiv_article
-
-bibliography: "references.bib" #references.bib
-biblio-style: unsrt
+# output: 
+#   rticles::arxiv_article
+bibliography: "references.bib"
+#biblio-style: unsrt
 ---
@@
 
-\label{sec:headings} The Adolescent Brain Cognitive Development (ABCD) Study® is the largest long-term investigation of neurodevelopment and child health in the United States.
+\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States.
 Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018].
@@
 Together, this presents a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes emerge and unfold across critical periods of development.
-
-\[section still be developed...\]
+In the following text, we describe methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents.
+We begin with a brief orientation to the ABCD Study and its accompanying [publicly available dataset](https://nda.nih.gov/abcd/ "Visit the ABCD Study data repository"){#ndardata}.
 
@@
 
-## Organization of current manuscript
+##
 
-The rich longitudinal nature of the ABCD Study dataset will allow researchers to perform analyses of both methodological and substantive interest.
-This report describes methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents.
-The manuscript is organized as follows:
+::: {.callout-note collapse="true" icon="false"}
 
-\[section still be developed...\]
+This manuscript will go through the following sections :xxxxxx, xxxx, xxxxx.
+# Part I: Longitudinal Research
+# Part II: Longitudinal Data
+# Part III: Longitudinal Analysis
+Overview of **concepts and considerations specific to developmentally-focused longitudinal research**
+
+## Organization and Aims
+
+Aims/Goals: Use this xxxxx during any xxxxx and you're virtually guaranteed to xxxxx
+
+-   Concepts and Considerations when working with large and multimodal longitudinal data
+
+-   Concepts and considerations of developmental research (longitudinal research via developmental perspective)
+
+-   Learn specific methods
+
+-   Use our open-source resources to get started more quickly
+
+-   Learn How to Apply xxxxx to large longitudinal data such as the ABCD Study
+
+identify common xxxxx of longitianl research
+
+1.3 Objectives and scope
+The focus of this xxxx is on the analysis of event history data using
+xxxx model with the objectives
+• to demonstrate the use of event history analysis in the analysis
+of xxxxx, xxxx and to identify the methodological problems in the analysis,
+• to propose procedures to choose a basic time scale,
+• to discuss the connections between the methods for modeling
+dual time scales and to perform quantitative comparisons between them,
+• to compare existing methods to deal with longitudinal measurements in the xxxx model with two proposed methods.
+
+:::
 
@@
 
-### \[may add section here about the "structure" of longitudinal data we're mostly focused on-- repeated measures of panel data...\]
-
-"traditional approaches": marginal models
-
-random effects: Growth Curves- SEM (lgcm) vs ME (lmm)
-
-(between- and within-person levels?)
+["traditional approaches": marginal models]
+[random effects: Growth Curves- SEM (lgcm) vs ME (lmm)]
+[(between- and within-person levels?)]
 
@@
 
-# Part III: Additional Section(s)
+# Part III: Longitudinal Analysis
 
-\label{sec:headings} \## In Progress: \[TBD\]
+\label{sec:headings}
 
-Section focused on different ways of considering/grouping/deciding on which specific analysis to choose (see curran&bauer slides)
+### Longitudinal Data Structures
 
-### Subsection TBD.
+An ideal longitudinal study integrates (a) a well-articulated theoretical model, (b) an appropriate longitudinal data structure, and (c) a statistical model that is an operationalization of the theoretical model [@collins2006].
+To accommodate various research questions and contexts, different types of longitudinal data and data structures have emerged (see Figure x).
+An understanding of these data structures is helpful, as they can warrant different types of longtiudinal data analysis.
+Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and Bauer (xxxx) provide a nice on-ramp in distinguishing between the use of "time-to-event" and "repeated measures" data.
+Although both model time, the former is concerned with *whether* and *when* an event occurs, whereas the later is focused on *growth* and *change* \[Curran and Bauer, xxxx\].
+Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest.
+This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models) and the time-to-event data can be based on a single assessment or iclude multiple recurrent or competing events.
+While much has been written about about "time-to-event" data (see xxx; xxx; xxxx), including several recent xxxx (see), our emphasis will be given to the modeling of "repeated measures" data.
 
-![test](./figures/Mindmap1.png)
+[![LDA Structural Diagram](./figures/LDAStructure2.png)](https://centerstat.org/resources/)
+
+When discussing longitudinal analysis, we are most often talking about data collected on the same unit (e.g., individuals) across multiple measurement occasions.
+However, repeated-measures analysis is not a monolith and it will serve us well to distinguish between a few of the most common types.
+One such approach to repeated measures analysis is the use of `time-series models`.
+These models generally consist of a long sequence of repeated measurements (≧ 50-100 measurements) on a single (or small number of) some variable of interest.
+Time-series analysis is often used to predict/forecast temporal trends and cyclic patterns and is geared toward making inferences about prospective outcomes within a population (with relatively with less focus on inferring individual-level mechanisms and risk factors).
+A related type of repeated measures analysis is `Intensive Longitudinal Data (ILD)`.
+Similar to time-series analysis, ILD models involve frequent measurements (\~ 30-40 measurements) of the same individuals in a relatively circumspect period of time (e.g., experience sampling to obtain time series on many individuals).
+Although ILD models may include slightly fewer measurement occasions than time-series data, ILD models tend to have more subjects than time series models (\~ 50-100 subjects).
+This allows ILD models to examine short-term patterns by incorporating a time series model that can sometimes fit parameter estimates to each individual's data in order to model individual difference outcomes.
+The final type of repeated measures analaysis that we will discuss is the longitudinal panel model.
+These models follow a group of individuals--- a panel (also referred to as a cohort) --- across relatively fewer measurement occassions (~ 5-15), and are often interested in examining change across both, within-individuals and between-individuals.
+
+While other longitudinal designs have their own unique strengths and applications, the longitudinal panel design is particularly well-suited for investigating developmental processes in the context of the ABCD Study.
+In the following sections, we will discuss various analytic methods commonly used to analyze longitudinal panel data, including growth models, mixed models, and a number of additional trajectory models.
+These methods provide valuable insights into within- and between-individual differences and are highly relevant for researchers working with the ABCD Study dataset.
+By focusing on these methods, we aim to equip readers with the knowledge necessary to conduct longitudinal research and perform analyses using the rich, longitudinal, and publicly available data from the ABCD Study.
+
+The following text includes collated information about each of the categories of longitudinal analyses that we have discussed during previous questions and responses that has been written up for publication in a scientific journal:
+
+#### Types of longitudinal panel models
+
+With the large and continually expanding body of research on statistical methods for longitudinal analyses, determining which longitudinal model to implement can be challenging.
+The aim of this section is to help researchers to navigate these many options in order to identify the statistical approach most appropriate to their unique research question.
+Notably, there are a myriad of viable ways one can go about grouping various types of longitudinal models for presentation.
+Common examples iclude grouping by linearity (linear vs nonlinear models; see Collins, 2006(?)), the number of measurement occassions (see King et al., 2018), and xxxxxx (see xxxx, xxx).
+The ordering/organzation we use below overlaps in a number of ways with these examples, and with Curran & Bauer (2019); however, it is important to note that in each case, the ordering/organization that is chosen is somewhat arbitrary and primarly intended to allow the reader to compare and contrast of various analytic approaches.
+In the following we will describe (briefly) and summarize the advantages/disadvantages for a series of 16(?) distinct longitudinal models organized into the following x# `fuzzy` groupings: Traditional Models, Modern GLM Extensions, SEM, Advanced SEM, xxx, xxxx.
+We note that this is note an exhaustive review of each of these methods, but for more in-depth detail we do provide the reader relevant resources.
+As aptly summarized by Curran & Bauer (2019), "there are many exceptions, alternatives, nuances, “what ifs”, and “but couldn’t yous” that aren’t addressed here."
+
+##### Traditional (linear and nonlinear)
+Traditional methods for longitudinal analysis primarily focus on modeling mean-level change. 
+For example, is there a difference in average math scores obtained across multiple assessments between boys and girls?
+Longitudinal models that focus on mean-level change are also referred to as `marginal models` and examples of specific methods include `repeated measures ANOVA` (and `MANOVA`), `ANCOVA`, and `Generalized Estimating Equations` (GEE).
+These methods are commonly used when data is only available from 2 measurement occassions.
+For example, computing a `difference score` (e.g., math scores at t2 - math scores at t1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups.
+We also note the longitudinal `signed-rank test` which is a nonparametric alternative to the paired t-test that can be a useful tool for analyzing non-normal paired data.
+Another common approach, often used in pre-/post-design studies, is to use `residualized change score analysis` to assess the degree of change in a variable, while controlling for its initial level.
+For example, regressing post-treatment scores on pre-treatment scores and then using the resulting residuals as a measure of change that is adjusted for baseline scores (ignoring any prior group assignments/differences).
+Similar to difference scores, the residualized change score is often included in subsequent analysis, such as for evaluating intervention effects in pre-test/post-test designs [@Kisbu-Sakarya2013].
+
+Traditional longitudinal methods can still be useful in some contexts (e.g., < 3 measurement occasions), but overall their practical utility for answering questions about developmental processes is limited.
+Perhaps most notably, these models do not allow for characterizing patterns of within-person change, which is often an emphasis of longitudinal research.
+In turn, these models often correspond poorly with most theoretical models of change.
+
+Moreover, even determining which of these procedures to use for comparing change over two time points across groups can be surprisingly complicated.
+A particularly vexing example is that of imbalanced baseline scores which can produce biased effects when using these methods.
+Indeed, it is well-documented that this can result in different conclusions between these approaches, which has led to many articles debating the merits of which procedures is most appropriate to use and when [see xxxxx for a review].
+Beyond these concerns, additional limitations include xxxxxx.
+Given these shortcomings and the complexity of the issues surrounding some of these methods, when possible, it is typically recommended that researchers make use of more modern approaches for analyzing longitudinal data (and preferably make use of data collected across 3 or more time points.
+
+##### Modern GLM Extensions
+
+Modern approaches to longitudinal data analysis have advanced beyond traditional methods by offering greater flexibility and a more in-depth understanding of within-person and between-person variability.
+`Generalized Estimating Equations (GEE)`, `Linear Mixed Models (LMM)`, `Generalized Linear Mixed Models (GLMM)`, and `Autoregressive Cross-Lagged Panel Models (ARCL)` are examples of such contemporary techniques.
+GEE is an extension of Generalized Linear Models.
+This model combines the generalized linear model for non-normal outcomes with repeated measures (marginal) model and is suitable for analyzing correlated longitudinal data and modeling population-averaged effects.
+LMMs, also known as multilevel or hierarchical linear models, facilitate the simultaneous analysis of within-person and between-person variability, making them ideal for nested data structures or repeated measures.
+GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data.
+Finally, ARCL models are uses to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects.
+
+The strengths of these modern methods lie in their ability to account for individual differences, within-person change, and time-varying predictors, thereby providing a more comprehensive understanding of complex relationships in longitudinal data.
+Despite these advantages, modern approaches may require more complex modeling assumptions and higher computational demands compared to traditional methods.
+Additionally, proper model specification and the interpretation of results can be more challenging, especially in cases of high multicollinearity or missing data.
+However, modern longitudinal analysis methods have generally surpassed traditional methods in addressing a wider range of research questions, accommodating diverse data structures, and elucidating the intricate dynamics of developmental processes.
+
+##### SEM
+
+Structural equation modeling (SEM) approaches have gained prominence in longitudinal data analysis due to their ability to estimate complex relationships among observed and latent variables while accounting for measurement error.
+SEM techniques can be categorized into variable-centered, person-centered, and hybrid approaches, each with unique strengths and limitations.
+The choice of method depends on the research question, data structure, and underlying assumptions.
+
+Variable-centered approaches, such as Latent Change Scores, Latent Growth Curve Models, Multivariate (Parallel Process) Latent Growth Curve Models, and Latent State-Trait Models, focus on examining relationships among variables and population-level patterns.
+These models offer a powerful means to estimate individual change trajectories and the relationships between growth parameters of different variables, while also decomposing observed measurements into latent state and trait components.
+However, these approaches may not adequately capture distinct subgroups of individuals who share similar patterns of change over time, which can be crucial for understanding heterogeneous developmental processes.
+
+Person-centered approaches, including Latent Transition Analysis and Latent Class Growth Analysis, address this limitation by identifying subgroups of individuals who share similar patterns of change.
+These models can reveal meaningful subpopulations and help researchers understand the factors that contribute to differences in developmental trajectories.
+Nevertheless, person-centered methods may overlook important relationships among variables, which can be essential for understanding the dynamics of change over time.
+
+Hybrid approaches, such as Growth Mixture Modeling, combine aspects of both variable-centered and person-centered models, allowing for the identification of latent subgroups while also modeling relationships among growth parameters.
+This combination provides a more comprehensive understanding of longitudinal data by capturing both within- and between-person variability.
+However, hybrid models can be more complex, necessitating careful model specification, selection, and interpretation.
+Additionally, these methods may require larger sample sizes to ensure the stability and accuracy of results.
+
+In summary, SEM approaches offer powerful tools for longitudinal data analysis, enabling researchers to investigate complex relationships, individual differences, and change dynamics over time.
+The choice between variable-centered, person-centered, and hybrid approaches depends on the research objectives and the nature of the data.
+Despite their limitations, these models have greatly advanced our understanding of developmental processes and the factors that contribute to individual differences in change trajectories.
+
+##### Advanced SEM
+
+Advanced structural equation modeling (SEM) approaches, such as the Random-Intercept Cross-Lagged Panel Model (RI-CLPM) and Latent Curve Models with Structured Residuals (LCM-SR), have emerged to address more complex research questions and data structures in longitudinal analysis.
+These advanced models extend traditional SEM techniques, enabling researchers to disentangle within-person and between-person effects, as well as capture additional time-specific dependencies and associations that may not be accounted for by the latent growth factors.
+
+The RI-CLPM enhances the traditional cross-lagged panel model by incorporating random intercepts, which allow for the separation of stable individual differences from the dynamic within-person associations between variables over time.
+This distinction is particularly valuable for understanding the time-specific effects of one variable on another, while accounting for the influence of individual differences.
+However, RI-CLPM may require larger sample sizes to ensure stability and accuracy of the estimates and can be computationally demanding.
+
+LCM-SR, on the other hand, extends the Latent Growth Curve Model by modeling structured residuals, such as autoregressive or cross-lagged effects among the residuals.
+This approach allows for a more comprehensive understanding of within-person change dynamics and factors influencing change over time.
+By including structured residuals, LCM-SR can capture additional time-specific relationships that are not explained by the latent growth factors.
+However, similar to RI-CLPM, LCM-SR also comes with increased model complexity and requires careful specification and interpretation.
+
+In conclusion, advanced SEM approaches for longitudinal data analysis provide valuable tools for addressing complex research questions and data structures.
+While they offer more nuanced insights into within-person change dynamics and the influence of individual differences, these models also come with certain limitations, such as increased complexity, computational demands, and the need for careful model specification and interpretation.
+As with any statistical method, researchers should carefully consider their research objectives, data characteristics, and the assumptions of each model when selecting the most appropriate advanced SEM approach for longitudinal analysis.
+
+![Longitudinal Models/Data Structures](./figures/Mindmap1.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="600" height="400" fig-align="center"}
+
+##### (possibly others xxxx?)
+(longitudinal neuroimaging?)
+(longitudinal twins/genetics?)
+
+### Longitudinal Analysis Section Wrap-up
+
+This section, along with associated code and additional resources made available as online supplements (), aims to serve as a resource for researchers seeking to understand and implement various longitudinal panel models.
+By providing an overview of different approaches, their strengths and limitations, and key considerations for their use, we hope to facilitate the selection of appropriate models tailored to specific research questions and data structures.
+It is essential for researchers to consider their research objectives, the characteristics of their data, and the assumptions underlying each model when choosing the most suitable approach for longitudinal analysis.
+In this vast and continually evolving field, with numerous models and approaches available to address a wide range of research questions, no single model is universally applicable or without limitations.
+The diversity of methods ensures that researchers can find an appropriate tool for their specific needs.
+By familiarizing themselves with the various types of longitudinal models, researchers can more effectively navigate the complexities of longitudinal data and contribute valuable insights into the developmental processes and individual differences that shape human experience.
+
+We encourage researchers to consult the cited literature and online supplements for further guidance in selecting and implementing longitudinal models when using the ABCD Study dataset.
+As the field continues to advance, we anticipate the emergence of new methods and refinements to existing approaches, further expanding the toolkit available to researchers for the analysis of longitudinal data.
+By staying informed about developments in this area and critically evaluating the appropriateness of different models for their research questions, researchers can ensure that their longitudinal analyses are both rigorous and informative.
+
+------------------------------------------------------------------------
 
@@
 
+{{< pagebreak >}}
+
 # Part V: References

### 2023-06-15 _data_wrangling.qmd
revised text

# **Data Wrangling** {.unnumbered}

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE

#Clear Environment
#rm(list=ls())
```

<style>
div.blue pre { background-color:ghostwhite; }
div.blue pre.r { background-color:seashell; }

.nav-tabs > li.active:nth-child(1) > a {
  background-color: ##964B00;
}

.nav-tabs > li > a {
  background-color: ##FFFF00;
  color: blue;
}

main-container {
   max-width: unset;
}
.descriptives {
   overflow: auto; /*do not use `resize`*/
}

.tableby {
   overflow: auto;
}

.descriptives {
   overflow: auto;
   resize: both;
}

table, th, td {
  padding-top: 0px !important;
  padding-bottom: 0px !important;
  vertical-align: middle;
}
</style>

::: {.column-body-outset}
## Preliminary Setup
::: {.panel-tabset}
### Load Libraries {.tabset .tabset-fade .tabset-pills}
example-text before code block
<div class = "blue">
```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Install necessary packages (if not already installed)
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("arrow" %in% installed.packages())) install.packages("arrow")

```
</div>
example-text after code block

### Load Packages
<div class = "blue">
```{r pckg-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| class.source: fold-hide

## Load packages
library(tidyverse)
library(arrow)
library(tippy)
```
</div>

### Config Options
<div class = "blue">
```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220), fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
</div>

```{r}
##| include: FALSE
##| evaluate: FALSE
#| message: FALSE
#| warning: FALSE
#| echo: FALSE

#df_ndaFile <- readRDS("/Users/shawes/Library/CloudStorage/GoogleDrive-swh004@gmail.com/My Drive/bio_Main/_data/nda4.0.rds")

#df_test <- readRDS("path_to_ndaFile/nda4.0.rds")

#Create temporary dataframe
#df_temp <- (df_ndaFile) %>%
#   select(subjectid, eventname, site_id_l, rel_family_id, sex_at_birth, 
#          demo_brthdat_p, screentime_13,screentime_wkdy_3, screentime_wknd_9,
#          cbcl_scr_syn_aggressive_r, cbcl_scr_syn_external_r)

# remove large dataframe object

#write_dataset(df_temp, "./df_temp", format = "csv")
df_temp<- read_csv("./df_temp.csv")

df_temp <- df_temp %>% rename("ids" = "subjectid",
                    "event" = "eventname",
                    "site_id" = "site_id_l",
                    "fam_id"="rel_family_id",
                    "sex"="sex_at_birth",
                    "age"="demo_brthdat_p",
                    "mature_vg" = "screentime_13",
                    "vg_wkdy" = "screentime_wkdy_3",
                    "vg_wknd" = "screentime_wknd_9",
                    "cbcl_agg"="cbcl_scr_syn_aggressive_r",
                    "cbcl_extern"="cbcl_scr_syn_external_r"
                    )

df_temp$ids <- as.factor(df_temp$ids)
df_temp$event <- as.factor(df_temp$event)
df_temp$site_id <- as.factor(df_temp$site_id)
df_temp$fam_id <- as.factor(df_temp$fam_id)
df_temp$sex <- as.factor(df_temp$sex)
df_temp$age <- as.numeric(df_temp$age)
df_temp$mature_vg <- as.numeric(df_temp$mature_vg)
df_temp$vg_wkdy <- as.numeric(df_temp$vg_wkdy)
df_temp$vg_wknd <- as.numeric(df_temp$vg_wknd)
df_temp$cbcl_agg <- as.numeric(df_temp$cbcl_agg)
df_temp$cbcl_extern <- as.numeric(df_temp$cbcl_extern)

## recode value of each level for factors
df_temp$sex<-recode(df_temp$sex, 'F' = 'Female', 'M' = 'Male')

df_temp$event<-recode(df_temp$event, 'baseline_year_1_arm_1' = 'Baseline',
                      '6_month_follow_up_arm_1' = '6_month', 
                      '1_year_follow_up_y_arm_1'= 'Year_1', 
                      '18_month_follow_up_arm_1' = '18_month',
                      '2_year_follow_up_y_arm_1'='Year_2', 
                      '30_month_follow_up_arm_1' = '30_month',
                      '3_year_follow_up_y_arm_1'='Year_3',
                      '42_month_follow_up_arm_1' = '42_month'
                      )

## relevel factor categories
df_temp$event <- fct_relevel(df_temp$event, "Baseline", "Year_1", "Year_2")

df_temp<-df_temp[!(df_temp$event=="6_month" | df_temp$event=="18_month" |
                     df_temp$event=="30_month" | df_temp$event=="42_month" | 
                     df_temp$event=="Year_3"),]

#drop unused factor levels
df_temp$event <- droplevels(df_temp$event)
levels(df_temp$event)

## Compute total video gaming hours variable
df_temp$vg_total <- (df_temp$vg_wkdy*5)+(df_temp$vg_wknd*2)
df_temp$vg_total
## Compute video gaming grouping variable
df_temp$vg_group <- ifelse(df_temp$vg_total>=7, "VG", NA)
df_temp$vg_group <- ifelse(df_temp$vg_total>0 & df_temp$vg_total<7, "NVG", df_temp$vg_group)
## set video gaming grouping variable as factor variable
df_temp$vg_group <- as.factor(df_temp$vg_group)

##############
#creating 3 separate dataframes for each "event" period (Baseline, Year_1, Year_2)
# due to uncertainty in best procedure for converting original "long-format" of the 
# dataset to wide format (as needed for SEM analyses) when filtering participants at each
# "event" based on their video game playing hours only at "event"== "baseline"

## create T0 dataframe, filtering by chosen criterion
df_t0 <- df %>%
  filter(df$event == "Baseline" & df$vg_total >= 7)
summary(df_t0)

## create T1 dataframe, filtering by chosen criterion
df_t1 <- df %>%
  filter(df$event == "Year_1")
df_t1 <- (df_t1) %>%
  select(ids, cbcl_agg, cbcl_extern)
summary(df_t1)

## create T2 dataframe, filtering by chosen criterion
df_t2 <- df %>%
  filter(df$event == "Year_2")
df_t2 <- (df_t2) %>%
  select(ids, cbcl_agg, cbcl_extern)
summary(df_t2)

df_merge <- merge(df_t0, df_t1, by='ids')
df_merge <- merge(df_merge, df_t2, by='ids')
df_merge <- df_merge %>% rename("cbcl_agg_T2" = "cbcl_agg",
                                "cbcl_agg_T0" = "cbcl_agg.x",
                                "cbcl_agg_T1" = "cbcl_agg.y",
                                "cbcl_extern_T2" = "cbcl_extern",
                                "cbcl_extern_T0" = "cbcl_extern.x",
                                "cbcl_extern_T1" = "cbcl_extern.y"
                                )
summary(df_merge)
print(df_merge)
#df_xxxxx <- df_xxxxx %>%
#filter(mature_vg > 1)

******************************

df_wide <- (df_merge) %>%
  select(ids, event, site_id, fam_id, sex, age, mature_vg, vg_total, vg_group,
         cbcl_agg_T0, cbcl_agg_T1, cbcl_agg_T2,
         cbcl_extern_T0, cbcl_extern_T1, cbcl_extern_T2)

###########

df_sub <- df %>% group_by(event) %>% sample_n(500)
print (df_sub)
###########

write_dataset(df, "./df", format = "csv")

rm(df_ndaFile, df_temp)
```

```{r}
#| message: FALSE
#| warning: FALSE
#| echo: FALSE

#create a new 'long-to-wide' tibble from original tibble

#df_wide <- pivot_wider(df, id_cols = ids,
#                       names_from = event, 
#                       values_from = cbcl_agg,
#                       )
#head(df_wide)

#mature_vg, vg_total, vg_group, cu_traits, DBD, DBDcu, 
#cbcl_rulbrk, cbcl_agg, cbcl_extern, cbcl_opp, cbcl_cd

#as_tibble(df_wide)
#tibble(df_wide)

#write_dataset(df_wide, "./", format = "feather")
#write_dataset(df_wide, "./", format = "csv")
```

```{r}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: FALSE

df_long <- df_wide %>%
  pivot_longer(
    cols = starts_with("cbcl_agg"),
    names_to = "time",
    names_prefix = "cbcl_agg_",
    values_to = "cbcl_agg"
  )

df_long <- (df_long) %>%
  select(ids, time, site_id, fam_id, sex, age, mature_vg, vg_total, vg_group, 
         cbcl_agg)

 df_long <- df_long %>% rename("event" = "time")
 df_long$event <- as.factor(df_long$event)
  df_long

df_long_sub <- df_long %>% sample_n(500)
print (df_long_sub)

#create a new 'wide-to-long' tibble from original tibble
# longer <- df_wide %>% 
#     pivot_longer(cols = -ids, 
#                  names_to = c(".value", "event2"), 
#                  names_pattern = "([^\\.]*)\\.*(\\{4})")
# 
# longer<-pivot_longer(df_wide, cols=-1, names_pattern = "(.*)(..)$", names_to = c("limit", "name")) %>% 
#      mutate(limit=ifelse(limit=="", "value", limit))
# 
# answer <-pivot_wider(longer, id_cols = c(event, name), names_from = limit, values_from = value, names_repair = "check_unique")

# df_long <- df_wide %>%
#   pivot_longer(
#     cols = starts_with("cu_traits"),
#     names_to = "time",
#     names_prefix = "cu_traits_",
#     values_to = "cu_traits"
#   )
# 
# df_long <- (df_long) %>%
#   select(ids, time, site_id, fam_id, sex, age, mature_vg, vg_total, vg_group,
#          DBD, DBDcu, cu_traits) 
# 
#  df_long <- df_long %>% rename("event" = "time")
#  df_long$event <- as.factor(df_long$event)
# 
# 
# df_long <- df_long
write_dataset(df_long, "./df_long", format = "csv")
write_dataset(df_long_sub, "./df_long_sub", format = "csv")
```

### 2023-06-15 manuscript.qmd
revised text

@@
 short-title: Longitudinal Analysis
-published: "Under review at *xxxxx Journal*"
 code-repo: "Access the code, data, and analysis at <http://swhawes.github.io/test/>"
@@
         country: USA
-  - name: Erin L. Thompson
-    email: erthomps@fiu.edu
-    affiliations:
-      #- id: terok
-        name: Florida International University
-        department: Center for Children and Families
-        city: Miami
-        region: FL
-        country: USA
-  - name: Raul Gonzalez
-    email: gonzara@fiu.edu
-    affiliations: 
-      #- id: ds9
-        name: Florida International University
-        department: Center for Children and Families
-        city: Miami
-        region: FL
-        country: USA
+  # - name: Erin L. Thompson
+  #   email: erthomps@fiu.edu
+  #   affiliations:
+  #     #- id: terok
+  #       name: Florida International University
+  #       department: Center for Children and Families
+  #       city: Miami
+  #       region: FL
+  #       country: USA
+  # - name: Raul Gonzalez
+  #   email: gonzara@fiu.edu
+  #   affiliations:
+  #     #- id: ds9
+  #       name: Florida International University
+  #       department: Center for Children and Families
+  #       city: Miami
+  #       region: FL
+  #       country: USA
   - name: Additional Co-authors
@@
 
+format:
+  docx:
+    toc: false
+  elsevier-pdf:
+    journal:
+      cite-style: number
+      formatting: preprint
+      model: 3p
+      layout: twocolumn
+      #cite-style: number
+    include-in-header: 
+      text: |
+        \newpageafter{author}
+  html:
+    page-layout: article
+    toc: true
+    toc-depth: 1
+    toc-expand: 2
+    toc-title: Table of Contents
+    code-tools:
+      source: true
+      toggle: false
+      caption: none
+    grid:
+      sidebar-width: 200px
+      margin-width: 0px
+
+#fontfamily: libertinus
+#fontsize: 20pt
+#colorlinks: true
+#format-links: [html, pdf]
+#execute:
+  #echo: false
+#bibliography: "references.bib"
+#biblio-style: unsrt
+
 abstract: |
   The Adolescent Brain Cognitive Development (ABCD) Study presents a unique opportunity for researchers to investigate developmental processes in a large, diverse cohort of children and adolescents.
-Given the complex nature of the longitudinal data collected in the ABCD Study, researcher are likely to encounter a myriad of methodological and analytic considerations and concerns.
-This review provides a comprehensive examination of key issues and techniques related to longitudinal data analysis, specifically focusing on the ABCD Study.
-The text discusses model assumptions, common violations (e.g., independent and identically distributed residuals, *heterogeneous* *variability*) and their implications for valid inference.
-The importance of appropriately modeling covariance structures, understandings trade-offs between model fit and parsimony, and challenges related to sample size, attrition, missing data are highlighted.
+Given the complex nature of the longitudinal data collected in the ABCD Study, researchers are likely to encounter a myriad of methodological and analytic considerations and concerns.
+This review provides a comprehensive examination of key issues and techniques related to longitudinal data analysis, with a specific focus on use with the ABCD Study dataset.
+The text discusses model assumptions, common violations (e.g., independent and identically distributed residuals, heterogeneous variability), and their implications for valid inference.
+The importance of appropriately modeling covariance structures, understandings trade-offs between model fit and parsimony, and challenges related to sample size, attrition, and missing data are highlighted.
 Consideration is given to the importance of selecting appropriate statistical models to account for correlations in repeated measurements and the assumptions underlying these models.
@@
 By addressing these complexities, the review seeks to equip researchers with the necessary knowledge and tools to make informed decisions as they navigate effectively analyzing and interpreting data available in the ABCD Study.
-# thanks: |
-#   Placeholder text generated at <https://vlad-saling.github.io/star-trek-ipsum/>.
-# additional-info: |
-#   We have no known conflict of interest to disclose.
+
 keywords:
@@
 - development
-format:
-  #hikmah-pdf: default
-  hikmah-manuscript-pdf: default
-  #html:
-    #page-layout: article
-    #toc: true
-    #toc-depth: 1
-    #toc-expand: 2
-    #toc-title: Table of Contents
-  #   code-tools:
-  #     source: true
-  #     toggle: false
-  #     caption: none
-  #   grid:
-  #     sidebar-width: 200px
-  #     margin-width: 0px
-  pdf:
-    geometry:
-      - top=30mm
-      - left=20mm
-  #   keep_tex: yes
-    #fontfamily: libertinus
-    #fontsize: 20pt
-#colorlinks: true
-#format-links: [html, pdf]
 
-execute:
- echo: false
-
-# output: 
-#   rticles::arxiv_article
-bibliography: "references.bib"
-#biblio-style: unsrt
 ---
@@
 
-\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States.
+\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largestz long-term investigation of neurodevelopment and child health in the United States.
 Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018].
@@
 
-Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians.
+Participants enrolled in the ABCD Study include a large cohort of youth (n=11,880) aged 9-10 years at baseline and their parents/guardians.
 The study sample was recruited from household populations in defined catchment areas for each of the 21 study sites across the United States (information regarding funding agencies, recruitment sites, investigators, and project organization can be obtained at the ABCD Study website).
-The ABCD Study is collecting longitudinal data on a rich variety of outcomes that will enable the construction of realistically-complex etiological models by incorporating factors from many domains simultaneously.
+The ABCD Study is collecting longitudinal data on a rich variety of outcomes that will enable the construction of complex etiological models by incorporating factors from many domains simultaneously.
 Each new wave of data collection provides the building blocks for conducting probing longitudinal analyses that allow us to characterize normative development, identify variables that presage deviations from prototypic development, and assess a range of outcomes associated with variables of interest.
-This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use \[add citation\], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020].
+This data includes a neurocognitive battery [@luciana2018a; @thompson2019], mental and physical health assessments [@barch2018], measures of culture and environment [@zucker2018], substance use [@lisdahl2021], biospecimens [@uban2018], structural and functional brain imaging [@casey2018; @hagler2019], geolocation-based environmental exposure data, wearables, and mobile technology [@bagot2018], and whole genome genotyping [@loughnan2020].
 Many of these measures are collected at in-person annual visits, with brain imaging collected at baseline and every other year going forward.
@@
 
-This manuscript will go through the following sections :xxxxxx, xxxx, xxxxx.
-# Part I: Longitudinal Research
-# Part II: Longitudinal Data
-# Part III: Longitudinal Analysis
-Overview of **concepts and considerations specific to developmentally-focused longitudinal research**
-
 ## Organization and Aims
 
-Aims/Goals: Use this xxxxx during any xxxxx and you're virtually guaranteed to xxxxx
+• Part I. Longitudinal Research
 
--   Concepts and Considerations when working with large and multimodal longitudinal data
+  -   Identify fundamental concepts
 
--   Concepts and considerations of developmental research (longitudinal research via developmental perspective)
+• Part II. Longitudinal Data
 
--   Learn specific methods
+  -   Highlight key challenges
 
--   Use our open-source resources to get started more quickly
+• Part III. Longitudinal Analysis
 
--   Learn How to Apply xxxxx to large longitudinal data such as the ABCD Study
+  -   Learn methods & analysis to apply
 
-identify common xxxxx of longitianl research
+• Part IV. Supplemental materials
 
-1.3 Objectives and scope
-The focus of this xxxx is on the analysis of event history data using
-xxxx model with the objectives
-• to demonstrate the use of event history analysis in the analysis
-of xxxxx, xxxx and to identify the methodological problems in the analysis,
-• to propose procedures to choose a basic time scale,
-• to discuss the connections between the methods for modeling
-dual time scales and to perform quantitative comparisons between them,
-• to compare existing methods to deal with longitudinal measurements in the xxxx model with two proposed methods.
-
+  -   Make use of linked open-source resources
 :::
 
-# Part I: Longitudinal Research
+# Longitudinal Research
 
@@
 These include different ways of thinking about developmental course, whether certain periods of development are relatively sensitive or insensitive to various types of insults or stressors, whether some time periods or situations inhibit the expression of individual differences due to extreme environmental pressures, and whether the same behavior manifested at different times represent the same phenomenon or different ones.
-Further, in the case of developmentally focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child's life course (and not just characterize growth during this period but also assesses the durability/chronicity of prior effects/consequences) but also brings with it greater methodological opportunities to exploit the statistical properties of longitudinal data in the furtherance of critical scientific questions.
-That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows, assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest.
+Further, in the case of developmentally focused longitudinal research, each new measurement occasion not only provides a more extended portrait of the child’s life course but also brings with it greater methodological opportunities to make use of statistical models that distinguish within and between subject effects and that loosen constraints that could be imposed on the furtherance of critical scientific questions.
 Appreciation of these and other issues can help to guide the analysis and interpretation of data and aid translation to clinical and public health applications.
 
-**Vulnerable periods**.
+### Vulnerable periods.
+
 Development normatively progresses from less mature to more mature levels of functioning.
 However, unique epochs and experiences can alter the course of this idealized form of development.
-Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development \[add citation\]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans)\[add citation\].
+Consider research that shows cannabis use during adolescence is associated with later psychosis to a greater degree than cannabis use initiated later in development [@arseneault2002; @bechtold2016; @hasan2020; @semple2005]; or, similarly, experimental research on rodents that shows rodent brains to be especially sensitive to the neurotoxic effects of alcohol on brain structure and learning early in development (corresponding to early adolescence in humans) [@spear2016; @crews2000; @ji2018].
 These examples highlight the importance of considering the role of vulnerable periods -- temporal windows of rapid brain development or remodeling during which the effects of environmental stimuli (e.g.
@@
 
-**Developmental disturbances**.
-Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times environmental pressures will tend to suppress stability and disrupt the orderly stochastic process of normative development (e.g., xxx-xxx).
+### Developmental disturbances.
+
+Whereas vulnerable periods heighten neurobiological susceptibility to environmental influences, at other times environmental pressures will tend to suppress stability and disrupt the orderly stochastic process of normative development [e.g., @schulenberg2019].
 This situation reflects a developmental disturbance in that the normal course of development is "disturbed" for some time by some time-limited process.
 In such cases, we might find that prediction of behavior in the period of the disturbance is reduced and/or, similarly, the behavior exhibited during the disturbance might have less predictive power with respect to distal outcomes compared to the behavior exhibited before and following the disrupted period.
-That is, once the environmental stimuli are removed (or the individual is removed from the environment), individual differences are again more freely expressed and the autoregressive effects increase to levels similar to those before entering the environment.
+That is, once the environmental stimuli are removed (or the individual is removed from the environment), patterns of individual differences (and autoregressive effects) increase to levels similar to those prior to entering the environment.
 
-**Developmental snares and cascade effects**.
+### Developmental snares and cascade effects.
+
 Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next.
-For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect, creating a "snare" that retards psychosocial development.
-Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified.
+For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect, creating a "snare" that retards psychosocial development such as maturing out of adolescent antisocial behavior [@moffitt2015].
+Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified [e.g., @masten2005; @rogosch2010].
 Although conceptually distinct from vulnerable periods, both of these types of developmental considerations highlight the importance of viewing behavior in the context of development and the importance of attempting to determine how various developmental pathways unfold.
 
-**Distinguishing developmental change from experience effects**.
+### Distinguishing developmental change from experience effects.
+
 One can often observe systematic changes over time in a variable of interest and assume this change is attributable to development.
@@
 While ABCD was not designed as an accelerated longitudinal design, the variability in age at the time of baseline recruitment (9 years, 0 months to 10 years, 11 months) allows some measures, collected every year, to be conceptualized as an accelerated longitudinal design.
-Moreover, it is possible that in later waves, patterns of longitudinal missing data will allow some analyses to assess the confounded effects of age and the number of prior assessments.
+It is possible that in later waves, analyses will allow for disaggregating the confounded effects of age and the number of prior assessments.
 However, ABCD is fundamentally a single-cohort, longitudinal design, where a number of prior assessments and age are highly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.
 
-# Part II: Longitudinal Data
+# Longitudinal Data
 
@@
 
-**Defining Features of Longitudinal Data Analysis.** The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
+The hallmark characteristic of longitudinal data analysis is its application to repeated measurements of the same assessment targets (e.g., individuals, families) across time.
 While the primary reason for collecting longitudinal data is in pursuit of addressing scientific questions, from a methodological perspective, having multiple observations over time allows researchers to identify potentially problematic observations when highly improbable longitudinal patterns are observed.
-That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest .
+That is, we can ask more nuanced questions and make stronger inferences as our number of time-ordered observations grows assuming we have assessed the "right" variables and the timings of our observations comport with the temporal dynamics of the mechanisms of interest.
 
-["traditional approaches": marginal models]
-[random effects: Growth Curves- SEM (lgcm) vs ME (lmm)]
-[(between- and within-person levels?)]
-
-### Modeling Data Across Two Time Points versus Three or More Time Points.
+### Two Time Points versus Three or More.
 
@@
 This type of "rank-order" stability evaluates between-individual change by focusing on the degree to which individuals retain their relative placement in a group across time.
-Consider, someone who is the shortest person in their class in 6th grade may grow considerably over the school year (i.e., exhibit mean level change), but remain the shortest person among their classmates.
+Consider someone who is the shortest person in their class in 6th grade may grow considerably over the school year (i.e., exhibit mean level change), but remain the shortest person among their classmates.
 That is, the individual is manifesting a type of rank-order stability.
@@
 For example, in some areas of development, there is considerable mean-level change that occurs over time (e.g., changes in Big 5 personality traits), but relatively high rank-order stability.
-Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of "within-individual" change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest.
+Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of "within-individual" change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest[@curran2011].
 
-There is growing recognition that statistical models commonly applied to longitudinal data often fail to comport with the developmental theory they are being used to assess (e.g., Curran, Lee, Howard, Lane, & MacCallum, 2012; Hoffman, 2015; Littlefield et al., 2021.
+![Types of Stability and Change](./figures/StabilityChange.png){#fig-StablityChange fig-alt="Stability & Change Diagram." width="300" height="300" fig-align="left"}
+
+There is growing recognition that statistical models commonly applied to longitudinal data often fail to align with the developmental theory they are being used to assess [e.g., @curran2012; @hoffman2015; @littlefield2021].
 Specifically, developmental studies typically involve the use of prospective data to inform theories that are concerned with clear within-person (i.e., intraindividual) processes (e.g., how phenotypes change or remain stable within individuals over time) [e.g., see @curran2011].
 Despite this, methods generally unsuited for disaggregating between- and within-person effects (e.g., cross-lagged panel models [CLPM]) remain common within various extant literatures.
-As a result, experts increasingly caution about the need to xxxxxxxx [add citation].
 Fortunately, there exists a range of models that have been proposed to tease apart between- and within-person sources of variance across time [see @littlefield2021; @orth2021].
@@
 
-This is illustrated well by @fig-LDAstruct.
-
 ### Model Assumptions
@@
 Many statistical models assume certain characteristics about the data to which they are being applied.
-As an example, common assumptions of parametric statistical models include normality, linearity, and equality of variances.
+Common assumptions of parametric statistical models include normality, linearity, and equality of variances.
 These assumptions must be carefully considered before conducting analysis so that valid inferences can be made from the data; that is, violation of a model's assumptions can substantively alter the interpretation of results.
-Similarly, statistical models employed in the analyses of longitudinal data often entail a range of assumptions that must be closely inspected.
-One central issue for repeated measurements on an individual is how to account for the correlated nature of the data; another common feature of longitudinal data is heterogeneous variability; that is, the variance of the response changes over the duration of the study.
+For example, longitudinal data can exhibit heterogeneous variability (i.e., the variance of the response changes over the duration of the study) that will need to be accounted for.
+Similarly, statistical models used to analyze longitudinal data often entail a range of assumptions that must be closely inspected.
+One common prerequisite for analyses of repeated measures data is the assessment of longitudinal measurement invariance, which helps to establish whether or not a construct is measured consistently over time [@liu2017; @van2015; @willoughby2012].
+Establishing longitudinal measurement invariance, in turn, provides researchers with greater confidence that any change over time identified for a construct is attributable to individual change rather than a measurement artifact.
+Another central issue for repeated measurements on an individual is how to account for the correlated nature of the data.
 Traditional techniques, such as a standard regression or ANOVA model, assume residuals are independent and thus are inappropriate for designs that assess (for example) the same individuals across time.
@@
 
-As an example, one alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design.
+One alternative structure that attempts to handle the reality that correlations between repeated assessments tend to diminish across time is the autoregressive design.
 As the name implies, the structure assumes a subsequent measurement occasion (e.g., assessment at Wave 2) is regressed onto (that is, is predicted by) a prior measurement occasion (e.g., assessment at Wave 1).
@@
 Diverting from compound symmetry, this model assumes the correlations between repeated assessments decline exponentially across time rather than remaining constant.
-For example, per the AR(1) structure, if the correlation between Time 1 and Time 2 data is thought to be .5, then the correlation between Time 1 and Time 3 data would be assumed to be .5*.5 = .25, and the correlation between Time 1 and Time 4 data would be assumed to be .5*.5\*.5 = .125.
+For example, per the AR(1) structure, if the correlation between Time 1 and Time 2 data is thought to be .5, then the correlation between Time 1 and Time 3 data would be assumed to be .5 × .5 = .25, and the correlation between Time 1 and Time 4 data would be assumed to be .5 × .5 × .5 = .125.
 As with compound symmetry, the basic AR(1) model is parsimonious in that it only requires two parameters (the variance of the assessments and the autoregressive coefficient).
@@
 within-person sources of variance [@hamaker2015].
+Finally, it is also worth noting that in many longitudinal contexts, the time intervals between assessments are not equidistant and researchers need to consider carefully how to appropriately model time in their model and what this model implies.
 
-### Linear vs non-linear models
+### Continuous and Discrete Indicators
 
@@
 These models can account for issues that can occur when working with discrete outcomes, including overdispersion (when the variance is higher than would be expected based on a given distribution) and zero-inflation [when more zeros occur than is expected based on a given distribution; see @lenz2016].
-Models involving zero-inflation parameters are referred to as two-part models, given one part of the model predicts the zero-inflation whereas the other part of the model predicts outcomes consistent with a given distribution [e.g., Poisson distribution; see @farewell2017, for a review of two-part models for longitudinal data].
-Although there exist several alternative models for discrete indicators, some more recent models that have been proposed for prospective data are only feasible in cases where indicators are assumed to be continuous rather than discrete [e.g., LCM-SR; @curran2014a].
+Models involving zero-inflation parameters are referred to as two-part models, given one part of the model primarily is used to predict zero-inflation whereas the other part of the model predicts outcomes consistent with a given distribution [e.g., Poisson distribution; see @farewell2017, for a review of different types of two-part models for longitudinal data].
 Given the sheer breadth of issues relevant to determining better models for discrete outcomes, it is not uncommon for texts on longitudinal data analysis to only cover models and approaches that assume continuous indicators [e.g., @little2013].
@@
 
-### Marginal vs Conditional Models (?)
-
-\[section still under development...\]
-
 ### Missing Data/Attrition
 
-As recently reviewed by Littlefield (in press), investigators of prospective data are confronted with study attrition (i.e., participants may not provide data at a given wave of assessment) and thus approaches are needed to confront the issue of missing data.
+As recently reviewed by Littlefield [in press; @littlefield2022], investigators of prospective data are confronted with study attrition (i.e., participants may not provide data at a given wave of assessment) and thus approaches are needed to confront the issue of missing data.
 Three models of missingness are typically considered in the literature [see @little1989].
@@
 LOCF replaces a participant's missing values after dropout with the last available measurement [@molnar2008].
-This approach assumes stability (i.e., a given participant's score is not anticipated to increase or decline after study dropout) and that the data are MCA R.
+This approach assumes stability (i.e., a given participant's score is not anticipated to increase or decline after study dropout) and that the data are MCAR.
 However, as described by @molnar2008, it is common for treatment groups to show higher attrition compared to control groups in studies of dementia drugs.
@@
 
-More modern approaches, such as using maximum likelihood or multiple imputation to estimate missing data, are thought to avoid some of the biases of older approaches [see @enders2010; @graham2009].
+More modern approaches, such as using maximum likelihood, propensity scoring, auxiliary variables, or multiple imputation to estimate missing data, are thought to avoid some of the biases of older approaches [see @enders2010; @graham2009].
 @graham2009 noted several "myths" regarding missing data.
@@
 Attrition from a longitudinal panel study such as ABCD is inevitable and represents a threat to the validity of longitudinal analyses and cross-sectional analyses conducted at later time points, especially since attrition can only be expected to grow over time.
-While, to date, attrition in ABCD has been minimal (some cite here), it remains an important focus for longitudinal analysis and its significance is likely to only grow as the cohort ages.
+The ABCD Retention Workgroup (RW) employs a data-driven approach to examine, track, and intervene in these issues and while preliminary findings show `participant race` and `parent education level` to be associated with late and missing visits, to date, attrition in ABCD has been minimal [@ewing2022].
 Ideally, one tries to minimize attrition through good retention practices from the outset via strategies designed to maintain engagement in the project [@cotter2005; @hill2016; @watson2018].
@@
 Perhaps the most key concern when dealing with data that is missing due to attrition is determining the degree of bias in retained variables that is a consequence of attrition.
+Such bias can attenuate generalizability, particularly if the pattern of missingness is not random (e.g., certain subsets of the population are more likely to drop out/not attend a visit).
 Assuming that the data are not missing completely at random, attention to the nature of the missingness and employing techniques designed to mitigate attrition-related biases need to be considered in all longitudinal analyses.
-Several different approaches can be considered and employed depending upon the nature of the intended analyses, the degree of missingness, and data available to help estimate missing and unobserved values.
 
@@
 Effect size can be defined as, "a population parameter (estimated in a sample) encapsulating the practical or clinical importance of a phenomenon under study." [@kraemer2014].
-Common effect size metrics include r (i.e., the standardized covariance, or correlation, between two variables) and Cohen's d [@cohen1988].
+Common effect size metrics include r (i.e., the standardized covariance, or correlation, between two variables) and Cohen's d [i.e., the standardized difference between two means; @cohen1988].
 Adjustments to common effect size calculations, such as Cohen's d, are required even when only two time points are considered [e.g., see @morris2002].
@@
 @feingold2019 reviews the approach for effect size metrics for analyses based on growth modeling, including when considering linear and non-linear (i.e., quadratic) growth factors.
-@morris2002 review various equations for effect size calculations relevant to when combining estimates in meta-analysis with repeated measures and independent-groups designs.
+@morris2002 review various equations for effect size calculations relevant to combining estimates in meta-analysis with repeated measures and independent-groups designs.
 Other approaches to quantifying effect sizes longitudinally may be based on standardized estimates from models that more optimally disentangle between- and within-person sources of variance (as reviewed above).
-As an example, within a RI-CLPM framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.
-
-# Part III: Longitudinal Analysis
-
-\label{sec:headings}
+As an example, within a random-intercept cross-lagged panel model (RI-CLPM) framework, standardized estimates between random intercepts (i.e., the correlation between two random intercepts for two different constructs assessed repeatedly) could be used to index the between-person relation, whereas standardized estimates among the structured residuals could be used as informing the effect sizes of within-person relations.
 
@@
 An understanding of these data structures is helpful, as they can warrant different types of longtiudinal data analysis.
-Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and Bauer (xxxx) provide a nice on-ramp in distinguishing between the use of "time-to-event" and "repeated measures" data.
-Although both model time, the former is concerned with *whether* and *when* an event occurs, whereas the later is focused on *growth* and *change* \[Curran and Bauer, xxxx\].
+Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and Bauer [2019; @bauer2019] provide a nice on-ramp in first distinguishing between the use of "time-to-event" and "repeated measures" data.
+Although both model time, the former is concerned with *whether* and *when* an event occurs, whereas the later is focused on *growth* and *change* [@bauer2019].
 Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest.
-This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models) and the time-to-event data can be based on a single assessment or iclude multiple recurrent or competing events.
-While much has been written about about "time-to-event" data (see xxx; xxx; xxxx), including several recent xxxx (see), our emphasis will be given to the modeling of "repeated measures" data.
+This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models) and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events.
+While much has been written about "time-to-event" data (see xxx; xxx; xxxx), including several recent xxxx (see), our emphasis will be given to the modeling of "repeated measures" data.
 
@@
 One such approach to repeated measures analysis is the use of `time-series models`.
-These models generally consist of a long sequence of repeated measurements (≧ 50-100 measurements) on a single (or small number of) some variable of interest.
+These models generally consist of a long sequence of repeated measurements (≧ 50-100 measurements) on a single (or small number of) variable of interest.
 Time-series analysis is often used to predict/forecast temporal trends and cyclic patterns and is geared toward making inferences about prospective outcomes within a population (with relatively with less focus on inferring individual-level mechanisms and risk factors).
@@
 
-The following text includes collated information about each of the categories of longitudinal analyses that we have discussed during previous questions and responses that has been written up for publication in a scientific journal:
+# Longitudinal Analysis
 
-#### Types of longitudinal panel models
+\label{sec:headings}
+
+### Types of longitudinal panel models
 
 With the large and continually expanding body of research on statistical methods for longitudinal analyses, determining which longitudinal model to implement can be challenging.
-The aim of this section is to help researchers to navigate these many options in order to identify the statistical approach most appropriate to their unique research question.
+The aim of this section is to help researchers navigate these many options in order to identify the statistical approach most appropriate to their unique research question.
 Notably, there are a myriad of viable ways one can go about grouping various types of longitudinal models for presentation.
-Common examples iclude grouping by linearity (linear vs nonlinear models; see Collins, 2006(?)), the number of measurement occassions (see King et al., 2018), and xxxxxx (see xxxx, xxx).
-The ordering/organzation we use below overlaps in a number of ways with these examples, and with Curran & Bauer (2019); however, it is important to note that in each case, the ordering/organization that is chosen is somewhat arbitrary and primarly intended to allow the reader to compare and contrast of various analytic approaches.
-In the following we will describe (briefly) and summarize the advantages/disadvantages for a series of 16(?) distinct longitudinal models organized into the following x# `fuzzy` groupings: Traditional Models, Modern GLM Extensions, SEM, Advanced SEM, xxx, xxxx.
-We note that this is note an exhaustive review of each of these methods, but for more in-depth detail we do provide the reader relevant resources.
-As aptly summarized by Curran & Bauer (2019), "there are many exceptions, alternatives, nuances, “what ifs”, and “but couldn’t yous” that aren’t addressed here."
+Common examples include grouping by linearity [linear vs nonlinear models; see @collins2006], the number of measurement occasions (see @king2018), and statistical equivalency (e.g., change scores vs.
+residualized change; see @castro-schilo2018).
+The ordering/organzation we use below overlaps in a number of ways with these examples, and in particular with Bauer and Curran (2019; @bauer2019).
+However, it is important to note that in each case, the ordering/organization that is chosen is somewhat arbitrary and primarly intended to allow the reader to compare and contrast various analytic approaches.
+In the following we describe (briefly) and summarize the advantages/disadvantages for a series of longitudinal models organized into the following `fuzzy` groupings: Traditional Models, Modern GLM Extensions, SEM, and Advanced SEM.
+We note that this is not an exhaustive review of each of these methods, but for more in-depth detail we do provide the reader relevant resources.
+As aptly summarized by Bauer and Curran (2019; @bauer2019), "there are many exceptions, alternatives, nuances, 'what ifs', and 'but couldn’t yous' that aren’t addressed here."
 
-##### Traditional (linear and nonlinear)
-Traditional methods for longitudinal analysis primarily focus on modeling mean-level change. 
+![Longitudinal Models/Data Structures](./figures/Mindmap1copy.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="300" height="300" fig-align="left"}
+<br>
+
+#### Traditional Models
+Traditional methods for longitudinal analysis primarily focus on modeling mean-level change, and how these changes may differ across groups or levels of some other variable.
 For example, is there a difference in average math scores obtained across multiple assessments between boys and girls?
@@
 For example, computing a `difference score` (e.g., math scores at t2 - math scores at t1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups.
-We also note the longitudinal `signed-rank test` which is a nonparametric alternative to the paired t-test that can be a useful tool for analyzing non-normal paired data.
+Additionally the longitudinal `signed-rank test`, a nonparametric alternative to the paired t-test, can be a useful tool for analyzing non-normal paired data.
 Another common approach, often used in pre-/post-design studies, is to use `residualized change score analysis` to assess the degree of change in a variable, while controlling for its initial level.
 For example, regressing post-treatment scores on pre-treatment scores and then using the resulting residuals as a measure of change that is adjusted for baseline scores (ignoring any prior group assignments/differences).
-Similar to difference scores, the residualized change score is often included in subsequent analysis, such as for evaluating intervention effects in pre-test/post-test designs [@Kisbu-Sakarya2013].
+Similar to difference scores, the residualized change score is often included in subsequent analysis, such as for evaluating intervention effects in pre-test/post-test designs [@kisbu2013].
 
 Traditional longitudinal methods can still be useful in some contexts (e.g., < 3 measurement occasions), but overall their practical utility for answering questions about developmental processes is limited.
-Perhaps most notably, these models do not allow for characterizing patterns of within-person change, which is often an emphasis of longitudinal research.
-In turn, these models often correspond poorly with most theoretical models of change.
+Perhaps most notably, these models do not allow for characterizing patterns of within-person change.
+This is a particularly important limitation since most psychological theories posit within-person processes (i.e., what will happen within a given individual).
+As such, traditional approaches often correspond poorly with most theoretical models of change and a failure to disaggregate between-person and within-person effects can result in consequential errors of inference (e.g., ecological fallacy)[@curran2011].
+Moreover, even determining which of these procedures to use for comparing change over two time points across groups can be surprisingly complicated.
+A particularly vexing example is that of imbalanced baseline scores (i.e., when baseline scores are correlated with a covariate of interest), which can produce different conclusions across methods (see Littlefield, 2023[@zotero-34098], for a review).
+Given these shortcomings, and the complexity of the issues surrounding some of these methods, it is typically recommended that researchers make use of more modern approaches for analyzing longitudinal data (and preferably make use of data collected across 3 or more time points).
 
-Moreover, even determining which of these procedures to use for comparing change over two time points across groups can be surprisingly complicated.
-A particularly vexing example is that of imbalanced baseline scores which can produce biased effects when using these methods.
-Indeed, it is well-documented that this can result in different conclusions between these approaches, which has led to many articles debating the merits of which procedures is most appropriate to use and when [see xxxxx for a review].
-Beyond these concerns, additional limitations include xxxxxx.
-Given these shortcomings and the complexity of the issues surrounding some of these methods, when possible, it is typically recommended that researchers make use of more modern approaches for analyzing longitudinal data (and preferably make use of data collected across 3 or more time points.
-
-##### Modern GLM Extensions
+#### Modern GLM Extensions
 
@@
 
-##### SEM
+#### SEM
 
@@
 
-Variable-centered approaches, such as Latent Change Scores, Latent Growth Curve Models, Multivariate (Parallel Process) Latent Growth Curve Models, and Latent State-Trait Models, focus on examining relationships among variables and population-level patterns.
+Variable-centered approaches, such as latent change scores, latent growth curve models, multivariate (parallel process) latent growth curve models, and latent state-trait models, focus on examining relationships among variables and population-level patterns.
 These models offer a powerful means to estimate individual change trajectories and the relationships between growth parameters of different variables, while also decomposing observed measurements into latent state and trait components.
@@
 
-Person-centered approaches, including Latent Transition Analysis and Latent Class Growth Analysis, address this limitation by identifying subgroups of individuals who share similar patterns of change.
+Person-centered approaches, including latent transition analysis and latent class growth analysis, address this limitation by identifying subgroups of individuals who share similar patterns of change.
 These models can reveal meaningful subpopulations and help researchers understand the factors that contribute to differences in developmental trajectories.
@@
 
-Hybrid approaches, such as Growth Mixture Modeling, combine aspects of both variable-centered and person-centered models, allowing for the identification of latent subgroups while also modeling relationships among growth parameters.
+Hybrid approaches, such as growth mixture modeling, combine aspects of both variable-centered and person-centered models, allowing for the identification of latent subgroups while also modeling relationships among growth parameters.
 This combination provides a more comprehensive understanding of longitudinal data by capturing both within- and between-person variability.
@@
 
-##### Advanced SEM
+#### Advanced SEM
 
-Advanced structural equation modeling (SEM) approaches, such as the Random-Intercept Cross-Lagged Panel Model (RI-CLPM) and Latent Curve Models with Structured Residuals (LCM-SR), have emerged to address more complex research questions and data structures in longitudinal analysis.
+Advanced structural equation modeling (SEM) approaches, such as the random-intercept cross-lagged panel model (RI-CLPM) and latent curve models with structured residuals (LCM-SR), have emerged to address more complex research questions and data structures in longitudinal analysis.
 These advanced models extend traditional SEM techniques, enabling researchers to disentangle within-person and between-person effects, as well as capture additional time-specific dependencies and associations that may not be accounted for by the latent growth factors.
@@
 The RI-CLPM enhances the traditional cross-lagged panel model by incorporating random intercepts, which allow for the separation of stable individual differences from the dynamic within-person associations between variables over time.
-This distinction is particularly valuable for understanding the time-specific effects of one variable on another, while accounting for the influence of individual differences.
+Within-person variance in these models are captured by a series of latent variables that reflect time specific variance (i.e., the residual variance from the random intercept).
+These time-specific variables are referred to as structured residuals.
+Distinguishing between-person variance subsumed by the random intercept from the structured residuals is particularly valuable for understanding the time-specific effects of one variable on another, while accounting for the influence of individual differences.
 However, RI-CLPM may require larger sample sizes to ensure stability and accuracy of the estimates and can be computationally demanding.
 
-LCM-SR, on the other hand, extends the Latent Growth Curve Model by modeling structured residuals, such as autoregressive or cross-lagged effects among the residuals.
+LCM-SR, on the other hand, extends the RI-CLPM by including additional growth factors, such as a random linear slope.
+That is, the LCM-SR is a hybrid between a latent growth model and CLPM.
 This approach allows for a more comprehensive understanding of within-person change dynamics and factors influencing change over time.
 By including structured residuals, LCM-SR can capture additional time-specific relationships that are not explained by the latent growth factors.
-However, similar to RI-CLPM, LCM-SR also comes with increased model complexity and requires careful specification and interpretation.
+However, even more so than the RI-CLPM, LCM-SR comes with increased model complexity and requires careful specification and interpretation.
 
 In conclusion, advanced SEM approaches for longitudinal data analysis provide valuable tools for addressing complex research questions and data structures.
-While they offer more nuanced insights into within-person change dynamics and the influence of individual differences, these models also come with certain limitations, such as increased complexity, computational demands, and the need for careful model specification and interpretation.
+While they offer more nuanced insights into within-person change dynamics and the influence of individual differences, these models also come with certain limitations, such as the necessity of multiple assessments (e.g., four or more for LCM-SR), increased complexity, computational demands, and the need for careful model specification and interpretation.
 As with any statistical method, researchers should carefully consider their research objectives, data characteristics, and the assumptions of each model when selecting the most appropriate advanced SEM approach for longitudinal analysis.
 
-![Longitudinal Models/Data Structures](./figures/Mindmap1.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="600" height="400" fig-align="center"}
+![LDA Table](./figures/Tabletemp.png)
 
-##### (possibly others xxxx?)
-(longitudinal neuroimaging?)
-(longitudinal twins/genetics?)
+# Discussion
 
-### Longitudinal Analysis Section Wrap-up
+\label{sec:headings}
 
-This section, along with associated code and additional resources made available as online supplements (), aims to serve as a resource for researchers seeking to understand and implement various longitudinal panel models.
+As we enter the era of large-scale longitudinal investigations, it is essential to critically examine the various analytical methods that can be employed to glean insights from these rich datasets.
+The complex nature of longitudinal data demands sophisticated and well-suited methodologies to accurately address research questions and minimize biases.
+This study aimed to provide a comprehensive review of diverse longitudinal analysis techniques, with a particular emphasis on their application to extensive longitudinal studies such as the ABCD Study.
+Beyond contributing to the ever-growing body of knowledge on longitudinal data analysis, we hope this manuscript also serves as a valuable resource for researchers seeking to optimize the use of large-scale longitudinal investigations in advancing our understanding of human development and behavior.
+In this discussion, we will focus on the key findings and recommendations of our review and discuss potential innovations that can further enhance the utility of these methods.
+
+We began by addressing fundamental concepts and considerations in longitudinal research that are essential for generating accurate and meaningful insights into developmental processes.
+Concepts such as vulnerable periods, developmental disturbances and snares, or cascade and experience effects (among many others), are instrumental in shaping the design, analysis, and interpretation of longitudinal studies.
+For example, understanding and accounting for vulnerable periods and developmental disturbances in research designs allows researchers to investigate the timing and magnitude of the impact of specific factors on development, while snares and cascade effects demonstrate the importance of considering the interconnectedness of developmental processes.
+Together, these concepts provide a framework for understanding the mechanisms underlying the course of development, while also accounting for the complex interplay between individual development and the influence of environmental factors.
+By considering the intricate relationships among these factors, researchers can better identify the critical time periods, situations, and contexts that contribute to individual differences in developmental outcomes.
+This awareness enables more precise inferences regarding the causal relationships between exposures and outcomes, ultimately leading to more robust and meaningful findings that can help facilitate the translation of research findings into practical applications in clinical and public health settings.
+
+We also discussed some of the opportunities, challenges, and pitfalls that arise when working with longitudinal data.
+Key issues include selecting appropriate methods to account for the intricacies of longitudinal data,
+addressing missing data in a way that minimizes biases, and determining suitable longitudinal data structures that align with research questions and context.
+To address these challenges, researchers should carefully consider issues such as study design, selection of methods that account for both within- and between-person sources of variance, and employing modern techniques, (e.g., FIML, multiple imputation) for handling missing data.
+By adhering to best practices in longitudinal research and remaining vigilant of potential pitfalls, researchers can effectively harness the power of longitudinal data to maximize the potential of their investigations and gain valuable insights into complex developmental processes, individual differences, and the underlying mechanisms that drive change over time.
+
+The final section, along with associated code and additional resources made available as online supplements (), aims to serve as a resource for researchers seeking to understand and implement various longitudinal panel models.
 By providing an overview of different approaches, their strengths and limitations, and key considerations for their use, we hope to facilitate the selection of appropriate models tailored to specific research questions and data structures.
 It is essential for researchers to consider their research objectives, the characteristics of their data, and the assumptions underlying each model when choosing the most suitable approach for longitudinal analysis.
-In this vast and continually evolving field, with numerous models and approaches available to address a wide range of research questions, no single model is universally applicable or without limitations.
-The diversity of methods ensures that researchers can find an appropriate tool for their specific needs.
-By familiarizing themselves with the various types of longitudinal models, researchers can more effectively navigate the complexities of longitudinal data and contribute valuable insights into the developmental processes and individual differences that shape human experience.
-
 We encourage researchers to consult the cited literature and online supplements for further guidance in selecting and implementing longitudinal models when using the ABCD Study dataset.
@@
 By staying informed about developments in this area and critically evaluating the appropriateness of different models for their research questions, researchers can ensure that their longitudinal analyses are both rigorous and informative.
-
-------------------------------------------------------------------------
-
-# Part IV: Discussion
-
-\label{sec:headings}
-
-### Subsection TBD.
+Notably, in this vast and continually evolving field, with numerous models and approaches available to address a wide range of research questions, no single model is universally applicable or without limitations.
+The diversity of methods ensures that researchers can find an appropriate tool for their specific needs.
+By familiarizing themselves with the various types of longitudinal models, researchers can more effectively navigate the complexities of longitudinal data and contribute valuable insights into the developmental processes and individual differences that shape human experience.
 

### 2023-06-15 test.qmd
revised text

@@
-######
-######
-######
-######
-
 ---
-title: "Longitudinal Analysis Manuscript: Working Draft"
+title: |
+  Longitudinal Analysis Manuscript: Working Draft
+subtitle: Longitudinal Analysis using data from the ABCD Study
+short-title: Longitudinal Analysis
+published: "Under review at *xxxxx Journal*"
+code-repo: "Access the code, data, and analysis at <http://swhawes.github.io/test/>"
 number-sections: true
 number-depth: 3
-subtitle: Longitudinal Analysis using data from the ABCD Study 
-authors:
-  - name: Samuel W. Hawes
+correspondence-prefix: "Correspondence concerning this article should be addressed to"
+author:
+  - name: Samuel Hawes
+    email: shawes@fiu.edu
+    title: Associate Research Professor
+    affiliations: 
+      #- id: ds9
+        name: Florida International University
+        department: Center for Children & Families
+        address: 11200 SW 8th St, AHC-IV
+        city: Miami
+        region: FL
+        country: USA
+        postal-code: 33199
+    attributes:
+      corresponding: true
+  - name: Andrew K. Littlefield
+    email: andrew.littlefield@ttu.edu
+    affiliations: 
+      #- id: ds9
+        name: Texas Tech University
+        department: Psychological Sciences
+        city: Lubbock
+        region: TX
+        country: USA
+  - name: Daniel A. Lopez
+    email: Daniel_Lopez2@URMC.Rochester.edu
+    affiliations:
+      #- id: terok
+        name: University of Rochester Medical Center
+        department: Department of Public Health
+        city: Rochester
+        region: NY
+        country: USA
+      #- ref: ds9
+  - name: Kenneth J. Sher
+    email: SherK@missouri.edu
+    affiliations: 
+      #- id: ds9
+        name: University of Missouri
+        department: Psychological Sciences
+        city: Columbia
+        region: MO
+        country: USA
+  - name: Erin L. Thompson
+    email: erthomps@fiu.edu
+    affiliations:
+      #- id: terok
+        name: Florida International University
         department: Center for Children and Families
-    affiliation: Florida International University
-    location: Miami, FL
-    email: shawes@fiu.edu
-  - name: Andrew K. Littlefield
-    department: Psychological Sciences
-    affiliation: Texas Tech University
-    location: Lubbock, TX
-    email: andrew.littlefield@ttu.edu
-  - name: Daniel A. Lopez
-    department: Department of Public Health
-    affiliation: University of Rochester Medical Center
-    location: Rochester, NY
-    email: Daniel_Lopez2@URMC.Rochester.edu
-  - name: Kenneth J. Sher
-    department: Psychological Sciences
-    affiliation: University of Missouri
-    location: Columbia, MO
-    email: SherK@missouri.edu
-  - name: Erin L. Thompson
+        city: Miami
+        region: FL
+        country: USA
+  - name: Raul Gonzalez
+    email: gonzara@fiu.edu
+    affiliations: 
+      #- id: ds9
+        name: Florida International University
         department: Center for Children and Families
-    affiliation: Florida International University
-    location: Miami, FL
-    email: erthomps@fiu.edu
-  - name: Raul Gonzalez
-    department: Center for Children and Families
-    affiliation: Florida International University
-    location: Miami, FL
-    email: gonzara@fiu.edu  
+        city: Miami
+        region: FL
+        country: USA
   - name: Additional Co-authors
-    department: xxxxxx
-    affiliation: University of xxxx
-    location: xxxxxxx
-    email: xxxxxx
+    email: xxxxx@fiu.edu
+    affiliations:
+      #- id: terok
+        name: xxxxxx University
+        department: xxxxxx Dept
+        city: xxxx
+        region: xxxxx
+        country: xxxxx
   - name: Wesley K. Thompson
+    email: wes.stat@gmail.com
+    affiliations:
+      #- id: terok
+        name: Laureate Institute for Brain Research
         department: Center for Population Neuroscience and Genetics
-    affiliation: Laureate Institute for Brain Research
-    location: Tulsa, OK
-    email: wes.stat@gmail.com
+        city: Tulsa
+        region: OK
+        country: USA
 
@@
 By addressing these complexities, the review seeks to equip researchers with the necessary knowledge and tools to make informed decisions as they navigate effectively analyzing and interpreting data available in the ABCD Study.
-
-author-summary: |
-  The Adolescent Brain Cognitive Development (ABCD) Study offers valuable insights into adolescent brain development.
-Researchers working with this data face methodological and analytic concerns, including sample size, attrition, missing data, and measurement.
-This review addresses complexities in analyzing longitudinal data, with a focus on statistical models accounting for correlations in repeated measurements and their assumptions.
-We emphasize the importance of choosing appropriate models and analytic functions based on the data and the specific scientific question to be addressed.
-This concise resource aims  to help researchers navigate longitudinal data analysis, make informed methodological decisions, and enhance the accuracy and reliability of their findings when working with data from the ABCD Study.
-
+# thanks: |
+#   Placeholder text generated at <https://vlad-saling.github.io/star-trek-ipsum/>.
+# additional-info: |
+#   We have no known conflict of interest to disclose.
 keywords:
-    *ABCD Study*,
-    *longitudinal analyses*,
-    *development*
+  - ABCD Study
+  - longitudinal analyses
+  - development
 
 format:
-  html:
-    page-layout: article
-    mermaid:
-      theme: forest
+  #hikmah-pdf: default
+  docx:
     toc: true
-    toc-depth: 1
-    toc-expand: 2
-    toc-title: Table of Contents
-    code-tools:
-      source: true
-      toggle: false
-      caption: none
-    grid:
-      sidebar-width: 200px
-      margin-width: 0px
-  pdf: 
-    geometry:
-      - top=30mm
-      - left=20mm
-    keep_tex: yes
-    #fontfamily: libertinus
-    #fontsize: 20pt
-    colorlinks: true
-format-links: [html, pdf]
-
-execute:
-  echo: false
-
-output: 
-  rticles::arxiv_article
-bibliography: "references.bib"
-biblio-style: unsrt
+#   hikmah-manuscript-pdf: default
+#   #html:
+#     #page-layout: article
+#     #toc: true
+#     #toc-depth: 1
+#     #toc-expand: 2
+#     #toc-title: Table of Contents
+#   #   code-tools:
+#   #     source: true
+#   #     toggle: false
+#   #     caption: none
+#   #   grid:
+#   #     sidebar-width: 200px
+#   #     margin-width: 0px
+#   pdf:
+#     geometry:
+#       - top=30mm
+#       - left=20mm
+#   #   keep_tex: yes
+#     #fontfamily: libertinus
+#     #fontsize: 20pt
+# #colorlinks: true
+# #format-links: [html, pdf]
+# 
+# execute:
+#  echo: false
+# 
+# # output: 
+# #   rticles::arxiv_article
+# bibliography: "references.bib"
+# #biblio-style: unsrt
 ---
@@
 Together, this presents a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes emerge and unfold across critical periods of development.
-
-\[section still be developed...\]
+In the following text, we describe methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents.
+We begin with a brief orientation to the ABCD Study and its accompanying [publicly available dataset](https://nda.nih.gov/abcd/ "Visit the ABCD Study data repository"){#ndardata}.
 
 ## The ABCD Study® Data
+
 Participants enrolled in the ABCD Study include a large cohort of youth (n=11880) aged 9-10 years at baseline and their parents/guardians.
@@
 
-## Organization of current manuscript
+##
 
-The rich longitudinal nature of the ABCD Study dataset will allow researchers to perform analyses of both methodological and substantive interest.
-This report describes methods for longitudinal analyses of ABCD Study data that can address its fundamental scientific aims, as well as challenges inherent in a large population-based long-term study of adolescents.
-The manuscript is organized as follows:
+::: {.callout-note collapse="true" icon="false"}
 
-\[section still be developed...\]
+## Organization and Aims
 
-::: {.callout-note  collapse="true" icon=false}
-## Organization `callout`
-Note that there are five types of callouts, including:
-`xxxxx`, `xxxxxx`, `xxxxxx`.
+• Longitudinal Research (Basic Concepts and Considerations)
+
+  -   Concepts and considerations of developmental research
+
+• Longitudinal Data (Interpretation / Issues / Pitfalls & Assumptions)
+
+  -   Considerations when working with large longitudinal datasets
+
+  -   xxxxx
+
+• Longitudinal Analysis (Methods/Procedures/Models/etc...)
+
+  -   Learn how to apply xxxxx to large longitudinal data such as the ABCD Study
+
+  -   Use linked open-source materials to get started more quickly
+
+most commonly used analysis for abcd data
+
+Learn
+
+• to demonstrate the use of various longitudinal models
+• to discuss the connections between the methods for modeling
+• to compare existing methods to deal with longitudinal xxxxxx
+
+**Outline:**
+Longitudinal Research (Basic Concepts and Considerations)
+Longitudinal Data (Interpretation / Issues / Pitfalls & Assumption)
+Longitudinal Analysis (Methods/Procedures/Models/etc...)
+
 :::
@@
 
-![LDA Structural Diagram](./figures/LDA Structure 2.png){#fig-LDAstruct  fig-alt="LDA Structure Diagram." width=600 height=400 fig-align="center"}
-
 **Defining Features of Longitudinal Data Analysis.** The hallmark characteristic of longitudinal data analysis is its application to repeated assessments of the same assessment targets (e.g., individuals, families) across time.
@@
 
-"traditional approaches": marginal models
-
-random effects: Growth Curves- SEM (lgcm) vs ME (lmm)
-
-(between- and within-person levels?)
+["traditional approaches": marginal models]
+[random effects: Growth Curves- SEM (lgcm) vs ME (lmm)]
+[(between- and within-person levels?)]
 
@@
 
-### Marginal vs Conditional Models (?)
-
-\[section still under development...\]
-
 ### Missing Data/Attrition
@@
 
-[psychometrics as a new "Part" below, or facets of psychometrics (e.g., invariance, reliability, etc...) inserted above?]
+### Longitudinal Data Structures
 
-# Part III: Additional Section(s)
+An ideal longitudinal study integrates (a) a well-articulated theoretical model, (b) an appropriate longitudinal data structure, and (c) a statistical model that is an operationalization of the theoretical model [@collins2006].
+To accommodate various research questions and contexts, different types of longitudinal data and data structures have emerged (see Figure x).
+An understanding of these data structures is helpful, as they can warrant different types of longtiudinal data analysis.
+Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and Bauer (xxxx) provide a nice on-ramp in first distinguishing between the use of "time-to-event" and "repeated measures" data.
+Although both model time, the former is concerned with *whether* and *when* an event occurs, whereas the later is focused on *growth* and *change* \[Curran and Bauer, xxxx\].
+Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest.
+This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models) and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events.
+While much has been written about about "time-to-event" data (see xxx; xxx; xxxx), including several recent xxxx (see), our emphasis will be given to the modeling of "repeated measures" data.
 
-\label{sec:headings} \## In Progress: \[TBD\]
+[![LDA Structural Diagram](./figures/LDAStructure2.png)](https://centerstat.org/resources/)
 
-Section focused on different ways of considering/grouping/deciding on which specific analysis to choose (see curran&bauer slides)
+When discussing longitudinal analysis, we are most often talking about data collected on the same unit (e.g., individuals) across multiple measurement occasions.
+However, repeated-measures analysis is not a monolith and it will serve us well to distinguish between a few of the most common types.
+One such approach to repeated measures analysis is the use of `time-series models`.
+These models generally consist of a long sequence of repeated measurements (≧ 50-100 measurements) on a single (or small number of) some variable of interest.
+Time-series analysis is often used to predict/forecast temporal trends and cyclic patterns and is geared toward making inferences about prospective outcomes within a population (with relatively with less focus on inferring individual-level mechanisms and risk factors).
+A related type of repeated measures analysis is `Intensive Longitudinal Data (ILD)`.
+Similar to time-series analysis, ILD models involve frequent measurements (\~ 30-40 measurements) of the same individuals in a relatively circumspect period of time (e.g., experience sampling to obtain time series on many individuals).
+Although ILD models may include slightly fewer measurement occasions than time-series data, ILD models tend to have more subjects than time series models (\~ 50-100 subjects).
+This allows ILD models to examine short-term patterns by incorporating a time series model that can sometimes fit parameter estimates to each individual's data in order to model individual difference outcomes.
+The final type of repeated measures analaysis that we will discuss is the longitudinal panel model.
+These models follow a group of individuals--- a panel (also referred to as a cohort) --- across relatively fewer measurement occassions (~ 5-15), and are often interested in examining change across both, within-individuals and between-individuals.
 
-### Subsection TBD.
+While other longitudinal designs have their own unique strengths and applications, the longitudinal panel design is particularly well-suited for investigating developmental processes in the context of the ABCD Study.
+In the following sections, we will discuss various analytic methods commonly used to analyze longitudinal panel data, including growth models, mixed models, and a number of additional trajectory models.
+These methods provide valuable insights into within- and between-individual differences and are highly relevant for researchers working with the ABCD Study dataset.
+By focusing on these methods, we aim to equip readers with the knowledge necessary to conduct longitudinal research and perform analyses using the rich, longitudinal, and publicly available data from the ABCD Study.
 
-![Longitudinal Models](./figures/Mindmap1.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width=600 height=400 fig-align="center"}
+# Part III: Longitudinal Analysis
+
+\label{sec:headings}
+
+## xxx types of analysis for abcd study data xxxx
+
+### Types of longitudinal panel models
+
+With the large and continually expanding body of research on statistical methods for longitudinal analyses, determining which longitudinal model to implement can be challenging.
+The aim of this section is to help researchers to navigate these many options in order to identify the statistical approach most appropriate to their unique research question.
+Notably, there are a myriad of viable ways one can go about grouping various types of longitudinal models for presentation.
+Common examples iclude grouping by linearity (linear vs nonlinear models; see Collins, 2006(?)), the number of measurement occassions (see King et al., 2018), and xxxxxx (see xxxx, xxx).
+The ordering/organzation we use below overlaps in a number of ways with these examples, and with Curran & Bauer (2019); however, it is important to note that in each case, the ordering/organization that is chosen is somewhat arbitrary and primarly intended to allow the reader to compare and contrast of various analytic approaches.
+In the following we will describe (briefly) and summarize the advantages/disadvantages for a series of 16(?) distinct longitudinal models organized into the following x# `fuzzy` groupings: Traditional Models, Modern GLM Extensions, SEM, Advanced SEM, xxx, xxxx.
+We note that this is note an exhaustive review of each of these methods, but for more in-depth detail we do provide the reader relevant resources.
+As aptly summarized by Curran & Bauer (2019), "there are many exceptions, alternatives, nuances, “what ifs”, and “but couldn’t yous” that aren’t addressed here."
+
+#### Traditional Models
+Traditional methods for longitudinal analysis primarily focus on modeling mean-level change. 
+For example, is there a difference in average math scores obtained across multiple assessments between boys and girls?
+Longitudinal models that focus on mean-level change are also referred to as `marginal models` and examples of specific methods include `repeated measures ANOVA` (and `MANOVA`), `ANCOVA`, and `Generalized Estimating Equations` (GEE).
+These methods are commonly used when data is only available from 2 measurement occassions.
+For example, computing a `difference score` (e.g., math scores at t2 - math scores at t1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups.
+We also note the longitudinal `signed-rank test` which is a nonparametric alternative to the paired t-test that can be a useful tool for analyzing non-normal paired data.
+Another common approach, often used in pre-/post-design studies, is to use `residualized change score analysis` to assess the degree of change in a variable, while controlling for its initial level.
+For example, regressing post-treatment scores on pre-treatment scores and then using the resulting residuals as a measure of change that is adjusted for baseline scores (ignoring any prior group assignments/differences).
+Similar to difference scores, the residualized change score is often included in subsequent analysis, such as for evaluating intervention effects in pre-test/post-test designs [@Kisbu-Sakarya2013].
+
+Traditional longitudinal methods can still be useful in some contexts (e.g., < 3 measurement occasions), but overall their practical utility for answering questions about developmental processes is limited.
+Perhaps most notably, these models do not allow for characterizing patterns of within-person change, which is often an emphasis of longitudinal research.
+In turn, these models often correspond poorly with most theoretical models of change.
+
+Moreover, even determining which of these procedures to use for comparing change over two time points across groups can be surprisingly complicated.
+A particularly vexing example is that of imbalanced baseline scores which can produce biased effects when using these methods.
+Indeed, it is well-documented that this can result in different conclusions between these approaches, which has led to many articles debating the merits of which procedures is most appropriate to use and when [see xxxxx for a review].
+Beyond these concerns, additional limitations include xxxxxx.
+Given these shortcomings and the complexity of the issues surrounding some of these methods, when possible, it is typically recommended that researchers make use of more modern approaches for analyzing longitudinal data (and preferably make use of data collected across 3 or more time points.
+
+#### Modern GLM Extensions
+
+Modern approaches to longitudinal data analysis have advanced beyond traditional methods by offering greater flexibility and a more in-depth understanding of within-person and between-person variability.
+`Generalized Estimating Equations (GEE)`, `Linear Mixed Models (LMM)`, `Generalized Linear Mixed Models (GLMM)`, and `Autoregressive Cross-Lagged Panel Models (ARCL)` are examples of such contemporary techniques.
+GEE is an extension of Generalized Linear Models.
+This model combines the generalized linear model for non-normal outcomes with repeated measures (marginal) model and is suitable for analyzing correlated longitudinal data and modeling population-averaged effects.
+LMMs, also known as multilevel or hierarchical linear models, facilitate the simultaneous analysis of within-person and between-person variability, making them ideal for nested data structures or repeated measures.
+GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data.
+Finally, ARCL models are uses to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects.
+
+The strengths of these modern methods lie in their ability to account for individual differences, within-person change, and time-varying predictors, thereby providing a more comprehensive understanding of complex relationships in longitudinal data.
+Despite these advantages, modern approaches may require more complex modeling assumptions and higher computational demands compared to traditional methods.
+Additionally, proper model specification and the interpretation of results can be more challenging, especially in cases of high multicollinearity or missing data.
+However, modern longitudinal analysis methods have generally surpassed traditional methods in addressing a wider range of research questions, accommodating diverse data structures, and elucidating the intricate dynamics of developmental processes.
+
+#### SEM
+
+Structural equation modeling (SEM) approaches have gained prominence in longitudinal data analysis due to their ability to estimate complex relationships among observed and latent variables while accounting for measurement error.
+SEM techniques can be categorized into variable-centered, person-centered, and hybrid approaches, each with unique strengths and limitations.
+The choice of method depends on the research question, data structure, and underlying assumptions.
+
+Variable-centered approaches, such as Latent Change Scores, Latent Growth Curve Models, Multivariate (Parallel Process) Latent Growth Curve Models, and Latent State-Trait Models, focus on examining relationships among variables and population-level patterns.
+These models offer a powerful means to estimate individual change trajectories and the relationships between growth parameters of different variables, while also decomposing observed measurements into latent state and trait components.
+However, these approaches may not adequately capture distinct subgroups of individuals who share similar patterns of change over time, which can be crucial for understanding heterogeneous developmental processes.
+
+Person-centered approaches, including Latent Transition Analysis and Latent Class Growth Analysis, address this limitation by identifying subgroups of individuals who share similar patterns of change.
+These models can reveal meaningful subpopulations and help researchers understand the factors that contribute to differences in developmental trajectories.
+Nevertheless, person-centered methods may overlook important relationships among variables, which can be essential for understanding the dynamics of change over time.
+
+Hybrid approaches, such as Growth Mixture Modeling, combine aspects of both variable-centered and person-centered models, allowing for the identification of latent subgroups while also modeling relationships among growth parameters.
+This combination provides a more comprehensive understanding of longitudinal data by capturing both within- and between-person variability.
+However, hybrid models can be more complex, necessitating careful model specification, selection, and interpretation.
+Additionally, these methods may require larger sample sizes to ensure the stability and accuracy of results.
+
+In summary, SEM approaches offer powerful tools for longitudinal data analysis, enabling researchers to investigate complex relationships, individual differences, and change dynamics over time.
+The choice between variable-centered, person-centered, and hybrid approaches depends on the research objectives and the nature of the data.
+Despite their limitations, these models have greatly advanced our understanding of developmental processes and the factors that contribute to individual differences in change trajectories.
+
+#### Advanced SEM
+
+Advanced structural equation modeling (SEM) approaches, such as the Random-Intercept Cross-Lagged Panel Model (RI-CLPM) and Latent Curve Models with Structured Residuals (LCM-SR), have emerged to address more complex research questions and data structures in longitudinal analysis.
+These advanced models extend traditional SEM techniques, enabling researchers to disentangle within-person and between-person effects, as well as capture additional time-specific dependencies and associations that may not be accounted for by the latent growth factors.
+
+The RI-CLPM enhances the traditional cross-lagged panel model by incorporating random intercepts, which allow for the separation of stable individual differences from the dynamic within-person associations between variables over time.
+This distinction is particularly valuable for understanding the time-specific effects of one variable on another, while accounting for the influence of individual differences.
+However, RI-CLPM may require larger sample sizes to ensure stability and accuracy of the estimates and can be computationally demanding.
+
+LCM-SR, on the other hand, extends the Latent Growth Curve Model by modeling structured residuals, such as autoregressive or cross-lagged effects among the residuals.
+This approach allows for a more comprehensive understanding of within-person change dynamics and factors influencing change over time.
+By including structured residuals, LCM-SR can capture additional time-specific relationships that are not explained by the latent growth factors.
+However, similar to RI-CLPM, LCM-SR also comes with increased model complexity and requires careful specification and interpretation.
+
+In conclusion, advanced SEM approaches for longitudinal data analysis provide valuable tools for addressing complex research questions and data structures.
+While they offer more nuanced insights into within-person change dynamics and the influence of individual differences, these models also come with certain limitations, such as increased complexity, computational demands, and the need for careful model specification and interpretation.
+As with any statistical method, researchers should carefully consider their research objectives, data characteristics, and the assumptions of each model when selecting the most appropriate advanced SEM approach for longitudinal analysis.
+
+#### (additional section(s)?)
+(longitudinal neuroimaging?)
+(longitudinal twins/genetics?)
+
+![Longitudinal Models/Data Structures](./figures/Mindmap1.png){#fig-LONGmod fig-alt="Longitudinal Models Diagram." width="600" height="400" fig-align="center"}
 
@@
 
-### Subsection TBD.
+[Opening paragraph of discussion xxxxx.]
 
-|                                           |                         |                          |                                      |                         |
-|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
-| Models                                    |     Form of Change      | Minimum \# of Timepoints | Variable-Centered vs Person-Centered | Time-varying Covariates |
-| Difference Scores                         |       mean-level        |            2+            |                  \-                  |          none           |
-| Change Scores                             |       rank-order        |            2+            |                  \-                  |          none           |
-| Signed-Rank Test                          |       rank-order        |            2+            |                  \-                  |          none           |
-| Generalized Estimating Equations          |       rank-order        |            2+            |                  \-                  |         allowed         |
-| Linear Mixed Models                       |   within and between    |            3+            |          Variable-Centered           |         allowed         |
-| Generalized Linear Mixed-Effects Models   |   within and between    |            3+            |                                      |         allowed         |
-| Autoregressive Crosslagged Panel Models   |       rank-order        |            3+            |          Variable-Centered           |         allowed         |
-| Latent Change Scores Models               |       rank-order        |            2+            |          Variable-Centered           |         allowed         |
-| Latent Growth Curve Models                |   within and between    |            3+            |          Variable-Centered           |         allowed         |
-| Multivariate Latent Growth Curve Model    |   within and between    |            3+            |          Variable-Centered           |         allowed         |
-| Latent Transition Analysis                |   within and between    |            3+            |                Mixed                 |         allowed         |
-| Latent Class Growth Analysis              |   within and between    |            3+            |           Person-Centered            |         allowed         |
-| Growth Mixture Models                     |   within and between    |            3+            |                Mixed                 |         allowed         |
-| State-Trait Models                        | continuous and discrete |            3+            |          Variable-Centered           |         allowed         |
-| Random-Intercept Crosslagged Panel Models |   within and between    |            4+            |          Variable-Centered           |         allowed         |
-| Latent Curve Models Structured Residuals  |   within and between    |            4+            |          Variable-Centered           |         allowed         |
-|                                           |                         |                          |                                      |                         |
+xxxxxx
 
-: LDA Models {#tbl-lda .striped .hover .responsive-xl}
+We began by addressing fundamental concepts and considerations in longitudinal research that are essential for generating accurate and meaningful insights into developmental processes.
+Concepts such as vulnerable periods, developmental disturbances and snares, or cascade and experience effects (among many others), are instrumental in shaping the design, analysis, and interpretation of longitudinal studies.
+For example, understanding and accounting for vulnerable periods and developmental disturbances in research designs allows researchers to investigate the timing and magnitude of the impact of specific factors on development, while snares and cascade effects demonstrate the importance of considering the interconnectedness of developmental processes.
+Together, these concepts provide a framework for understanding the mechanisms underlying the course of development, while also accounting for the complex interplay between individual development and the influence of environmental factors.
+By considering the intricate relationships among these factors, researchers can better identify the critical time periods, situations, and contexts that contribute to individual differences in developmental outcomes.
+This awareness enables more precise inferences regarding the causal relationships between exposures and outcomes, ultimately leading to more robust and meaningful findings that can help facilitate the translation of research findings into practical applications in clinical and public health settings.
 
-axasx
+We also discussed some of the opportunities, challenges, and pitfalls that arise when working with longitudinal data.
+Key issues include selecting appropriate methods to account for the intricacies of longitudinal data,
+addressing missing data in a way that minimizes biases, and determining suitable longitudinal data structures that align with research questions and context.
+To address these challenges, researchers should carefully consider issues such as study design, selection of methods that account for both within- and between-person sources of variance, and employing modern techniques, (e.g., FIML, multiple imputation) for handling missing data.
+By adhering to best practices in longitudinal research and remaining vigilant of potential pitfalls, researchers can effectively harness the power of longitudinal data to maximize the potential of their investigationss and gain valuable insights into complex developmental processes, individual differences, and the underlying mechanisms that drive change over time.
 
-See @tbl-lda.
+The final section, along with associated code and additional resources made available as online supplements (), aims to serve as a resource for researchers seeking to understand and implement various longitudinal panel models.
+By providing an overview of different approaches, their strengths and limitations, and key considerations for their use, we hope to facilitate the selection of appropriate models tailored to specific research questions and data structures.
+It is essential for researchers to consider their research objectives, the characteristics of their data, and the assumptions underlying each model when choosing the most suitable approach for longitudinal analysis.
+In this vast and continually evolving field, with numerous models and approaches available to address a wide range of research questions, no single model is universally applicable or without limitations.
+The diversity of methods ensures that researchers can find an appropriate tool for their specific needs.
+By familiarizing themselves with the various types of longitudinal models, researchers can more effectively navigate the complexities of longitudinal data and contribute valuable insights into the developmental processes and individual differences that shape human experience.
 
-xxxxx
+We encourage researchers to consult the cited literature and online supplements for further guidance in selecting and implementing longitudinal models when using the ABCD Study dataset.
+As the field continues to advance, we anticipate the emergence of new methods and refinements to existing approaches, further expanding the toolkit available to researchers for the analysis of longitudinal data.
+By staying informed about developments in this area and critically evaluating the appropriateness of different models for their research questions, researchers can ensure that their longitudinal analyses are both rigorous and informative.
+
+------------------------------------------------------------------------
+
+{{< pagebreak >}}
 

### 2023-06-15 delete.js

@@ -1,0 +1,3 @@
+function generateRandomNumberBetween1And10()
+
+//q: what does SOLID stand for in oo programming?

### 2023-06-15 manuscript.qmd
testing

@@
 
-\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largestz long-term investigation of neurodevelopment and child health in the United States.
+\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largestz long-termzz investigation of neurodevelopment and child health in the United States.
 Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018].

### 2023-06-15 manuscript.qmd
author list updates

--- 2023-06-15 manuscript.qmd
@@
         country: USA
-  # - name: Erin L. Thompson
-  #   email: erthomps@fiu.edu
-  #   affiliations:
-  #     #- id: terok
-  #       name: Florida International University
-  #       department: Center for Children and Families
-  #       city: Miami
-  #       region: FL
-  #       country: USA
-  # - name: Raul Gonzalez
-  #   email: gonzara@fiu.edu
-  #   affiliations:
-  #     #- id: ds9
-  #       name: Florida International University
-  #       department: Center for Children and Families
-  #       city: Miami
-  #       region: FL
-  #       country: USA
+  - name: Erin L. Thompson
+    email: erthomps@fiu.edu
+    affiliations:
+      #- id: terok
+        name: Florida International University
+        department: Center for Children and Families
+        city: Miami
+        region: FL
+        country: USA
+  - name: Raul Gonzalez
+    email: gonzara@fiu.edu
+    affiliations:
+      #- id: ds9
+        name: Florida International University
+        department: Center for Children and Families
+        city: Miami
+        region: FL
+        country: USA
   - name: Additional Co-authors
@@
 
-\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largestz long-term investigation of neurodevelopment and child health in the United States.
+\label{sec:headings} [The Adolescent Brain Cognitive Development (ABCD) Study®](https://abcdstudy.org/ "Visit the ABCD Study website"){#abcdwebsite} is the largest long-term investigation of neurodevelopment and child health in the United States.
 Conceived and initiated by the National Institutes of Health (NIH), this landmark prospective longitudinal study aims to transform our understanding of the genetic and environmental influences on brain development and their roles in behavioral and health outcomes in adolescents [@volkow2018].

### 2023-06-16 manuscript.qmd
updated citations

@@
 With the large and continually expanding body of research on statistical methods for longitudinal analyses, determining which longitudinal model to implement can be challenging.
-The aim of this section is to help researchers navigate these many options in order to identify the statistical approach most appropriate to their unique research question.
+The aim of this section is to help researchers navigate these many options to identify the statistical approach most appropriate to their unique research question.
 Notably, there are a myriad of viable ways one can go about grouping various types of longitudinal models for presentation.
@@
 residualized change; see @castro-schilo2018).
-The ordering/organzation we use below overlaps in a number of ways with these examples, and in particular with Bauer and Curran (2019; @bauer2019).
-However, it is important to note that in each case, the ordering/organization that is chosen is somewhat arbitrary and primarly intended to allow the reader to compare and contrast various analytic approaches.
-In the following we describe (briefly) and summarize the advantages/disadvantages for a series of longitudinal models organized into the following `fuzzy` groupings: Traditional Models, Modern GLM Extensions, SEM, and Advanced SEM.
+The organization we use below overlaps in a number of ways with these examples, and in particular with Bauer and Curran (2019; @bauer2019).
+However, it is important to note that in each case, the ordering that is chosen is somewhat arbitrary and primarily intended to allow the reader to compare and contrast various analytic approaches.
+In the following we describe (briefly) and summarize the advantages/disadvantages of a series of longitudinal models organized into the following groupings: Traditional Models, Modern GLM Extensions, SEM, and Advanced SEM.
 We note that this is not an exhaustive review of each of these methods, but for more in-depth detail we do provide the reader relevant resources.
@@
 Longitudinal models that focus on mean-level change are also referred to as `marginal models` and examples of specific methods include `repeated measures ANOVA` (and `MANOVA`), `ANCOVA`, and `Generalized Estimating Equations` (GEE).
-These methods are commonly used when data is only available from 2 measurement occassions.
+These methods are commonly used when data is only available from 2 measurement occasions.
 For example, computing a `difference score` (e.g., math scores at t2 - math scores at t1) that can be used as an outcome in a subsequent GLM analysis (e.g., paired-samples t-test, repeated measures ANOVA) to test for differences in patterns of change over time and between groups.
-Additionally the longitudinal `signed-rank test`, a nonparametric alternative to the paired t-test, can be a useful tool for analyzing non-normal paired data.
+Additionally, the longitudinal `signed-rank test`, a nonparametric alternative to the paired t-test, can be a useful tool for analyzing non-normal paired data.
 Another common approach, often used in pre-/post-design studies, is to use `residualized change score analysis` to assess the degree of change in a variable, while controlling for its initial level.
@@
 GLMMs further extend the LMM framework to accommodate non-normal response variables, such as binary, count, or ordinal data.
-Finally, ARCL models are uses to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects.
+Finally, ARCL models are used to investigate reciprocal relationships between variables over time, as they estimate both autoregressive and cross-lagged effects (although, ARCL models are relatively less useful for teasing apart between-person and within-person sources of variances; see @curran2021a).
 

### 2023-06-18 manuscript.qmd
revisions

@@
 If one were to try to sum up what development in a living organism is exactly, one could plausibly argue it's the characterization of stability and change as the organism traverses the life course.
-There are a few different ways to think of stability (and change).
+There are a few different ways to think about patterns of stability and change (see @fig-StablityChange).
 Consider we measure the height of all youth in a 6th-grade class, once in the fall at the beginning of the school year and once again in the spring at the end of the school year.
@@
 For example, in some areas of development, there is considerable mean-level change that occurs over time (e.g., changes in Big 5 personality traits), but relatively high rank-order stability.
-Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of "within-individual" change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest[@curran2011].
+Despite the useful information afforded by examining mean-level and rank-order change, these approaches are limited in that they provide little information about patterns of "within-individual" change and, in turn, can result in fundamental misinterpretations about substantial or meaningful changes in an outcome of interest [@curran2011].
 
